### 2019-10-04 main_text/main_text.tex
initial commit of Kathrin's draft

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\cline{#1}%
\noalign{\vskip\arrayrulewidth}%
\noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\usepackage{refcount}
\usepackage{textgreek}
\usepackage{bm}

\newcommand{\unit}[1]{\,\mathrm{#1}}
\newcommand{\n}[1]{\mathrm{#1}}
\newcommand{\dd}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}
\newcommand*{\defeq}{\mathrel{\vcenter{\baselineskip0.5ex \lineskiplimit0pt
			\hbox{\scriptsize.}\hbox{\scriptsize.}}}%
	=}
\newcommand\subref[2]{%
	\def\myref{\getrefnumber{#1}}% extract the reference number
	\hyperref[#1]{\myref\mbox{#2}}% print the reference number with argument
}

\hyphenation{XapA}
\hyphenation{XapB}
\hyphenation{XapR}
\hyphenation{XapAB}
\hyphenation{XapABR}
\hyphenation{xapA}
\hyphenation{xapB}
\hyphenation{xapR}
\hyphenation{xapAB}
\hyphenation{xapABR}

\usepackage[obeyFinal]{todonotes} % todo notes and missing figures
\reversemarginpar

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
	{\Large
		\textbf\newline{Theoretical investigation of a genetic switch for metabolic adaptation} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
	}
	\newline
	% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
	\\
	Kathrin Laxhauber\textsuperscript{1},
	Muir J. Morrison\textsuperscript{2},
	Rob Phillips\textsuperscript{2,3$\ast$}
	\\
	\bigskip
	\textbf{1} Dept???, ETHZ, Zurich, State???, Switzerland\todo{MJM: Add dept \& complete affiliation}
	\\
	\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
	\\
	\textbf{3} Department of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
	\\
	\bigskip

	% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
	% 
	% Remove or comment out the author notes below if they aren't used.
	%
	% Primary Equal Contribution Note
	% Yinyang These authors contributed equally to this work.

	% Additional Equal Contribution Note
	% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
	% \ddag These authors also contributed equally to this work.

	% Current address notes
	% \textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
	% \textcurrency b Insert second current address 
	% \textcurrency c Insert third current address

	% Use the asterisk to denote corresponding authorship and provide email address in note below.
	$\ast$ phillips@pboc.caltech.edu

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Membrane transporters carry important molecules through the cell membrane
and, from a resource standpoint, should only be produced when necessary. The
expression of membrane transporters in metabolic pathways is often
upregulated by the transporter substrate. In \emph{E. coli}, such systems
include for example the \emph{lacY}, \emph{araFGH}, and \emph{xylFGH} genes,
which encode for lactose, arabinose, and xylose transporters, respectively.
As a case study, we build a physical model of the \emph{xapABR} genetic
circuit, which features a regulatory feedback loop through membrane
transport (positive feedback) and enzymatic degradation (negative feedback)
of an inducer. Dynamical systems analysis and stochastic simulations show
that the membrane transport makes the model system bistable in certain
parameter regimes. Thus, it serves as a genetic “on-off” switch, enabling
the system to only produce a set of metabolic enzymes when the corresponding
metabolite is present in large amounts. We find that the negative feedback
from the degradation enzyme does not significantly disturb the positive
feedback from the membrane transporter. We investigate hysteresis in the
switching and discuss the role of cooperativity and multiple binding sites
in the model circuit. Altogether, this work explores how a stable genetic
switch for a set of enzymes is obtained from transcriptional auto-activation
of a membrane transporter through its substrate.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
% For figure citations, please use "Fig" instead of "Figure".
% Place figure captions after the first paragraph in which they are cited.
% Results and Discussion can be combined.
% Place tables after the first paragraph in which they are cited.
% PLOS does not support heading levels beyond the 3rd (no 4th level headings).
% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.

\section*{Introduction}
Genetic regulatory circuits are fundamental building blocks of functioning
cells and organisms. One abundant class of these circuits are genetic
switches. They can be constructed in various ways and have different
functions. What all of them have in common is that the system is bistable:
there are two stable steady-states that it can be in. The distribution of
gene expression in a cell culture can then be bimodal. This is not to be
confused with mere stochastic bimodality, where the system is not stable,
but the gene expression in each cell is fluctuating between the two levels.

One classic example of a genetic switch is a system where two repressor
proteins each regulate the transcription of the
other~\cite{Jacob1961,Gardner2000} (illustrated schematically in
Fig~\ref{fig1:switches}). Here, the two stable states are high expression of
the first protein and low expression of the second one or the opposite. This
switch enables the system to have a memory: if something induces expression
of either one of the proteins, the system will remain in this state until a
significant perturbation occurs. Another well known and even simpler example
is an auto-activating circuit in which a protein activates its own
transcription~\cite{Wolf1998}. This gives the system an ``on-off''-switch.
%\todo{make eps}

\begin{figure}%[h!]
	\centering
	\includegraphics{media/Circuits.eps}
	\caption{{\bf A schematic of different genetic switches.}
	(A) and (B) show the two most well-known genetic switches that are
	mentioned in the text. In (C), a very much simplified version of the
	circuit that we investigate in this paper can be seen, where the
	similarity to the switch in (B) becomes obvious. A complete version of
	the model circuit can be found in Fig~\ref{fig2:model}.}
	\label{fig1:switches}
\end{figure}

Through physical and mathematical modeling, we investigate a more complex
but common switch system where the bistability is due, as we will show, to a
membrane transport protein. Its key feature is the indirect activation of
the transporter gene by the transporter substrate, leading to positive
feedback similar to the aforementioned ``on-off'' switch. An example for
such an architecture is the \emph{lac} operon, where lactose indirectly
activates the expression of lactose permease. Other examples in \emph{E.
coli} include the \emph{araFGH} and \emph{xylFGH} genes, which encode for
arabinose and xylose transporters, respectively. For \emph{lac} and
\emph{araFGH}, bistability has indeed been observed and attributed to such a
positive feedback
loop~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
A eukaryotic example is the glucose transporter GLUT-2 in liver and
\textbeta-cells~\cite{Bae2010,Tiedge1991}, though this system is much more
complex than the following analysis.

It is quite conceivable that this auto-activation process is common to many
substances that a cell would want to accumulate. Such a switch enables the
cell to sense and respond to its environment: if the substrate enters the
cell, it activates the production of membrane transporters. The cell then
starts accumulating the substrate, thereby ``testing'' the substrate's
presence in the extracellular environment. If there is enough, the
expression stabilizes at an ``on'' point and the cell has, in a short-term
sense, adapted. When there is not enough substrate anymore, the gene, which
often encodes for a whole set of enzymes for this one metabolite, switches
``off'' again. Such a mechanism could be involved in various cases of
short-term adaptation.

For our modeling, we focus on the \emph{xapABR} genetic circuit from
\emph{E. coli} as a case study. It is similar to \emph{lac}, but less
complex. Instead of lactose, its purpose is to make use of the nucleoside
xanthosine as an energy source~\cite{Buxton1980,Hammer-Jespersen1980}. The
circuit is made up of two genes: one that encodes for XapR and another one
that encodes for XapA and XapB. XapR appears to be a transcription factor
that is induced by xanthosine and activates the \emph{xapAB} promoter. It
seems to have two binding sites on the promoter, but the promoter
architecture and function is not fully understood up to now. The
transcription of \emph{xapR} seems to be constitutive and not
auto-regulated~\cite{Seeger1995}. Structural homology to other transcription
factors suggests that XapR appears in dimers where one dimer can bind two
xanthosine molecules~\cite{Joergensen1999}. The protein XapA is a purine
nucleoside phosphorylase that degrades xanthosine into components (ribose
and xanthine) that can be fed into metabolic
pathways~\cite{Buxton1980,Hammer-Jespersen1980}. XapB on the other hand is a
membrane transporter of xanthosine~\cite{Seeger1995,Norholm2001}.

Experimentally it was found that the expression level of \emph{xapAB} among
cells follows a bimodal distribution and that the system seems to be
bistable~\cite{Chure2019}. We aim to understand which of the circuit's
features are necessary for bistability and investigate its behavior in
different parameter regimes. In the following section, we will discuss the
details of our model. After estimating the free parameters we then present
the observations we made through phase diagrams, followed by the results
from stochastic simulations.

\section*{Model}
\subsection*{Step by step modeling of the system}
In this section, we present our model of the \emph{xapAB} genetic circuit.
Fig~\ref{fig2:model} shows an overview of this model. It consists of the
following steps: 
\begin{itemize}
	\item Small amounts of xanthosine are transported into and out of cell in various ways (discussed in more detail later on).
	\item XapR is induced by xanthosine.
	\item The induced XapR binds to the \emph{xapAB} promoter, leading to transcription.
	\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
	\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
	\item At some point, the proteins decay.
\end{itemize}
Because xanthosine induces the transcription factor XapR, we have a positive
and negative feedback loop due to XapB and XapA, respectively. The remainder
of this subsection discusses each of the above steps in detail, leading us
to a set of two coupled ODE's. More in-depth explanations can be found
in~\nameref{S1_Text}.

\begin{figure}[h!]
	\centering
	\includegraphics{media/XapSystem.eps}
	\caption{{\bf Model of the \emph{xapAB} circuit.}
		The XapR dimers are induced by xanthosine and the induced XapR binds
		cooperatively as an activator to the \emph{xapAB} promoter. For
		these two steps, quasi-equilibrium is assumed. If both XapR binding
		sites are occupied and the polymerase is bound, the gene is
		transcribed at rate $r_{\n{m}}$. The mRNA decays at rate
		$\gamma_{\n{m}}$, and both proteins are translated at rate
		$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
		xanthosine with Michaelis-Menten parameters $k_{a}$ and $K_a$.
		Similarly, XapB imports and exports xanthosine with parameters
		$k_{\n{b,i}}, K_{\n{b,i}}$ and $k_{\n{b,e}}, K_{\n{b,e}}$,
		respectively. Furthermore, xanthosine enters and leaves the cell in
		other ways, proportional to rates $k_{\n{nup}}$ and $\xi
		k_{\n{nup}}$, respectively.}
	\label{fig2:model}
\end{figure}

\paragraph*{Induction of XapR.} 
We treat dimers as the only form of XapR that appears in the cell. Each
dimer can bind two xanthosine molecules. The Monod-Wyman-Changeux (MWC)
model is used to describe the fraction of XapR dimers in the active state: 
\begin{eqnarray}
\label{eq:MWC}
\n{[XapR]_A} = \n{[XapR]_{tot}} \frac{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2}{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2 + e^{\beta \Delta E_{\n{x}}} \left(1+\frac{\n{[x]}}{K_{\n{xA}}} \frac{K_{\n{xA}}}{K_{\n{xI}}}\right)^2}
\end{eqnarray}
Here, $\n{[x]}$ is the xanthosine concentration, and $\n{[XapR]_A}$ and
$\n{[XapR]_{tot}}$ denote the concentration of active and total XapR dimers,
respectively. Furthermore, $K_{\n{xI}}$ and $K_{\n{xA}}$ are the
dissociation constants of xanthosine to the inactive and the active XapR
dimer, respectively, and $\Delta E_{\n{x}}$ stands for the energy difference
between the inactive and the active state. We expect $\Delta E_{\n{x}} > 0$
and $K_{\n{xA}} < K_{\n{xI}}$ for inducible activation. This corresponds to
XapR being mainly inactive in the absence of xanthosine and becoming mostly
active at high concentrations of xanthosine. A detailed discussion of the
MWC model can be found in~\cite{Marzen2013}.

\paragraph*{Transcription.} 
Transcription and translation of the \emph{xapAB} gene, regulated by the
induced XapR, produce the two proteins XapA and XapB. We start with
transcription and assume that the binding of XapR and polymerase to the
promoter is at quasi-equilibrium. The polymerase binding is modeled as
independent of that of XapR, and all influence of the activator is pushed
into the transcription rate. Furthermore, the binding energy of XapR to each
of its two sites is assumed to be the same. A discussion of these
simplifications can be found in~\nameref{S1_Text}.

In Fig~\ref{fig3:states}, all possible states of the promoter in our model
and their corresponding thermodynamic weights are shown. There, $\n{[P]}$
denotes the polymerase concentration and $\Delta E_{\n{coop}}$ stands for
the interaction energy of the two XapR dimers. If cooperativity in
transcription factor binding is neglected, this is set to zero. Furthermore,
$K_{\n{XapR}}$ and $K_{\n{P}}$ denote the dissociation constant of XapR and
polymerase to the promoter, respectively. In statistical mechanics language
this is equivalent to $\frac{N_{\mathrm{NS}}}{V} e^{\beta \Delta E_{i}}$
with $N_{\mathrm{NS}}$ being the number of non-specific binding sites on the
DNA, $V$ the volume of the cell, and $\Delta E_{i}$ the interaction energy
of $i=$XapR or $i=$polymerase with the promoter.

\begin{figure}%[h!]
	\centering
	\includegraphics{media/States.eps}
	\caption{{\bf The promoter states.} We consider only the completely
	occupied state as active and all other states (faded out in the figure)
	as completely inactive. The parameters are the interaction energy of
	the two XapR dimers $\Delta E_{\n{coop}}$ and the dissociation
	constants $K_{\n{XapR}}$ and $K_{\n{P}}$ of XapR and polymerase
	to the promoter, respectively. The concentrations of polymerase
	and active XapR are denoted by $\n{[P]}$ and $\n{[XapR]}$.}
	\label{fig3:states}
\end{figure}

We consider only the state where both XapR binding sites are occupied as
active and all other states as inactive, meaning they have transcription
rate $0$. Experiments show that the expression becomes very weak when one of
the XapR binding sites is removed from the promoter, suggesting that this
simplification is reasonable~\cite{Chure2019}. Furthermore, we find that in
the bistable parameter range, considering the single occupancy states as
active instead has almost no influence on the results (see
also~\nameref{S1_Text}).

With $\n{[m]}$ being the mRNA concentration, $r_{\n{m}}$ the transcription
rate, $\gamma_{\n{m}}$ the mRNA decay rate, and $p_{\n{active}}$ the
probability of the promoter being in the active state, we obtain
\begin{eqnarray}
\label{eq:mbasic}
\dd{\n{[m]}}{t} &=& r_{\n{m}} p_{\n{active}} - \gamma_{\n{m}} \n{[m]} \\
p_{\n{active}} &=& \frac{w_8}{\sum_{i=1}^{8} w_i} = 
\frac{\n{[P]}}{K_{\n{P}}+\n{[P]}} 
\frac{
	\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 
	e^{- \beta \Delta E_{\n{coop}}}
}{
	1 + 
	2 \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} +
	\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 e^{- \beta \Delta E_{\n{coop}}}
}
\end{eqnarray}
Here, $w_i$ stands for the thermodynamic weight of the ith state in the
order in which they are listed in Fig~\ref{fig3:states}. As written above,
the partition function factorizes into a polymerase and a XapR term because
of our assumption of independent binding, which is further discussed
in~\nameref{S1_Text}. Note that because $r_{\n{m}}$ implicitly contains the
gene copy number per cell, it has units of $\n{M^{-1} \ s^{-1}}$ and not
just $\n{s^{-1}}$. This rate equation gives the mean mRNA concentration
$\langle\n{[mRNA]}\rangle = \frac{r_{\n{m}}}{\gamma_{\n{m}}}
p_{\n{active}}$, which we will need in the next paragraph.
In~\nameref{S1_Text}, the full chemical master equation is shown, which
leads to a Poisson distribution of the mRNA concentration with the same
mean.

\paragraph*{Translation.} 
The next step in our modeling progression is translation. As a
simplification, we write $\n{[p]}=\n{[XapA]=[XapB]}$ for the general protein
concentration. This assumes that the rates of transcription, mRNA decay,
translation, and protein decay are the same for both proteins, which, as
discussed in \nameref{S1_Text}, does not have a significant influence on the
results. We write the following rate equation for the protein concentration,
where $r_{\n{p}}$ denotes the translation rate, $\gamma_{\n{p}}$ the protein
decay rate, and $\langle \n{[mRNA]} \rangle$ the mean mRNA concentration:
\begin{eqnarray}
\label{eq:pbasic}
\dd{\n{[p]}}{t} = r_{\n{p}} \langle \n{[mRNA]} \rangle - \gamma_{\n{p}} \n{[p]}.
\end{eqnarray}

\paragraph*{Xanthosine dynamics.}
We have now discussed how xanthosine activates the synthesis of XapA and
XapB through XapR. In this part of the section, we close the feedback loop
by setting up a xanthosine rate equation.

Transport of xanthosine across the cell membrane happens in two steps which
we will discuss now. All statements in this paragraph are based on data
from~\cite{Norholm2001}. In a first transport step, xanthosine passes the
outer membrane through porins like Tsx, OmpF and OmpC. This does not seem to
be a rate-limiting step. Then, XapB actively transports it across the inner
membrane, powered by the proton motive force. There are two other active
nucleoside transporters that seem to be transporting xanthosine with a very
low affinity: NupC and NupG. It was found that \textDelta\emph{nupC}
\textDelta\emph{nupG} strains cannot grow on xanthosine. Hence, these seem
to be necessary to ``start'' the system by importing small amounts of
xanthosine that then activate \emph{xapB}, and there appears to be no other
significant way in which xanthosine can enter the cell.

We assume the kinetic scheme of all three transporters to be similar to that
of \emph{lac} permease (as it is described in~\cite{Kaback2015}). One
transporter can then do two things to change the intracellular xanthosine
concentration $\n{[x]}$: it can actively transport one xanthosine into the
cell, powered by the proton gradient, or it can transport one out of the
cell, against the proton gradient. For net transport, a proton and a
substrate need to bind on one side of the membrane and detach from the
transporter on the other side of the membrane before the transporter goes
back to the first side again (see also~\nameref{S1_Text}). Because of the
proton gradient, influx is overall thermodynamically more likely than export
at low intracellular xanthosine concentrations, which leads to a net influx.
For much higher intra- than extracellular xanthosine concentrations, the
difference in the chemical potential of xanthosine across the membrane can
dominate that of the protons and there is net efflux.

We model influx and efflux separately and use Michaelis-Menten kinetics for
XapB. For the turnover rate and Michaelis constant for influx we write
$k_{\n{b, i}}$ and $K_{\n{b, i}}$, respectively, and $k_{\n{b, e}}, K_{\n{b,
e}}$ for efflux. These parameters implicitly include the proton gradient.
For NupC and NupG, the Michaelis constant is very large, due to their low
xanthosine affinity. Thus, the general $k_{\n{cat}} \n{[E]_0}
\frac{\n{[S]}}{K_{\n{M}} + \n{[S]}}$ can be approximated as
$\frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0} \n{[S]}$. In the following,
$\frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0}$ is denoted by $k_{\n{nup}}$ for
influx, and the efflux rate is written as $\xi k_{\n{nup}}$. For the XapB
kinetics, no approximations can be made, because the xanthosine
concentration in the dynamic system can range from far below the respective
$K_{\n{M}}$ value to far above. The final terms for Nup and XapB are shown
below.

After transport into the cell, XapA degrades xanthosine. We model this using
standard Michaelis-Menten kinetics, with parameters $k_{\n{a}}, K_{\n{a}}$
(corresponding to turnover rate and Michaelis constant, respectively). All
the above then leads to the xanthosine rate equation
\begin{eqnarray}
\dd{\n{[x]}}{t} = \biggl(\underbrace{k_{\n{b,i}} \frac{c}{K_{\n{b,i}} + c} - k_{\n{b,e}} \frac{\n{[x]}}{K_{\n{b,e}} + \n{[x]}}}_{\n{XapB}} - \underbrace{k_{\n{a}} \frac{\n{[x]}}{K_{\n{a}} + \n{[x]}}}_{\n{XapA}}\biggr) \n{[p]} + \underbrace{k_{\n{nup}} \left(c- \xi \n{[x]}\right)}_{\n{NupC\ \& \ NupG}}.
\end{eqnarray}
We recall that $\n{[x]}$ is the intracellular xanthosine concentration, and
$c$ denotes the extracellular one. Because $k_{\n{b,i}} > k_{\n{b,e}}$ and
$K_{\n{b,i}} < K_{\n{b,e}}$, influx dominates at low intracellular
xanthosine concentrations. At much higher intra- than extracellular
xanthosine concentrations, the efflux term takes over. More details on the
aforementioned steps and a discussion of passive diffusion can be found
in~\nameref{S1_Text}.

\subsection*{Nondimensionalization}
We have now formulated the behavior of the system in terms of the rate
equations for mRNA, protein, and xanthosine. These equations can be
nondimensionalized, which reduces the dimension of parameter space. We
measure time in units of $\gamma_{\n{p}}^{-1}$ and concentrations in units
of $K_{\n{a}}$ (except XapR, where the equations make it more natural to use
$K_{\n{XapR}}$). In Table~\ref{table1:nondim}, all the nondimensional
parameters and their definitions are listed. Furthermore, we define
$\n{[m]_a} \defeq \frac{\n{[m]}}{K_{\n{a}}}$, $\n{[p]_a} \defeq
\frac{\n{[p]}}{K_{\n{a}}}$, $\n{[x]_a} \defeq \frac{\n{[x]}}{K_{\n{a}}}$,
and $\tau \defeq \gamma_{\n{p}} t$. Using these definitions, the following
equations are obtained
\begin{alignat}{2}
\dd{\mathrm{[m]_a}}{\tau} &= \ &&
\rho_{\n{m}} 
\frac{
	\mathrm{[XapR]_{R,A}}^2 
	e^{- \Delta \epsilon_{\text{coop}}}
}{
	1 + 
	2 \mathrm{[XapR]_{R,A}} +
	\mathrm{[XapR]_{R,A}}^2 e^{- \Delta \epsilon_{\text{coop}}}
}
- \gamma_{\n{mp}} \mathrm{[m]_a}
\\
\dd{\mathrm{[p]_a}}{\tau} &= \ &&
\rho_{\n{p}} \mathrm{[m]_a}
- \mathrm{[p]_a}
\\
\dd{\mathrm{[x]_a}}{\tau} &= \ && \left(k_{\n{\beta,i}} \frac{\n{[c]_a}}{K_{\n{\beta,i}} + \n{[c]_a}} - k_{\n{\beta,e}} \frac{\n{[x]_a}}{K_{\n{\beta,e}} + \n{[x]_a}} - k_{\n{\alpha}} \frac{\n{[x]_a}}{1 + \n{[x]_a}}\right) \mathrm{[p]_a} \\ & && + k_{\n{\eta}} \left(\mathrm{[c]_a} - \xi \mathrm{[x]_a}\right)
\nonumber \\
\text{with} & && \raggedleft [\n{XapR}]_{\mathrm{R,\n{A}}} = [\n{XapR}]_{\mathrm{R}} \frac{\left(1 + \frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}}\right)^2}{\left(1 + \frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}}\right)^2 + e^{\Delta \epsilon_{\n{x}}} \left(1+\frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}} \frac{1}{K_{\n{IA}}}\right)^2} \nonumber
\end{alignat}

Very little is known about the \emph{xap} system, and thus, there are no
measured values for the free parameters. Nevertheless, we were able to
estimate a reasonable range by using values from similar, well studied
systems and by finding physical constraints or relations between parameters.
The results of these estimates are shown in Table~\ref{table1:nondim}. They
are based on a choice of $\gamma_{\n{p}} = 5 \cdot 10^{-4} \unit{s^{-1}}$
and $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$. A detailed derivation can be
found in~\nameref{S1_Text}.

\begin{table}%[h!]
	\centering
	\caption{
		{\bf Nondimensional parameters and their estimated values.}}
	\begin{tabular}{rllr}
		\thickhline
		Param. & Definition & Estimated range & Value used \\
		\hline
		$\rho_{\n{m}}$ & $\defeq \frac{r_{\n{m}}}{\gamma_{\n{p}} K_{\n{a}}} \frac{\n{[P]}}{K_{\n{P}}+\n{[P]}}$ & $\approx 10^{-3 \pm 2}$ & $10^{-3}$ \\
		$\gamma_\n{{mp}}$ & $\defeq \frac{\gamma_{\n{m}}}{\gamma_{\n{p}}}$ & $\approx 10^{1 \pm 0.5}$ & $10^{1}$ \\
		$\rho_{\n{p}}$ & $\defeq \frac{r_{\n{p}}}{\gamma_{\n{p}}}$ & $\approx 10^{2 \pm 0.5}$ & $10^{2}$ \\
		$\n{[XapR]_R}$ & $\defeq\frac{\n{[XapR]_{\n{tot}}}}{K_{\n{XapR}}}$ & $\approx 10^{0 \pm 2}$ & $10^0$ \\
		$\n{[c]_{a}}$ & $\defeq \frac{c}{K_{\n{a}}}$ & ($\in [ 0, 10^3 ]$) & $13$ \\
		$k_{\n{\beta,i}}$ & $\defeq\frac{k_{\n{b,i}}}{\gamma_{\n{p}}}$ & $\approx 10^{4 \pm 1}$ & $5 \cdot 10^4$ \\
		$k_{\n{\beta,e}}$ & $\defeq\frac{k_{\n{b,e}}}{\gamma_{\n{p}}}$ & $\approx 10^{3 \pm 2}$ & $10^3$ \\
		$k_{\n{\alpha}}$ & $\defeq\frac{k_{\n{a}}}{\gamma_{\n{p}}}$ & $\approx 10^{2 \pm 0.8}$ & $10^2$ \\
		$k_{\n{\eta}}$ & $\defeq\frac{k_{\n{nup}}}{\gamma_{\n{p}}}$ & $\approx 10^{0 \pm 3}$ & $5 \cdot 10^{-1}$ \\
		$\xi$ & $=\xi$ & $\approx 0.8 \pm 0.1$ & $0.8$ \\
		$K_{\n{\beta,i}}$ & $\defeq\frac{K_{\n{b,i}}}{K_{\n{a}}}$ & $\approx 10^{1 \pm 2}$ & $10^1$ \\
		$K_{\n{\beta,e}}$ & $\defeq\frac{K_{\n{b,e}}}{K_{\n{a}}}$ & $\approx 10^{2 \pm 2}$ & $10^2$ \\
		$K_{\n{\chi A}}$ & $\defeq\frac{K_{\n{xA}}}{K_{\n{a}}}$ & $\approx 10^{2 \pm 1} \cdot 10^{\Delta \epsilon_{\n{x}} - 5}$ & $10^2$ \\
		$K_{\n{IA}}$ & $\defeq\frac{K_{\n{xI}}}{K_{\n{xA}}}$ & $\approx 10^{2 \pm 1}$ & $10^2$  \\
		$\Delta \epsilon_{\n{x}}$ & $\defeq\beta \Delta E_{\n{x}}$ & $\approx 2$ to $2 \left( \ln\left(K_{\n{IA}}\right) - 1\right) < 12$ & $5$ \\
		$\Delta \epsilon_{\n{coop}}$ & $\defeq \beta \Delta E_{\n{coop}}$ & $\approx 0 - 10$ & $5$ \\
		\thickhline
	\end{tabular}
	\begin{flushleft} 
		The left column shows all nondimensional parameters that appear in
		the final equations. In the middle are their definition and
		estimated values. They are based on $\gamma_{\n{p}} = 5 \cdot
		10^{-4} \unit{s^{-1}}$ and $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$.
		Note that the range of the three MWC parameters depends on each
		other, but they can still be chosen independently. The range given
		for $\n{[c]_{a}}$ denotes the estimated ``interesting'' range in
		which switching happens, but $\n{[c]_{a}}$ can of course exceed
		these values. Details on the parameters and their estimation can be
		found in~\nameref{S1_Text}. Finally, the last column shows the value
		that we use for the rest of this paper, unless otherwise noted. An
		explanation of this choice will follow in the next section.
	\end{flushleft}
	\label{table1:nondim}
\end{table}

\section*{Results and discussion}
In the modeling process in the previous section, we have obtained three
coupled differential equations. In this section, we will analyze these
equations with deterministic methods and stochastic simulations. Analytical
closed-form solutions could not be obtained and would, if they existed,
probably not be helpful due to their large complexity. Finding such
solutions requires solving a fifth order algebraic equation.

\subsection*{Deterministic phase portraits}
A standard way to analyze dynamical systems like ours deterministically is
to plot phase portraits. In the following, we present such plots where the
state variables are the mRNA, the protein, and the xanthosine concentration.

\paragraph*{From a 3D to a 2D system.} 
Fig~\subref{fig4:bistability}{A} shows the 3D phase portrait for a
representative set of parameters (that from Table~\ref{table1:nondim}),
whose choice is explained below. The plot looks rather complicated at first
but can be understood intuitively. The three planes are the nullclines and
the gray lines point in the direction in which the dynamical system moves at
each point. The planes intersect in three points, which are the steady-state
solutions of the dynamical system. For this choice of parameters, the system
first flows towards the mRNA nullcline, then it moves along that plane to
the intersection with the protein nullcline, and lastly, it moves along that
intersection line to one of the three intersection points of all three
planes.

\begin{figure}%[h!]
	\centering
	\includegraphics[width=1\textwidth]{media/Bistability.eps}
	\caption{{\bf Phase portraits showing bistability.}
		3D and 2D phase portraits for one set of parameters that leads to
		bistability. The parameter values are listed in
		Table~\ref{table1:nondim}. Note that all the concentrations
		($\n{[m]_a}, \n{[p]_a}, \n{[x]_a}$) are measured in units of
		$K_{\n{a}} = 5 \cdot 10^{4} \unit{nM}$. The planes in (A) and the
		lines in (B) are the nullclines of the state variables, and their
		intersection points, marked in red in (B), are the steady-state
		solutions of the system. The region shaded in gray in (B) leads to
		negative concentrations and is unphysical. A vector plot of (B) that
		also shows the magnitude of flow at each point can be found
		in~\nameref{S1_Text}.}
	\label{fig4:bistability}
\end{figure}
%\todo{edge,size}

It is important to point out that for a different set of parameters, the
dynamics can be quite different. There are, for example, scenarios where the
xanthosine kinetics are roughly as fast as the mRNA kinetics and the
dynamics unfolds in two steps: first to the intersection of the mRNA and the
xanthosine nullcline, then along that line to the protein nullcline and
thereby to a fixed point.

A usual simplification with genetic circuits like this is to assume the mRNA
concentration to be at steady-state, i.e. to write $\dd{\n{[m]_a}}{\tau}=0$
and solve this for $\n{[m]_a}(\n{[p]_a},\n{[x]_a})$ to simplify the 3D to a
2D system. This restricts the dynamics to the green plane in our plot, which
is reasonable here because as explained above, the system first flows
towards that plane before either the protein or the xanthosine concentration
changes significantly. However, as already pointed out, this is different
for other parameter values, and thus, this assumption generally is not a
good one. If the xanthosine dynamics are faster than the mRNA dynamics, the
system first flows towards the xanthosine nullcline. In that case, forcing
it onto the mRNA nullcline leads to significant changes in the dynamics.

Nevertheless, the steady-state solutions and the qualitative features that
we address in this paper remain the same. Because the 3D plots are rather
hard to read, we will, in the following, make the compromise to show a 2D
version of the phase portraits but ensure that all of our statements also
hold true in 3D space. As explained above, it makes the most sense here to
do this by setting $\dd{\n{[m]_a}}{\tau}=0$. The resulting equations can be
found in~\nameref{S1_Text}. In particular, we define $\rho \defeq
\frac{\rho_{\n{m}} \rho_{\n{p}}}{\gamma_{\n{mp}}}$ for everything that
follows.

\paragraph*{Bistability.} 
From the 2D plot in Fig~\subref{fig4:bistability}{B}, it can clearly be seen
that for the chosen parameters, there are three steady-state solutions.
Because the 2D plane that we are in now is the mRNA nullcline, these
solutions are the same fixed points as in the 3D plot
($\dd{\n{[m]_a}}{\tau}=0$ on the nullcline and $\dd{\n{[p]_a}}{\tau}=0$,
$\dd{\n{[x]_a}}{\tau}=0$ for the 2D fixed points). One can see from the
vector field that the two outer points (labeled 1 and 3) are stable and the
middle one (labeled 2) is unstable and serves as a sort of ``switch-point''
between the other two. This means that there are two stable states the cell
can be in, one at high (point 3) and one at low (point 1) expression. As a
result, there is bistability and the distribution of expression among cells
is bimodal.

This corresponds to the experimental observations, so the model passes this
sanity check. Furthermore, the xanthosine and protein concentrations at the
upper fixed point have the expected order of magnitude: the xanthosine
concentration is roughly $10-100 \unit{nM}$, and there are roughly $500$
proteins, which is just a bit lower than what was measured for the number of
Nup transporters~\cite{Li2014} which fulfill a similar purpose. We do not
have well founded expectations for the other fixed points, so no comparison
can be made here. Nevertheless, the orders of magnitude at the lower fixed
point -- roughly $1-10 \unit{nM}$ of xanthosine and around $5$ proteins --
seem perfectly reasonable. Note that $\n{[x]_a} \approx \n{[c]_a}$ at the
lower fixed point because there is only weak accumulation due to Nup and a
few XapB transporters.

As already mentioned, we are working with one specific set of parameters
here and we will now explain this choice of values. Firstly, they were
picked such that they are well within their estimated range. Secondly, we
wanted clear bistability to occur in the phase portraits as well as in the
stochastic simulations (see later). Thirdly, by the corresponding choice of
parameters it was ensured that the mRNA number per cell at the
``switch-point'' is around 1: this is large enough to enable the system to
clearly resolve the two stable fixed points (as we will see from the
stochastic simulations later on), but is low enough to lead to mean mRNA
numbers that are very reasonable (see~\cite{Milo2016} for the average mRNA
numbers in cells, which are rather low). The protein and xanthosine
concentrations followed from this, but with some variation in the parameters
they could still be tuned to a certain extent.

As a remark we point out that we have not observed any oscillations in the
system. Intuitively, they might be expected when the XapA rate is
significantly larger than the XapB rate, but it turns out that oscillations
cannot be obtained. This can be understood when looking at the regions that
are bounded by all three nullclines: within these regions, the streamlines
on the nullclines point into the regions, so deterministically, the system
is trapped in there and can only flow towards the fixed points.

For a different set of parameters, the orders of magnitude in the plots and
even the qualitative behavior can change. In the following, we will discuss
some interesting features of the system that can be observed through the
phase portraits.

\paragraph*{The extracellular xanthosine concentration.}
The parameter that is the experimentally most easily tunable and
biologically the most relevant is the extracellular xanthosine
concentration. When it is increased in the experiment, the cells go from all
being in the low expression state to being in either the low or the high
expression state (all-or-none phenomenon) to all being in the high
expression state~\cite{Chure2019}. If our model is correct, it should
exhibit the same qualitative behavior. Indeed we find exactly this: as can
be seen in Fig~\ref{fig5:extraxanth}, increasing $\n{[c]_a}$ makes the high
stable fixed point appear and then, for even higher $\n{[c]_a}$, the lower
one disappear. Thus, for low $\n{[c]_a}$ the only stable point of the system
is at low expression, and for high $\n{[c]_a}$ there is only high
expression. In between, there are two deterministically stable expression
levels.

\begin{figure}%[h!]
	\centering
	\includegraphics[width=1\textwidth]{media/VaryingC.eps}
	\caption{{\bf Phase portraits for different extracellular xanthosine concentrations.}
		All parameters but $\n{[c]_a}$ are as presented in
		Table~\ref{table1:nondim}. The extracellular xanthosine
		concentration in these plots is $\n{[c]_a}=10^{1}$ in (A) and
		$\n{[c]_a}= 5 \cdot 10^{2}$ in (B) (recall $\n{[c]_{a}} \defeq
		\frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$).
		Tuning $\n{[c]_a}$ moves the orange line (xanthosine nullcline) up
		or down, the blue line (mRNA nullcline) is unchanged (see
		also~\nameref{S1_Text}). It can clearly be seen that in (A) there is
		only the lower fixed point (fixed point number 1), whereas in (B)
		there is only the upper one (fixed point number 3). In between lies
		the bistable case that was shown in Fig~\ref{fig4:bistability}.}
	\label{fig5:extraxanth}
\end{figure}
%\todo{size of plot}

Furthermore, by setting $\n{[c]_a}=0$ (not shown here) we found that in the
absence of xanthosine, there are roughly 2-3 proteins, which agrees very
well with measurements, where around 2 proteins per cell were
found~\cite{Li2014}. In addition, the parameter $K_{\n{\chi A}}$
(dissociation constant of xanthosine from active XapR) can be tuned such
that the extracellular xanthosine concentration $\n{[c]_a}$ in the
switching-regime is similar to that in the experiment. It was found that the
cell only adapts at very high xanthosine concentrations of almost a
millimolar~\cite{Norholm2001} which is not completely unexpected when
recalling that for \emph{lac}, cells also limit themselves to glucose as
long as possible. Interestingly, because there is no parameter other than
$K_{\n{\chi A}}$ that tunes the critical value of $\n{[c]_a}$, this tells us
that $K_{\n{\chi A}}$ is large as argued in the estimation of $K_{\n{\chi
A}}$ in~\nameref{S1_Text}. Thus, the interaction between xanthosine and XapR
should be weak.

\paragraph*{The roles of XapA and XapB.}
While it is clear that the bistability in the model system is due to the
feedback loop from XapA and XapB, it is not obvious if both XapA and XapB
are necessary. It turns out that the bistability is due to XapB only. XapA
is neither sufficient nor necessary and, within the estimated parameter
regime, does not even have a significant influence on the system. This can
be seen from the plots in figure~\ref{fig6:xapAB}. XapA lowers the upper
solution by a little and could, in principle, thereby make the
high-expression solution vanish. For our choice of all other parameters,
this requires $k_{\n{\alpha}} > 10^4$ which is far from what has been
measured. However, a higher rate could, in principle, be achieved by
different translation rates of XapA and XapB (see simplifications of the
model in~\nameref{S1_Text}). Hence, we cannot exclude the possibility that
XapA gets so strong that it makes bistability impossible, but this is an
extreme case. XapB, on the other hand, is essential; without it the system
only has the one fixed point at low expression.

\begin{figure}%[h!]
	\includegraphics[width=1\textwidth]{media/XapAB.eps}
	\centering
	\caption{{\bf Phase portraits without XapA or XapB.}
		All parameters are as presented in Table~\ref{table1:nondim}. In
		(A), the XapA term was removed from the kinetic equations. In (B),
		the equations lack the two terms from XapB. These plots clearly show
		that XapA has almost no influence on the qualitative behavior of the
		system (i.e. bistability and the order of magnitudes), but XapB is
		the essential feature for bistability.}
	\label{fig6:xapAB}
\end{figure}
%\todo{size of plot}

For a cell, the above is a useful feature: by coupling XapA and XapB on an
operon, XapA is switched on and off together with XapB but it does not
significantly disturb this adaptation mechanism, while its kinetic
parameters can be chosen somewhat freely as necessary for metabolism. By
having a membrane transporter gene on an operon whose expression is
activated by the transporter substrate, the expression of a whole set of
enzymes can be turned on and off depending on the presence of the substrate.
It seems likely that this way of short-term adaptation of a single cell to
its environment may be used by cells for many metabolic processes.

\paragraph*{The role of cooperativity.} 
The model has two cooperatively interacting binding sites for XapR on the
\emph{xapAB} promoter and two cooperative binding sites on XapR for
xanthosine. An interesting question to pose is whether the cooperativity is
a necessary feature for bistability. This is motivated by the importance of
cooperativity in typical ``genetic switches''\cite{Gardner2000,Cherry2000}.

If, as a purely theoretical consideration, we remove either the second
xanthosine binding site on XapR or the second XapR binding site on the
promoter, we find that the system still has a bistable parameter regime.
However, this is only a rather narrow parameter range which makes the system
less stable. When the second binding site is removed in both places, the
equations are insufficiently non-linear to produce bistability. An example
of these three scenarios can be seen in Fig~\ref{fig7:coop}. It follows that
there need to be either two xanthosine binding sites on XapR or two XapR
binding sites on the promoter (or both) in order to obtain a switch-like
behavior.

\begin{figure}%[h!]
	\centering
	\includegraphics[width=1\textwidth]{media/FewerSites.eps}
	\caption{{\bf Phase portraits for less or no cooperativity.}
		Most parameters are as presented in Table~\ref{table1:nondim},
		changes are mentioned below. Fixed points are marked in red. In (A),
		there is only one xanthosine binding site on XapR and everything
		unchanged for the XapR-promoter binding. Two parameters are changed:
		$\rho = 0.07$ and $\n{[c]_a} = 6$. This is necessary to compensate
		for the weaker induction such that the system is bistable. In (B),
		there is only one XapR binding site on the promoter and everything
		is unchanged for the xanthosine-XapR binding. Two parameters are
		changed: $\rho = 0.13$ and $\n{[c]_a} = 3$. In (C), there is only
		one xanthosine binding site on XapR and also only one XapR binding
		site on the the promoter. Two parameters are changed: $\rho = 0.1$
		and $\n{[XapR]_R} = 5$. Whereas bistability is retained in (A) and
		(B), it cannot be obtained anymore in (C).}
	\label{fig7:coop}
\end{figure}
%\todo{size of plot}

One can also ask how much cooperative interaction is needed between the two
binding sites. For the promoter, this is given by $\Delta E_{\n{coop}}$ in
our model, and we find that setting $\Delta E_{\n{coop}} = 0$ has almost no
influence on the phase diagrams. For XapR, the two binding sites interact
indirectly, because the active state is much likelier if two xanthosine
molecules are bound. This cooperative interaction can, however, not be tuned
like in the case of the promoter with $\Delta E_{\n{coop}}$.

Note that we are not writing Hill equations and measuring cooperativity in
terms of the Hill coefficient. If Hill equations were to be used for the
modeling, the Hill coefficient could have values between 1 and 2, which
would yield bistability for large enough values, but not for lower ones.
This could be investigated more rigorously similar to the analysis of a
simple genetic switch in~\cite{Cherry2000}. However, we refrain from looking
for a minimal Hill coefficient in our system, because we do not find this
very insightful. Hill equations only describe scenarios with multiple,
non-interacting binding sites where any other state but the fully occupied
and active or the fully empty and inactive state is not significantly
populated. In that sense, the Hill coefficient simply denotes the number of
binding sites and does not reflect how cooperative these binding sites are.
We suggest that cooperativity should be explored more in-depth and a more
rigorous analysis of the role of cooperativity in simple genetic switches
should be done before returning to more complex systems like this one.

\subsection*{Stochastic simulations}
Stochastic simulations of the full 3-dimensional system of mRNA, protein,
and xanthosine were run for comparison with the deterministic results.
In~\nameref{S1_Text}, we present the underlying chemical master equation of
the system. Because of the two different fixed points at low and high
expression, the copy numbers in the problem vary from very low to very high.
For large amounts of mRNA, protein, and xanthosine, the number of reactions
is large too, and thus, Gillespie's classical algorithm has a large
computational cost and his \texttau-leap algorithm would be ideal. On the
other hand, \texttau-leaping cannot be used for the small copy numbers. For
these reasons, we chose to work with the algorithm described
in~\cite{Cao2006}, a hybrid form between Gillespie's classical and his
\texttau-leap algorithm. We gratefully worked with the Python implementation
of this algorithm in \emph{StochPy}, version 2.3~\cite{Maarleveld2015}.

\paragraph*{Bimodality and the extracellular xanthosine concentration.}
This approach results in the same bimodal distributions that were already
seen in the deterministic investigation and experimental studies.
Fig~\ref{fig8:stochC} shows the distribution of protein expression found in
the simulations for different values of the extracellular xanthosine
concentration. The parameters that were used are the same as in the previous
section (listed in Table~\ref{table1:nondim}). To obtain the distributions,
we ran the simulation 5000 times for a simulated time of $10^6 \unit{s}$
each and started at a mRNA, protein and intracellular xanthosine count of 0.

\begin{figure}%[h!]
	\centering
	\includegraphics[width=1\textwidth]{media/HistogramsC.eps}
	\caption{{\bf Distributions from stochastic simulations and the corresponding phase portraits.}
		Apart from $\n{[c]_a}$, the parameters are the same as in
		Table~\ref{table1:nondim}. For the distributions, the simulations
		were run 5000 times for $10^6 \unit{s}$ each (simulated time) and
		started at a mRNA, protein and intracellular xanthosine count of 0.
		We show the two cases of unimodality (low expression in (A) and high
		expression in (C)) as well as the case of bimodality in (B). The
		values of $\n{[c]_a}$ are 12 in (A), 18.5 in (B), and 25 in (C)
		(recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5
		\cdot 10^{-5} \unit{M}$). The output from the stochastic simulations
		is in good agreement with the concentrations at the fixed points in
		the deterministic phase portraits.}
	\label{fig8:stochC}
\end{figure}

The results agree very well with the deterministic fixed points and
experiments: the mean numbers of mRNA, protein, and xanthosine in the
stochastic results are as predicted from the phase portraits. It does,
however, become clear that the phase portraits do not tell whether the cells
will actually populate both the high and the low expression state: as can be
seen in Fig~\subref{fig8:stochC}{(A)}, the cells did not switch to the high
expression state for the waiting time in our simulations, even though one
would naively expect this from the phase portrait.

We found that the two lower fixed points (marked as 1 and 2 in
Fig~\ref{fig4:bistability}) need to be very close like in
Fig~\subref{fig8:stochC}{(B)} to give bimodality. For lower $\n{[c]_a}$,
meaning larger distance between the first and second fixed point, no or
almost no switching was observed. Of course, this is also a matter of the
waiting time and stochastic effects: if one waits for long enough, switching
should eventually occur. However, switching times of more than several hours
are not at the center of this investigation and would mean that switching is
extremely unlikely. There are two aspects that become relevant in this
context that we neglect in our analysis but shortly mention here:
transcription and translation bursts lead to higher stochasticity and cell
division leads to some discontinuity in the process.

Note that while the deterministic analysis assumes the variables to be
continuous, the simulations work with discrete numbers of mRNA, protein, and
xanthosine. This per se is no problem, because the deterministic analysis
describes the mean values and the simulation fluctuates around this mean.
If, however, the mRNA number at the third, high fixed point is too low, the
system will not be able to resolve the two points anymore. This becomes a
problem when the distance between the first and the third fixed point
becomes as low as the stochastic fluctuations in the system, which is around
3 mRNA molecules. In Fig~\ref{fig8:stochC}, the mRNA number at the low fixed
point was fluctuating between 0 and 2 with a mean significantly smaller than
1, and at the high fixed point it was varying between 2 and 6.

\paragraph*{Time evolution and switching times.}
Fig~\ref{fig:StochT} shows the time evolution of one specific example.
Again, the simulation was started with a mRNA, protein and intracellular
xanthosine count of 0 and was run for a simulated time of $5 \cdot 10^5
\unit{s}$. In this specific example, switching occurred after $10^5
\unit{s}$.

Comparing this to the experimentally expected timescales is difficult,
because the switching time strongly depends on the extracellular xanthosine
concentration. Experiments were always stopped after a few hours, and in
this time, the cell population might not reach its steady-state expression
distribution. Hence, the distribution could be bimodal when the experiment
is stopped but become unimodal after further waiting. That way,
extracellular xanthosine concentrations that are too high for deterministic
bistability could lead to experimental bimodality if the experiment is
stopped too early. In this case, the observed switching time would be
shorter, which makes the comparison to our simulations even harder. Thus, we
cannot say if it is problematic that the $10^5 \unit{s}$ is larger than what
was found in the experiment.

Nevertheless, we do warn the reader that the timescales in the simulations
should be taken with reservations. Cell divisions are not considered here,
and neither is the burstiness of transcription and translation. This means
that stochasticity may be larger in the real system which should have an
influence on the timescales and probably shorten the time until switching
occurs.

\begin{figure}%[h!]
	\centering
	\includegraphics[width=0.7\textwidth]{media/TimeEvolution.eps}
	\caption{{\bf Time evolution of protein (XapA/XapB) from one run of the simulation.}
		The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
		at an mRNA, protein and intracellular xanthosine count of 0. In this
		case the system switched into the high-expression state after $10^5
		\unit{s}$. The parameters that were used are the same as in
		Table~\ref{table1:nondim}, the only exception being the
		extracellular xanthosine concentration, which was chosen to be
		$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$) just as in
		Fig~\subref{fig8:stochC}{(B)}.}
	\label{fig:StochT}
\end{figure}
%\todo{size of plot}

\paragraph*{Hysteresis.} 
In the previous two paragraphs, the simulations were started at initial
intracellular concentrations of 0 to investigate what happens if xanthosine
is suddenly added to the cell's environment. We can now ask the opposite
question: what happens when xanthosine is removed from the extracellular
environment? To answer this, the simulation was started with initially fully
induced cells, i.e. at the mRNA, protein and intracellular xanthosine counts
of the high fixed point in the corresponding phase portrait.

The resulting distributions can be found in~\nameref{S1_Text}. They show
clear hysteresis effects: at concentrations where all cells remained in the
low expression state when being started there before, all cells now remain
at high expression. Only when the second and the third fixed point are very
close does switching to the uninduced state happen. Interestingly, this
behavior is symmetric to the ``switching on'' in the previous paragraphs,
where the lower two fixed points needed to be about as close as the upper
two need to be now for ``switching off''. The extracellular xanthosine
concentration (for the other parameters being as in
Table~\ref{table1:nondim}) needed to be as low as $\n{[c]_a} = 10$ to
observe bimodality and $\n{[c]_a}=9$ to have all cells in the low expression
state (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5 \cdot
10^{-5} \unit{M}$). \todo{Is this understandable without seeing the data? Do
you have any suggestions to make it clearer?}

We have learned that cells only change their metabolism to xanthosine if
enough of the latter is around, but after they have switched, this metabolic
state is stable even if the xanthosine concentration decreases to a certain
extent. This explains what has been observed by Novick and
Weiner~\cite{Novick1957} for the \emph{lac} operon: when induced cells were
transferred into lower concentrations of lactose, they remained induced,
even though uninduced cells could not become induced at these
concentrations.

\section*{Conclusion}
\paragraph*{A genetic switch for metabolic adaptation with remarkable features.}
In this paper, we propose a model for biological pathway systems containing
a membrane transporter whose gene expression is, directly or indirectly,
activated by its substrate. We have shown that such a system can be bistable
and thus work as a genetic switch which reacts to the extracellular
concentration of the relevant metabolite. This switch has very useful
biological features: first, coupling of the transporter with, for example, a
degradation enzyme creates a genetic switch that enables short-term
adaptation of the cell's metabolism to its environment. Second, the switch
is stabilized by hysteresis effects when the extracellular xanthosine
concentration decreases, which explains previous experimental findings.

\paragraph*{Required properties for this system.}
We have found that no bistability can emerge from the genetic circuit unless
at least one component has two binding sites for its activator. Additional
binding sites or cooperativity seem to increase the stability of the switch.
In addition, simply knowing the experimental switching concentration of
xanthosine let us, for example, infer the approximate value of the
dissociation constant between the transcription factor XapR and the inducer
xanthosine. We found that their interaction needs to be rather weak.

\paragraph*{Possibilities for further investigation.}
Phase diagrams, showing for which parameters the system is bistable and for
which there is only the lower or the upper stable fixed point, could be
calculated from arguments made in~\cite{Cherry2000}. However, the
simulations have shown that deterministic bistability does not mean that
bimodality occurs, which is why we have refrained from showing such
diagrams. Furthermore, the timescales in the problem could be investigated
more thoroughly, for example the dependence of the switching time on
$\n{[c]_a}$. Having said that, the simulation for such an analysis should
account for the burstiness of transcription and translation as well as for
cell divisions, which is anything but straightforward. Lastly, some rigorous
work could be done to identify the most and least relevant parameters of the
system to create a simpler, minimal model that captures the most relevant
features.

\paragraph*{General lessons learned for the modeling of other systems.}
We point out again that assuming the mRNA concentration to be at
steady-state can have a significant impact on the apparent dynamics of a
system, as we have seen in the 3-dimensional phase portraits. Moreover, we
remind the reader that deterministic phase portraits do not show which
states are actually populated, and thus hide essential effects like
hysteresis. Lastly, we have exemplified issues with the common picture of
measuring cooperativity in terms of the Hill coefficient: the Hill
coefficient simply denotes the number of binding sites and does not reflect
how cooperative these binding sites are.

\paragraph*{A successful model.}
All model parameters could be estimated fairly well despite the lack of deep
experimental knowledge about the model system. In the results, every
variable and parameter has its expected order of magnitude, which suggests
that the model captures all relevant features and is able to describe the
dynamics of the \emph{xap} system. With the framework given in this text, it
should be straightforward to model other promoters, regulatory pathways or
enzymes and thereby adapt the model to other genes and metabolites. Examples
include \emph{lac}, \emph{ara}, and \emph{xyl}, but we suspect that many if
not most metabolic processes involve the adaptation mechanism that we have
investigated here, and that a lot can be understood about them through our
model. This apparent success demonstrates once more that even for broadly
unknown systems, rigorous physical modeling can potentially offer an
efficient way to gain a very thorough understanding of the behavior of the
system.

\todo{Is the conclusion too long?}

\section*{Supporting information}
\paragraph*{S1 Text.}
\label{S1_Text}
{\bf The aforementioned further information.} Discussion of simplifications
in the model, parameter estimation, elaborations on the results, and the
chemical master equation of this circuit.

\section*{Acknowledgments}
We thank Griffin Chure, Charlotte Strandkvist, and Kyle Naughton for
providing their data on the \emph{xap} system and pointing out the
bistability. We also thank Griffin Chure for answering any biological
questions that arose, Nathan Belliveau for bringing similar systems to our
attention, and Jane Kondev and Jin Wang for interesting discussions. Plots
were done in \emph{Mathematica} using the packages MaTeX and
CurvesGraphics6. K.S.L. acknowledges financial support from the
Werner~Siemens-Foundation through the Swiss Study Foundation.
\todo[inline]{I don't really know what/who is usually listed here... Did I list too many people (I don't know at what ``extent'' of help people are usually named here) or did I forget anyone? Should I list everyone who worked on the xap system last summer and provided the data even though they are already cited, but only as Griffin et al.? Is it enough to list packages like this or should I try to find some website to cite? Should I list Python \& StochPy again it's already mentioned in the text? SHould I thank someone for the providing the elements for the figures and whom? Is there funding that should be listed?}

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.

\begin{thebibliography}{10}

	\bibitem{Jacob1961}
	Jacob F, Monod J.
	\newblock Genetic regulatory mechanisms in the synthesis of proteins.
	\newblock Journal of Molecular Biology. 1961;3(3):318--356.

	\bibitem{Gardner2000}
	Gardner TS, Cantor CR, Collins JJ.
	\newblock Construction of a genetic toggle switch in \emph{Escherichia coli}.
	\newblock Nature. 2000;403:339.

	\bibitem{Wolf1998}
	Wolf DM, Eeckman FH.
	\newblock On the Relationship between Genomic Regulatory Element Organization
	and Gene Regulatory Dynamics.
	\newblock Journal of Theoretical Biology. 1998;195(2):167 -- 186.
	\newblock doi:{https://doi.org/10.1006/jtbi.1998.0790}.

	\bibitem{Novick1957}
	Novick A, Weiner M.
	\newblock Enzyme induction as an all-or-none phenomenon.
	\newblock Proceedings of the National Academy of Sciences of the United States
	of America. 1957;43(16590055):553--566.

	\bibitem{Narang2008}
	Narang A, Pilyugin SS.
	\newblock Bistability of the lac Operon During Growth of Escherichia coli on
	Lactose and Lactose + Glucose.
	\newblock Bulletin of Mathematical Biology. 2008;70(4):1032--1064.

	\bibitem{Choi2008}
	Choi PJ, Cai L, Frieda K, Xie XS.
	\newblock A Stochastic Single-Molecule Event Triggers Phenotype Switching of a
	Bacterial Cell.
	\newblock Science. 2008;322(5900):442--446.
	\newblock doi:{10.1126/science.1161427}.

	\bibitem{Ozbudak2004}
	Ozbudak EM, Thattai M, Lim HN, Shraiman BI, van Oudenaarden A.
	\newblock Multistability in the lactose utilization network of
	\emph{Escherichia coli}.
	\newblock Nature. 2004;427:737.

	\bibitem{Fritz2014}
	Fritz G, Megerle JA, Westermayer SA, Brick D, Heermann R, Jung K, et~al.
	\newblock Single Cell Kinetics of Phenotypic Switching in the Arabinose
	Utilization System of E. coli.
	\newblock PLOS ONE. 2014;9(2):e89532.
	\newblock doi:{10.1371/journal.pone.0089532}.

	\bibitem{Jenkins2017}
	Jenkins A, Macauley M.
	\newblock Bistability and Asynchrony in a Boolean Model of the l-arabinose
	Operon in Escherichia coli.
	\newblock Bulletin of Mathematical Biology. 2017;79(8):1778--1795.

	\bibitem{Siegele1997}
	Siegele DA, Hu JC.
	\newblock Gene expression from plasmids containing the araBAD promoter at
	subsaturating inducer concentrations represents mixed populations.
	\newblock Proceedings of the National Academy of Sciences of the United States
	of America. 1997;94(9223333):8168--8172.

	\bibitem{Bae2010}
	Bae JS, Kim TH, Kim MY, Park JM, Ahn YH.
	\newblock Transcriptional Regulation of Glucose Sensors in Pancreatic
	beta-Cells and Liver: An Update.
	\newblock Sensors. 2010;10(5):5031–5053.
	\newblock doi:{10.3390/s100505031}.

	\bibitem{Tiedge1991}
	Tiedge M, Lenzen S.
	\newblock Regulation of glucokinase and GLUT-2 glucose-transporter gene
	expression in pancreatic beta-cells.
	\newblock The Biochemical journal. 1991;279 (Pt 3)(1953686):899--901.

	\bibitem{Buxton1980}
	Buxton RS, Hammer-Jespersen K, Valentin-Hansen P.
	\newblock A second purine nucleoside phosphorylase in \emph{Escherichia coli}
	K-12.
	\newblock Molecular and General Genetics MGG. 1980;179(2):331--340.

	\bibitem{Hammer-Jespersen1980}
	Hammer-Jespersen K, Buxton RS, Hansen TD.
	\newblock A second purine nucleoside phosphorylase in \emph{Escherichia coli}
	K-12.
	\newblock Molecular and General Genetics MGG. 1980;179(2):341--348.

	\bibitem{Seeger1995}
	Seeger C, Poulsen C, Dandanell G.
	\newblock Identification and characterization of genes (xapA, xapB, and xapR)
	involved in xanthosine catabolism in \emph{Escherichia coli}.
	\newblock Journal of bacteriology. 1995;177(7559336):5506--5516.

	\bibitem{Joergensen1999}
	Jorgensen C, Dandanell G.
	\newblock Isolation and characterization of mutations in the \emph{Escherichia
		coli} regulatory protein XapR.
	\newblock Journal of bacteriology. 1999;181(10400599):4397--4403.

	\bibitem{Norholm2001}
	Norholm MH, Dandanell G.
	\newblock Specificity and topology of the \emph{Escherichia coli} xanthosine
	permease, a representative of the NHS subfamily of the major facilitator
	superfamily.
	\newblock Journal of bacteriology. 2001;183(11466294):4900--4904.

	\bibitem{Chure2019}
	Chure G, et~al.
	\newblock Measurements on the \emph{xapABR} system in \emph{E. coli.};
	Unpublished.

	\bibitem{Marzen2013}
	Marzen S, Garcia HG, Phillips R.
	\newblock Statistical Mechanics of Monod--Wyman--Changeux (MWC) Models.
	\newblock Journal of Molecular Biology. 2013;425(9):1433 -- 1460.
	\newblock doi:{https://doi.org/10.1016/j.jmb.2013.03.013}.

	\bibitem{Kaback2015}
	Kaback HR.
	\newblock A chemiosmotic mechanism of symport.
	\newblock Proceedings of the National Academy of Sciences.
	2015;112(5):1259--1264.
	\newblock doi:{10.1073/pnas.1419325112}.

	\bibitem{Li2014}
	Li GW, Burkhardt D, Gross C, Weissman J.
	\newblock Quantifying Absolute Protein Synthesis Rates Reveals Principles
	Underlying Allocation of Cellular Resources.
	\newblock Cell. 2014;157(3):624 -- 635.
	\newblock doi:{https://doi.org/10.1016/j.cell.2014.02.033}.

	\bibitem{Milo2016}
	Milo R, Phillips R.
	\newblock In: Cell biology by the numbers. Garland Science. 2016; 217.

	\bibitem{Cherry2000}
	Cherry JL, Adler FR.
	\newblock How to make a Biological Switch.
	\newblock Journal of Theoretical Biology. 2000;203(2):117--133.

	\bibitem{Cao2006}
	Cao Y, Gillespie DT, Petzold LR.
	\newblock Efficient step size selection for the tau-leaping simulation method.
	\newblock J Chem Phys. 2006;124(4):044109.
	\newblock doi:{10.1063/1.2159468}.

	\bibitem{Maarleveld2015}
	Maarleveld TR, Olivier BG, Bruggeman FJ. StochPy, Stochastic Modeling in
	Python. 2015.
	\newblock Available from: \url{http://stochpy.sourceforge.net/}.

\end{thebibliography}

\todo[inline]{Should I also list the references that are cited in the appendix?}

\todo[inline]{The capitalization style is different for the titles of the references. Should I list them as they are on the corresponding paper or should I do it uniformly? Does it matter at all? Otherwise I leave it as it is.}

\end{document}

### 2019-10-04 s1_text/s1_text.tex
initial commit of Kathrin's draft

\documentclass[10pt,letterpaper]{article}
\usepackage[top=1.2in,left=1.65in,right=1.65in,footskip=0.6in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
	\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
	\cline{#1}%
	\noalign{\vskip\arrayrulewidth}%
	\noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
	\hline
	\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Figures
\usepackage{graphicx}

%% Include all macros below

\usepackage{placeins}
\usepackage{refcount}
\usepackage{textgreek}
\usepackage{bm}
\usepackage{url} % url in Quellen verlinken

\renewcommand\thesection{\Alph{section}}

\newcommand{\unit}[1]{\,\mathrm{#1}}
\newcommand{\n}[1]{\mathrm{#1}}
\newcommand{\dd}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}
\newcommand*{\defeq}{\mathrel{\vcenter{\baselineskip0.5ex \lineskiplimit0pt
			\hbox{\scriptsize.}\hbox{\scriptsize.}}}%
	=}
\newcommand\subref[2]{%
	\def\myref{\getrefnumber{#1}}% extract the reference number
	\hyperref[#1]{\myref\mbox{#2}}% print the reference number with argument
}

\hyphenation{XapA}
\hyphenation{XapB}
\hyphenation{XapR}
\hyphenation{XapAB}
\hyphenation{XapABR}
\hyphenation{xapA}
\hyphenation{xapB}
\hyphenation{xapR}
\hyphenation{xapAB}
\hyphenation{xapABR}

\usepackage[obeyFinal]{todonotes} % todo notes and missing figures

%% END MACROS SECTION

\begin{document}

\begin{flushleft}
	{\Large
		\textbf\newline{Supporting information for ``theoretical investigation of a genetic switch for metabolic adaptation''}
	}
	\newline
	% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
	\\
	Kathrin Laxhauber\textsuperscript{1},
	Muir J. Morrison\textsuperscript{2},
	Rob Phillips\textsuperscript{2,3$\ast$}
	\\
	\bigskip
	\textbf{1} Dept???, ETHZ, Zurich, State???, Switzerland\todo{MJM: Add dept \& complete affiliation}
	\\
	\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
	\\
	\textbf{3} Department of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
	\\
	\bigskip

	% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
	% 
	% Remove or comment out the author notes below if they aren't used.
	%
	% Primary Equal Contribution Note
	% Yinyang These authors contributed equally to this work.

	% Additional Equal Contribution Note
	% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
	% \ddag These authors also contributed equally to this work.

	% Current address notes
	% \textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
	% \textcurrency b Insert second current address 
	% \textcurrency c Insert third current address

	% Use the asterisk to denote corresponding authorship and provide email address in note below.
	$\ast$ phillips@pboc.caltech.edu

\end{flushleft}

\newpage
\linenumbers

\section{Model choices, simplifications and assumptions}
This part of supporting information text offers a more detailed discussion
of the assumptions that were made in the model and explains the choices that
were made.

\paragraph*{Simplistic modeling of transcription.}
The precise mechanism at the promoter level is not the key feature of this
system, and we expect the same qualitative results for slightly different
transcription models. This work aims to gain an intuition from the results,
and the more unknown parameters there are, the harder this becomes.
Furthermore, the mechanistic details of transcription are not entirely known
and would make the model too complex: for example, transcription probably is
not a Poisson process, but comes in bursts. Additionally, very little is
known about the function of the \emph{xapAB} promoter. For these reasons, we
do not model every detail, but focus on the most relevant features of the
system, while still trying to be physically correct.

Firstly, the equilibrium assumption is not completely accurate, but has
proven successful in previous work. An example for this is the very detailed
analysis of a gene with one repressor in~\cite{Phillips2018}. Secondly,
different binding energies of XapR to its two binding sites would change the
equations as follows: the factor $\left(\frac{1}{K_{\n{XapR}}}\right)^2$
would become $\frac{1}{K_{\n{XapR,1}} K_{\n{XapR,2}}}$ and the factor
$\frac{2}{K_{\n{XapR}}}$ would be replaced by $\frac{K_{\n{XapR,1}} +
K_{\n{XapR,2}}}{K_{\n{XapR,1}} K_{\n{XapR,2}}}$. For most choices of
$K_{\n{XapR}}$, this does not give the exact same results quantitatively,
but we do observe the same qualitative outcome (see also next paragraph).
Thirdly, we consider all but one state as inactive, which is not strictly
correct. Still, as mentioned in the main text, experiments suggest that this
is reasonable. The fourth main simplification, which is the independence of
polymerase and XapR binding, will be discussed in a separate subsection
below.

In Fig~\ref{figS1:transcription}, we have plotted phase portraits where some
of the above assumptions have not been made. For that, we have chosen some
extreme values for the new parameters to demonstrate that these
simplifications do not have a significant influence. The two additional
transcription rates are large compared to experimental observations.
Moreover, the chosen values of $\n{[XapR]}_{\n{R},i}$ correspond to a
difference of two orders of magnitude between $K_{\n{XapR,1}}$ and
$K_{\n{XapR,2}}$. The difference between Oid and O3 in \emph{lac} are two to
three orders of magnitude~\cite{RazoMejia2018}. The values were chosen such
that $K_{\n{XapR,1}} K_{\n{XapR,2}} = K_{\n{XapR}}^2$. Unsurprisingly, this
is necessary to not observe large deviations from the original equations.
The results lead to the conclusion that, for the purposes of this paper, the
aforementioned simplifications are appropriate.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{FigSI1.pdf}
	\caption{{\bf Results for different models of transcription.}
		The parameters are the same as the ones that were used for all other
		phase portraits in this report and that are listed in a table in the
		main part. In (A), the phase portrait after addition of non-zero
		transcription rates to all other polymerase-bound states is shown
		(third assumption lifted). The new parameter values are $r_{\n{I}} =
		10^{-2} r_{\n{p}}$ for the transcription rate when only one XapR is
		bound and $r_{\n{0}} = 10^{-3} r_{\n{p}}$ for the basal rate. In
		(B), the phase portrait after introducing different XapR
		dissociation constants can be seen (second assumption lifted). The
		new parameter values are $\n{[XapR]_{R,1}} = 10$ and
		$\n{[XapR]_{R,2}} = 0.1$.}
	\label{figS1:transcription}
\end{figure}

\paragraph*{Activator and polymerase interaction.}
A mechanistically more precise model of activation would include an
interaction energy between polymerase and XapR. In that case,
$p_{\n{active}}$ cannot be factored into separate terms for the polymerase
and the XapR binding probability. However, the precise activation mechanism
is unknown.

Moreover, introducing an interaction energy between polymerase and XapR
would make the equations overly complex. This would not only make
factorization impossible, it would also be just as consequential to have
different dissociation constants and additional rates as in the previous
paragraph. This leads to a huge number of terms and parameters because of
different interaction energies for every state, which can be foreseen from
the states and weights shown in the main text. The precise transcription
mechanism is not one of the central parts of the circuit we are
investigating (see also discussion in the previous paragraph). Thus, making
it the by far most complex and precise part of the equations would be
unreasonable and hinder the understanding of the more relevant components of
the system.

\begin{table}
	\centering
	\caption{
		{\bf States and weights for a more precise model of transcription.}}
	\begin{tabular}{llll}
		\thickhline
		XapR & Pol. & Transcription & Weight \\
		\hline
		none & no & $0$ & $1$ \\
		first & no & $0$ & $w_{\n{X1}}$ \\
		second & no & $0$ & $w_{\n{X2}}$ \\
		both & no & $0$ & $w_{\n{X1}} w_{\n{X2}} e^{\Delta \epsilon_{\n{coop}}}$ \\
		none & yes & $r_0$ & $w_{\n{p}}$ \\
		first & yes & $r_{\n{X1}}$ & $w_{\n{p}} w_{\n{X1}} e^{\Delta \epsilon_{\n{X1}}}$ \\
		second & yes & $r_{\n{X2}}$ & $w_{\n{p}} w_{\n{X2}} e^{\Delta \epsilon_{\n{X2}}}$ \\
		both & yes & $r_2$ & $w_{\n{p}} w_{\n{X1}} w_{\n{X2}} e^{\Delta \epsilon_{\n{X1}} + \Delta \epsilon_{\n{X2}} + \Delta \epsilon_{\n{coop}}}$ \\
		\thickhline
	\end{tabular}
	\begin{flushleft} 
		The first column shows which XapR sites are occupied in this state,
		the second one whether polymerase is bound to the promoter, and the
		third the transcription rate. In the weights, the notation is as
		follows: $w_{\n{X1}}$ is the weight for XapR binding to the first
		binding site on the promoter, $w_{\n{X2}}$ is that to the second
		site, $w_{\n{p}}$ is the weight of polymerase binding to the
		promoter, $e^{\Delta \epsilon_{\n{X1}}}$ is the interaction energy
		between one XapR on the first binding site and the polymerase that
		is bound, $e^{\Delta \epsilon_{\n{X2}}}$ is the same for the second
		binding site, and finally, $e^{\Delta \epsilon_{\n{coop}}}$ is the
		interaction energy of the two XapR at both binding sites as in the
		main part.
	\end{flushleft}
	\label{tableSI1:statesPromoter}
\end{table}

Furthermore, we do not expect our approximation to have a large influence:
the interaction energy will make the doubly bound state dominate, and the
polymerase-XapR interaction should be weak in the singly bound state,
because very weak activation is observed in the experiment when one of the
XapR binding sites is removed~\cite{Chure2019}. If $e^{\Delta
\epsilon_{\n{X1}}}$ and $e^{\Delta \epsilon_{\n{X2}}}$ are small, we have
$w_{\n{p}} w_{\n{X1}} e^{\Delta \epsilon_{\n{X1}}} \approx w_{\n{p}}
w_{\n{X2}} e^{\Delta \epsilon_{\n{X2}}} \approx w_{\n{p}} w_{\n{X}}$. If, in
addition, the transcription rates for the singly bound states are much
smaller than those for the doubly bound state, we can neglect the singly
bound states in the nominator of $p_{\n{active}}$ (which, as argued in the
previous paragraph ``Simplistic modeling of transcription'', is a good
assumption). Thereby, we can obtain the same form of $p_{\n{active}}$ that
we are working with in the main part by including the interaction between
polymerase and XapR into $\epsilon_{\n{coop}}$.

\paragraph*{Working with only one protein species.}
The proteins are transcribed together, and thus, their rate of transcription
and mRNA decay should be the same. The protein decay is assumed to be
dominated by cell division, making the decay rates roughly equal as well.
What could differ is the translation rate. Changes in this rate rescale the
protein synthesis rate by the same factor and thus, in steady-state, change
the total number of proteins. The only other place where the proteins appear
in the ODE's is in their respective Michaelis-Menten term, and the latter is
proportional to $k_{\n{cat}} \n{[E]_0}$. Hence, different numbers of XapA
and XapB can be accounted for by changing their respective values of
$k_{\n{cat}}$.

\paragraph*{Exponential protein decay.}
The protein decay should be dominated by dilution through cell division,
which is a discrete process. However, the cell grows before division,
resulting in a continuous dilution. Furthermore, replication leads to a
varying number of gene copies and thereby also a varying transcription rate.
Trying to capture all these mechanistic details would lead to an
overcomplicated model. Treating the protein decay as a Poisson process has
produced correct results in past research, which is why we expect this
assumption to be reasonable.

\paragraph*{Kinetics of XapB, NupC, and NupG.}
All transporters in our system are driven by the proton
gradient~\cite{Norholm2001}, and we assume their kinetic scheme to be like
that of \emph{lac} permease. This scheme is shown schematically in
Fig~\ref{figS2:permease}. Even though it includes several steps, the
essential part is substrate binding, conformational change of the
transporter, and substrate release on the other side of the membrane. We
assume the proton gradient across the membrane to be constant and proton
binding and unbinding to the transporter to be fast. Hence, the latter does
not need to be included explicitly in the kinetic model (it is implicitly
part of the turnover rate).

\begin{figure}
	\centering
	\includegraphics[width=0.7\textwidth]{FigSI2.pdf}
	\caption{{\bf Kinetic scheme for transport by proton-driven pumps.}
		The mechanism of lac permease as for example described
		in~\cite{Kaback2015} is used as a model for xanthosine transport.
		The enzyme binds a proton and a substrate (steps 1 and 2), undergoes
		conformational change (steps 3 and 4), and then releases substrate
		and proton on the other side of the membrane (steps 5 and 6). When
		the enzyme is empty, it can change its conformation again to repeat
		the transport (steps 7 and 8). Instead, it could also bind a new
		proton and substrate and transport them in the opposite direction
		(steps 6, 5, 4, 3, 2, 1). Alternatively, it could also keep the
		proton and just cycle back and forth, exchanging substrates from
		outside and inside the cell (steps 5, 4, 3, 2).}
	\label{figS2:permease}
\end{figure}

Substrate release should be much faster than conformational change, which is
analogous to ignoring the ``EP'' state in Michaelis-Menten (enzyme with
product bound, here the conformationally changed transporter with substrate
still bound). Also, it should be reasonable to neglect the immediate
backwards reaction (before proton release), and thus, Michaelis-Menten
kinetics is obtained. Data from~\cite{Kaczorowski1979}
and~\cite{Viitanen1984} actually show that, at a physiologically reasonable
proton gradient, both, influx and efflux, can be described by the
Michaelis-Menten equation very well.

What is neglected in this modeling is the fact that the transporter needs to
release the proton and undergo a conformational change again before being
able to conduct the same reaction again. However, by using experimentally
determined apparent values to estimate the kinetic constants, this is
already included in these parameter values. Another simplification we make
is to treat influx and efflux as two independent reactions instead of
thinking of it as one reversible kinetic process. This seems reasonable,
though, because the transporter needs to lose its proton between reactions
in order to obtain a net transport. If the transporter has for example done
steps 1-6 in Fig~\ref{figS2:permease}, it can from there on really enter two
entirely different reactions: it can do its conformational change in its
empty state (steps 7 and 8) to then transport a substrate again (steps 1-6)
or it can bind a new proton and a new substrate directly and transport them
back (steps 6-1). The former would give influx and the latter efflux, so the
two of them can be modeled as separate reactions.

The NupC and NupG reactions are approximated as linearly dependent on the
substrate. This should be appropriate, since their affinity for xanthosine
is so low that it cannot even be measured properly~\cite{Norholm2001}. This
makes $K_{\n{M}}$ much larger than $c$ or $\n{[x]}$ will ever be. For XapB,
this is not the case, because it has much higher specificity for xanthosine
and $c$ and $\n{[x]}$ take values around the $K_{\n{M}}$ values of XapB.

\paragraph*{Passive transport across the membrane.}  
Most small molecules diffuse passively through the membrane, but at
physiological pH, xanthosine is mostly negatively charged and charged
substances do not significantly diffuse across the membrane on their own.
Nevertheless, from the $\n{pK_a}$ of xanthosine we find that at
physiological pH, about 1.5\% of it is uncharged and could pass the
membrane. This could be modeled as follows:
\begin{eqnarray}
\label{eq:diff}
\left(\dd{\n{[x]}}{t}\right)_{\n{diff}} = J A = p (c-\n{[x]}) A = \underbrace{p A}_{\n{rate}} (c-\n{[x]})
\end{eqnarray}
Here, $J$ is the flux through the membrane, which follows directly from
Fick's first law (if steady-state diffusion is assumed). The other new
parameters are the membrane area $A$ and the membrane permeability
coefficient $p$ of the diffusing substrate.

The diffusion rate $k_{\n{d}} = p A$ can be estimated with the help of
Overton's rule. In our model, we find that this passive diffusion alone
(i.e. without NupC and NupG) is enough to obtain bistability in the
deterministic analysis as well as bimodality in the stochastic simulations.
However, experiments show that \textDelta\emph{nupC} \textDelta\emph{nupG}
strains cannot grow on xanthosine~\cite{Norholm2001}. This suggests that
passive diffusion alone is not enough to make the cells switch. Experiments
with cells in minimal media containing some glucose and a xanthosine
concentration in the bimodal regime show that the switching time can be
several generations. The same is observed in our simulations. The xanthosine
concentration that was used for the double knockout experiments, where no
glucose was present, is just above the switching threshold, which is when
switching is slow. For this reason, we suspect that without NupC and NupG
and with only xanthosine as an energy resource, the cells starve before they
have switched. This probably is the reason why passive diffusion is, in
reality, not enough to observe switching. NupC and NupG can probably keep
the cells going at very low intracellular xanthosine concentrations. Because
of the active transport, they can accumulate a higher intracellular than
extracellular concentration of xanthosine.

Since NupC and NupG transport seems to be much more significant than
diffusion, we do not include the latter explicitly in our model. However,
because it takes the form given in Eq~\ref{eq:diff}, it can easily be
accounted for implicitly by slightly changing the values of $k_{\n{nup}}$
and $\xi$.

\paragraph*{Other mechanisms for xanthosine degradation.}
There probably are some ways in which xanthosine is degraded, other than by
XapA. Otherwise, initially uninduced cells that are transferred to only
xanthosine as an energy source might not survive until they have switched.
As mentioned previously, it was found that they do~\cite{Norholm2001}, so
they need to be able to metabolize xanthosine somehow. Nevertheless, these
degradation pathways cannot have a big effect: as argued later on when
estimating the Nup parameters, the intracellular xanthosine concentration in
the uninduced state must be a bit larger, but not too much larger, than the
extracellular one. Thus, degradation of xanthosine has to be small because
import through Nup is weak. Due to this, it should be appropriate to
linearize the equations describing these effects. Then, they can simply be
accounted for in the term for Nup transport by making slight changes in the
values of $\xi$ and $k_{\n{nup}}$. It could, however, also be that the basal
level of XapA is enough to obtain the small degradation effect without any
other pathways.

\section{Parameter estimation}
In this section we present a detailed estimation of the values of all
nondimensional parameters. Two dimensionful parameters appear in most
nondimensional ones, namely $\gamma_{\n{p}}$ and $K_{\n{a}}$. For this
reason, we start by estimating these.

\paragraph*{The protein decay rate $\bm{\gamma_{\n{p}}}$:}
Protein decay is generally dominated by dilution through cell growth and
division. The average doubling time of \emph{E. coli} is 20-30 minutes. We
take this as an estimate for the protein half life, which leads us to a
decay rate of $\gamma_{\n{p}} = \frac{\ln 2}{t_{1/2}} \approx 5\substack{+2
\\ -3} \cdot 10^{-4}~\n{s^{-1}}$. Note that these upper and lower bounds
correspond to half-lifes between 15 and 60 minutes, which should be
reasonable. For the following nondimensionalization, we work with a fixed
value of $\gamma_{\n{p}} = 5 \cdot 10^{-4}~\n{s^{-1}}$.

\paragraph*{The Michaelis-Menten parameters for XapA ($K_{\n{a}}$, $k_{\n{\alpha}}$):}
The Michaelis-Menten parameters of XapA kinetics have been measured in two
independent publications~\cite{Dandanell2005,Koszalka1988}. Both have found
similar values and we conclude $K_{\n{a}} \approx 6 \pm 3 \cdot
10^{-5}~\n{M}$ and $k_{\n{a}} \approx 10^{-1 \pm 0.8}~\n{s^{-1}}$. This
gives $k_{\n{\alpha}} \approx 10^{2 \pm 0.8}$. We choose to work with
$K_{\n{a}} = 5 \cdot 10^{-5}~\n{M}$ in the following.

\paragraph*{The extracellular xanthosine concentration $\n{[c]_{a}}$:} In
the experiment, switching occurred if cells were grown in a solution with a
concentration of xanthosine of roughly a few mM~\cite{Norholm2001}.
According to these experimental observations, the interesting range of
xanthosine should be $\n{[c]_{a}} \in [ 0, 10^3 ]$ (nondimensionalized),
which corresponds to $\n{c} \in [ 0, 5 \cdot 10^{-2} ]~\n{M}$. However, we
are careful because there are issues with the solubility of xanthosine.

\paragraph*{The MWC parameters $K_{\n{\chi A}}$, $K_{\n{IA}}$, and $\Delta \epsilon_{\n{x}}$:}
The three MWC parameters are the energy barrier between the active and
inactive state of XapR and the equilibrium constants of xanthosine binding
to XapR in the two states. In Fig~\ref{figS2:induction}, the induction curve
as a function of the nondimensional parameters is shown. We assume that
evolution has tuned the induction curve such that less than $10 \%$ of XapR
is active in the absence of xanthosine, and more than $90 \%$ is active for
$\n{[x]}$ approaching infinity. This is a relatively weak assumption,
otherwise XapR would not really be fulfilling its job. It helps us by
putting a constraint on $K_{\n{IA}}$ and $\Delta \epsilon_{\n{x}}$:
\begin{alignat*}{1}
p_{\n{active}}(\n{[x]_a} = 0) &= \frac{1}{1 + e^{\Delta \epsilon_{\n{x}}}} < 0.1 \\
1-p_{\n{active}}(\n{[x]_a} \to \infty) &= \frac{e^{\Delta \epsilon_{\n{x}}}}{K_{\n{IA}}^2 + e^{\Delta \epsilon_{\n{x}}}} < 0.1
\end{alignat*}
Thus, 
\begin{eqnarray*}
	\Delta \epsilon_{\n{x}} > 2.2 \hspace{2 em} \n{and} \hspace{2 em} K_{\n{IA}} > 3 e^{\Delta \epsilon_{\n{x}} / 2}.
\end{eqnarray*}

A further constraint is found from $K_{\n{IA}}=\frac{K_{\n{x I}}}{K_{\n{x
A}}} = e^{\beta \Delta E}$, where $\Delta E$ is the energy difference
between xanthosine binding to the two states. It is unlikely that this
energy difference is more than $7~k_{\n{B}} T$, which corresponds to roughly
three strong hydrogen bonds. This gives us an upper bound $K_{\n{IA}} <
10^3$ and thereby $\Delta \epsilon_{\n{x}} < 10^1$. For $K_{\n{IA}} = 10^2$,
the upper bound is $\Delta \epsilon_{\n{x}} < 7$. This gives us a range of
$K_{\n{IA}} \approx 10^1 - 10^3$ and $\Delta \epsilon_{\n{x}} \approx 2 -
10$ or, more precisely, $\Delta \epsilon_{\n{x}} \approx 2$ to
$\ln\left(\frac{1}{9} \left(K_{\n{IA}}\right)^2 \right) \approx 2 \left(
\ln\left(K_{\n{IA}}\right) - 1\right) < 12$. Compared with other systems,
these values do seem reasonable.

\begin{figure}
	\centering
	\includegraphics[width=0.85\textwidth]{FigSI3.pdf}
	\caption{{\bf Induction curves given by MWC model.}
	The main plot shows the probability of XapR being in the active state as
	a function of the logarithm of the xanthosine concentration for $\Delta
	\epsilon_{\n{x}}$ as shown, $K_{\n{\chi A}} = 10^{2}$, and $K_{\n{IA}} =
	10^{2}$. The shaded areas indicate where the curve lies for changes in
	$K_{\n{IA}}$ of one order of magnitude. Changing $K_{\n{\chi A}}$ by one
	order of magnitude, on the other hand, shifts all the curves by one
	order of magnitude in the corresponding direction to the left or right.
	The inset shows how the allowed regimes of $\Delta \epsilon_{\n{x}}$ and
	$K_{\n{IA}}$ depend on each other. This explains why for $\Delta
	\epsilon_{\n{x}} = 7$ and $K_{\n{IA}}=10$ (lower green line in the main
	plot), $p_{\n{active}}$ at infinite xanthosine concentrations becomes
	far too low.}
	\label{figS2:induction}
\end{figure}

To find $K_{\n{\chi A}}$ we furthermore assume that the steep part of the
induction curve lies in the biologically relevant xanthosine regime, which
estimated experimentally, see the paragraph about $\n{[c]_a}$ above. Again,
this is a rather weak assumption. As can be seen from
Fig~\ref{figS2:induction}, this assumption sets the value range of
$K_{\n{\chi A}}$. The line for $\Delta \epsilon_{\n{x}} = 5$ in
Fig~\ref{figS2:induction} is at the experimentally expected position in the
xanthosine regime. It becomes clear from the figure that the position of the
steep range is strongly affected by $\Delta \epsilon_{\n{x}}$, and a change
of 1 in the latter leads to roughly a change of $10^1$ in the former. Thus,
we estimate $K_{\n{\chi A}} \approx 10^{2 \pm 1} \cdot 10^{\Delta
\epsilon_{\n{x}}-5}$ and thereby finish our estimation of the MWC
parameters.

\paragraph*{The XapR concentration $\n{[XapR]_R}$:} 
It is experimentally known that the XapR copy number is very low, on the
order of $10^{-8}~\n{M}$ ($10^1$ molecules per cell)~\cite{Li2014}. We
estimate the XapR -- DNA dissociation constant from the values for
\emph{lac} that were found in~\cite{RazoMejia2018} and obtain $10^{-8 \pm
2}~\n{M}$, where we have added one additional order of magnitude as
insecurity to the upper and lower bound. This gives
$\n{[XapR]_R}=\frac{\n{[XapR]_{\n{tot}}}}{K_{\n{XapR}}} \approx 10^{0 \pm
2}$.

\paragraph*{The XapR cooperative energy $\Delta \epsilon_{\n{coop}}$:}
Any cooperative energy should come mainly from an interaction between the
two XapR molecules. It would be surprising if it was larger than $10 \
k_{\n{B}} T$, which corresponds to roughly 4 strong hydrogen bonds. Hence,
we estimate $\Delta \epsilon_{\n{coop}} \approx 0 - 10$.

\paragraph*{The transcription rate parameter $\rho_{\n{m}}$:} The
transcription rate $r_{\n{m}} \frac{\n{[P]}}{\n{[P]} + K_{\n{P}}}$ in the
presence of two activator molecules should be on the order of $10^{-2 \pm
1}~\n{nM\ s^{-1}}$, thus $\rho_{\n{m}} \defeq
\frac{r_{\n{m}}}{\gamma_{\n{p}} K_{\n{a}}} \frac{\n{[P]}}{K_{\n{P}}+\n{[P]}}
\approx 10^{-3 \pm 2}$. Note that the units are $\n{nM\ s^{-1}}$ and not
just $\n{s^{-1}}$ because the corresponding rate equation is of the form
$\dd{\n{[m]}}{t} = r \cdot p$ where the probability $p$ is dimensionless and
the left-hand side has dimensions of mRNA concentration per time.

This estimate comes from the assumption that this rate should be roughly
$10^{1} - 10^{2}$ times larger than a normal basal rate (of other genes),
for which common values are around $10^{-3 \pm 1}~\n{nM\
s^{-1}}$~\cite{Milo2016}. Because experiments show that compared to the
activated promoter, the basal transcription rate is almost zero for
\emph{xapAB} (probably due to the weak polymerase binding site), this
actually means a $10^{2}$- to $10^{3}$-fold increase in the transcription
rate due to activation by XapR.

Keep in mind that the rate includes not only the effect of XapR on the pure
speed of transcription, bt also any effect of XapR on the polymerase
dissociation constant. Still, we do not expect higher values such as
$10^{0}~\n{nM\ s^{-1}}$ since this is roughly the rate for transcription of
rRNA and the latter should be significantly faster. To further validate our
estimate, we can compare it to measured values for \emph{lac}, which should
be roughly the same because the gene function is similar. The transcription
rate for \emph{lac} in the absence of any repression that was found
in~\cite{RazoMejia2019} corresponds well with our estimate.

\paragraph*{The mRNA decay rate parameter $\gamma_{\n{mp}}$.}
The mRNA life time is usually on the order of 1 to 8 minutes which gives a
decay rate of $\gamma_{\n{m}} \approx 10^{-3} - 10^{-2}~\n{s^{-1}}$. Thus,
we have $\gamma_{\n{mp}} \defeq \frac{\gamma_{\n{m}}}{\gamma_{\n{p}}}
\approx 10^{1 \pm 0.5}$.

\paragraph*{The translation rate parameter $\rho_{\n{p}}$.}
The translation rate should be on the order of $r_{\n{p}} \approx 10^{-2} -
10^{-1}~\n{s^{-1}}$~\cite{Milo2016}. This gives $\rho_{\n{p}} \defeq
\frac{r_{\n{p}}}{\gamma_{\n{p}}} \approx 10^{2 \pm 0.5}$.

%\com{do growth rate thing}

\paragraph*{The Michaelis-Menten parameters for XapB ($k_{\n{\beta,i}}$, $K_{\n{\beta,i}}$, $k_{\n{\beta,e}}$, $K_{\n{\beta,e}}$):}
To estimate the parameters for both influx and efflux, we need data from
experiments where these two processes have been measured independently.
Lactose permease, which we expect to function in a similar way as XapB, is
comparatively well studied such that these rates have been measured in
multiple ways and for different conditions. Some data from several
publications is collected in~\cite{Viitanen1984} and we will base our
estimates on this and the corresponding references.

Comparing the effective rates and Michaelis constants of lactose permease
and the nucleoside transporters NupC and NupG shows that their orders of
magnitude are similar: from~\cite{Dornmair1989,Wright1985} we find
$k_{\n{cat}} \approx 10^1~\n{s^{-1}}$, $K_{\n{M}} \approx 10^1-10^2~\n{\mu
M}$ for LacY and from~\cite{Norholm2001,Komatsu1973,MunchPetersen1979}
$k_{\n{cat}} \approx 10^2~\n{s^{-1}}$, $K_{\n{M}} \approx 10^0-10^1~\n{\mu
M}$ for pyrimidines through NupC and NupG. Note that especially for Nup,
these are only rough measurements, and it is known that NupC behaves
different with purines (thus also with xanthosine). Given these values, we
expect lactose permease to be similar enough to nucleoside transporters so
that we can use it as a starting point for our estimation of the XapB
kinetic values.

We use the measured influx values that are summarized in~\cite{Viitanen1984}
for a membrane potential of $100~\n{mV}$ and obtain $k_{\n{b,i}} \approx
10^{1 \pm 1}~\n{s^{-1}}$ and $K_{\n{b,i}} \approx 10^{-4 \pm 2}~\n{M}$.
Here, the uncertainty is estimated from the distribution of values of many
kinds of enzymes that is presented in~\cite{Milo2016}. The bounds also
reflect the difference between LacY and Nup that was found above. For the
dimensionless quantities, this gives $k_{\n{\beta,i}} \approx 10^{4 \pm 1}$
and $K_{\n{\beta,i}} \approx 10^{1 \pm 2}$.

For the efflux parameters, we proceed similarly and obtain $k_{\n{b,e}}
\approx 10^{0 \pm 2}~\n{s^{-1}}$ and $K_{\n{b,e}} \approx 10^{-3 \pm
3}~\n{M}$, therefore $k_{\n{\beta,e}} \approx 10^{3 \pm 2}$ and
$K_{\n{\beta,e}} \approx 10^{2 \pm 3}$. Because of disagreement in
literature, difficulty in measuring these parameters, and a lack of
knowledge about what value range should be expected, we put larger error
bounds than for the influx.

\paragraph*{The non-specific transport parameters $k_{\n{\eta}}$, $\xi$:}
Estimating the rates of the Nup transporters for xanthosine is difficult
since there are no experimental values that could be used as a reference.
Because the rate cannot even be measured~\cite{Norholm2001}, it must be much
lower than that of XapB. The kinetic parameters should, however, be such
that the steady-state intracellular xanthosine concentration is
significantly larger than the extracellular one. Otherwise, there would be
no relevant difference to passive diffusion, which was shown to be
insufficient for switching~\cite{Norholm2001}. This essentially means $\xi <
1$, which on the other hand implies that the Nup rate is larger than or at
least equal to the diffusion rate: If Nup transport were slower than
diffusion, the latter would rule and $\xi$ would be close to~1 (because this
is the $\xi$ value of diffusion).

In addition, we know from \emph{lac} permease that influx is not much faster
than efflux, which implies that $\xi$ also cannot be too much smaller
than~1. We guess $\xi \approx 0.8 \pm 0.1$. The number of Nup transporters
per cell is roughly $10^2-10^3$~\cite{Li2014}. We expect
$\frac{k_{\n{cat}}}{K_{\n{M}}}$ to be at least two or three orders of
magnitude lower for Nup than for saturated XapB, where we have
$\frac{k_{\n{b}}}{K_{\n{b}}} \approx 10^{5 \pm 3}~\n{M^{-1}\ s^{-1}}$. These
arguments set an estimated upper bound on the non-specific transport rate:
$k_{\n{nup}} = \frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0} \lesssim 10^{-3 \pm
3}~\n{s^{-1}}$.

To find a lower bound, we estimate the diffusion rate. As described in
Eq~\ref{eq:diff} in the previous section, the diffusion rate can be written
as $10^{-2} \cdot p \cdot A$, where $p$ is the permeability coefficient of
the membrane for xanthosine, $A$ is the total cell surface area, and
$10^{-2}$ is the fraction of uncharged xanthosine at physiological pH. From
the permeabilities of various substances that are collected
in~\cite{Milo2016} and from the measured permeability values for purine
nucleosides in~\cite{Wei2011}, we estimate a permeability coefficient of
$\tilde{p}=10^{-8 \pm 3}~\n{m\ s^{{-1}}}$. We write the tilde because this
value is given in a convention where $\tilde{p} \cdot A = \dd{x}{t}$ with
$x$ being the number of xanthosine molecules and not the concentration.
Since we work in concentrations, we divide by an estimated cell volume of
$10^{-18}~\n{m^3}$ and, with a cell surface area of $10^{-11}~\n{m^2}$, find
a diffusion rate of roughly $10^{-3 \pm 3}~\n{s^{-1}}$. This sets the lower
bound of $k_{\n{nup}}$.
%\com{Check if Overton's rule gives the samediffusion coefficient}

Interestingly, the upper and lower bound turn out to be the same. Thus, we
obtain the range $k_{\n{nup}} \approx 10^{-3 \pm 3}~\n{s^{-1}}$, meaning
$k_{\n{\eta}} \approx 10^{-1 \pm 3}$.

\section{Additional plots and explanations of the results}
\paragraph*{The 2D set of equations.}
If the mRNA concentration is set to steady-state as discussed in the main
text, the dynamical system is reduced to the two equations shown below:
\begin{alignat}{2}
\dd{\mathrm{[p]_a}}{\tau} &= \ &&
\rho 
\frac{
	\mathrm{[XapR]_{R,A}}^2 
	e^{- \Delta \epsilon_{\text{coop}}}
}{
	1 + 
	2 \mathrm{[XapR]_{R,A}} +
	\mathrm{[XapR]_{R,A}}^2 e^{- \Delta \epsilon_{\text{coop}}}
}
- \mathrm{[p]_a}
\\
\dd{\mathrm{[x]_a}}{\tau} &= \ && \left(k_{\n{\beta,i}} \frac{\n{[c]_a}}{K_{\n{\beta,i}} + \n{[c]_a}} - k_{\n{\beta,e}} \frac{\n{[x]_a}}{K_{\n{\beta,e}} + \n{[x]_a}} - k_{\n{\alpha}} \frac{\n{[x]_a}}{1 + \n{[x]_a}}\right) \mathrm{[p]_a} \\ & && + k_{\n{\eta}} \left(\mathrm{[c]_a} - \xi \mathrm{[x]_a}\right)
\nonumber \\
\text{with} & && \raggedleft [\n{XapR}]_{\mathrm{R,\n{A}}} = [\n{XapR}]_{\mathrm{R}} \frac{\left(1 + \frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}}\right)^2}{\left(1 + \frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}}\right)^2 + e^{\Delta \epsilon_{\n{x}}} \left(1+\frac{\mathrm{[x]_{\n{a}}}}{K_{\n{\chi A}}} \frac{1}{K_{\n{IA}}}\right)^2}. \nonumber
\end{alignat}
Here, several of the parameters from the 3D system could now be combined
into one by defining $\rho \defeq \frac{\rho_{\n{m}}
\rho_{\n{p}}}{\gamma_{\n{mp}}}$.

\paragraph*{Magnitude of flow in the phase portraits.}
The plots in the main part do not show the magnitude of the vector fields.
For additional information, we show a 2D vector plot with the magnitudes in
Fig~\ref{figS4:vector}.

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{FigSI4.pdf}
	\caption{{\bf Vector plot of a standard case of bistability.}
		The 2D phase portraits as a vector plot for for the standard set of
		parameters that was used in the main text.}
	\label{figS4:vector} 
\end{figure}

\paragraph*{Increasing $\n{[c]_a}$.}
Fig~\ref{figS5:incC} shows a family of nullclines for increasing
extracellular xanthosine concentration. It shows the transition from
monostability, to bistability, and back to monostability again.

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{FigSI5.pdf}
	\caption{{\bf A family of curves for increasing $\n{[c]_a}$.}
		The blue curve is the protein nullcline, which does not change with
		$\n{[c]_a}$. The other lines are the xanthosine nullclines for
		different values of the extracellular xanthosine concentration.}
	\label{figS5:incC}
\end{figure}

\paragraph*{Hysteresis effects.}
In Fig~\ref{figS8:hysteresis}, the results mentioned in the hysteresis part
of the main text are shown. As mentioned there,  the simulation was started
with initially fully induced cells. At concentrations where no cells
switched from the low to the high expression state before, cells here do not
switch back to low expression.

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{FigSI8.pdf}
	\caption{{\bf Decreasing $\n{[c]_a}$ in the fully induced system.}
		Apart from $\n{[c]_a}$, the parameters are the same as in the table
		in the main text. For the distributions, the simulations were run
		1000 times for $10^6 \unit{s}$ each and started at the mRNA, protein
		and intracellular xanthosine counts of the high fixed point in the
		corresponding phase portraits. The values of $\n{[c]_a}$ are 12 in
		(A), 10 in (B), and 9 in (C). When comparing to the distributions in
		the main text where the simulation had initial concentrations of 0,
		a clear hysteresis effect can be observed.}
	\label{figS8:hysteresis}
\end{figure}

\section{Chemical master equation}
Here we provide some further details on our stochastic simulations by
writing the chemical master equation for our model. The state of our system
is described by the number of mRNA $m$, protein $p$, and xanthosine
molecules $x$, and we write $P(m,p,x)$ for the probability of that state.
The chemical master equation for the system is:
\begin{alignat}{2}
\dd{P(m,p,x)}{t} &= \ && - P(m,p,x) [ r_{\n{m}} p_{\n{active}}(x) + \gamma_{\n{m}} m + r_{\n{p}} m + \gamma_{\n{p}} p \nonumber \\ & && + k_{\n{b,i}} \frac{c}{K_{\n{b,i}} + c} \n{p} + k_{\n{b,e}} \frac{x}{K_{\n{b,e}} + x} p + k_{\n{a}} \frac{x}{K_{\n{a}} + x} p \nonumber \\ & && + k_{\n{nup}} c + k_{\n{nup}} \xi x ] \nonumber \\ & && 
+ P(m+1,p,x) \gamma_{\n{m}} (m + 1) \nonumber \\ & &&  + P(m-1,p,x) r_{\n{m}} p_{\n{active}}(x) \nonumber \\ & && 
+ P(m,p+1,x) \gamma_{\n{p}} (p + 1) \nonumber \\ & &&  + P(m,p-1,x) r_{\n{p}} m \nonumber \\ & && 
+ P(m,p,x+1) [ k_{\n{b,e}} \frac{x+1}{K_{\n{b,e}} + x+1} p \nonumber \\ & && + k_{\n{a}} \frac{x+1}{K_{\n{a}} + x+1} p + k_{\n{nup}} \xi (x+1) ] \nonumber \\ & &&
+ P(m,p,x-1) [ k_{\n{b,i}} \frac{c}{K_{\n{b,i}} + c} p + k_{\n{nup}} c ]  
\end{alignat}
Here, all the parameters of course have to be measured in counts, not in
concentrations. As can be seen from the equation, none of the propensities
depends explicitly on time and hence, we can use the hybrid algorithm
between classical Gillespie and \texttau-leaping that we relied on.

\FloatBarrier

\begin{thebibliography}{10}

	\bibitem{Phillips2018}
	{Phillips} R, {Belliveau} NM, {Chure} G, {Garcia} HG, {Razo-Mejia} M, {Scholes}
	C.
	\newblock {Figure 1 Theory Meets Figure 2 Experiments in the Study of Gene
		Expression}.
	\newblock arXiv e-prints. 2018; p. arXiv:1812.11627.

	\bibitem{RazoMejia2018}
	Razo-Mejia M, Barnes SL, Belliveau NM, Chure G, Einav T, Lewis M, et~al.
	\newblock Tuning Transcriptional Regulation through Signaling: A Predictive
	Theory of Allosteric Induction.
	\newblock Cell Systems. 2018;6(4):456 -- 469.e10.
	\newblock doi:{10.1016/j.cels.2018.02.004}.

	\bibitem{Chure2019}
	Chure G, et~al.
	\newblock Measurements on the \emph{xapABR} system in \emph{E. coli.};
	Unpublished.

	\bibitem{Norholm2001}
	Norholm MH, Dandanell G.
	\newblock Specificity and topology of the \emph{Escherichia coli} xanthosine
	permease, a representative of the NHS subfamily of the major facilitator
	superfamily.
	\newblock Journal of bacteriology. 2001;183(11466294):4900--4904.

	\bibitem{Kaback2015}
	Kaback HR.
	\newblock A chemiosmotic mechanism of symport.
	\newblock Proceedings of the National Academy of Sciences.
	2015;112(5):1259--1264.
	\newblock doi:{10.1073/pnas.1419325112}.

	\bibitem{Kaczorowski1979}
	Kaczorowski GJ, Robertson DE, Kaback HR.
	\newblock Mechanism of lactose translocation in membrane vesicles from
	\emph{Escherichia coli}. 2. Effect of imposed $\Delta \Psi$, $\Delta$ pH, and
	$\Delta \mu_{\mathrm{H^+}}$.
	\newblock Biochemistry. 1979;18(17):3697--3704.
	\newblock doi:{10.1021/bi00584a010}.

	\bibitem{Viitanen1984}
	Viitanen P, Garcia ML, Kaback HR.
	\newblock Purified reconstituted lac carrier protein from \emph{Escherichia
		coli} is fully functional.
	\newblock Proceedings of the National Academy of Sciences of the United States
	of America. 1984;81(6324209):1629--1633.

	\bibitem{Dandanell2005}
	Dandanell G, Szczepanowski RH, Kierdaszuk B, Shugar D, Bochtler M.
	\newblock \emph{Escherichia coli} Purine Nucleoside Phosphorylase II, the
	Product of the xapA Gene.
	\newblock Journal of Molecular Biology. 2005;348(1):113--125.

	\bibitem{Koszalka1988}
	Koszalka GW, Vanhooke J, Short SA, Hall WW.
	\newblock Purification and properties of inosine-guanosine phosphorylase from
	\emph{Escherichia coli} K-12.
	\newblock Journal of bacteriology. 1988;170(3042752):3493--3498.

	\bibitem{Li2014}
	Li GW, Burkhardt D, Gross C, Weissman J.
	\newblock Quantifying Absolute Protein Synthesis Rates Reveals Principles
	Underlying Allocation of Cellular Resources.
	\newblock Cell. 2014;157(3):624 -- 635.
	\newblock doi:{https://doi.org/10.1016/j.cell.2014.02.033}.

	\bibitem{Milo2016}
	Milo R, Phillips R.
	\newblock In: Cell biology by the numbers. Garland Science; 2016. p. 217.

	\bibitem{RazoMejia2019}
	Razo-Mejia M, Phillips R.
	\newblock First-principles prediction of the information processing capacity of
	a simple genetic circuit.
	\newblock bioRxiv. 2019;doi:{10.1101/594325}.

	\bibitem{Dornmair1989}
	Dornmair K, Overath P, Jähnig F.
	\newblock Fast measurement of galactoside transport by lactose permease.
	\newblock The Journal of biological chemistry. 1989;264:342--6.

	\bibitem{Wright1985}
	Wright J K, Seckler R.
	\newblock The lactose/H+ carrier of \emph{Escherichia coli}: lac YUN mutation
	decreases the rate of active transport and mimics an energy-uncoupled
	phenotype.
	\newblock The Biochemical journal. 1985;227:287--97.
	\newblock doi:{10.1042/bj2270287}.

	\bibitem{Komatsu1973}
	Komatsu Y, Tanaka K.
	\newblock Deoxycytidine uptake by isolated membrain vesicles from
	\emph{Escherichia coli} K 12.
	\newblock Biochimica et Biophysica Acta (BBA) - Biomembranes.
	1973;311(4):496--506.

	\bibitem{MunchPetersen1979}
	Munch-Petersen A, Mygind B, Nicolaisen AK, Pihl NJ.
	\newblock Nucleoside transport in cells and membrane vesicles from
	\emph{Escherichia coli} K12.
	\newblock The Journal of biological chemistry. 1979;254 10:3730--7.

	\bibitem{Wei2011}
	Wei C, Pohorille A.
	\newblock Permeation of Nucleosides through Lipid Bilayers.
	\newblock J Phys Chem B. 2011;115(13):3681--3688.
	\newblock doi:{10.1021/jp112104r}.

\end{thebibliography}

\end{document}

### 2019-10-07 main_text/main_text.tex
Edits to intro

@@
 cells and organisms. One abundant class of these circuits are genetic
-switches. They can be constructed in various ways and have different
-functions. What all of them have in common is that the system is bistable:
-there are two stable steady-states that it can be in. The distribution of
+switches. Although their construction and function may differ,
+their common feature is bistability: their output gene expression
+will flow to and remain at one of two steady-state levels.
+The distribution of
 gene expression in a cell culture can then be bimodal. This is not to be
 confused with mere stochastic bimodality, where the system is not stable,
-but the gene expression in each cell is fluctuating between the two levels.
+and the gene expression in each cell can fluctuate between the two levels.
 
@@
 other~\cite{Jacob1961,Gardner2000} (illustrated schematically in
-Fig~\ref{fig1:switches}). Here, the two stable states are high expression of
-the first protein and low expression of the second one or the opposite. This
+Fig~\ref{fig1:switches}). Here, one stable state is high expression of
+the first protein and low expression of the second,
+and the second stable state is the opposite. This
 switch enables the system to have a memory: if something induces expression
@@
 is an auto-activating circuit in which a protein activates its own
-transcription~\cite{Wolf1998}. This gives the system an ``on-off''-switch.
-%\todo{make eps}
+transcription~\cite{Wolf1998}. This gives the system an ``on-off'' switch.
 
@@
 	\caption{{\bf A schematic of different genetic switches.}
-	(A) and (B) show the two most well-known genetic switches that are
-	mentioned in the text. In (C), a very much simplified version of the
+	(A) and (B) show the two most well-known genetic switches:
+	(A) two mutual repressors and (B) a self-activating gene.
+	In (C), a very much simplified version of the
 	circuit that we investigate in this paper can be seen, where the
@@
 
+\textit{MM: I like the above paragraph, but we should also add something
+about the TF \& frame it as resourse utility: xapR, lacI, and I suspect
+others of these TFs happen to be constitutively expressed at low copy
+number. They act as ``always on'' sensors that trigger a switch to produce
+massive numbers of metabolic enzymes only when their substrate is
+available.}
+
 For our modeling, we focus on the \emph{xapABR} genetic circuit from
@@
 xanthosine as an energy source~\cite{Buxton1980,Hammer-Jespersen1980}. The
-circuit is made up of two genes: one that encodes for XapR and another one
+circuit is made up of two operons: one that encodes for XapR and another
 that encodes for XapA and XapB. XapR appears to be a transcription factor
-that is induced by xanthosine and activates the \emph{xapAB} promoter. It
-seems to have two binding sites on the promoter, but the promoter
-architecture and function is not fully understood up to now. The
+that is induced by xanthosine and activates the \emph{xapAB} promoter.
+This promoter seems to have two binding sites for XapR, but the promoter
+architecture and function is not yet fully understood. The
 transcription of \emph{xapR} seems to be constitutive and not

### 2019-10-11 main_text/main_text.tex
A few more edits to intro

@@
 activates the expression of lactose permease. Other examples in \emph{E.
-coli} include the \emph{araFGH} and \emph{xylFGH} genes, which encode for
+coli} include the \emph{araFGH} and \emph{xylFGH} operons, which contain genes for
 arabinose and xylose transporters, respectively. For \emph{lac} and
@@
 expression stabilizes at an ``on'' point and the cell has, in a short-term
-sense, adapted. When there is not enough substrate anymore, the gene, which
+sense, adapted. When there is not enough substrate anymore, the operon, which
 often encodes for a whole set of enzymes for this one metabolite, switches
@@
 
-\textit{MM: I like the above paragraph, but we should also add something
-about the TF \& frame it as resourse utility: xapR, lacI, and I suspect
-others of these TFs happen to be constitutively expressed at low copy
-number. They act as ``always on'' sensors that trigger a switch to produce
-massive numbers of metabolic enzymes only when their substrate is
-available.}
+A key element of this mechanism is the presence of a transcription factor
+which is always expressed at a low level (often at copy numbers
+of order $\sim10$) and which binds to the transporter substrate.
+This is resource efficient for the cell, as this low copy number
+transcription factor acts as an ``always on'' sensor to detect
+the substrate, allowing high copy numbers of the membrane transporter
+and its attendant operon to be expressed only when
+their substrate is actually present. The transcription factors
+LacI, AraC, and XylR all appear to fill this
+role~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
+\todo{do we have a cite for xylR too?}
 
@@
 that encodes for XapA and XapB. XapR appears to be a transcription factor
-that is induced by xanthosine and activates the \emph{xapAB} promoter.
-This promoter seems to have two binding sites for XapR, but the promoter
-architecture and function is not yet fully understood. The
+that is induced by xanthosine and activates the \emph{xapAB} promoter,
+in close analogy to LacI, AraC, and XylR.
+The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
+but the promoter architecture and function is not yet fully understood. The
 transcription of \emph{xapR} seems to be constitutive and not

### 2019-10-12 main_text/main_text.tex
Edits to model section

@@
 role~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
-\todo{do we have a cite for xylR too?}
+\todo{MJM: do we have a cite for xylR too?}
 
@@
 In this section, we present our model of the \emph{xapAB} genetic circuit.
-Fig~\ref{fig2:model} shows an overview of this model. It consists of the
-following steps: 
+Fig~\ref{fig2:model} shows an overview of this model.
+The qualitative picture of the circuit switching its state is as follows:
 \begin{itemize}
-	\item Small amounts of xanthosine are transported into and out of cell in various ways (discussed in more detail later on).
+	\item In the initial absense of XapB, small amounts of xanthosine are
+		nonspecifically transported into and out of the cell (discussed in more detail below).
+	\todo{MJM: ``nonspecific transport'' is a bit misleading:
+	it's actually weakly specific transporters, but the precise mechanism is unimportant
+	\& I think it better to emphasize generality of the model. Does this seem fine?}
 	\item XapR is induced by xanthosine.
@@
 	\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
-	\item At some point, the proteins decay.
+	\item Production of more XapA \& XapB is balanced by decay or dilution through cell divisions.
 \end{itemize}
-Because xanthosine induces the transcription factor XapR, we have a positive
-and negative feedback loop due to XapB and XapA, respectively. The remainder
+Because xanthosine induces the transcription factor XapR, we have positive
+and negative feedback loops due to XapB and XapA, respectively. The remainder
 of this subsection discusses each of the above steps in detail, leading us
-to a set of two coupled ODE's. More in-depth explanations can be found
+to a set of two coupled ODEs. More in-depth explanations can be found
 in~\nameref{S1_Text}.
@@
 		$k_{\n{b,i}}, K_{\n{b,i}}$ and $k_{\n{b,e}}, K_{\n{b,e}}$,
-		respectively. Furthermore, xanthosine enters and leaves the cell in
-		other ways, proportional to rates $k_{\n{nup}}$ and $\xi
+		respectively. Furthermore, xanthosine enters and leaves the cell
+		through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
 		k_{\n{nup}}$, respectively.}
@@
 dimer can bind two xanthosine molecules. The Monod-Wyman-Changeux (MWC)
-model is used to describe the fraction of XapR dimers in the active state: 
+model is used to describe the fraction of XapR dimers in the active state,
+which has the form
 \begin{eqnarray}
 \label{eq:MWC}
-\n{[XapR]_A} = \n{[XapR]_{tot}} \frac{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2}{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2 + e^{\beta \Delta E_{\n{x}}} \left(1+\frac{\n{[x]}}{K_{\n{xA}}} \frac{K_{\n{xA}}}{K_{\n{xI}}}\right)^2}
+\n{[XapR]_A} = \n{[XapR]_{tot}} \frac{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2}{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2 + e^{\beta \Delta E_{\n{x}}} \left(1+\frac{\n{[x]}}{K_{\n{xA}}} \frac{K_{\n{xA}}}{K_{\n{xI}}}\right)^2}.
 \end{eqnarray}
@@
 In Fig~\ref{fig3:states}, all possible states of the promoter in our model
-and their corresponding thermodynamic weights are shown. There, $\n{[P]}$
-denotes the polymerase concentration and $\Delta E_{\n{coop}}$ stands for
+and their corresponding thermodynamic weights are shown. $\n{[P]}$
+denotes the polymerase concentration, and $\Delta E_{\n{coop}}$ stands for
 the interaction energy of the two XapR dimers. If cooperativity in
@@
 polymerase to the promoter, respectively. In statistical mechanics language
-this is equivalent to $\frac{N_{\mathrm{NS}}}{V} e^{\beta \Delta E_{i}}$
+these dissociation constants are equivalent to
+$\frac{N_{\n{NS}}}{V} e^{\beta \Delta E_{\n{XapR}}}$
+and $\frac{N_{\n{NS}}}{V} e^{\beta \Delta E_{\n{P}}}$, respectively,
 with $N_{\mathrm{NS}}$ being the number of non-specific binding sites on the
-DNA, $V$ the volume of the cell, and $\Delta E_{i}$ the interaction energy
-of $i=$XapR or $i=$polymerase with the promoter.
+DNA, $V$ the volume of the cell, and $\Delta E_{\n{XapR}}$ and
+$\Delta E_{\n{XapR}}$, respectively, the interaction energies
+of XapR or polymerase with the promoter.
 
@@
 active and all other states as inactive, meaning they have transcription
-rate $0$. Experiments show that the expression becomes very weak when one of
+rate equal zero. Experiments show that the expression becomes very weak when one of
 the XapR binding sites is removed from the promoter, suggesting that this
@@
 In~\nameref{S1_Text}, the full chemical master equation is shown, which
-leads to a Poisson distribution of the mRNA concentration with the same
+leads to a \todo{MJM: I worry referees will give us grief: ``Real dists aren't Poisson,'' even though it's irrelevant}
+Poisson distribution of the mRNA concentration with the same
 mean.

### 2019-10-13 main_text/main_text.tex
Edits to model section & results/discussion

@@
 \paragraph*{Xanthosine dynamics.}
-We have now discussed how xanthosine activates the synthesis of XapA and
-XapB through XapR. In this part of the section, we close the feedback loop
+Having described how xanthosine activates the synthesis of XapA and
+XapB through XapR, we now close the feedback loop
 by setting up a xanthosine rate equation.
 
+\todo{MJM: Is this para \& next excessively detailed for main text?
+maybe replace w/ 1-2 sentences about low affinity Mich-Ment transporters
+that supplement XapB, \& shift all this to supplement?
+If we want to emphasize the general class of model rather than the particular
+biology details of Xap}
 Transport of xanthosine across the cell membrane happens in two steps which
@@
 \end{eqnarray}
-We recall that $\n{[x]}$ is the intracellular xanthosine concentration, and
-$c$ denotes the extracellular one. Because $k_{\n{b,i}} > k_{\n{b,e}}$ and
+We recall that $\n{[x]}$ is the intracellular xanthosine concentration, while
+$c$ denotes the extracellular. Because $k_{\n{b,i}} > k_{\n{b,e}}$ and
 $K_{\n{b,i}} < K_{\n{b,e}}$, influx dominates at low intracellular
@@
 
-Very little is known about the \emph{xap} system, and thus, there are no
+Very little is known about the \emph{xap} system, and thus, there are almost no
 measured values for the free parameters. Nevertheless, we were able to
@@
 \subsection*{Deterministic phase portraits}
-A standard way to analyze dynamical systems like ours deterministically is
+A standard way to analyze dynamical systems deterministically is
 to plot phase portraits. In the following, we present such plots where the
@@
 Fig~\subref{fig4:bistability}{A} shows the 3D phase portrait for a
-representative set of parameters (that from Table~\ref{table1:nondim}),
+representative set of parameters (shown in Table~\ref{table1:nondim}),
 whose choice is explained below. The plot looks rather complicated at first
@@
 
-It is important to point out that for a different set of parameters, the
+It is important to point out that, for a different set of parameters, the
 dynamics can be quite different. There are, for example, scenarios where the
@@
 A usual simplification with genetic circuits like this is to assume the mRNA
-concentration to be at steady-state, i.e. to write $\dd{\n{[m]_a}}{\tau}=0$
+concentration to be at steady-state, i.e., to write $\dd{\n{[m]_a}}{\tau}=0$
 and solve this for $\n{[m]_a}(\n{[p]_a},\n{[x]_a})$ to simplify the 3D to a
@@
 changes significantly. However, as already pointed out, this is different
-for other parameter values, and thus, this assumption generally is not a
-good one. If the xanthosine dynamics are faster than the mRNA dynamics, the
+for other parameter values, and thus, this assumption does not hold in general.
+If the xanthosine dynamics are faster than the mRNA dynamics, the
 system first flows towards the xanthosine nullcline. In that case, forcing
@@
 result, there is bistability and the distribution of expression among cells
-is bimodal.
-
-This corresponds to the experimental observations, so the model passes this
+can be bimodal, depending on initial conditions.
+
+\todo{MJM: Do any of the older papers have data for this too,
+or were they all just colony growth screens on plates?}
+This corresponds to the experimental observations\cite{Chure2019}, so the model passes this
 sanity check. Furthermore, the xanthosine and protein concentrations at the
 upper fixed point have the expected order of magnitude: the xanthosine
-concentration is roughly $10-100 \unit{nM}$, and there are roughly $500$
+concentration is roughly $10-100 \unit{mM}$, and there are roughly $500$
 proteins, which is just a bit lower than what was measured for the number of
@@
 cannot be obtained. This can be understood when looking at the regions that
-are bounded by all three nullclines: within these regions, the streamlines
-on the nullclines point into the regions, so deterministically, the system
-is trapped in there and can only flow towards the fixed points.
+are bounded by all three nullclines: on these boundaries, the streamlines
+point into the regions, so deterministically, they serve as
+trapping regions from which the system cannot escape. Once inside, the only
+possible trajectory is non-oscillatory flow towards the stable fixed point.
 
@@
 biologically the most relevant is the extracellular xanthosine
-concentration. When it is increased in the experiment, the cells go from all
+concentration. When it is increased in experiments, the cells go from all
 being in the low expression state to being in either the low or the high
@@
 
-Furthermore, by setting $\n{[c]_a}=0$ (not shown here) we found that in the
-absence of xanthosine, there are roughly 2-3 proteins, which agrees very
+Furthermore, we found that in the absence of xanthosine, i.e., setting
+$\n{[c]_a}=0$ (not shown here), there are roughly 2-3 proteins, which agrees very
 well with measurements, where around 2 proteins per cell were
@@
 that $K_{\n{\chi A}}$ is large as argued in the estimation of $K_{\n{\chi
-A}}$ in~\nameref{S1_Text}. Thus, the interaction between xanthosine and XapR
+A}}$ in~\nameref{S1_Text}. Thus, we predict that the interaction between xanthosine and XapR
 should be weak.
@@
 be seen from the plots in figure~\ref{fig6:xapAB}. XapA lowers the upper
-solution by a little and could, in principle, thereby make the
+solution by a small amount and could, in principle, thereby make the
 high-expression solution vanish. For our choice of all other parameters,
 this requires $k_{\n{\alpha}} > 10^4$ which is far from what has been
-measured. However, a higher rate could, in principle, be achieved by
+measured. However, a higher effective rate could, in principle, be achieved by
 different translation rates of XapA and XapB (see simplifications of the
 model in~\nameref{S1_Text}). Hence, we cannot exclude the possibility that
-XapA gets so strong that it makes bistability impossible, but this is an
+XapA becomes so strong that it makes bistability impossible, but this is an
 extreme case. XapB, on the other hand, is essential; without it the system
-only has the one fixed point at low expression.
+only has the one fixed point at low expression.\footnote{
+Seeger et.\ al.\cite{Seeger1995} observed that $\Delta xapB$ mutants could
+survive, but grew extremely slowly, with xanthosine as the only carbon
+source, which makes sense in light of Fig~\ref{fig6:xapAB}.
+With \textit{xapB} removed, the switch never activates and the cells are
+forced to survive with an extremely meager quantity of XapA to metabolize
+the abundant xanthosine.}
 
@@
 significantly disturb this adaptation mechanism, while its kinetic
-parameters can be chosen somewhat freely as necessary for metabolism. By
+parameters and expression levels can be chosen 
+somewhat freely as necessary for metabolism. By
 having a membrane transporter gene on an operon whose expression is
@@
 enzymes can be turned on and off depending on the presence of the substrate.
-It seems likely that this way of short-term adaptation of a single cell to
+It seems likely that this mechanism of short-term adaptation of a single cell to
 its environment may be used by cells for many metabolic processes.
@@
 a necessary feature for bistability. This is motivated by the importance of
-cooperativity in typical ``genetic switches''\cite{Gardner2000,Cherry2000}.
+cooperativity in ``typical'' genetic switches\cite{Gardner2000,Cherry2000}.
 
@@
 promoter, we find that the system still has a bistable parameter regime.
-However, this is only a rather narrow parameter range which makes the system
-less stable. When the second binding site is removed in both places, the
+However, this is only a rather narrow parameter range, which makes the system
+\todo{MJM: the (hypothetical) utility of this switch seems to be resource efficiency,
+which $\Rightarrow$ pushing the fixed points to extreme highs \& lows is good
+$\Rightarrow$ stronger coop is good. maybe we should work this in}
+less bistable. When the second binding site is removed in both places, the
 equations are insufficiently non-linear to produce bistability. An example
@@
 very insightful. Hill equations only describe scenarios with multiple,
+\todo{MJM: but that already sounds like an effective interaction,
+I'm not sure I like this phrasing, or the next sentence's.}
 non-interacting binding sites where any other state but the fully occupied

### 2019-10-14 main_text/main_text.tex
Conclusions edits and more

@@
 
-\todo{MJM: Is this para \& next excessively detailed for main text?
+\todo[inline]{MJM: Are the following 2 paragraphs excessively detailed for main text?
 maybe replace w/ 1-2 sentences about low affinity Mich-Ment transporters
@@
 will actually populate both the high and the low expression state: as can be
-seen in Fig~\subref{fig8:stochC}{(A)}, the cells did not switch to the high
-expression state for the waiting time in our simulations, even though one
-would naively expect this from the phase portrait.
+seen in Fig~\subref{fig8:stochC}{(A)}, the cells never switched to the high
+expression state during the run time of our simulations.
+\todo{MJM: depends on an effective barrier height, hard to intuit from phase port.
+Bistability rather than bimodality means it may very well practically never switch.}
 
@@
 describes the mean values and the simulation fluctuates around this mean.
-If, however, the mRNA number at the third, high fixed point is too low, the
+If, however, the mRNA number at the third (high) fixed point is too low, the
 system will not be able to resolve the two points anymore. This becomes a
@@
 
-Comparing this to the experimentally expected timescales is difficult,
+Comparing this to the experimentally expected timescales~\cite{Chure2019} is difficult,
 because the switching time strongly depends on the extracellular xanthosine
@@
 \paragraph*{Hysteresis.} 
-In the previous two paragraphs, the simulations were started at initial
+In Figs~\ref{fig8:stochC} and~\ref{fig:StochT}, the simulations were started at initial
 intracellular concentrations of 0 to investigate what happens if xanthosine
@@
 The resulting distributions can be found in~\nameref{S1_Text}. They show
-clear hysteresis effects: at concentrations where all cells remained in the
-low expression state when being started there before, all cells now remain
-at high expression. Only when the second and the third fixed point are very
-close does switching to the uninduced state happen. Interestingly, this
+clear hysteresis effects: there exist extracellular xanthosine concentrations where
+initially uninduced cells remained uninduced and initially induced cells remain induced.
+Only when the second and the third fixed point are very close can initially
+induced cells switch to the uninduced state. Interestingly, this
 behavior is symmetric to the ``switching on'' in the previous paragraphs,
 where the lower two fixed points needed to be about as close as the upper
-two need to be now for ``switching off''. The extracellular xanthosine
-concentration (for the other parameters being as in
-Table~\ref{table1:nondim}) needed to be as low as $\n{[c]_a} = 10$ to
-observe bimodality and $\n{[c]_a}=9$ to have all cells in the low expression
-state (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5 \cdot
-10^{-5} \unit{M}$). \todo{Is this understandable without seeing the data? Do
-you have any suggestions to make it clearer?}
-
-We have learned that cells only change their metabolism to xanthosine if
+two need to be now for ``switching off''. %The extracellular xanthosine
+% concentration (for the other parameters being as in
+% Table~\ref{table1:nondim}) needed to be as low as $\n{[c]_a} = 10$ to
+% observe bimodality and $\n{[c]_a}=9$ to have all cells in the low expression
+% state (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5 \cdot
+% 10^{-5} \unit{M}$).
+\todo[inline]{Is this understandable without seeing the data? Do
+you have any suggestions to make it clearer?
+MJM: I think the quantitative details can go to the supplement where they
+have more context, the qualitative observation of hysteresis is enough here.}
+
+In other words, cells only change their metabolism to xanthosine if
 enough of the latter is around, but after they have switched, this metabolic
@@
 \paragraph*{A genetic switch for metabolic adaptation with remarkable features.}
-In this paper, we propose a model for biological pathway systems containing
+In this paper, we propose a model for genetic circuits containing
 a membrane transporter whose gene expression is, directly or indirectly,
@@
 concentration of the relevant metabolite. This switch has very useful
-biological features: first, coupling of the transporter with, for example, a
-degradation enzyme creates a genetic switch that enables short-term
+biological features. First, coupling of the transporter with, for example, an
+enzyme which metabolizes the transporter substrate creates a genetic switch that enables short-term
 adaptation of the cell's metabolism to its environment. Second, the switch
-is stabilized by hysteresis effects when the extracellular xanthosine
+is stabilized by hysteresis effects when the extracellular substrate
 concentration decreases, which explains previous experimental findings.
@@
 more thoroughly, for example the dependence of the switching time on
-$\n{[c]_a}$. Having said that, the simulation for such an analysis should
+$\n{[c]_a}$, but such an analysis would probably need to
 account for the burstiness of transcription and translation as well as for
@@
 features.
+\todo[inline]{MJM: I think this last sentence sells us short: qualitative changes result
+if we remove any present feature of the model.
+That sounds to me like a distillation of the essential.}
 
@@
 how cooperative these binding sites are. 
+\todo[inline]{MJM: I'm not sure we've really proven the asserted points re 3D \& Hill.
+I would advocate removing this whole paragraph.}
 
 \paragraph*{A successful model.}
-All model parameters could be estimated fairly well despite the lack of deep
+All model parameters could be reasonably estimated despite the paucity of
 experimental knowledge about the model system. In the results, every
-variable and parameter has its expected order of magnitude, which suggests
+variable and parameter \todo{MJM: I'm confused,
+we estimated all the model parameters, so what expectations are we comparing to?}
+has its expected order of magnitude, which suggests
 that the model captures all relevant features and is able to describe the
@@
 not most metabolic processes involve the adaptation mechanism that we have
-investigated here, and that a lot can be understood about them through our
+investigated here, and that much can be understood about them through our
 model. This apparent success demonstrates once more that even for broadly
@@
 
-\todo{Is the conclusion too long?}
+\todo{Is the conclusion too long? MJM: not necessarily}
 
@@
 providing their data on the \emph{xap} system and pointing out the
-bistability. We also thank Griffin Chure for answering any biological
+bistability. We also thank Griffin Chure for answering the many biological
 questions that arose, Nathan Belliveau for bringing similar systems to our
-attention, and Jane Kondev and Jin Wang for interesting discussions. Plots
-were done in \emph{Mathematica} using the packages MaTeX and
-CurvesGraphics6. K.S.L. acknowledges financial support from the
+attention, and Jane Kondev and Jin Wang for interesting discussions.
+For assistance with illustrations, we also thank Nigel Orme and Griffin Chure.
+\todo{MJM: this kinda reads like GC should be in the author list...}
+K.S.L. acknowledges financial support from the
 Werner~Siemens-Foundation through the Swiss Study Foundation.
-\todo[inline]{I don't really know what/who is usually listed here...
-Did I list too many people (I don't know at what ``extent'' of help people are usually named here) or did I forget anyone?
-Should I list everyone who worked on the xap system last summer and provided the data even though they are already cited, but only as Griffin et al.?
-Is it enough to list packages like this or should I try to find some website to cite?
-Should I list Python \& StochPy again it's already mentioned in the text?
-SHould I thank someone for the providing the elements for the figures and whom?
-Is there funding that should be listed?}
+This material is based upon work supported by the National Science Foundation
+Graduate Research Fellowship under Grant No.\ DGE-114446.
+\todo{MJM: need to double-check my NSF \#}
+\todo[inline]{MJM: Software/code acknowledgements usually go in a separate section,
+and if we want we could set up a code repo to cite along with that.}
 
@@
 
-\todo[inline]{Should I also list the references that are cited in the appendix?}
+\todo[inline]{Should I also list the references that are cited in the appendix? MJM: no, I think the supplement has its own separate list.}
 
@@
 Does it matter at all?
-Otherwise I leave it as it is.}
+Otherwise I leave it as it is.
+MJM: I think the copy editors will sort it out.}
 

### 2019-10-28 main_text/main_text.tex
Fix funding acknowledgement
Edits by KL using RP & MM feedback

@@
 		\\
-	Kathrin Laxhauber\textsuperscript{1},
+		Kathrin S. Laxhuber\textsuperscript{1},
 		Muir J. Morrison\textsuperscript{2},
@@
 		\bigskip
-	\textbf{1} Dept???, ETHZ, Zurich, State???, Switzerland\todo{MJM: Add dept \& complete affiliation}
+		\textbf{1} Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland
 		\\
@@
 	which encode for lactose, arabinose, and xylose transporters, respectively.
-As a case study, we build a physical model of the \emph{xapABR} genetic
+	As a case study of a minimal system, we build a generizable physical model of the \emph{xapABR} genetic
 	circuit, which features a regulatory feedback loop through membrane
@@
 	switching and discuss the role of cooperativity and multiple binding sites
-in the model circuit. Altogether, this work explores how a stable genetic
+	in the model circuit. Fundamentally, this work explores how a stable genetic
 	switch for a set of enzymes is obtained from transcriptional auto-activation
@@
 
-\section*{Introduction}
+	\section*{Introduction} \todo{remove images \& todonotes everywhere for submission}
 	Genetic regulatory circuits are fundamental building blocks of functioning
@@
 			circuit that we investigate in this paper can be seen, where the
-	similarity to the switch in (B) becomes obvious. A complete version of
+			similarity to the switch in (B) is clear. A complete version of
 			the model circuit can be found in Fig~\ref{fig2:model}.}
@@
 	Through physical and mathematical modeling, we investigate a more complex
-but common switch system where the bistability is due, as we will show, to a
-membrane transport protein. Its key feature is the indirect activation of
+	switch system where the bistability is due, as we will show, to a
+	membrane transport protein.
+Such a switch is common for metabolic processes in biology, for reasons discussed below.
+So far, models we have found are either very extensive descriptions of specific, complicated systems, or they are very basic.
+To enable an intuitive and fundamental understanding of the behavior, we develop a simple model of a minimal system that only has the essential components and interactions for the questions we pose, while still modeling every component explicitly and discussing the necessary model complexity for a physically correct model.
+
+	The key feature of the type of system we investigate is the indirect activation of
 	the transporter gene by the transporter substrate, leading to positive
@@
 	positive feedback
-loop~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
+	loop~\cite{Novick1957,Santillan2008,Ozbudak2004,Narang2008,Choi2008,Fritz2014,Jenkins2017,Siegele1997}.
 	A eukaryotic example is the glucose transporter GLUT-2 in liver and
@@
 	expression stabilizes at an ``on'' point and the cell has, in a short-term
-sense, adapted. When there is not enough substrate anymore, the operon, which
+	sense, adapted. When there is not enough substrate, the operon, which
 	often encodes for a whole set of enzymes for this one metabolite, switches
@@
 	LacI, AraC, and XylR all appear to fill this
-role~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
-\todo{MJM: do we have a cite for xylR too?}
+	role~\cite{Novick1957,Santillan2008,Ozbudak2004,Narang2008,Choi2008,Fritz2014,Jenkins2017,Siegele1997,Song1997,Schmidt2015}.
 
@@
 	that is induced by xanthosine and activates the \emph{xapAB} promoter,
-in close analogy to LacI, AraC, and XylR.
+	in close analogy to AraC, XylR, and also LacI, although the latter acts via repression.
 	The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
@@
 	\begin{itemize}
-	\item In the initial absense of XapB, small amounts of xanthosine are
-		nonspecifically transported into and out of the cell (discussed in more detail below).
-	\todo{MJM: ``nonspecific transport'' is a bit misleading:
-	it's actually weakly specific transporters, but the precise mechanism is unimportant
-	\& I think it better to emphasize generality of the model. Does this seem fine?}
+		\item In the initial absense of XapB, small amounts of xanthosine are nonspecifically transported\todo{``permeate'' wouldn't include active transport (NupC \& NupG), would it?} into and out of the cell (discussed in more detail below).
 		\item XapR is induced by xanthosine.
@@
 			$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
-		xanthosine with Michaelis-Menten parameters $k_{a}$ and $K_a$.
-		Similarly, XapB imports and exports xanthosine with parameters
-		$k_{\n{b,i}}, K_{\n{b,i}}$ and $k_{\n{b,e}}, K_{\n{b,e}}$,
-		respectively. Furthermore, xanthosine enters and leaves the cell
+			xanthosine with Michaelis-Menten parameters $k_{\n{a}}$ and $K_{\n{a}}$.
+			Similarly, XapB imports ($k_{\n{b,i}}, K_{\n{b,i}}$) and exports ($k_{\n{b,e}}, K_{\n{b,e}}$) xanthosine. Furthermore, xanthosine enters and leaves the cell
 			through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
@@
 	$\langle\n{[mRNA]}\rangle = \frac{r_{\n{m}}}{\gamma_{\n{m}}}
-p_{\n{active}}$, which we will need in the next paragraph.
-In~\nameref{S1_Text}, the full chemical master equation is shown, which
-leads to a \todo{MJM: I worry referees will give us grief: ``Real dists aren't Poisson,'' even though it's irrelevant}
-Poisson distribution of the mRNA concentration with the same
-mean.
+	p_{\n{active}}$, which we will need in the next paragraph. The mean can also be found from the full chemical master equation, which is shown in~\nameref{S1_Text}.
 
@@
 
-\todo[inline]{MJM: Are the following 2 paragraphs excessively detailed for main text?
-maybe replace w/ 1-2 sentences about low affinity Mich-Ment transporters
-that supplement XapB, \& shift all this to supplement?
-If we want to emphasize the general class of model rather than the particular
-biology details of Xap}
-Transport of xanthosine across the cell membrane happens in two steps which
-we will discuss now. All statements in this paragraph are based on data
-from~\cite{Norholm2001}. In a first transport step, xanthosine passes the
-outer membrane through porins like Tsx, OmpF and OmpC. This does not seem to
-be a rate-limiting step. Then, XapB actively transports it across the inner
-membrane, powered by the proton motive force. There are two other active
-nucleoside transporters that seem to be transporting xanthosine with a very
-low affinity: NupC and NupG. It was found that \textDelta\emph{nupC}
-\textDelta\emph{nupG} strains cannot grow on xanthosine. Hence, these seem
-to be necessary to ``start'' the system by importing small amounts of
-xanthosine that then activate \emph{xapB}, and there appears to be no other
-significant way in which xanthosine can enter the cell.
-
-We assume the kinetic scheme of all three transporters to be similar to that
-of \emph{lac} permease (as it is described in~\cite{Kaback2015}). One
-transporter can then do two things to change the intracellular xanthosine
-concentration $\n{[x]}$: it can actively transport one xanthosine into the
-cell, powered by the proton gradient, or it can transport one out of the
-cell, against the proton gradient. For net transport, a proton and a
-substrate need to bind on one side of the membrane and detach from the
-transporter on the other side of the membrane before the transporter goes
-back to the first side again (see also~\nameref{S1_Text}). Because of the
-proton gradient, influx is overall thermodynamically more likely than export
-at low intracellular xanthosine concentrations, which leads to a net influx.
-For much higher intra- than extracellular xanthosine concentrations, the
-difference in the chemical potential of xanthosine across the membrane can
-dominate that of the protons and there is net efflux.
-
-We model influx and efflux separately and use Michaelis-Menten kinetics for
-XapB. For the turnover rate and Michaelis constant for influx we write
-$k_{\n{b, i}}$ and $K_{\n{b, i}}$, respectively, and $k_{\n{b, e}}, K_{\n{b,
-e}}$ for efflux. These parameters implicitly include the proton gradient.
-For NupC and NupG, the Michaelis constant is very large, due to their low
-xanthosine affinity. Thus, the general $k_{\n{cat}} \n{[E]_0}
-\frac{\n{[S]}}{K_{\n{M}} + \n{[S]}}$ can be approximated as
-$\frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0} \n{[S]}$. In the following,
-$\frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0}$ is denoted by $k_{\n{nup}}$ for
-influx, and the efflux rate is written as $\xi k_{\n{nup}}$. For the XapB
-kinetics, no approximations can be made, because the xanthosine
-concentration in the dynamic system can range from far below the respective
-$K_{\n{M}}$ value to far above. The final terms for Nup and XapB are shown
-below.
+	There are two significant ways of transport of xanthosine across the cell membrane.
+In the induced system, the main transporter is XapB, whereas in the uninduced system, there is almost no XapB.
+Instead, xanthosine can enter the cell through the two nucleoside transporters NupC and NupG, which have a very low affinity for xanthosine~\cite{Norholm2001}.
+All these transporters, XapB, NupC, and NupG, are powered by the proton gradient across the membrane~\cite{Norholm2001}, which is why we assume their kinetic scheme to be similar to that of the \emph{lac} permease (as it is described in~\cite{Kaback2015}).
+There can be import and export of xanthosine, and which one dominates depends on the proton and xanthosine concentrations on the two sides of the membrane.
+In both cases, a proton and a substrate need to bind to the transporter on one side of the membrane and detach from it on the other side before the empty transporter moves back to the other side again.
+We refer the reader to~\nameref{S1_Text} for a detailed description of the transport and its modeling, and just state the result here.
+We model influx and efflux separately.
+For XapB, we use Michaelis-Menten kinetics with parameters $k_{\n{b, i}}$, $K_{\n{b, i}}$ for influx and $k_{\n{b, e}}, K_{\n{b,e}}$ for efflux.
+For the Nup transporters, we can linearize because of their low affinity for xanthosine and are left with the rate parameters $k_{\n{nup}}$ for influx and $\xi k_{\n{nup}}$ for efflux.
 
@@
 	standard Michaelis-Menten kinetics, with parameters $k_{\n{a}}, K_{\n{a}}$
-(corresponding to turnover rate and Michaelis constant, respectively). All
-the above then leads to the xanthosine rate equation
+	(corresponding to turnover rate and Michaelis constant, respectively). Transport and degradation then leads to the xanthosine rate equation
 	\begin{eqnarray}
@@
 	\end{eqnarray}
-We recall that $\n{[x]}$ is the intracellular xanthosine concentration, while
+	Recall that $\n{[x]}$ is the intracellular xanthosine concentration, while
 	$c$ denotes the extracellular. Because $k_{\n{b,i}} > k_{\n{b,e}}$ and
@@
 	whose choice is explained below. The plot looks rather complicated at first
-but can be understood intuitively. The three planes are the nullclines and
+	but can be understood intuitively. The three surfaces are the nullcline surfaces and
 	the gray lines point in the direction in which the dynamical system moves at
-each point. The planes intersect in three points, which are the steady-state
+	each point. The surfaces intersect in three points, which are the steady-state
 	solutions of the dynamical system. For this choice of parameters, the system
-first flows towards the mRNA nullcline, then it moves along that plane to
+	first flows towards the mRNA nullcline, then it moves along that surface to
 	the intersection with the protein nullcline, and lastly, it moves along that
 	intersection line to one of the three intersection points of all three
-planes.
+	surfaces.
 
@@
 		\centering
-	\includegraphics[width=1\textwidth]{media/Bistability.eps}
+		\includegraphics{media/Bistability.eps}
 		\caption{{\bf Phase portraits showing bistability.}
@@
 			($\n{[m]_a}, \n{[p]_a}, \n{[x]_a}$) are measured in units of
-		$K_{\n{a}} = 5 \cdot 10^{4} \unit{nM}$. The planes in (A) and the
+			$K_{\n{a}} = 5 \cdot 10^{4} \unit{nM}$. The surfaces in (A) and the
 			lines in (B) are the nullclines of the state variables, and their
@@
 	\end{figure}
-%\todo{edge,size}
 
@@
 	and solve this for $\n{[m]_a}(\n{[p]_a},\n{[x]_a})$ to simplify the 3D to a
-2D system. This restricts the dynamics to the green plane in our plot, which
+	2D system. This restricts the dynamics to the green surface in our plot, which
 	is reasonable here because as explained above, the system first flows
-towards that plane before either the protein or the xanthosine concentration
+	towards that surface before either the protein or the xanthosine concentration
 	changes significantly. However, as already pointed out, this is different
@@
 	\paragraph*{Bistability.} 
-From the 2D plot in Fig~\subref{fig4:bistability}{B}, it can clearly be seen
+	We map the mRNA nullcline surface (green in Fig~\subref{fig4:bistability}{A}) onto a plane to show it as the 2D plot in Fig~\subref{fig4:bistability}{B}.
+	From this 2D plot, it can clearly be seen
 	that for the chosen parameters, there are three steady-state solutions.
-Because the 2D plane that we are in now is the mRNA nullcline, these
-solutions are the same fixed points as in the 3D plot
+	Because the system is restricted to the mRNA nullcline surface, these steady-state solutions are the same as those in the 3D plot
 	($\dd{\n{[m]_a}}{\tau}=0$ on the nullcline and $\dd{\n{[p]_a}}{\tau}=0$,
 	$\dd{\n{[x]_a}}{\tau}=0$ for the 2D fixed points). One can see from the
-vector field that the two outer points (labeled 1 and 3) are stable and the
+	vector field that the two outer fixed points (labeled 1 and 3) are stable and the
 	middle one (labeled 2) is unstable and serves as a sort of ``switch-point''
@@
 
-\todo{MJM: Do any of the older papers have data for this too,
-or were they all just colony growth screens on plates?}
-This corresponds to the experimental observations\cite{Chure2019}, so the model passes this
+	The bistability corresponds to the experimental observations\cite{Chure2019}, so the model passes this
 	sanity check. Furthermore, the xanthosine and protein concentrations at the
@@
 	point -- roughly $1-10 \unit{nM}$ of xanthosine and around $5$ proteins --
-seem perfectly reasonable. Note that $\n{[x]_a} \approx \n{[c]_a}$ at the
+	seem quite reasonable. Note that $\n{[x]_a} \approx \n{[c]_a}$ at the
 	lower fixed point because there is only weak accumulation due to Nup and a
@@
 	here and we will now explain this choice of values. Firstly, they were
-picked such that they are well within their estimated range. Secondly, we
-wanted clear bistability to occur in the phase portraits as well as in the
-stochastic simulations (see later). Thirdly, by the corresponding choice of
+	picked such that they are well within the parameter range that was estimated beforehand (see Table~\ref{table1:nondim}). Secondly, we
+	chose parameters that allow clear bistability in the phase portraits as well as in the
+	stochastic simulations (see later), which, of course, is not the case for any possible choice of parameters. Thirdly, by the corresponding choice of
 	parameters it was ensured that the mRNA number per cell at the
@@
 	numbers that are very reasonable (see~\cite{Milo2016} for the average mRNA
-numbers in cells, which are rather low). The protein and xanthosine
+	numbers in cells). The protein and xanthosine
 	concentrations followed from this, but with some variation in the parameters
@@
 	significantly larger than the XapB rate, but it turns out that oscillations
-cannot be obtained. This can be understood when looking at the regions that
+	cannot be obtained. Why they do not occur can be understood when looking at the regions that
 	are bounded by all three nullclines: on these boundaries, the streamlines
-point into the regions, so deterministically, they serve as
+	point into the bounded regions, so deterministically, they serve as
 	trapping regions from which the system cannot escape. Once inside, the only
@@
 	biologically the most relevant is the extracellular xanthosine
-concentration. When it is increased in experiments, the cells go from all
-being in the low expression state to being in either the low or the high
-expression state (all-or-none phenomenon) to all being in the high
+	concentration. When it is increased in experiments, the cells go from (1.) all
+	being in the low expression state to (2.) the population being in a mixed state with some cells in a low expression state and others in a high expression state (all-or-none phenomenon) and then to (3.) all being in the high
 	expression state~\cite{Chure2019}. If our model is correct, it should
@@
 	stable fixed point appear and then, for even higher $\n{[c]_a}$, the lower
-one disappear. Thus, for low $\n{[c]_a}$ the only stable point of the system
+	one disappears. Thus, for low $\n{[c]_a}$ the only stable point of the system
 	is at low expression, and for high $\n{[c]_a}$ there is only high
@@
 		\centering
-	\includegraphics[width=1\textwidth]{media/VaryingC.eps}
+		\includegraphics{media/VaryingC.eps}
 		\caption{{\bf Phase portraits for different extracellular xanthosine concentrations.}
@@
 			concentration in these plots is $\n{[c]_a}=10^{1}$ in (A) and
-		$\n{[c]_a}= 5 \cdot 10^{2}$ in (B) (recall $\n{[c]_{a}} \defeq
-		\frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$).
+			$\n{[c]_a}= 5 \cdot 10^{2}$ in (B) (recall that $\n{[c]_{a}} \defeq
+			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$ is dimensionless).
 			Tuning $\n{[c]_a}$ moves the orange line (xanthosine nullcline) up
-		or down, the blue line (mRNA nullcline) is unchanged (see
+			or down, the blue curve (mRNA nullcline) is unchanged (see
 			also~\nameref{S1_Text}). It can clearly be seen that in (A) there is
@@
 	\end{figure}
-%\todo{size of plot}
 
 	Furthermore, we found that in the absence of xanthosine, i.e., setting
-$\n{[c]_a}=0$ (not shown here), there are roughly 2-3 proteins, which agrees very
-well with measurements, where around 2 proteins per cell were
+	$\n{[c]_a}=0$ (not shown here), there are roughly 2-3 copies of XapA and XapB, which agrees very
+	well with measurements, where around 2 copies per cell were
 	found~\cite{Li2014}. In addition, the parameter $K_{\n{\chi A}}$
@@
 	regime, does not even have a significant influence on the system. This can
-be seen from the plots in figure~\ref{fig6:xapAB}. XapA lowers the upper
-solution by a small amount and could, in principle, thereby make the
+	be seen from the plots in figure~\ref{fig6:xapAB}. The degradatory effect of XapA lowers the xanthosine and protein concentration at the upper
+	fixed point by a small amount and could, in principle, thereby make the
 	high-expression solution vanish. For our choice of all other parameters,
-this requires $k_{\n{\alpha}} > 10^4$ which is far from what has been
+	it only vanishes for $k_{\n{\alpha}} > 10^4$ which is far from what has been
 	measured. However, a higher effective rate could, in principle, be achieved by
@@
 	\begin{figure}%[h!]
-	\includegraphics[width=1\textwidth]{media/XapAB.eps}
+		\includegraphics{media/XapAB.eps}
 		\centering
@@
 	\end{figure}
-%\todo{size of plot}
-
-For a cell, the above is a useful feature: by coupling XapA and XapB on an
+
+	For a cell, the insignificance of the effect of XapA is a useful feature: by coupling XapA and XapB on an
 	operon, XapA is switched on and off together with XapB but it does not
@@
 	xanthosine. An interesting question to pose is whether the cooperativity is
-a necessary feature for bistability. This is motivated by the importance of
+	a necessary feature for bistability. This question is motivated by the importance of
 	cooperativity in ``typical'' genetic switches\cite{Gardner2000,Cherry2000}.
@@
 	xanthosine binding site on XapR or the second XapR binding site on the
-promoter, we find that the system still has a bistable parameter regime.
-However, this is only a rather narrow parameter range, which makes the system
-\todo{MJM: the (hypothetical) utility of this switch seems to be resource efficiency,
-which $\Rightarrow$ pushing the fixed points to extreme highs \& lows is good
-$\Rightarrow$ stronger coop is good. maybe we should work this in}
-less bistable. When the second binding site is removed in both places, the
+	promoter, leaving cooperativity in one place of the system, we find that the system still has a bistable parameter regime.
+	However, this bistable parameter range is only narrow, which makes the system
+	less stable: small stochastic fluctuations in the parameter values can collapse the system to monostability, possibly leaving it in the wrong state and without its ability to adapt.
+But only when the second binding site is removed in both places, leaving no cooperativity in the system, the
 	equations are insufficiently non-linear to produce bistability. An example
-of these three scenarios can be seen in Fig~\ref{fig7:coop}. It follows that
+	of the three scenarios (only cooperative XapR, only cooperative promoter, no cooperativity) can be seen in Fig~\ref{fig7:coop}. It follows that
 	there need to be either two xanthosine binding sites on XapR or two XapR
@@
 		\centering
-	\includegraphics[width=1\textwidth]{media/FewerSites.eps}
+		\includegraphics{media/FewerSites.eps}
 		\caption{{\bf Phase portraits for less or no cooperativity.}
@@
 	\end{figure}
-%\todo{size of plot}
 
 	One can also ask how much cooperative interaction is needed between the two
-binding sites. For the promoter, this is given by $\Delta E_{\n{coop}}$ in
+	binding sites. For the promoter, the amount of cooperativity is given by $\Delta E_{\n{coop}}$ in
 	our model, and we find that setting $\Delta E_{\n{coop}} = 0$ has almost no
-influence on the phase diagrams. For XapR, the two binding sites interact
+	influence on the phase diagrams. For XapR, we cannot test how much interaction is needed: the two binding sites interact
 	indirectly, because the active state is much likelier if two xanthosine
-molecules are bound. This cooperative interaction can, however, not be tuned
-like in the case of the promoter with $\Delta E_{\n{coop}}$.
+	molecules are bound, and thus there is no tuning parameter for the cooperative interaction
+	like $\Delta E_{\n{coop}}$ in the case of the promoter.
 
@@
 	for a minimal Hill coefficient in our system, because we do not find this
-very insightful. Hill equations only describe scenarios with multiple,
-\todo{MJM: but that already sounds like an effective interaction,
-I'm not sure I like this phrasing, or the next sentence's.}
-non-interacting binding sites where any other state but the fully occupied
-and active or the fully empty and inactive state is not significantly
-populated. In that sense, the Hill coefficient simply denotes the number of
-binding sites and does not reflect how cooperative these binding sites are.
+	very insightful.
+Hill equations only describe some specific limit cases of cooperative systems, but for example do not account for interaction energies and assume the partially bound states (e.g.
+only one XapR bound to the promoter) to never be populated.
 	We suggest that cooperativity should be explored more in-depth and a more
@@
 	\paragraph*{Bimodality and the extracellular xanthosine concentration.}
-This approach results in the same bimodal distributions that were already
+	This stochastic approach results in the same bimodal distributions that were already
 	seen in the deterministic investigation and experimental studies.
@@
 		\centering
-	\includegraphics[width=1\textwidth]{media/HistogramsC.eps}
+		\includegraphics{media/HistogramsC.eps}
 		\caption{{\bf Distributions from stochastic simulations and the corresponding phase portraits.}
@@
 	however, become clear that the phase portraits do not tell whether the cells
-will actually populate both the high and the low expression state: as can be
-seen in Fig~\subref{fig8:stochC}{(A)}, the cells never switched to the high
+	will actually populate both the high and the low expression state, because they do not show the barrier height between the two states.
+In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
 	expression state during the run time of our simulations.
-\todo{MJM: depends on an effective barrier height, hard to intuit from phase port.
-Bistability rather than bimodality means it may very well practically never switch.}
 
@@
 	Fig~\subref{fig8:stochC}{(B)} to give bimodality. For lower $\n{[c]_a}$,
-meaning larger distance between the first and second fixed point, no or
-almost no switching was observed. Of course, this is also a matter of the
-waiting time and stochastic effects: if one waits for long enough, switching
+	meaning larger distance between the first and second fixed point,
+	almost no switching was observed. Of course, switching is also a matter of the
+	waiting time and stochastic effects: if one waits for long enough, it
 	should eventually occur. However, switching times of more than several hours
@@
 	If, however, the mRNA number at the third (high) fixed point is too low, the
-system will not be able to resolve the two points anymore. This becomes a
-problem when the distance between the first and the third fixed point
+	system will not be able to resolve the two points anymore. It is too low when the distance between the first and the third fixed point
 	becomes as low as the stochastic fluctuations in the system, which is around
@@
 	\paragraph*{Time evolution and switching times.}
-Fig~\ref{fig:StochT} shows the time evolution of one specific example.
+	Fig~\ref{fig:StochT} shows the time evolution of the protein concentration for one exemplary run of the simulation.
 	Again, the simulation was started with a mRNA, protein and intracellular
@@
 		\centering
-	\includegraphics[width=0.7\textwidth]{media/TimeEvolution.eps}
+		\includegraphics{media/TimeEvolution.eps}
 		\caption{{\bf Time evolution of protein (XapA/XapB) from one run of the simulation.}
@@
 	\end{figure}
-%\todo{size of plot}
 
@@
 	where the lower two fixed points needed to be about as close as the upper
-two need to be now for ``switching off''. %The extracellular xanthosine
-% concentration (for the other parameters being as in
-% Table~\ref{table1:nondim}) needed to be as low as $\n{[c]_a} = 10$ to
-% observe bimodality and $\n{[c]_a}=9$ to have all cells in the low expression
-% state (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5 \cdot
-% 10^{-5} \unit{M}$).
-\todo[inline]{Is this understandable without seeing the data? Do
-you have any suggestions to make it clearer?
-MJM: I think the quantitative details can go to the supplement where they
-have more context, the qualitative observation of hysteresis is enough here.}
+	two need to be now for ``switching off''.
 
@@
 	state is stable even if the xanthosine concentration decreases to a certain
-extent. This explains what has been observed by Novick and
+	extent. This stability explains what has been observed by Novick and
 	Weiner~\cite{Novick1957} for the \emph{lac} operon: when induced cells were
@@
 	\paragraph*{A genetic switch for metabolic adaptation with remarkable features.}
-In this paper, we propose a model for genetic circuits containing
+	In this paper, we propose a simple model for genetic circuits containing
 	a membrane transporter whose gene expression is, directly or indirectly,
@@
 	dissociation constant between the transcription factor XapR and the inducer
-xanthosine. We found that their interaction needs to be rather weak.
+	xanthosine. We found that their interaction needs to be rather weak compared to usual interactions between an inducer and its target.
 
@@
 	account for the burstiness of transcription and translation as well as for
-cell divisions, which is anything but straightforward. Lastly, some rigorous
-work could be done to identify the most and least relevant parameters of the
-system to create a simpler, minimal model that captures the most relevant
-features.
-\todo[inline]{MJM: I think this last sentence sells us short: qualitative changes result
-if we remove any present feature of the model.
-That sounds to me like a distillation of the essential.}
-
-\paragraph*{General lessons learned for the modeling of other systems.}
-We point out again that assuming the mRNA concentration to be at
-steady-state can have a significant impact on the apparent dynamics of a
-system, as we have seen in the 3-dimensional phase portraits. Moreover, we
-remind the reader that deterministic phase portraits do not show which
-states are actually populated, and thus hide essential effects like
-hysteresis. Lastly, we have exemplified issues with the common picture of
-measuring cooperativity in terms of the Hill coefficient: the Hill
-coefficient simply denotes the number of binding sites and does not reflect
-how cooperative these binding sites are. 
-\todo[inline]{MJM: I'm not sure we've really proven the asserted points re 3D \& Hill.
-I would advocate removing this whole paragraph.}
-
-\paragraph*{A successful model.}
+	cell divisions, which is anything but straightforward.
+
+	\paragraph*{A successful generalizable model.}
 	All model parameters could be reasonably estimated despite the paucity of
-experimental knowledge about the model system. In the results, every
-variable and parameter \todo{MJM: I'm confused,
-we estimated all the model parameters, so what expectations are we comparing to?}
-has its expected order of magnitude, which suggests
-that the model captures all relevant features and is able to describe the
-dynamics of the \emph{xap} system. With the framework given in this text, it
+	experimental knowledge about the model system.
+The concentrations of mRNA, protein, and xanthosine at the fixed points as well as all qualitative features are as expected from similar systems and the few experiments on the \emph{xap} circuit, which suggests that the model captures the relevant components of the system correctly and is able to describe its
+	dynamics.
+Furthermore, the modeling results let us, to some extent, understand why the biological system is constructed the way it is.
+By keeping the model as minimal as possible, but still modeling every part explicitly with an appropriate complexity, we can investigate the interesting features while still being able to understand the influence of all parameters and their interplay intuitively.
+
+	With the framework given in this text, it
 	should be straightforward to model other promoters, regulatory pathways or
@@
 	system.
-
-\todo{Is the conclusion too long? MJM: not necessarily}
 
@@
 	\todo{MJM: need to double-check my NSF \#}
-\todo[inline]{MJM: Software/code acknowledgements usually go in a separate section,
-and if we want we could set up a code repo to cite along with that.}
+	\todo{remove funding again for submission}
 
@@
 
+		\bibitem{Santillan2008}
+		Santillán M, Mackey MC.
+		\newblock Quantitative approaches to the study of bistability in the \emph{lac}
+		operon of \emph{Escherichia coli}.
+		\newblock Journal of The Royal Society Interface. 2008;5(suppl\_1):S29--S39.
+		\newblock doi:{10.1098/rsif.2008.0086.focus}.
+
 		\bibitem{Narang2008}
@@
 
+		\bibitem{Song1997}
+		Song S, Park C.
+		\newblock Organization and regulation of the D-xylose operons in \emph{Escherichia
+		coli} K-12: XylR acts as a transcriptional activator.
+		\newblock Journal of Bacteriology. 1997;179(22):7025--7032.
+		\newblock doi:{10.1128/jb.179.22.7025-7032.1997}.
+
+		\bibitem{Schmidt2015}
+		Schmidt A, Kochanowski K, Vedelaar S, Ahrné E, Volkmer B, Callipo L, et~al.
+		\newblock The quantitative and condition-dependent \emph{Escherichia coli} proteome.
+		\newblock Nature Biotechnology. 2015;34:104.
+
 		\bibitem{Buxton1980}
@@
 
-\todo[inline]{Should I also list the references that are cited in the appendix? MJM: no, I think the supplement has its own separate list.}
-
-\todo[inline]{The capitalization style is different for the titles of the references.
-Should I list them as they are on the corresponding paper or should I do it uniformly?
-Does it matter at all?
-Otherwise I leave it as it is.
-MJM: I think the copy editors will sort it out.}
-
 \end{document}

### 2019-10-29 main_text/main_text.tex
MM edits to main text

@@
 	switch system where the bistability is due, as we will show, to a
-	membrane transport protein.
-Such a switch is common for metabolic processes in biology, for reasons discussed below.
-So far, models we have found are either very extensive descriptions of specific, complicated systems, or they are very basic.
-To enable an intuitive and fundamental understanding of the behavior, we develop a simple model of a minimal system that only has the essential components and interactions for the questions we pose, while still modeling every component explicitly and discussing the necessary model complexity for a physically correct model.
+	membrane transport protein. Such a switch is common for
+	metabolic processes in biology, for reasons discussed below.
+	Existing models in the literature tend towards one of two extremes:
+	either highly detailed descriptions of specific, complicated systems (e.g.,~\cite{Wong1997}),
+	or Hill function descriptions that coarse-grain all complexity
+	into a few parameters with inscrutable microscopic physical meaning.
+	We aim for a middle ground in this work.
+	We seek an intuitive understanding through a simple model of a minimal system,
+	with only the essential components and interactions for the questions we pose.
+	Yet we still model these components explicitly and
+	discuss the necessary model complexity for a physically correct model.
 
@@
 	that is induced by xanthosine and activates the \emph{xapAB} promoter,
-	in close analogy to AraC, XylR, and also LacI, although the latter acts via repression.
+	in close analogy to AraC, XylR, and also LacI.
+	\footnote{One might object that LacI represses its target operon, while
+	XapR, AraC, and XylR activate their target operons. However, the analogy
+	we wish to draw is that the qualitative logic of their inducers are all
+	identical, i.e., the presence of their respective inducer causes their
+	target operon to be transcribed.}
 	The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
@@
 	\begin{itemize}
-		\item In the initial absense of XapB, small amounts of xanthosine are nonspecifically transported\todo{``permeate'' wouldn't include active transport (NupC \& NupG), would it?} into and out of the cell (discussed in more detail below).
+		\item In the initial absense of XapB, small amounts of xanthosine permeate\todo{``permeate'' wouldn't include active transport (NupC \& NupG), would it?
+MM: I think it's sufficiently vague to capture all this, details are too subtle to worry about here} into and out of the cell (discussed in more detail below).
 		\item XapR is induced by xanthosine.
@@
 			xanthosine with Michaelis-Menten parameters $k_{\n{a}}$ and $K_{\n{a}}$.
-			Similarly, XapB imports ($k_{\n{b,i}}, K_{\n{b,i}}$) and exports ($k_{\n{b,e}}, K_{\n{b,e}}$) xanthosine. Furthermore, xanthosine enters and leaves the cell
+			Similarly, XapB is treated as a Michaelis-Menten enzyme which imports ($k_{\n{b,i}}, K_{\n{b,i}}$) and exports ($k_{\n{b,e}}, K_{\n{b,e}}$) xanthosine.
+Furthermore, xanthosine enters and leaves the cell
 			through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
@@
 
-	There are two significant ways of transport of xanthosine across the cell membrane.
-In the induced system, the main transporter is XapB, whereas in the uninduced system, there is almost no XapB.
-Instead, xanthosine can enter the cell through the two nucleoside transporters NupC and NupG, which have a very low affinity for xanthosine~\cite{Norholm2001}.
-All these transporters, XapB, NupC, and NupG, are powered by the proton gradient across the membrane~\cite{Norholm2001}, which is why we assume their kinetic scheme to be similar to that of the \emph{lac} permease (as it is described in~\cite{Kaback2015}).
-There can be import and export of xanthosine, and which one dominates depends on the proton and xanthosine concentrations on the two sides of the membrane.
-In both cases, a proton and a substrate need to bind to the transporter on one side of the membrane and detach from it on the other side before the empty transporter moves back to the other side again.
-We refer the reader to~\nameref{S1_Text} for a detailed description of the transport and its modeling, and just state the result here.
-We model influx and efflux separately.
-For XapB, we use Michaelis-Menten kinetics with parameters $k_{\n{b, i}}$, $K_{\n{b, i}}$ for influx and $k_{\n{b, e}}, K_{\n{b,e}}$ for efflux.
-For the Nup transporters, we can linearize because of their low affinity for xanthosine and are left with the rate parameters $k_{\n{nup}}$ for influx and $\xi k_{\n{nup}}$ for efflux.
+	\todo{MM: do these paragraphs on xanthosine dynamics have an appropriate level of detail now, or too much, or not enough?}
+	There are two significant mechanisms for transport of xanthosine across
+	the cell membrane. In the induced system, the main transporter is XapB,
+	whereas in the uninduced system, there is almost no XapB. Instead,
+	xanthosine can enter the cell through the two nucleoside transporters
+	NupC and NupG, which have a very low affinity for
+	xanthosine~\cite{Norholm2001}. All these transporters, XapB, NupC, and
+	NupG, are powered by the proton gradient across the
+	membrane~\cite{Norholm2001}, which is why we assume their kinetic scheme
+	to be similar to that of the \emph{lac} permease (as it is described
+	in~\cite{Kaback2015}). There can be import and export of xanthosine, and
+	which one dominates depends on the proton and xanthosine concentrations
+	on the two sides of the membrane. In both cases, a proton and a
+	substrate need to bind to the transporter on one side of the membrane
+	and detach from it on the other side before the empty transporter moves
+	back to the other side again. We refer the reader to~\nameref{S1_Text}
+	for a detailed description of the transport and its modeling, and just
+	state the result here. We model influx and efflux separately. For XapB,
+	we use Michaelis-Menten kinetics with parameters $k_{\n{b, i}}$,
+	$K_{\n{b, i}}$ for influx and $k_{\n{b, e}}, K_{\n{b,e}}$ for efflux.
+	For the Nup transporters, we can linearize because of their low affinity
+	for xanthosine and are left with the rate parameters $k_{\n{nup}}$ for
+	influx and $\xi k_{\n{nup}}$ for efflux.
 
@@
 			$\n{[c]_a}= 5 \cdot 10^{2}$ in (B) (recall that $\n{[c]_{a}} \defeq
-			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$ is dimensionless).
+			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$,
+			so $\n{[c]_a}$ is dimensionless).
 			Tuning $\n{[c]_a}$ moves the orange line (xanthosine nullcline) up
-			or down, the blue curve (mRNA nullcline) is unchanged (see
+			or down, but the blue curve (mRNA nullcline) is unchanged (see
 			also~\nameref{S1_Text}). It can clearly be seen that in (A) there is
@@
 	regime, does not even have a significant influence on the system. This can
-	be seen from the plots in figure~\ref{fig6:xapAB}. The degradatory effect of XapA lowers the xanthosine and protein concentration at the upper
+	be seen from the plots in figure~\ref{fig6:xapAB}.
+	Degradation of xanthosine by XapA lowers the xanthosine and protein concentration at the upper
 	fixed point by a small amount and could, in principle, thereby make the
 	high-expression solution vanish. For our choice of all other parameters,
-	it only vanishes for $k_{\n{\alpha}} > 10^4$ which is far from what has been
+	bistability only vanishes for $k_{\n{\alpha}} > 10^4$ which is far from what has been
 	measured. However, a higher effective rate could, in principle, be achieved by
@@
 
-	For a cell, the insignificance of the effect of XapA is a useful feature: by coupling XapA and XapB on an
+	For a cell, the minimal effect of XapA on bistability is a useful feature: by coupling XapA and XapB on an
 	operon, XapA is switched on and off together with XapB but it does not
@@
 	xanthosine binding site on XapR or the second XapR binding site on the
-	promoter, leaving cooperativity in one place of the system, we find that the system still has a bistable parameter regime.
-	However, this bistable parameter range is only narrow, which makes the system
-	less stable: small stochastic fluctuations in the parameter values can collapse the system to monostability, possibly leaving it in the wrong state and without its ability to adapt.
-But only when the second binding site is removed in both places, leaving no cooperativity in the system, the
+	promoter, leaving cooperativity in only one component of the system,
+	we find that the system still has a bistable parameter regime.
+	However, this bistable parameter range is smaller than in the original
+	model, which makes the system less stable: small stochastic fluctuations
+	in the parameter values can collapse the system to monostability,
+	possibly leaving it in the wrong state and without its ability to adapt.
+	But only when the second binding site is removed in both places, leaving
+	no cooperativity in the system, are the
 	equations are insufficiently non-linear to produce bistability. An example
@@
 	indirectly, because the active state is much likelier if two xanthosine
-	molecules are bound, and thus there is no tuning parameter for the cooperative interaction
+	molecules are bound, and thus there is no continuous tuning parameter for the cooperative interaction
 	like $\Delta E_{\n{coop}}$ in the case of the promoter.
@@
 	however, become clear that the phase portraits do not tell whether the cells
-	will actually populate both the high and the low expression state, because they do not show the barrier height between the two states.
+	will actually populate both the high and the low expression state,
+	because they do not show the effective barrier height between the two states.
 	In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
@@
 
+		\bibitem{Wong1997}
+		Wong P, Gladney S, Keasling JD.
+		\newblock Mathematical Model of the \textit{lac} Operon: Inducer Exclusion,
+		Catabolite Repression, and Diauxic Growth on Glucose and Lactose.
+		\newblock Biotechnology Progress. 1997;13(2):132 -- 143.
+		\newblock doi:{https://doi.org/10.1021/bp970003o}.
+
 		\bibitem{Novick1957}

### 2019-11-15 s1_text/s1_text.tex
MJM additions to S1 supplement on memb transporter details

@@
 
-\usepackage[obeyFinal]{todonotes} % todo notes and missing figures
-
 %% END MACROS SECTION
@@
 		\\
-	Kathrin Laxhauber\textsuperscript{1},
+	Kathrin S. Laxhuber\textsuperscript{1},
 	Muir J. Morrison\textsuperscript{2},
@@
 	\bigskip
-	\textbf{1} Dept???, ETHZ, Zurich, State???, Switzerland\todo{MJM: Add dept \& complete affiliation}
+	\textbf{1} Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland
 	\\
@@
 	% Primary Equal Contribution Note
-	% Yinyang These authors contributed equally to this work.
+	%\Yinyang These authors contributed equally to this work.
 
@@
 	% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-	% \ddag These authors also contributed equally to this work.
+	%\ddag These authors also contributed equally to this work.
 
 	% Current address notes
-	% \textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+	%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
 	% \textcurrency b Insert second current address 
 	% \textcurrency c Insert third current address
+
+	% Deceased author note
+	%\dag Deceased
+
+	% Group/Consortium Author Note
+	%\textpilcrow Membership list can be found in the Acknowledgments section.
 
@@
 \section{Model choices, simplifications and assumptions}
-This part of supporting information text offers a more detailed discussion
-of the assumptions that were made in the model and explains the choices that
-were made.
+This part of supporting information text offers a more detailed discussion of the assumptions that were made in the model and explains the choices that were made.
 
-\paragraph*{Simplistic modeling of transcription.}
-The precise mechanism at the promoter level is not the key feature of this
-system, and we expect the same qualitative results for slightly different
-transcription models. This work aims to gain an intuition from the results,
-and the more unknown parameters there are, the harder this becomes.
-Furthermore, the mechanistic details of transcription are not entirely known
-and would make the model too complex: for example, transcription probably is
-not a Poisson process, but comes in bursts. Additionally, very little is
-known about the function of the \emph{xapAB} promoter. For these reasons, we
-do not model every detail, but focus on the most relevant features of the
-system, while still trying to be physically correct.
+\paragraph*{Simplistic modeling of transcription.} The precise mechanism at the promoter level is not the key feature of this system, and we expect the same qualitative results for slightly different transcription models.
+This work aims to gain an intuition from the results, and the more unknown parameters there are, the harder this becomes.
+Furthermore, the mechanistic details of transcription are not entirely known and would make the model too complex: for example, transcription probably is not a Poisson process, but comes in bursts.
+Additionally, very little is known about the function of the \emph{xapAB} promoter.
+For these reasons, we do not model every detail, but focus on the most relevant features of the system, while still trying to be physically correct.
 
-Firstly, the equilibrium assumption is not completely accurate, but has
-proven successful in previous work. An example for this is the very detailed
-analysis of a gene with one repressor in~\cite{Phillips2018}. Secondly,
-different binding energies of XapR to its two binding sites would change the
-equations as follows: the factor $\left(\frac{1}{K_{\n{XapR}}}\right)^2$
-would become $\frac{1}{K_{\n{XapR,1}} K_{\n{XapR,2}}}$ and the factor
-$\frac{2}{K_{\n{XapR}}}$ would be replaced by $\frac{K_{\n{XapR,1}} +
-K_{\n{XapR,2}}}{K_{\n{XapR,1}} K_{\n{XapR,2}}}$. For most choices of
-$K_{\n{XapR}}$, this does not give the exact same results quantitatively,
-but we do observe the same qualitative outcome (see also next paragraph).
-Thirdly, we consider all but one state as inactive, which is not strictly
-correct. Still, as mentioned in the main text, experiments suggest that this
-is reasonable. The fourth main simplification, which is the independence of
-polymerase and XapR binding, will be discussed in a separate subsection
-below.
+Firstly, the equilibrium assumption is not completely accurate, but has proven successful in previous work.
+An example for this is the very detailed analysis of a gene with one repressor in~\cite{Phillips2018}.
+Secondly, different binding energies of XapR to its two binding sites would change the equations as follows: the factor $\left(\frac{1}{K_{\n{XapR}}}\right)^2$ would become $\frac{1}{K_{\n{XapR,1}} K_{\n{XapR,2}}}$ and the factor $\frac{2}{K_{\n{XapR}}}$ would be replaced by $\frac{K_{\n{XapR,1}} + K_{\n{XapR,2}}}{K_{\n{XapR,1}} K_{\n{XapR,2}}}$.
+For most choices of $K_{\n{XapR}}$, this does not give the exact same results quantitatively, but we do observe the same qualitative outcome (see also next paragraph).
+Thirdly, we consider all but one state as inactive, which is not strictly correct.
+Still, as mentioned in the main text, experiments suggest that this is reasonable.
+The fourth main simplification, which is the independence of polymerase and XapR binding, will be discussed in a separate subsection below.
 
-In Fig~\ref{figS1:transcription}, we have plotted phase portraits where some
-of the above assumptions have not been made. For that, we have chosen some
-extreme values for the new parameters to demonstrate that these
-simplifications do not have a significant influence. The two additional
-transcription rates are large compared to experimental observations.
-Moreover, the chosen values of $\n{[XapR]}_{\n{R},i}$ correspond to a
-difference of two orders of magnitude between $K_{\n{XapR,1}}$ and
-$K_{\n{XapR,2}}$. The difference between Oid and O3 in \emph{lac} are two to
-three orders of magnitude~\cite{RazoMejia2018}. The values were chosen such
-that $K_{\n{XapR,1}} K_{\n{XapR,2}} = K_{\n{XapR}}^2$. Unsurprisingly, this
-is necessary to not observe large deviations from the original equations.
-The results lead to the conclusion that, for the purposes of this paper, the
-aforementioned simplifications are appropriate.
+In Fig~\ref{figS1:transcription}, we have plotted phase portraits where some of the above assumptions have not been made.
+For that, we have chosen some extreme values for the new parameters to demonstrate that these simplifications do not have a significant influence.
+The two additional transcription rates are large compared to experimental observations.
+Moreover, the chosen values of $\n{[XapR]}_{\n{R},i}$ correspond to a difference of two orders of magnitude between $K_{\n{XapR,1}}$ and $K_{\n{XapR,2}}$.
+The difference between Oid and O3 in \emph{lac} are two to three orders of magnitude~\cite{RazoMejia2018}.
+The values were chosen such that $K_{\n{XapR,1}} K_{\n{XapR,2}} = K_{\n{XapR}}^2$.
+Unsurprisingly, this is necessary to not observe large deviations from the original equations.
+The results lead to the conclusion that, for the purposes of this paper, the aforementioned simplifications are appropriate.
 
@@
 	\caption{{\bf Results for different models of transcription.}
-		The parameters are the same as the ones that were used for all other
-		phase portraits in this report and that are listed in a table in the
-		main part. In (A), the phase portrait after addition of non-zero
-		transcription rates to all other polymerase-bound states is shown
-		(third assumption lifted). The new parameter values are $r_{\n{I}} =
-		10^{-2} r_{\n{p}}$ for the transcription rate when only one XapR is
-		bound and $r_{\n{0}} = 10^{-3} r_{\n{p}}$ for the basal rate. In
-		(B), the phase portrait after introducing different XapR
-		dissociation constants can be seen (second assumption lifted). The
-		new parameter values are $\n{[XapR]_{R,1}} = 10$ and
-		$\n{[XapR]_{R,2}} = 0.1$.}
+		The parameters are the same as the ones that were used for all other phase portraits in this report and that are listed in a table in the main part.
+In (A), the phase portrait after addition of non-zero transcription rates to all other polymerase-bound states is shown (third assumption lifted).
+The new parameter values are $r_{\n{I}} = 10^{-2} r_{\n{p}}$ for the transcription rate when only one XapR is bound and $r_{\n{0}} = 10^{-3} r_{\n{p}}$ for the basal rate.
+In (B), the phase portrait after introducing different XapR dissociation constants can be seen (second assumption lifted).
+The new parameter values are $\n{[XapR]_{R,1}} = 10$ and $\n{[XapR]_{R,2}} = 0.1$.}
 	\label{figS1:transcription}
@@
 
-\paragraph*{Activator and polymerase interaction.}
-A mechanistically more precise model of activation would include an
-interaction energy between polymerase and XapR. In that case,
-$p_{\n{active}}$ cannot be factored into separate terms for the polymerase
-and the XapR binding probability. However, the precise activation mechanism
-is unknown.
+\paragraph*{Activator and polymerase interaction.} A mechanistically more precise model of activation would include an interaction energy between polymerase and XapR.
+In that case, $p_{\n{active}}$ cannot be factored into separate terms for the polymerase and the XapR binding probability.
+However, the precise activation mechanism is unknown.
 
-Moreover, introducing an interaction energy between polymerase and XapR
-would make the equations overly complex. This would not only make
-factorization impossible, it would also be just as consequential to have
-different dissociation constants and additional rates as in the previous
-paragraph. This leads to a huge number of terms and parameters because of
-different interaction energies for every state, which can be foreseen from
-the states and weights shown in the main text. The precise transcription
-mechanism is not one of the central parts of the circuit we are
-investigating (see also discussion in the previous paragraph). Thus, making
-it the by far most complex and precise part of the equations would be
-unreasonable and hinder the understanding of the more relevant components of
-the system.
+Moreover, introducing an interaction energy between polymerase and XapR would make the equations overly complex.
+This would not only make factorization impossible, it would also be just as consequential to have different dissociation constants and additional rates as in the previous paragraph.
+This leads to a huge number of terms and parameters because of different interaction energies for every state, which can be foreseen from the states and weights shown in the main text.
+The precise transcription mechanism is not one of the central parts of the circuit we are investigating (see also discussion in the previous paragraph).
+Thus, making it the by far most complex and precise part of the equations would be unreasonable and hinder the understanding of the more relevant components of the system.
 
@@
 	\begin{flushleft} 
-		The first column shows which XapR sites are occupied in this state,
-		the second one whether polymerase is bound to the promoter, and the
-		third the transcription rate. In the weights, the notation is as
-		follows: $w_{\n{X1}}$ is the weight for XapR binding to the first
-		binding site on the promoter, $w_{\n{X2}}$ is that to the second
-		site, $w_{\n{p}}$ is the weight of polymerase binding to the
-		promoter, $e^{\Delta \epsilon_{\n{X1}}}$ is the interaction energy
-		between one XapR on the first binding site and the polymerase that
-		is bound, $e^{\Delta \epsilon_{\n{X2}}}$ is the same for the second
-		binding site, and finally, $e^{\Delta \epsilon_{\n{coop}}}$ is the
-		interaction energy of the two XapR at both binding sites as in the
-		main part.
+		The first column shows which XapR sites are occupied in this state, the second one whether polymerase is bound to the promoter, and the third the transcription rate.
+In the weights, the notation is as follows: $w_{\n{X1}}$ is the weight for XapR binding to the first binding site on the promoter, $w_{\n{X2}}$ is that to the second site, $w_{\n{p}}$ is the weight of polymerase binding to the promoter, $e^{\Delta \epsilon_{\n{X1}}}$ is the interaction energy between one XapR on the first binding site and the polymerase that is bound, $e^{\Delta \epsilon_{\n{X2}}}$ is the same for the second binding site, and finally, $e^{\Delta \epsilon_{\n{coop}}}$ is the interaction energy of the two XapR at both binding sites as in the main part.
 	\end{flushleft}
@@
 
-Furthermore, we do not expect our approximation to have a large influence:
-the interaction energy will make the doubly bound state dominate, and the
-polymerase-XapR interaction should be weak in the singly bound state,
-because very weak activation is observed in the experiment when one of the
-XapR binding sites is removed~\cite{Chure2019}. If $e^{\Delta
-\epsilon_{\n{X1}}}$ and $e^{\Delta \epsilon_{\n{X2}}}$ are small, we have
-$w_{\n{p}} w_{\n{X1}} e^{\Delta \epsilon_{\n{X1}}} \approx w_{\n{p}}
-w_{\n{X2}} e^{\Delta \epsilon_{\n{X2}}} \approx w_{\n{p}} w_{\n{X}}$. If, in
-addition, the transcription rates for the singly bound states are much
-smaller than those for the doubly bound state, we can neglect the singly
-bound states in the nominator of $p_{\n{active}}$ (which, as argued in the
-previous paragraph ``Simplistic modeling of transcription'', is a good
-assumption). Thereby, we can obtain the same form of $p_{\n{active}}$ that
-we are working with in the main part by including the interaction between
-polymerase and XapR into $\epsilon_{\n{coop}}$.
+Furthermore, we do not expect our approximation to have a large influence: the interaction energy will make the doubly bound state dominate, and the polymerase-XapR interaction should be weak in the singly bound state, because very weak activation is observed in the experiment when one of the XapR binding sites is removed~\cite{Chure2019}.
+If $e^{\Delta \epsilon_{\n{X1}}}$ and $e^{\Delta \epsilon_{\n{X2}}}$ are small, we have $w_{\n{p}} w_{\n{X1}} e^{\Delta \epsilon_{\n{X1}}} \approx w_{\n{p}} w_{\n{X2}} e^{\Delta \epsilon_{\n{X2}}} \approx w_{\n{p}} w_{\n{X}}$.
+If, in addition, the transcription rates for the singly bound states are much smaller than those for the doubly bound state, we can neglect the singly bound states in the nominator of $p_{\n{active}}$ (which, as argued in the previous paragraph ``Simplistic modeling of transcription'', is a good assumption).
+Thereby, we can obtain the same form of $p_{\n{active}}$ that we are working with in the main part by including the interaction between polymerase and XapR into $\epsilon_{\n{coop}}$.
 
 \paragraph*{Working with only one protein species.}
-The proteins are transcribed together, and thus, their rate of transcription
-and mRNA decay should be the same. The protein decay is assumed to be
-dominated by cell division, making the decay rates roughly equal as well.
-What could differ is the translation rate. Changes in this rate rescale the
-protein synthesis rate by the same factor and thus, in steady-state, change
-the total number of proteins. The only other place where the proteins appear
-in the ODE's is in their respective Michaelis-Menten term, and the latter is
-proportional to $k_{\n{cat}} \n{[E]_0}$. Hence, different numbers of XapA
-and XapB can be accounted for by changing their respective values of
-$k_{\n{cat}}$.
+The proteins are transcribed together, and thus, their rate of transcription and mRNA decay should be the same.
+The protein decay is assumed to be dominated by cell division, making the decay rates roughly equal as well.
+What could differ is the translation rate.
+Changes in this rate rescale the protein synthesis rate by the same factor and thus, in steady-state, change the total number of proteins.
+The only other place where the proteins appear in the ODE's is in their respective Michaelis-Menten term, and the latter is proportional to $k_{\n{cat}} \n{[E]_0}$.
+Hence, different numbers of XapA and XapB can be accounted for by changing their respective values of $k_{\n{cat}}$.
 
 \paragraph*{Exponential protein decay.}
-The protein decay should be dominated by dilution through cell division,
-which is a discrete process. However, the cell grows before division,
-resulting in a continuous dilution. Furthermore, replication leads to a
-varying number of gene copies and thereby also a varying transcription rate.
-Trying to capture all these mechanistic details would lead to an
-overcomplicated model. Treating the protein decay as a Poisson process has
-produced correct results in past research, which is why we expect this
-assumption to be reasonable.
+The protein decay should be dominated by dilution through cell division, which is a discrete process.
+However, the cell grows before division, resulting in a continuous dilution.
+Furthermore, replication leads to a varying number of gene copies and thereby also a varying transcription rate.
+Trying to capture all these mechanistic details would lead to an overcomplicated model.
+Treating the protein decay as a Poisson process has produced correct results in past research, which is why we expect this assumption to be reasonable.
 
 \paragraph*{Kinetics of XapB, NupC, and NupG.}
-All transporters in our system are driven by the proton
-gradient~\cite{Norholm2001}, and we assume their kinetic scheme to be like
-that of \emph{lac} permease. This scheme is shown schematically in
-Fig~\ref{figS2:permease}. Even though it includes several steps, the
-essential part is substrate binding, conformational change of the
-transporter, and substrate release on the other side of the membrane. We
-assume the proton gradient across the membrane to be constant and proton
-binding and unbinding to the transporter to be fast. Hence, the latter does
-not need to be included explicitly in the kinetic model (it is implicitly
-part of the turnover rate).
+Entering the cell consists of two steps. In a first transport step, xanthosine passes the
+outer membrane through porins like Tsx, OmpF and OmpC. This does not seem to
+be a rate-limiting step. Then, XapB actively transports it across the inner
+membrane, powered by the proton motive force. There are two other active
+nucleoside transporters that seem to be transporting xanthosine with a very
+low affinity: NupC and NupG. It was found that \textDelta\emph{nupC}
+\textDelta\emph{nupG} strains cannot grow on xanthosine. Hence, these seem
+to be necessary to ``start'' the system by importing small amounts of
+xanthosine that then activate \emph{xapB}, and there appears to be no other
+significant way in which xanthosine can enter the cell. All statements in this paragraph are based on data
+from~\cite{Norholm2001}.
+
+As mentioned in the main text, we assume the kinetic scheme of all three transporters to be similar to that
+of \emph{lac} permease (as it is described in~\cite{Kaback2015}), which is discussed in more detail in the next subsection.
+One transporter of this type can do two things to change the intracellular xanthosine
+concentration $\n{[x]}$: it can actively transport one xanthosine into the
+cell, powered by the proton gradient, or it can transport one out of the
+cell, against the proton gradient. For net transport, a proton and a
+substrate need to bind on one side of the membrane and detach from the
+transporter on the other side of the membrane before the transporter goes
+back to the first side again (see next subsection). Because of the
+proton gradient, influx is overall thermodynamically more likely than export
+at low intracellular xanthosine concentrations, which leads to a net influx.
+For much higher intra- than extracellular xanthosine concentrations, the
+difference in the chemical potential of xanthosine across the membrane can
+dominate that of the protons and there is net efflux.
+
+We model influx and efflux separately and use Michaelis-Menten kinetics for
+XapB (this is discussed in more detail in the next subsection). For the turnover rate and Michaelis constant for influx we write
+$k_{\n{b, i}}$ and $K_{\n{b, i}}$, respectively, and $k_{\n{b, e}}, K_{\n{b,
+		e}}$ for efflux. These parameters implicitly include the proton gradient.
+
+For NupC and NupG, we approximate the transport reactions as linearly dependent on the substrate.
+This should be appropriate, since their affinity for xanthosine is so low that it cannot even be measured properly~\cite{Norholm2001}, which makes $K_{\n{M}}$ much larger than $c$ or $\n{[x]}$ will ever be.
+Thus, the general $k_{\n{cat}} \n{[E]_0}
+\frac{\n{[S]}}{K_{\n{M}} + \n{[S]}}$ can be approximated as
+$\frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0} \n{[S]}$. In our equations,
+$\frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0}$ is denoted by $k_{\n{nup}}$ for
+influx, and the efflux rate is written as $\xi k_{\n{nup}}$. For the XapB
+kinetics, no such approximation can be made, because the xanthosine
+concentration in the dynamic system can range from far below the respective
+$K_{\n{M}}$ value to far above.
+
+\paragraph*{The kinetics of \emph{lac} permease.}
+All transporters in our system are driven by the proton gradient~\cite{Norholm2001}, and we assume their kinetic scheme to be like that of \emph{lac} permease.
+This scheme is shown schematically in Fig~\ref{figS2:permease}.
+Even though it includes several steps, the essential part is substrate binding, conformational change of the transporter, and substrate release on the other side of the membrane.
+We assume the proton gradient across the membrane to be constant and proton binding and unbinding to the transporter to be fast.
+Hence, the latter does not need to be included explicitly in the kinetic model (it is implicitly part of the turnover rate).
 
@@
 	\caption{{\bf Kinetic scheme for transport by proton-driven pumps.}
-		The mechanism of lac permease as for example described
-		in~\cite{Kaback2015} is used as a model for xanthosine transport.
-		The enzyme binds a proton and a substrate (steps 1 and 2), undergoes
-		conformational change (steps 3 and 4), and then releases substrate
-		and proton on the other side of the membrane (steps 5 and 6). When
-		the enzyme is empty, it can change its conformation again to repeat
-		the transport (steps 7 and 8). Instead, it could also bind a new
-		proton and substrate and transport them in the opposite direction
-		(steps 6, 5, 4, 3, 2, 1). Alternatively, it could also keep the
-		proton and just cycle back and forth, exchanging substrates from
-		outside and inside the cell (steps 5, 4, 3, 2).}
+		The mechanism of lac permease as for example described in~\cite{Kaback2015} is used as a model for xanthosine transport.
+The enzyme binds a proton and a substrate (steps 1 and 2), undergoes conformational change (steps 3 and 4), and then releases substrate and proton on the other side of the membrane (steps 5 and 6).
+When the enzyme is empty, it can change its conformation again to repeat the transport (steps 7 and 8).
+Instead, it could also bind a new proton and substrate and transport them in the opposite direction (steps 6, 5, 4, 3, 2, 1).
+Alternatively, it could also keep the proton and just cycle back and forth, exchanging substrates from outside and inside the cell (steps 5, 4, 3, 2).}
 	\label{figS2:permease}
@@
 
-Substrate release should be much faster than conformational change, which is
-analogous to ignoring the ``EP'' state in Michaelis-Menten (enzyme with
-product bound, here the conformationally changed transporter with substrate
-still bound). Also, it should be reasonable to neglect the immediate
-backwards reaction (before proton release), and thus, Michaelis-Menten
-kinetics is obtained. Data from~\cite{Kaczorowski1979}
-and~\cite{Viitanen1984} actually show that, at a physiologically reasonable
-proton gradient, both, influx and efflux, can be described by the
-Michaelis-Menten equation very well.
+Substrate release should be much faster than conformational change, which is analogous to ignoring the ``EP'' state in Michaelis-Menten (enzyme with product bound, here the conformationally changed transporter with substrate still bound).
+Also, it should be reasonable to neglect the immediate backwards reaction (before proton release), and thus, Michaelis-Menten kinetics is obtained.
+Data from~\cite{Kaczorowski1979} and~\cite{Viitanen1984} actually show that, at a physiologically reasonable proton gradient, both, influx and efflux, can be described by the Michaelis-Menten equation very well.
 
-What is neglected in this modeling is the fact that the transporter needs to
-release the proton and undergo a conformational change again before being
-able to conduct the same reaction again. However, by using experimentally
-determined apparent values to estimate the kinetic constants, this is
-already included in these parameter values. Another simplification we make
-is to treat influx and efflux as two independent reactions instead of
-thinking of it as one reversible kinetic process. This seems reasonable,
-though, because the transporter needs to lose its proton between reactions
-in order to obtain a net transport. If the transporter has for example done
-steps 1-6 in Fig~\ref{figS2:permease}, it can from there on really enter two
-entirely different reactions: it can do its conformational change in its
-empty state (steps 7 and 8) to then transport a substrate again (steps 1-6)
-or it can bind a new proton and a new substrate directly and transport them
-back (steps 6-1). The former would give influx and the latter efflux, so the
-two of them can be modeled as separate reactions.
-
-The NupC and NupG reactions are approximated as linearly dependent on the
-substrate. This should be appropriate, since their affinity for xanthosine
-is so low that it cannot even be measured properly~\cite{Norholm2001}. This
-makes $K_{\n{M}}$ much larger than $c$ or $\n{[x]}$ will ever be. For XapB,
-this is not the case, because it has much higher specificity for xanthosine
-and $c$ and $\n{[x]}$ take values around the $K_{\n{M}}$ values of XapB.
+What is neglected in this modeling is the fact that the transporter needs to release the proton and undergo a conformational change again before being able to conduct the same reaction again.
+However, by using experimentally determined apparent values to estimate the kinetic constants, this is already included in these parameter values.
+Another simplification we make is to treat influx and efflux as two independent reactions instead of thinking of it as one reversible kinetic process.
+This seems reasonable, though, because the transporter needs to lose its proton between reactions in order to obtain a net transport.
+If the transporter has for example done steps 1-6 in Fig~\ref{figS2:permease}, it can from there on really enter two entirely different reactions: it can do its conformational change in its empty state (steps 7 and 8) to then transport a substrate again (steps 1-6) or it can bind a new proton and a new substrate directly and transport them back (steps 6-1).
+The former would give influx and the latter efflux, so the two of them can be modeled as separate reactions.
 
 \paragraph*{Passive transport across the membrane.}  
-Most small molecules diffuse passively through the membrane, but at
-physiological pH, xanthosine is mostly negatively charged and charged
-substances do not significantly diffuse across the membrane on their own.
-Nevertheless, from the $\n{pK_a}$ of xanthosine we find that at
-physiological pH, about 1.5\% of it is uncharged and could pass the
-membrane. This could be modeled as follows:
+Most small molecules diffuse passively through the membrane, but at physiological pH, xanthosine is mostly negatively charged and charged substances do not significantly diffuse across the membrane on their own.
+Nevertheless, from the $\n{pK_a}$ of xanthosine we find that at physiological pH, about 1.5\% of it is uncharged and could pass the membrane.
+This could be modeled as follows:
 \begin{eqnarray}
@@
 \end{eqnarray}
-Here, $J$ is the flux through the membrane, which follows directly from
-Fick's first law (if steady-state diffusion is assumed). The other new
-parameters are the membrane area $A$ and the membrane permeability
-coefficient $p$ of the diffusing substrate.
+Here, $J$ is the flux through the membrane, which follows directly from Fick's first law (if steady-state diffusion is assumed).
+The other new parameters are the membrane area $A$ and the membrane permeability coefficient $p$ of the diffusing substrate.
 
-The diffusion rate $k_{\n{d}} = p A$ can be estimated with the help of
-Overton's rule. In our model, we find that this passive diffusion alone
-(i.e. without NupC and NupG) is enough to obtain bistability in the
-deterministic analysis as well as bimodality in the stochastic simulations.
-However, experiments show that \textDelta\emph{nupC} \textDelta\emph{nupG}
-strains cannot grow on xanthosine~\cite{Norholm2001}. This suggests that
-passive diffusion alone is not enough to make the cells switch. Experiments
-with cells in minimal media containing some glucose and a xanthosine
-concentration in the bimodal regime show that the switching time can be
-several generations. The same is observed in our simulations. The xanthosine
-concentration that was used for the double knockout experiments, where no
-glucose was present, is just above the switching threshold, which is when
-switching is slow. For this reason, we suspect that without NupC and NupG
-and with only xanthosine as an energy resource, the cells starve before they
-have switched. This probably is the reason why passive diffusion is, in
-reality, not enough to observe switching. NupC and NupG can probably keep
-the cells going at very low intracellular xanthosine concentrations. Because
-of the active transport, they can accumulate a higher intracellular than
-extracellular concentration of xanthosine.
+The diffusion rate $k_{\n{d}} = p A$ can be estimated with the help of Overton's rule.
+In our model, we find that this passive diffusion alone (i.e.
+without NupC and NupG) is enough to obtain bistability in the deterministic analysis as well as bimodality in the stochastic simulations.
+However, experiments show that \textDelta\emph{nupC} \textDelta\emph{nupG} strains cannot grow on xanthosine~\cite{Norholm2001}.
+This suggests that passive diffusion alone is not enough to make the cells switch.
+Experiments with cells in minimal media containing some glucose and a xanthosine concentration in the bimodal regime show that the switching time can be several generations.
+The same is observed in our simulations.
+The xanthosine concentration that was used for the double knockout experiments, where no glucose was present, is just above the switching threshold, which is when switching is slow.
+For this reason, we suspect that without NupC and NupG and with only xanthosine as an energy resource, the cells starve before they have switched.
+This probably is the reason why passive diffusion is, in reality, not enough to observe switching.
+NupC and NupG can probably keep the cells going at very low intracellular xanthosine concentrations.
+Because of the active transport, they can accumulate a higher intracellular than extracellular concentration of xanthosine.
 
-Since NupC and NupG transport seems to be much more significant than
-diffusion, we do not include the latter explicitly in our model. However,
-because it takes the form given in Eq~\ref{eq:diff}, it can easily be
-accounted for implicitly by slightly changing the values of $k_{\n{nup}}$
-and $\xi$.
+Since NupC and NupG transport seems to be much more significant than diffusion, we do not include the latter explicitly in our model.
+However, because it takes the form given in Eq~\ref{eq:diff}, it can easily be accounted for implicitly by slightly changing the values of $k_{\n{nup}}$ and $\xi$.
 
 \paragraph*{Other mechanisms for xanthosine degradation.}
-There probably are some ways in which xanthosine is degraded, other than by
-XapA. Otherwise, initially uninduced cells that are transferred to only
-xanthosine as an energy source might not survive until they have switched.
-As mentioned previously, it was found that they do~\cite{Norholm2001}, so
-they need to be able to metabolize xanthosine somehow. Nevertheless, these
-degradation pathways cannot have a big effect: as argued later on when
-estimating the Nup parameters, the intracellular xanthosine concentration in
-the uninduced state must be a bit larger, but not too much larger, than the
-extracellular one. Thus, degradation of xanthosine has to be small because
-import through Nup is weak. Due to this, it should be appropriate to
-linearize the equations describing these effects. Then, they can simply be
-accounted for in the term for Nup transport by making slight changes in the
-values of $\xi$ and $k_{\n{nup}}$. It could, however, also be that the basal
-level of XapA is enough to obtain the small degradation effect without any
-other pathways.
+There probably are some ways in which xanthosine is degraded, other than by XapA.
+Otherwise, initially uninduced cells that are transferred to only xanthosine as an energy source might not survive until they have switched.
+As mentioned previously, it was found that they do~\cite{Norholm2001}, so they need to be able to metabolize xanthosine somehow.
+Nevertheless, these degradation pathways cannot have a big effect: as argued later on when estimating the Nup parameters, the intracellular xanthosine concentration in the uninduced state must be a bit larger, but not too much larger, than the extracellular one.
+Thus, degradation of xanthosine has to be small because import through Nup is weak.
+Due to this, it should be appropriate to linearize the equations describing these effects.
+Then, they can simply be accounted for in the term for Nup transport by making slight changes in the values of $\xi$ and $k_{\n{nup}}$.
+It could, however, also be that the basal level of XapA is enough to obtain the small degradation effect without any other pathways.
 
 \section{Parameter estimation}
-In this section we present a detailed estimation of the values of all
-nondimensional parameters. Two dimensionful parameters appear in most
-nondimensional ones, namely $\gamma_{\n{p}}$ and $K_{\n{a}}$. For this
-reason, we start by estimating these.
+In this section we present a detailed estimation of the values of all nondimensional parameters.
+Two dimensionful parameters appear in most nondimensional ones, namely $\gamma_{\n{p}}$ and $K_{\n{a}}$.
+For this reason, we start by estimating these.
 
 \paragraph*{The protein decay rate $\bm{\gamma_{\n{p}}}$:}
-Protein decay is generally dominated by dilution through cell growth and
-division. The average doubling time of \emph{E. coli} is 20-30 minutes. We
-take this as an estimate for the protein half life, which leads us to a
-decay rate of $\gamma_{\n{p}} = \frac{\ln 2}{t_{1/2}} \approx 5\substack{+2
-\\ -3} \cdot 10^{-4}~\n{s^{-1}}$. Note that these upper and lower bounds
-correspond to half-lifes between 15 and 60 minutes, which should be
-reasonable. For the following nondimensionalization, we work with a fixed
-value of $\gamma_{\n{p}} = 5 \cdot 10^{-4}~\n{s^{-1}}$.
+Protein decay is generally dominated by dilution through cell growth and division.
+The average doubling time of \emph{E.
+coli} is 20-30 minutes.
+We take this as an estimate for the protein half life, which leads us to a decay rate of $\gamma_{\n{p}} = \frac{\ln 2}{t_{1/2}} \approx 5\substack{+2 \\ -3} \cdot 10^{-4}~\n{s^{-1}}$.
+Note that these upper and lower bounds correspond to half-lifes between 15 and 60 minutes, which should be reasonable.
+For the following nondimensionalization, we work with a fixed value of $\gamma_{\n{p}} = 5 \cdot 10^{-4}~\n{s^{-1}}$.
 
 \paragraph*{The Michaelis-Menten parameters for XapA ($K_{\n{a}}$, $k_{\n{\alpha}}$):}
-The Michaelis-Menten parameters of XapA kinetics have been measured in two
-independent publications~\cite{Dandanell2005,Koszalka1988}. Both have found
-similar values and we conclude $K_{\n{a}} \approx 6 \pm 3 \cdot
-10^{-5}~\n{M}$ and $k_{\n{a}} \approx 10^{-1 \pm 0.8}~\n{s^{-1}}$. This
-gives $k_{\n{\alpha}} \approx 10^{2 \pm 0.8}$. We choose to work with
-$K_{\n{a}} = 5 \cdot 10^{-5}~\n{M}$ in the following.
+The Michaelis-Menten parameters of XapA kinetics have been measured in two independent publications~\cite{Dandanell2005,Koszalka1988}.
+Both have found similar values and we conclude $K_{\n{a}} \approx 6 \pm 3 \cdot 10^{-5}~\n{M}$ and $k_{\n{a}} \approx 10^{-1 \pm 0.8}~\n{s^{-1}}$.
+This gives $k_{\n{\alpha}} \approx 10^{2 \pm 0.8}$.
+We choose to work with $K_{\n{a}} = 5 \cdot 10^{-5}~\n{M}$ in the following.
 
-\paragraph*{The extracellular xanthosine concentration $\n{[c]_{a}}$:} In
-the experiment, switching occurred if cells were grown in a solution with a
-concentration of xanthosine of roughly a few mM~\cite{Norholm2001}.
-According to these experimental observations, the interesting range of
-xanthosine should be $\n{[c]_{a}} \in [ 0, 10^3 ]$ (nondimensionalized),
-which corresponds to $\n{c} \in [ 0, 5 \cdot 10^{-2} ]~\n{M}$. However, we
-are careful because there are issues with the solubility of xanthosine.
+\paragraph*{The extracellular xanthosine concentration $\n{[c]_{a}}$:} In the experiment, switching occurred if cells were grown in a solution with a concentration of xanthosine of roughly a few mM~\cite{Norholm2001}.
+According to these experimental observations, the interesting range of xanthosine should be $\n{[c]_{a}} \in [ 0, 10^3 ]$ (nondimensionalized), which corresponds to $\n{c} \in [ 0, 5 \cdot 10^{-2} ]~\n{M}$.
+However, we are careful because there are issues with the solubility of xanthosine.
 
 \paragraph*{The MWC parameters $K_{\n{\chi A}}$, $K_{\n{IA}}$, and $\Delta \epsilon_{\n{x}}$:}
-The three MWC parameters are the energy barrier between the active and
-inactive state of XapR and the equilibrium constants of xanthosine binding
-to XapR in the two states. In Fig~\ref{figS2:induction}, the induction curve
-as a function of the nondimensional parameters is shown. We assume that
-evolution has tuned the induction curve such that less than $10 \%$ of XapR
-is active in the absence of xanthosine, and more than $90 \%$ is active for
-$\n{[x]}$ approaching infinity. This is a relatively weak assumption,
-otherwise XapR would not really be fulfilling its job. It helps us by
-putting a constraint on $K_{\n{IA}}$ and $\Delta \epsilon_{\n{x}}$:
+The three MWC parameters are the energy barrier between the active and inactive state of XapR and the equilibrium constants of xanthosine binding to XapR in the two states. 
+In Fig~\ref{figS2:induction}, the induction curve as a function of the nondimensional parameters is shown.
+We assume that evolution has tuned the induction curve such that less than $10 \%$ of XapR is active in the absence of xanthosine, and more than $90 \%$ is active for $\n{[x]}$ approaching infinity.
+This is a relatively weak assumption, otherwise XapR would not really be fulfilling its job.
+It helps us by putting a constraint on $K_{\n{IA}}$ and $\Delta \epsilon_{\n{x}}$:
 \begin{alignat*}{1}
@@
 
-A further constraint is found from $K_{\n{IA}}=\frac{K_{\n{x I}}}{K_{\n{x
-A}}} = e^{\beta \Delta E}$, where $\Delta E$ is the energy difference
-between xanthosine binding to the two states. It is unlikely that this
-energy difference is more than $7~k_{\n{B}} T$, which corresponds to roughly
-three strong hydrogen bonds. This gives us an upper bound $K_{\n{IA}} <
-10^3$ and thereby $\Delta \epsilon_{\n{x}} < 10^1$. For $K_{\n{IA}} = 10^2$,
-the upper bound is $\Delta \epsilon_{\n{x}} < 7$. This gives us a range of
-$K_{\n{IA}} \approx 10^1 - 10^3$ and $\Delta \epsilon_{\n{x}} \approx 2 -
-10$ or, more precisely, $\Delta \epsilon_{\n{x}} \approx 2$ to
-$\ln\left(\frac{1}{9} \left(K_{\n{IA}}\right)^2 \right) \approx 2 \left(
-\ln\left(K_{\n{IA}}\right) - 1\right) < 12$. Compared with other systems,
-these values do seem reasonable.
+A further constraint is found from $K_{\n{IA}}=\frac{K_{\n{x I}}}{K_{\n{x A}}} = e^{\beta \Delta E}$, where $\Delta E$ is the energy difference between xanthosine binding to the two states.
+It is unlikely that this energy difference is more than $7~k_{\n{B}} T$, which corresponds to roughly three strong hydrogen bonds.
+This gives us an upper bound $K_{\n{IA}} < 10^3$ and thereby $\Delta \epsilon_{\n{x}} < 10^1$.
+For $K_{\n{IA}} = 10^2$, the upper bound is $\Delta \epsilon_{\n{x}} < 7$.
+This gives us a range of $K_{\n{IA}} \approx 10^1 - 10^3$ and $\Delta \epsilon_{\n{x}} \approx 2 - 10$ or, more precisely, $\Delta \epsilon_{\n{x}} \approx 2$ to $\ln\left(\frac{1}{9} \left(K_{\n{IA}}\right)^2 \right) \approx 2 \left( \ln\left(K_{\n{IA}}\right) - 1\right) < 12$.
+Compared with other systems, these values do seem reasonable.
 
@@
 	\includegraphics[width=0.85\textwidth]{FigSI3.pdf}
-	\caption{{\bf Induction curves given by MWC model.}
-	The main plot shows the probability of XapR being in the active state as
-	a function of the logarithm of the xanthosine concentration for $\Delta
-	\epsilon_{\n{x}}$ as shown, $K_{\n{\chi A}} = 10^{2}$, and $K_{\n{IA}} =
-	10^{2}$. The shaded areas indicate where the curve lies for changes in
-	$K_{\n{IA}}$ of one order of magnitude. Changing $K_{\n{\chi A}}$ by one
-	order of magnitude, on the other hand, shifts all the curves by one
-	order of magnitude in the corresponding direction to the left or right.
-	The inset shows how the allowed regimes of $\Delta \epsilon_{\n{x}}$ and
-	$K_{\n{IA}}$ depend on each other. This explains why for $\Delta
-	\epsilon_{\n{x}} = 7$ and $K_{\n{IA}}=10$ (lower green line in the main
-	plot), $p_{\n{active}}$ at infinite xanthosine concentrations becomes
-	far too low.}
+	\caption{{\bf Induction curves given by MWC model.} The main plot shows the probability of XapR being in the active state as a function of the logarithm of the xanthosine concentration for $\Delta \epsilon_{\n{x}}$ as shown, $K_{\n{\chi A}} = 10^{2}$, and $K_{\n{IA}} = 10^{2}$.
+The shaded areas indicate where the curve lies for changes in $K_{\n{IA}}$ of one order of magnitude.
+Changing $K_{\n{\chi A}}$ by one order of magnitude, on the other hand, shifts all the curves by one order of magnitude in the corresponding direction to the left or right.
+The inset shows how the allowed regimes of $\Delta \epsilon_{\n{x}}$ and $K_{\n{IA}}$ depend on each other.
+This explains why for $\Delta \epsilon_{\n{x}} = 7$ and $K_{\n{IA}}=10$ (lower green line in the main plot), $p_{\n{active}}$ at infinite xanthosine concentrations becomes far too low.}
 	\label{figS2:induction}
@@
 
-To find $K_{\n{\chi A}}$ we furthermore assume that the steep part of the
-induction curve lies in the biologically relevant xanthosine regime, which
-estimated experimentally, see the paragraph about $\n{[c]_a}$ above. Again,
-this is a rather weak assumption. As can be seen from
-Fig~\ref{figS2:induction}, this assumption sets the value range of
-$K_{\n{\chi A}}$. The line for $\Delta \epsilon_{\n{x}} = 5$ in
-Fig~\ref{figS2:induction} is at the experimentally expected position in the
-xanthosine regime. It becomes clear from the figure that the position of the
-steep range is strongly affected by $\Delta \epsilon_{\n{x}}$, and a change
-of 1 in the latter leads to roughly a change of $10^1$ in the former. Thus,
-we estimate $K_{\n{\chi A}} \approx 10^{2 \pm 1} \cdot 10^{\Delta
-\epsilon_{\n{x}}-5}$ and thereby finish our estimation of the MWC
-parameters.
+To find $K_{\n{\chi A}}$ we furthermore assume that the steep part of the induction curve lies in the biologically relevant xanthosine regime, which estimated experimentally, see the paragraph about $\n{[c]_a}$ above.
+Again, this is a rather weak assumption.
+As can be seen from Fig~\ref{figS2:induction}, this assumption sets the value range of $K_{\n{\chi A}}$.
+The line for $\Delta \epsilon_{\n{x}} = 5$ in Fig~\ref{figS2:induction} is at the experimentally expected position in the xanthosine regime.
+It becomes clear from the figure that the position of the steep range is strongly affected by $\Delta \epsilon_{\n{x}}$, and a change of 1 in the latter leads to roughly a change of $10^1$ in the former.
+Thus, we estimate $K_{\n{\chi A}} \approx 10^{2 \pm 1} \cdot 10^{\Delta \epsilon_{\n{x}}-5}$ and thereby finish our estimation of the MWC parameters.
 
 \paragraph*{The XapR concentration $\n{[XapR]_R}$:} 
-It is experimentally known that the XapR copy number is very low, on the
-order of $10^{-8}~\n{M}$ ($10^1$ molecules per cell)~\cite{Li2014}. We
-estimate the XapR -- DNA dissociation constant from the values for
-\emph{lac} that were found in~\cite{RazoMejia2018} and obtain $10^{-8 \pm
-2}~\n{M}$, where we have added one additional order of magnitude as
-insecurity to the upper and lower bound. This gives
-$\n{[XapR]_R}=\frac{\n{[XapR]_{\n{tot}}}}{K_{\n{XapR}}} \approx 10^{0 \pm
-2}$.
+It is experimentally known that the XapR copy number is very low, on the order of $10^{-8}~\n{M}$ ($10^1$ molecules per cell)~\cite{Li2014}.
+We estimate the XapR -- DNA dissociation constant from the values for \emph{lac} that were found in~\cite{RazoMejia2018} and obtain $10^{-8 \pm 2}~\n{M}$, where we have added one additional order of magnitude as insecurity to the upper and lower bound.
+This gives $\n{[XapR]_R}=\frac{\n{[XapR]_{\n{tot}}}}{K_{\n{XapR}}} \approx 10^{0 \pm 2}$.
 
 \paragraph*{The XapR cooperative energy $\Delta \epsilon_{\n{coop}}$:}
-Any cooperative energy should come mainly from an interaction between the
-two XapR molecules. It would be surprising if it was larger than $10 \
-k_{\n{B}} T$, which corresponds to roughly 4 strong hydrogen bonds. Hence,
-we estimate $\Delta \epsilon_{\n{coop}} \approx 0 - 10$.
+Any cooperative energy should come mainly from an interaction between the two XapR molecules.
+It would be surprising if it was larger than $10 \ k_{\n{B}} T$, which corresponds to roughly 4 strong hydrogen bonds.
+Hence, we estimate $\Delta \epsilon_{\n{coop}} \approx 0 - 10$.
 
-\paragraph*{The transcription rate parameter $\rho_{\n{m}}$:} The
-transcription rate $r_{\n{m}} \frac{\n{[P]}}{\n{[P]} + K_{\n{P}}}$ in the
-presence of two activator molecules should be on the order of $10^{-2 \pm
-1}~\n{nM\ s^{-1}}$, thus $\rho_{\n{m}} \defeq
-\frac{r_{\n{m}}}{\gamma_{\n{p}} K_{\n{a}}} \frac{\n{[P]}}{K_{\n{P}}+\n{[P]}}
-\approx 10^{-3 \pm 2}$. Note that the units are $\n{nM\ s^{-1}}$ and not
-just $\n{s^{-1}}$ because the corresponding rate equation is of the form
-$\dd{\n{[m]}}{t} = r \cdot p$ where the probability $p$ is dimensionless and
-the left-hand side has dimensions of mRNA concentration per time.
+\paragraph*{The transcription rate parameter $\rho_{\n{m}}$:} The transcription rate $r_{\n{m}} \frac{\n{[P]}}{\n{[P]} + K_{\n{P}}}$ in the presence of two activator molecules should be on the order of $10^{-2 \pm 1}~\n{nM\ s^{-1}}$, thus $\rho_{\n{m}} \defeq \frac{r_{\n{m}}}{\gamma_{\n{p}} K_{\n{a}}} \frac{\n{[P]}}{K_{\n{P}}+\n{[P]}} \approx 10^{-3 \pm 2}$.
+Note that the units are $\n{nM\ s^{-1}}$ and not just $\n{s^{-1}}$ because the corresponding rate equation is of the form $\dd{\n{[m]}}{t} = r \cdot p$ where the probability $p$ is dimensionless and the left-hand side has dimensions of mRNA concentration per time.
 
-This estimate comes from the assumption that this rate should be roughly
-$10^{1} - 10^{2}$ times larger than a normal basal rate (of other genes),
-for which common values are around $10^{-3 \pm 1}~\n{nM\
-s^{-1}}$~\cite{Milo2016}. Because experiments show that compared to the
-activated promoter, the basal transcription rate is almost zero for
-\emph{xapAB} (probably due to the weak polymerase binding site), this
-actually means a $10^{2}$- to $10^{3}$-fold increase in the transcription
-rate due to activation by XapR.
+This estimate comes from the assumption that this rate should be roughly $10^{1} - 10^{2}$ times larger than a normal basal rate (of other genes), for which common values are around $10^{-3 \pm 1}~\n{nM\ s^{-1}}$~\cite{Milo2016}.
+Because experiments show that compared to the activated promoter, the basal transcription rate is almost zero for \emph{xapAB} (probably due to the weak polymerase binding site), this actually means a $10^{2}$- to $10^{3}$-fold increase in the transcription rate due to activation by XapR.
 
-Keep in mind that the rate includes not only the effect of XapR on the pure
-speed of transcription, bt also any effect of XapR on the polymerase
-dissociation constant. Still, we do not expect higher values such as
-$10^{0}~\n{nM\ s^{-1}}$ since this is roughly the rate for transcription of
-rRNA and the latter should be significantly faster. To further validate our
-estimate, we can compare it to measured values for \emph{lac}, which should
-be roughly the same because the gene function is similar. The transcription
-rate for \emph{lac} in the absence of any repression that was found
-in~\cite{RazoMejia2019} corresponds well with our estimate.
+Keep in mind that the rate includes not only the effect of XapR on the pure speed of transcription, bt also any effect of XapR on the polymerase dissociation constant.
+Still, we do not expect higher values such as $10^{0}~\n{nM\ s^{-1}}$ since this is roughly the rate for transcription of rRNA and the latter should be significantly faster.
+To further validate our estimate, we can compare it to measured values for \emph{lac}, which should be roughly the same because the gene function is similar.
+The transcription rate for \emph{lac} in the absence of any repression that was found in~\cite{RazoMejia2019} corresponds well with our estimate.
 
 \paragraph*{The mRNA decay rate parameter $\gamma_{\n{mp}}$.}
-The mRNA life time is usually on the order of 1 to 8 minutes which gives a
-decay rate of $\gamma_{\n{m}} \approx 10^{-3} - 10^{-2}~\n{s^{-1}}$. Thus,
-we have $\gamma_{\n{mp}} \defeq \frac{\gamma_{\n{m}}}{\gamma_{\n{p}}}
-\approx 10^{1 \pm 0.5}$.
+The mRNA life time is usually on the order of 1 to 8 minutes which gives a decay rate of $\gamma_{\n{m}} \approx 10^{-3} - 10^{-2}~\n{s^{-1}}$.
+Thus, we have $\gamma_{\n{mp}} \defeq \frac{\gamma_{\n{m}}}{\gamma_{\n{p}}} \approx 10^{1 \pm 0.5}$.
 
 \paragraph*{The translation rate parameter $\rho_{\n{p}}$.}
-The translation rate should be on the order of $r_{\n{p}} \approx 10^{-2} -
-10^{-1}~\n{s^{-1}}$~\cite{Milo2016}. This gives $\rho_{\n{p}} \defeq
-\frac{r_{\n{p}}}{\gamma_{\n{p}}} \approx 10^{2 \pm 0.5}$.
+The translation rate should be on the order of $r_{\n{p}} \approx 10^{-2} - 10^{-1}~\n{s^{-1}}$~\cite{Milo2016}.
+This gives $\rho_{\n{p}} \defeq \frac{r_{\n{p}}}{\gamma_{\n{p}}} \approx 10^{2 \pm 0.5}$.
 
@@
 \paragraph*{The Michaelis-Menten parameters for XapB ($k_{\n{\beta,i}}$, $K_{\n{\beta,i}}$, $k_{\n{\beta,e}}$, $K_{\n{\beta,e}}$):}
-To estimate the parameters for both influx and efflux, we need data from
-experiments where these two processes have been measured independently.
-Lactose permease, which we expect to function in a similar way as XapB, is
-comparatively well studied such that these rates have been measured in
-multiple ways and for different conditions. Some data from several
-publications is collected in~\cite{Viitanen1984} and we will base our
-estimates on this and the corresponding references.
+To estimate the parameters for both influx and efflux, we need data from experiments where these two processes have been measured independently.
+Lactose permease, which we expect to function in a similar way as XapB, is comparatively well studied such that these rates have been measured in multiple ways and for different conditions.
+Some data from several publications is collected in~\cite{Viitanen1984} and we will base our estimates on this and the corresponding references.
 
-Comparing the effective rates and Michaelis constants of lactose permease
-and the nucleoside transporters NupC and NupG shows that their orders of
-magnitude are similar: from~\cite{Dornmair1989,Wright1985} we find
-$k_{\n{cat}} \approx 10^1~\n{s^{-1}}$, $K_{\n{M}} \approx 10^1-10^2~\n{\mu
-M}$ for LacY and from~\cite{Norholm2001,Komatsu1973,MunchPetersen1979}
-$k_{\n{cat}} \approx 10^2~\n{s^{-1}}$, $K_{\n{M}} \approx 10^0-10^1~\n{\mu
-M}$ for pyrimidines through NupC and NupG. Note that especially for Nup,
-these are only rough measurements, and it is known that NupC behaves
-different with purines (thus also with xanthosine). Given these values, we
-expect lactose permease to be similar enough to nucleoside transporters so
-that we can use it as a starting point for our estimation of the XapB
-kinetic values.
+Comparing the effective rates and Michaelis constants of lactose permease and the nucleoside transporters NupC and NupG shows that their orders of magnitude are similar: from~\cite{Dornmair1989,Wright1985} we find $k_{\n{cat}} \approx 10^1~\n{s^{-1}}$, $K_{\n{M}} \approx 10^1-10^2~\n{\mu M}$ for LacY and from~\cite{Norholm2001,Komatsu1973,MunchPetersen1979} $k_{\n{cat}} \approx 10^2~\n{s^{-1}}$, $K_{\n{M}} \approx 10^0-10^1~\n{\mu M}$ for pyrimidines through NupC and NupG.
+Note that especially for Nup, these are only rough measurements, and it is known that NupC behaves different with purines (thus also with xanthosine).
+Given these values, we expect lactose permease to be similar enough to nucleoside transporters so that we can use it as a starting point for our estimation of the XapB kinetic values.
 
-We use the measured influx values that are summarized in~\cite{Viitanen1984}
-for a membrane potential of $100~\n{mV}$ and obtain $k_{\n{b,i}} \approx
-10^{1 \pm 1}~\n{s^{-1}}$ and $K_{\n{b,i}} \approx 10^{-4 \pm 2}~\n{M}$.
-Here, the uncertainty is estimated from the distribution of values of many
-kinds of enzymes that is presented in~\cite{Milo2016}. The bounds also
-reflect the difference between LacY and Nup that was found above. For the
-dimensionless quantities, this gives $k_{\n{\beta,i}} \approx 10^{4 \pm 1}$
-and $K_{\n{\beta,i}} \approx 10^{1 \pm 2}$.
+We use the measured influx values that are summarized in~\cite{Viitanen1984} for a membrane potential of $100~\n{mV}$ and obtain $k_{\n{b,i}} \approx 10^{1 \pm 1}~\n{s^{-1}}$ and $K_{\n{b,i}} \approx 10^{-4 \pm 2}~\n{M}$.
+Here, the uncertainty is estimated from the distribution of values of many kinds of enzymes that is presented in~\cite{Milo2016}.
+The bounds also reflect the difference between LacY and Nup that was found above.
+For the dimensionless quantities, this gives $k_{\n{\beta,i}} \approx 10^{4 \pm 1}$ and $K_{\n{\beta,i}} \approx 10^{1 \pm 2}$.
 
-For the efflux parameters, we proceed similarly and obtain $k_{\n{b,e}}
-\approx 10^{0 \pm 2}~\n{s^{-1}}$ and $K_{\n{b,e}} \approx 10^{-3 \pm
-3}~\n{M}$, therefore $k_{\n{\beta,e}} \approx 10^{3 \pm 2}$ and
-$K_{\n{\beta,e}} \approx 10^{2 \pm 3}$. Because of disagreement in
-literature, difficulty in measuring these parameters, and a lack of
-knowledge about what value range should be expected, we put larger error
-bounds than for the influx.
+For the efflux parameters, we proceed similarly and obtain $k_{\n{b,e}} \approx 10^{0 \pm 2}~\n{s^{-1}}$ and $K_{\n{b,e}} \approx 10^{-3 \pm 3}~\n{M}$, therefore $k_{\n{\beta,e}} \approx 10^{3 \pm 2}$ and $K_{\n{\beta,e}} \approx 10^{2 \pm 3}$.
+Because of disagreement in literature, difficulty in measuring these parameters, and a lack of knowledge about what value range should be expected, we put larger error bounds than for the influx.
 
-\paragraph*{The non-specific transport parameters $k_{\n{\eta}}$, $\xi$:}
-Estimating the rates of the Nup transporters for xanthosine is difficult
-since there are no experimental values that could be used as a reference.
-Because the rate cannot even be measured~\cite{Norholm2001}, it must be much
-lower than that of XapB. The kinetic parameters should, however, be such
-that the steady-state intracellular xanthosine concentration is
-significantly larger than the extracellular one. Otherwise, there would be
-no relevant difference to passive diffusion, which was shown to be
-insufficient for switching~\cite{Norholm2001}. This essentially means $\xi <
-1$, which on the other hand implies that the Nup rate is larger than or at
-least equal to the diffusion rate: If Nup transport were slower than
-diffusion, the latter would rule and $\xi$ would be close to~1 (because this
-is the $\xi$ value of diffusion).
+\paragraph*{The non-specific transport parameters $k_{\n{\eta}}$, $\xi$:} Estimating the rates of the Nup transporters for xanthosine is difficult since there are no experimental values that could be used as a reference.
+Because the rate cannot even be measured~\cite{Norholm2001}, it must be much lower than that of XapB.
+The kinetic parameters should, however, be such that the steady-state intracellular xanthosine concentration is significantly larger than the extracellular one.
+Otherwise, there would be no relevant difference to passive diffusion, which was shown to be insufficient for switching~\cite{Norholm2001}.
+This essentially means $\xi < 1$, which on the other hand implies that the Nup rate is larger than or at least equal to the diffusion rate: If Nup transport were slower than diffusion, the latter would rule and $\xi$ would be close to~1 (because this is the $\xi$ value of diffusion).
 
-In addition, we know from \emph{lac} permease that influx is not much faster
-than efflux, which implies that $\xi$ also cannot be too much smaller
-than~1. We guess $\xi \approx 0.8 \pm 0.1$. The number of Nup transporters
-per cell is roughly $10^2-10^3$~\cite{Li2014}. We expect
-$\frac{k_{\n{cat}}}{K_{\n{M}}}$ to be at least two or three orders of
-magnitude lower for Nup than for saturated XapB, where we have
-$\frac{k_{\n{b}}}{K_{\n{b}}} \approx 10^{5 \pm 3}~\n{M^{-1}\ s^{-1}}$. These
-arguments set an estimated upper bound on the non-specific transport rate:
-$k_{\n{nup}} = \frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0} \lesssim 10^{-3 \pm
-3}~\n{s^{-1}}$.
+In addition, we know from \emph{lac} permease that influx is not much faster than efflux, which implies that $\xi$ also cannot be too much smaller than~1.
+We guess $\xi \approx 0.8 \pm 0.1$.
+The number of Nup transporters per cell is roughly $10^2-10^3$~\cite{Li2014}.
+We expect $\frac{k_{\n{cat}}}{K_{\n{M}}}$ to be at least two or three orders of magnitude lower for Nup than for saturated XapB, where we have $\frac{k_{\n{b}}}{K_{\n{b}}} \approx 10^{5 \pm 3}~\n{M^{-1}\ s^{-1}}$.
+These arguments set an estimated upper bound on the non-specific transport rate: $k_{\n{nup}} = \frac{k_{\n{cat}}}{K_{\n{M}}} \n{[E]_0} \lesssim 10^{-3 \pm 3}~\n{s^{-1}}$.
 
-To find a lower bound, we estimate the diffusion rate. As described in
-Eq~\ref{eq:diff} in the previous section, the diffusion rate can be written
-as $10^{-2} \cdot p \cdot A$, where $p$ is the permeability coefficient of
-the membrane for xanthosine, $A$ is the total cell surface area, and
-$10^{-2}$ is the fraction of uncharged xanthosine at physiological pH. From
-the permeabilities of various substances that are collected
-in~\cite{Milo2016} and from the measured permeability values for purine
-nucleosides in~\cite{Wei2011}, we estimate a permeability coefficient of
-$\tilde{p}=10^{-8 \pm 3}~\n{m\ s^{{-1}}}$. We write the tilde because this
-value is given in a convention where $\tilde{p} \cdot A = \dd{x}{t}$ with
-$x$ being the number of xanthosine molecules and not the concentration.
-Since we work in concentrations, we divide by an estimated cell volume of
-$10^{-18}~\n{m^3}$ and, with a cell surface area of $10^{-11}~\n{m^2}$, find
-a diffusion rate of roughly $10^{-3 \pm 3}~\n{s^{-1}}$. This sets the lower
-bound of $k_{\n{nup}}$.
-%\com{Check if Overton's rule gives the samediffusion coefficient}
+To find a lower bound, we estimate the diffusion rate.
+As described in Eq~\ref{eq:diff} in the previous section, the diffusion rate can be written as $10^{-2} \cdot p \cdot A$, where $p$ is the permeability coefficient of the membrane for xanthosine, $A$ is the total cell surface area, and $10^{-2}$ is the fraction of uncharged xanthosine at physiological pH.
+From the permeabilities of various substances that are collected in~\cite{Milo2016} and from the measured permeability values for purine nucleosides in~\cite{Wei2011}, we estimate a permeability coefficient of $\tilde{p}=10^{-8 \pm 3}~\n{m\ s^{{-1}}}$.
+We write the tilde because this value is given in a convention where $\tilde{p} \cdot A = \dd{x}{t}$ with $x$ being the number of xanthosine molecules and not the concentration.
+Since we work in concentrations, we divide by an estimated cell volume of $10^{-18}~\n{m^3}$ and, with a cell surface area of $10^{-11}~\n{m^2}$, find a diffusion rate of roughly $10^{-3 \pm 3}~\n{s^{-1}}$.
+This sets the lower bound of $k_{\n{nup}}$.
+%\com{Check if Overton's rule gives the same diffusion coefficient}
 
-Interestingly, the upper and lower bound turn out to be the same. Thus, we
-obtain the range $k_{\n{nup}} \approx 10^{-3 \pm 3}~\n{s^{-1}}$, meaning
-$k_{\n{\eta}} \approx 10^{-1 \pm 3}$.
+Interestingly, the upper and lower bound turn out to be the same. Thus, we obtain the range $k_{\n{nup}} \approx 10^{-3 \pm 3}~\n{s^{-1}}$, meaning $k_{\n{\eta}} \approx 10^{-1 \pm 3}$.
 
@@
 \paragraph*{The 2D set of equations.}
-If the mRNA concentration is set to steady-state as discussed in the main
-text, the dynamical system is reduced to the two equations shown below:
+If the mRNA concentration is set to steady-state as discussed in the main text, the dynamical system is reduced to the two equations shown below:
 \begin{alignat}{2}
@@
 \end{alignat}
-Here, several of the parameters from the 3D system could now be combined
-into one by defining $\rho \defeq \frac{\rho_{\n{m}}
-\rho_{\n{p}}}{\gamma_{\n{mp}}}$.
+Here, several of the parameters from the 3D system could now be combined into one by defining $\rho \defeq \frac{\rho_{\n{m}} \rho_{\n{p}}}{\gamma_{\n{mp}}}$.
 
-\paragraph*{Magnitude of flow in the phase portraits.}
-The plots in the main part do not show the magnitude of the vector fields.
-For additional information, we show a 2D vector plot with the magnitudes in
-Fig~\ref{figS4:vector}.
+\paragraph*{Magnitude of flow in the phase portraits.} The plots in the main part do not show the magnitude of the vector fields.
+For additional information, we show a 2D vector plot with the magnitudes in Fig~\ref{figS4:vector}.
 
@@
 	\caption{{\bf Vector plot of a standard case of bistability.}
-		The 2D phase portraits as a vector plot for for the standard set of
-		parameters that was used in the main text.}
+		The 2D phase portraits as a vector plot for for the standard set of parameters that was used in the main text.}
 	\label{figS4:vector} 
@@
 \paragraph*{Increasing $\n{[c]_a}$.}
-Fig~\ref{figS5:incC} shows a family of nullclines for increasing
-extracellular xanthosine concentration. It shows the transition from
-monostability, to bistability, and back to monostability again.
+Fig~\ref{figS5:incC} shows a family of nullclines for increasing extracellular xanthosine concentration. It shows the transition from monostability, to bistability, and back to monostability again.
 
@@
 	\caption{{\bf A family of curves for increasing $\n{[c]_a}$.}
-		The blue curve is the protein nullcline, which does not change with
-		$\n{[c]_a}$. The other lines are the xanthosine nullclines for
-		different values of the extracellular xanthosine concentration.}
+		The blue curve is the protein nullcline, which does not change with $\n{[c]_a}$. The other lines are the xanthosine nullclines for different values of the extracellular xanthosine concentration.}
 	\label{figS5:incC}
@@
 \paragraph*{Hysteresis effects.}
-In Fig~\ref{figS8:hysteresis}, the results mentioned in the hysteresis part
-of the main text are shown. As mentioned there,  the simulation was started
-with initially fully induced cells. At concentrations where no cells
-switched from the low to the high expression state before, cells here do not
-switch back to low expression.
+In Fig~\ref{figS8:hysteresis}, the results mentioned in the hysteresis part of the main text are shown.
+As mentioned there,  the simulation was started with initially fully induced cells.
+At concentrations where no cells switched from the low to the high expression state before, cells here do not switch back to low expression.
+The extracellular xanthosine concentration (for the other parameters being as in the Table in the main text) needed to be as low as $\n{[c]_a} = 10$ to observe bimodality and $\n{[c]_a}=9$ to have all cells in the low expression state (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5 \cdot
+10^{-5} \unit{M}$).
 
@@
 	\caption{{\bf Decreasing $\n{[c]_a}$ in the fully induced system.}
-		Apart from $\n{[c]_a}$, the parameters are the same as in the table
-		in the main text. For the distributions, the simulations were run
-		1000 times for $10^6 \unit{s}$ each and started at the mRNA, protein
-		and intracellular xanthosine counts of the high fixed point in the
-		corresponding phase portraits. The values of $\n{[c]_a}$ are 12 in
-		(A), 10 in (B), and 9 in (C). When comparing to the distributions in
-		the main text where the simulation had initial concentrations of 0,
-		a clear hysteresis effect can be observed.}
+		Apart from $\n{[c]_a}$, the parameters are the same as in the table in the main text.
+For the distributions, the simulations were run 1000 times for $10^6 \unit{s}$ each and started at the mRNA, protein and intracellular xanthosine counts of the high fixed point in the corresponding phase portraits.
+The values of $\n{[c]_a}$ are 12 in (A), 10 in (B), and 9 in (C).
+When comparing to the distributions in the main text where the simulation had initial concentrations of 0, a clear hysteresis effect can be observed.}
 	\label{figS8:hysteresis}
@@
 \section{Chemical master equation}
-Here we provide some further details on our stochastic simulations by
-writing the chemical master equation for our model. The state of our system
-is described by the number of mRNA $m$, protein $p$, and xanthosine
-molecules $x$, and we write $P(m,p,x)$ for the probability of that state.
+Here we provide some further details on our stochastic simulations by writing the chemical master equation for our model.
+The state of our system is described by the number of mRNA $m$, protein $p$, and xanthosine molecules $x$, and we write $P(m,p,x)$ for the probability of that state.
 The chemical master equation for the system is:
@@
 \end{alignat}
-Here, all the parameters of course have to be measured in counts, not in
-concentrations. As can be seen from the equation, none of the propensities
-depends explicitly on time and hence, we can use the hybrid algorithm
-between classical Gillespie and \texttau-leaping that we relied on.
+Here, all the parameters of course have to be measured in counts, not in concentrations.
+As can be seen from the equation, none of the propensities depends explicitly on time and hence, we can use the hybrid algorithm between classical Gillespie and \texttau-leaping that we relied on.
 

### 2019-11-15 main_text/main_text.tex
RP edits plus few minor edits by KL

@@
 		\\
-		\textbf{3} Department of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
+		\textbf{3} Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
 		\\
@@
 	\section*{Abstract}
-	Membrane transporters carry important molecules through the cell membrane
-	and, from a resource standpoint, should only be produced when necessary. The
+	Membrane transporters carry key metabolites through the cell membrane
+	and, from a resource standpoint, are hypothesized to be produced when necessary. The
 	expression of membrane transporters in metabolic pathways is often
@@
 	which encode for lactose, arabinose, and xylose transporters, respectively.
-	As a case study of a minimal system, we build a generizable physical model of the \emph{xapABR} genetic
+	As a case study of a minimal system, we build a
+	generalizable physical model of the \emph{xapABR} genetic
 	circuit, which features a regulatory feedback loop through membrane
@@
 	that the membrane transport makes the model system bistable in certain
-	parameter regimes. Thus, it serves as a genetic “on-off” switch, enabling
+	parameter regimes. Thus, it serves as a genetic ``on-off'' switch, enabling
 	the system to only produce a set of metabolic enzymes when the corresponding
@@
 
-	\section*{Introduction} \todo{remove images \& todonotes everywhere for submission}
+	\section*{Introduction} \todo{remove images, todonotes, and funding for submission}
 	Genetic regulatory circuits are fundamental building blocks of functioning
@@
 	of either one of the proteins, the system will remain in this state until a
-	significant perturbation occurs. Another well known and even simpler example
+	significant perturbation occurs. Another well-known and even simpler example
 	is an auto-activating circuit in which a protein activates its own
@@
 	Existing models in the literature tend towards one of two extremes:
-	either highly detailed descriptions of specific, complicated systems (e.g.,~\cite{Wong1997}),
+	either highly detailed descriptions of specific, complicated networks (e.g.,~\cite{Wong1997}),
 	or Hill function descriptions that coarse-grain all complexity
@@
 	positive feedback
-	loop~\cite{Novick1957,Santillan2008,Ozbudak2004,Narang2008,Choi2008,Fritz2014,Jenkins2017,Siegele1997}.
+	loop, for example in the well-known study of Novick and Weiner,
+	among other works~\cite{Novick1957,Santillan2008,Ozbudak2004,Narang2008,Choi2008,Fritz2014,Jenkins2017,Siegele1997}.
 	A eukaryotic example is the glucose transporter GLUT-2 in liver and
@@
 	presence in the extracellular environment. If there is enough, the
-	expression stabilizes at an ``on'' point and the cell has, in a short-term
+	expression stabilizes at an ``on'' state and the cell has, in a short-term
 	sense, adapted. When there is not enough substrate, the operon, which
@@
 	\begin{itemize}
-		\item In the initial absense of XapB, small amounts of xanthosine permeate\todo{``permeate'' wouldn't include active transport (NupC \& NupG), would it?
-MM: I think it's sufficiently vague to capture all this, details are too subtle to worry about here} into and out of the cell (discussed in more detail below).
+		\item In the initial absense of XapB, small amounts of xanthosine permeate
+				into and out of the cell (discussed in more detail below).
 		\item XapR is induced by xanthosine.
@@
 		\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
-		\item Production of more XapA \& XapB is balanced by decay or dilution through cell divisions.
+		\item Production of more XapA and XapB is balanced by decay or dilution through cell divisions.
 	\end{itemize}
@@
 	We treat dimers as the only form of XapR that appears in the cell. Each
-	dimer can bind two xanthosine molecules. The Monod-Wyman-Changeux (MWC)
+	dimer can bind two xanthosine molecules~\cite{Joergensen1999}. The Monod-Wyman-Changeux (MWC)
 	model is used to describe the fraction of XapR dimers in the active state,
@@
 	\end{eqnarray}
-	Here, $\n{[x]}$ is the xanthosine concentration, and $\n{[XapR]_A}$ and
+	A detailed discussion of the
+	MWC model can be found in~\cite{Marzen2013}. In Eq~\ref{eq:MWC},
+	$\n{[x]}$ is the xanthosine concentration, and $\n{[XapR]_A}$ and
 	$\n{[XapR]_{tot}}$ denote the concentration of active and total XapR dimers,
@@
 	dimer, respectively, and $\Delta E_{\n{x}}$ stands for the energy difference
-	between the inactive and the active state. We expect $\Delta E_{\n{x}} > 0$
+	between the inactive and the active states of the protein. We expect $\Delta E_{\n{x}} > 0$
 	and $K_{\n{xA}} < K_{\n{xI}}$ for inducible activation. This corresponds to
 	XapR being mainly inactive in the absence of xanthosine and becoming mostly
-	active at high concentrations of xanthosine. A detailed discussion of the
-	MWC model can be found in~\cite{Marzen2013}.
+	active at high concentrations of xanthosine.
 
@@
 	active and all other states as inactive, meaning they have transcription
-	rate equal zero. Experiments show that the expression becomes very weak when one of
+	rate equal to zero. Experiments show that the expression becomes very weak when one of
 	the XapR binding sites is removed from the promoter, suggesting that this
@@
 	where $r_{\n{p}}$ denotes the translation rate, $\gamma_{\n{p}}$ the protein
-	decay rate, and $\langle \n{[mRNA]} \rangle$ the mean mRNA concentration:
+	decay rate, and $\langle \n{[m]} \rangle$ the mean mRNA concentration:
 	\begin{eqnarray}
 	\label{eq:pbasic}
-	\dd{\n{[p]}}{t} = r_{\n{p}} \langle \n{[mRNA]} \rangle - \gamma_{\n{p}} \n{[p]}.
+	\dd{\n{[p]}}{t} = r_{\n{p}} \langle \n{[m]} \rangle - \gamma_{\n{p}} \n{[p]}.
 	\end{eqnarray}
@@
 
-	\todo{MM: do these paragraphs on xanthosine dynamics have an appropriate level of detail now, or too much, or not enough?}
 	There are two significant mechanisms for transport of xanthosine across
@@
 	$K_{\n{b, i}}$ for influx and $k_{\n{b, e}}, K_{\n{b,e}}$ for efflux.
-	For the Nup transporters, we can linearize because of their low affinity
-	for xanthosine and are left with the rate parameters $k_{\n{nup}}$ for
+	For the Nup transporters, we also use Michaelis-Menten kinetics but can linearize the Michaelis-Menten term $k_{\n{cat}} \frac{\n{[x]}}{K_{\n{M}} + \n{[x]}} \approx \tilde{k} \n{[x]}$ because of the transporter's low affinity for xanthosine (i.e.
+$K_{\n{M}} \gg \n{[x]}$).
+For the rate parameters $\tilde{k}$, we write $k_{\n{nup}}$ for
 	influx and $\xi k_{\n{nup}}$ for efflux.
@@
 	Recall that $\n{[x]}$ is the intracellular xanthosine concentration, while
-	$c$ denotes the extracellular. Because $k_{\n{b,i}} > k_{\n{b,e}}$ and
+	$c$ denotes the extracellular concentration. Because $k_{\n{b,i}} > k_{\n{b,e}}$ and
 	$K_{\n{b,i}} < K_{\n{b,e}}$, influx dominates at low intracellular
@@
 	estimate a reasonable range by using values from similar, well studied
-	systems and by finding physical constraints or relations between parameters.
+	systems and by exploiting physical constraints or relations between parameters.
 	The results of these estimates are shown in Table~\ref{table1:nondim}. They
@@
 	solutions of the dynamical system. For this choice of parameters, the system
-	first flows towards the mRNA nullcline, then it moves along that surface to
+	first flows towards the mRNA nullcline (independent of the initial condition),
+	then it moves along that surface to
 	the intersection with the protein nullcline, and lastly, it moves along that
@@
 			$K_{\n{a}} = 5 \cdot 10^{4} \unit{nM}$. The surfaces in (A) and the
-			lines in (B) are the nullclines of the state variables, and their
+			curves in (B) are the nullclines of the state variables, and their
 			intersection points, marked in red in (B), are the steady-state
@@
 	dynamics unfolds in two steps: first to the intersection of the mRNA and the
-	xanthosine nullcline, then along that line to the protein nullcline and
+	xanthosine nullcline, then along that curve to the protein nullcline and
 	thereby to a fixed point.
@@
 	here and we will now explain this choice of values. Firstly, they were
-	picked such that they are well within the parameter range that was estimated beforehand (see Table~\ref{table1:nondim}). Secondly, we
+	picked roughly in the middle of the range that was estimated beforehand for this parameter (see Table~\ref{table1:nondim} and~\nameref{S1_Text}). Secondly, we
 	chose parameters that allow clear bistability in the phase portraits as well as in the
@@
 	numbers that are very reasonable (see~\cite{Milo2016} for the average mRNA
-	numbers in cells). The protein and xanthosine
+	numbers in bacterial cells). The protein and xanthosine
 	concentrations followed from this, but with some variation in the parameters
@@
 			so $\n{[c]_a}$ is dimensionless).
-			Tuning $\n{[c]_a}$ moves the orange line (xanthosine nullcline) up
-			or down, but the blue curve (mRNA nullcline) is unchanged (see
+			Tuning $\n{[c]_a}$ moves the orange line (xanthosine nullcline), but the blue curve (mRNA nullcline) is unchanged (see
 			also~\nameref{S1_Text}). It can clearly be seen that in (A) there is
@@
 	While it is clear that the bistability in the model system is due to the
-	feedback loop from XapA and XapB, it is not obvious if both XapA and XapB
-	are necessary. It turns out that the bistability is due to XapB only. XapA
+	feedback loop from XapA and XapB, it is not intuitively clear if both XapA and XapB
+	are necessary. The model implies that the bistability is due to XapB only. XapA
 	is neither sufficient nor necessary and, within the estimated parameter
@@
 	\emph{xapAB} promoter and two cooperative binding sites on XapR for
-	xanthosine. An interesting question to pose is whether the cooperativity is
+	xanthosine. It is interesting to consider whether the cooperativity is
 	a necessary feature for bistability. This question is motivated by the importance of
@@
 	But only when the second binding site is removed in both places, leaving
-	no cooperativity in the system, are the
-	equations are insufficiently non-linear to produce bistability. An example
+	no cooperativity in the system, do we find that it is insufficiently non-linear to produce bistability. An example
 	of the three scenarios (only cooperative XapR, only cooperative promoter, no cooperativity) can be seen in Fig~\ref{fig7:coop}. It follows that
@@
 	the system. Because of the two different fixed points at low and high
-	expression, the copy numbers in the problem vary from very low to very high.
-	For large amounts of mRNA, protein, and xanthosine, the number of reactions
-	is large too, and thus, Gillespie's classical algorithm has a large
-	computational cost and his \texttau-leap algorithm would be ideal. On the
-	other hand, \texttau-leaping cannot be used for the small copy numbers. For
+	expression, the copy numbers in the problem vary from less than five to several thousand.
+	The larger the amounts of mRNA, protein, and xanthosine, the larger the number of reactions, and thus, the larger the computational cost of Gillespie's classical algorithm.
+This makes Gillespie's \texttau-leap algorithm ideal for the high expression state.
+On the
+	other hand, \texttau-leaping cannot be used for the small copy numbers in the low expression state. For
 	these reasons, we chose to work with the algorithm described
@@
 			values of $\n{[c]_a}$ are 12 in (A), 18.5 in (B), and 25 in (C)
-			(recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$, $K_{\n{a}} = 5
-			\cdot 10^{-5} \unit{M}$). The output from the stochastic simulations
+			(recall that $\n{[c]_{a}} \defeq
+			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$,
+			so $\n{[c]_a}$ is dimensionless). The output from the stochastic simulations
 			is in good agreement with the concentrations at the fixed points in
@@
 	extremely unlikely. There are two aspects that become relevant in this
-	context that we neglect in our analysis but shortly mention here:
+	context that we neglect in our analysis but briefly mention here:
 	transcription and translation bursts lead to higher stochasticity and cell
@@
 	Nevertheless, we do warn the reader that the timescales in the simulations
-	should be taken with reservations. Cell divisions are not considered here,
+	should be taken with reservation. Cell divisions are not considered here,
 	and neither is the burstiness of transcription and translation. This means
@@
 	state is stable even if the xanthosine concentration decreases to a certain
-	extent. This stability explains what has been observed by Novick and
+	extent. This stability explains what was observed by Novick and
 	Weiner~\cite{Novick1957} for the \emph{lac} operon: when induced cells were
@@
 	\section*{Conclusion}
-	\paragraph*{A genetic switch for metabolic adaptation with remarkable features.}
 	In this paper, we propose a simple model for genetic circuits containing
@@
 
-	\paragraph*{Required properties for this system.}
 	We have found that no bistability can emerge from the genetic circuit unless
@@
 	In addition, simply knowing the experimental switching concentration of
-	xanthosine let us, for example, infer the approximate value of the
+	xanthosine permits us, for example, to infer the approximate value of the
 	dissociation constant between the transcription factor XapR and the inducer
-	xanthosine. We found that their interaction needs to be rather weak compared to usual interactions between an inducer and its target.
-
-	\paragraph*{Possibilities for further investigation.}
+	xanthosine.
+The value we found is roughly one to two orders of magnitude larger than what has been measured for LacI and IPTG~\cite{RazoMejia2018}, meaning the interaction of XapR and xanthosine is rather weak.
+
 	Phase diagrams, showing for which parameters the system is bistable and for
@@
 	calculated from arguments made in~\cite{Cherry2000}. However, the
-	simulations have shown that deterministic bistability does not mean that
+	simulations showed that deterministic bistability does not mean that
 	bimodality occurs, which is why we have refrained from showing such
@@
 
-	\paragraph*{A successful generalizable model.}
 	All model parameters could be reasonably estimated despite the paucity of
@@
 	For assistance with illustrations, we also thank Nigel Orme and Griffin Chure.
-	\todo{MJM: this kinda reads like GC should be in the author list...}
-	K.S.L. acknowledges financial support from the
-	Werner~Siemens-Foundation through the Swiss Study Foundation.
-	This material is based upon work supported by the National Science Foundation
-	Graduate Research Fellowship under Grant No.\ DGE-114446.
-	\todo{MJM: need to double-check my NSF \#}
-	\todo{remove funding again for submission}
+	This work was supported by 1R35 GM118043-01 Maximizing Investigators' Research Award (MIRA) (to R.P.), the National Science Foundation
+	Graduate Research Fellowship under Grant No.\ DGE-114446 (M.J.M.). and the
+	Werner Siemens Foundation through the Swiss Study Foundation (K.S.L.).
 
@@
 
+		\bibitem{RazoMejia2018}
+		Razo-Mejia M, Barnes SL, Belliveau NM, Chure G, Einav T, Lewis M, et~al.
+		\newblock Tuning Transcriptional Regulation through Signaling: A Predictive
+		Theory of Allosteric Induction.
+		\newblock Cell Systems. 2018;6(4):456 -- 469.e10.
+		\newblock doi:{10.1016/j.cels.2018.02.004}.
+
 	\end{thebibliography}

### 2019-11-16 main_text/main_text.tex
Another round of edits by MJM

@@
 	$K_{\n{b, i}}$ for influx and $k_{\n{b, e}}, K_{\n{b,e}}$ for efflux.
-	For the Nup transporters, we also use Michaelis-Menten kinetics but can linearize the Michaelis-Menten term $k_{\n{cat}} \frac{\n{[x]}}{K_{\n{M}} + \n{[x]}} \approx \tilde{k} \n{[x]}$ because of the transporter's low affinity for xanthosine (i.e.
-$K_{\n{M}} \gg \n{[x]}$).
+	For the Nup transporters, we also use Michaelis-Menten kinetics but,
+	because of the transporter's low affinity for xanthosine,
+	we can linearize the Michaelis-Menten term as
+	$k_{\n{cat}} {\n{[x]}}/(K_{\n{M}} + \n{[x]}) \approx \tilde{k} \n{[x]}$
+	(i.e., $K_{\n{M}}~\gg~\n{[x]}$ across the physiologically relevant range for $\n{[x]}$).
 	For the rate parameters $\tilde{k}$, we write $k_{\n{nup}}$ for
@@
 	the system. Because of the two different fixed points at low and high
-	expression, the copy numbers in the problem vary from less than five to several thousand.
-	The larger the amounts of mRNA, protein, and xanthosine, the larger the number of reactions, and thus, the larger the computational cost of Gillespie's classical algorithm.
-This makes Gillespie's \texttau-leap algorithm ideal for the high expression state.
-On the
-	other hand, \texttau-leaping cannot be used for the small copy numbers in the low expression state. For
+	expression, the protein copy numbers in the problem vary from less than
+	five to several thousand. Even worse, xanthosine copy numbers may range
+	as high as $10^7$ at the high expression fixed point. For such large
+	copy numbers, the number of reaction firings that must be simulated with
+	Gillespie's classical algorithm leads to an impractical computational
+	cost. This would make Gillespie's \texttau-leap algorithm ideal for the high
+	expression state. On the other hand, \texttau-leaping cannot be used for
+	the small protein copy numbers in the low expression state, or the mRNA
+	copy number which remains of order ten or less in both states. For
 	these reasons, we chose to work with the algorithm described
@@
 	dissociation constant between the transcription factor XapR and the inducer
-	xanthosine.
-The value we found is roughly one to two orders of magnitude larger than what has been measured for LacI and IPTG~\cite{RazoMejia2018}, meaning the interaction of XapR and xanthosine is rather weak.
+	xanthosine. The value we infer is roughly one to two orders of magnitude
+	larger than what has been measured for LacI and
+	IPTG~\cite{RazoMejia2018}, meaning the interaction of XapR and
+	xanthosine is rather weak.
 
@@
 	For assistance with illustrations, we also thank Nigel Orme and Griffin Chure.
-	This work was supported by 1R35 GM118043-01 Maximizing Investigators' Research Award (MIRA) (to R.P.), the National Science Foundation
-	Graduate Research Fellowship under Grant No.\ DGE-114446 (M.J.M.). and the
-	Werner Siemens Foundation through the Swiss Study Foundation (K.S.L.).
+	This work was supported by 1R35 GM118043-01 Maximizing Investigators'
+	Research Award (MIRA) (to R.P.), and the Werner Siemens Foundation
+	through the Swiss Study Foundation (K.S.L.). This material is based upon
+	work supported by the National Science Foundation Graduate Research
+	Fellowship under Grant No.\ DGE-1745301 (M.J.M.).
 

### 2020-02-19 main_text/main_text.tex
Main text major edits (KL) + minor edits (MM)
1st pass at incorporating changes/additions
suggested by and in response to reviewer.

@@
 
-\usepackage[obeyFinal]{todonotes} % todo notes and missing figures
 \reversemarginpar
@@
 	Membrane transporters carry key metabolites through the cell membrane
-	and, from a resource standpoint, are hypothesized to be produced when necessary. The
+	and, from a resource standpoint, are hypothesized to be produced only when necessary. The
 	expression of membrane transporters in metabolic pathways is often
@@
 	parameter regimes. Thus, it serves as a genetic ``on-off'' switch, enabling
-	the system to only produce a set of metabolic enzymes when the corresponding
+	the cell to only produce a set of metabolic enzymes when the corresponding
 	metabolite is present in large amounts. We find that the negative feedback
@@
 
-	\section*{Introduction} \todo{remove images, todonotes, and funding for submission}
+	\section*{Introduction} 
 	Genetic regulatory circuits are fundamental building blocks of functioning
@@
 	other~\cite{Jacob1961,Gardner2000} (illustrated schematically in
-	Fig~\ref{fig1:switches}). Here, one stable state is high expression of
+	Fig~\ref{fig0:switches}). Here, one stable state is high expression of
 	the first protein and low expression of the second,
@@
 		\centering
-		\includegraphics{media/Circuits.eps}
+		\includegraphics{media/Fig0_switches.eps}
 		\caption{{\bf A schematic of different genetic switches.}
@@
 			the model circuit can be found in Fig~\ref{fig2:model}.}
-		\label{fig1:switches}
+		\label{fig0:switches}
 	\end{figure}
@@
 	that is induced by xanthosine and activates the \emph{xapAB} promoter,
-	in close analogy to AraC, XylR, and also LacI.
-	\footnote{One might object that LacI represses its target operon, while
+	in close analogy to AraC, XylR, and also LacI.\footnote{One might object that LacI represses its target operon, while
 	XapR, AraC, and XylR activate their target operons. However, the analogy
@@
 	identical, i.e., the presence of their respective inducer causes their
-	target operon to be transcribed.}
-	The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
+	target operon to be transcribed.} The \textit{xapAB} promoter has been suggested to have two binding sites for XapR\cite{Seeger1995},
 	but the promoter architecture and function is not yet fully understood. The
@@
 
-	Experimentally it was found that the expression level of \emph{xapAB} among
+	Experimentally, we found that the expression level of \emph{xapAB} among
 	cells follows a bimodal distribution and that the system seems to be
-	bistable~\cite{Chure2019}. We aim to understand which of the circuit's
+	bistable (see next section). We aim to understand which of the circuit's
 	features are necessary for bistability and investigate its behavior in
-	different parameter regimes. In the following section, we will discuss the
-	details of our model. After estimating the free parameters we then present
+	different parameter regimes. After presenting some experimental background on the \emph{xap} genes, we will discuss the
+	details of our model. Next, we estimate the free parameters and then present
 	the observations we made through phase diagrams, followed by the results
 	from stochastic simulations.
+
+	\section*{Experimental motivation} \label{sec:experiment}
+	Our work was motivated by the experimental observation of bimodality in the \emph{xap} circuit, which is shown in Fig~\ref{fig1:data}.
+We focus on the essential findings here and refer the reader to~\nameref{S1_Text} for more experimental details.
+Briefly, we placed a fluorescent reporter under the control of the wild type (wt) \emph{xapAB} promoter.
+This construct was placed in three different backgrounds: \textDelta\emph{xapABR}, \textDelta\emph{xapAB}, and wt.
+Fig~\subref{fig1:data}{A} shows that for increasing xanthosine concentration, expression in the wt-background increases in a switch-like way: it is nearly zero when there is little xanthosine (``uninduced state''), but increases drastically when there is more (``induced state'').
+In between, the aforementioned bimodal distribution is obtained, where some bacteria are in the uninduced expression state and others are induced.
+\textcolor{red}{maybe remove: In addition, we observed that the property of being induced seems to be passed on to daughter cells, suggesting that there is not merely bimodality, but bistability.}
+
+	\begin{figure}[h!]
+		\centering
+		\includegraphics[width=0.4\textwidth]{media/Fig1_data.pdf}
+		\caption{{\bf Experimental data on the \emph{xap} circuit.} The expression of the \emph{xapAB} promoter was measured for different concentrations of xanthosine in the bacteria's environment.
+(A) shows the wild type circuit while in (B) the genes \emph{xapABR} and in (C) the genes \emph{xapAB} are removed.
+The wild type circuit behaves like a switch.}
+		\label{fig1:data}
+	\end{figure}
+
+	When all genes of the \emph{xapABR} circuit are removed, the xanthosine response of our reporter construct disappears (see Fig~\subref{fig1:data}{B}, where the last line is autofluorescence).
+This is the result of removing the transcription factor XapR that is induced by xanthosine.
+When only the genes \emph{xapAB} are removed but \emph{xapR} is kept, a response to the xanthosine concentration is regained but it is no longer switch-like (see Fig~\subref{fig1:data}{C}).
+Instead, the distribution remains unimodal and simply shifts to higher expression as the xanthosine concentration is increased.
+As we will see later on, this is because the circuit now lacks the positive feedback loop due to the xanthosine membrane transporter XapB.
+
+	Clearly, XapR regulates the \emph{xapAB} promoter.
+As mentioned in the introduction, there are two binding sites for XapR on the \emph{xapAB} promoter: one directly at the polymerase binding site, and one further upstream.
+When the upstream binding site is removed, bimodality is lost and the expression of \emph{xapAB} becomes several orders of magnitude weaker, as can be seen in Fig~\ref{}.
+	\textcolor{red}{data missing}
+
+	Already, we believe this data, combined with the prior knowledge of the field, clearly suggest a minimal model of the \emph{xapABR} system.
+The construction and exploration of this model occupies the remainder of this work.
+We leave it for future work to quantitatively dissect and test the model, in the manner of, for example, \textcolor{red}{citations}.
 
@@
 		\centering
-		\includegraphics{media/XapSystem.eps}
+		\includegraphics{media/Fig2_model.eps}
 		\caption{{\bf Model of the \emph{xapAB} circuit.}
@@
 		\centering
-		\includegraphics{media/States.eps}
+		\includegraphics{media/Fig3_states.eps}
 		\caption{{\bf The promoter states.} We consider only the completely
@@
 	active and all other states as inactive, meaning they have transcription
-	rate equal to zero. Experiments show that the expression becomes very weak when one of
-	the XapR binding sites is removed from the promoter, suggesting that this
-	simplification is reasonable~\cite{Chure2019}. Furthermore, we find that in
+	rate equal to zero. Our experiments show that the expression becomes very weak when one of
+	the XapR binding sites is removed from the promoter (see~\nameref{S1_Text}), suggesting that this
+	simplification is reasonable. Furthermore, we find that in
 	the bistable parameter range, considering the single occupancy states as
@@
 	state variables are the mRNA, the protein, and the xanthosine concentration.
+Note that for the low copy numbers that can occur in our system, a deterministic analysis is not necessarily meaningful.
+Nevertheless, in our case we find that stochastic simulations agree well with the deterministic results.
+Thus, deterministic phase portraits are a good starting point.
 
@@
 		\centering
-		\includegraphics{media/Bistability.eps}
+		\includegraphics{media/Fig4_bistability.eps}
 		\caption{{\bf Phase portraits showing bistability.}
@@
 
-	The bistability corresponds to the experimental observations\cite{Chure2019}, so the model passes this
+	The bistability corresponds to the experimental observations, so the model passes this
 	sanity check. Furthermore, the xanthosine and protein concentrations at the
@@
 	being in the low expression state to (2.) the population being in a mixed state with some cells in a low expression state and others in a high expression state (all-or-none phenomenon) and then to (3.) all being in the high
-	expression state~\cite{Chure2019}. If our model is correct, it should
+	expression state (see section ``experimental motivation''). If our model is correct, it should
 	exhibit the same qualitative behavior. Indeed we find exactly this: as can
@@
 	expression. In between, there are two deterministically stable expression
-	levels.
+	levels. Later on in this text, we show this result in form of a bifurcation diagram.
 
@@
 		\centering
-		\includegraphics{media/VaryingC.eps}
+		\includegraphics{media/Fig5_extraxanth.eps}
 		\caption{{\bf Phase portraits for different extracellular xanthosine concentrations.}
@@
 			Table~\ref{table1:nondim}. The extracellular xanthosine
-			concentration in these plots is $\n{[c]_a}=10^{1}$ in (A) and
-			$\n{[c]_a}= 5 \cdot 10^{2}$ in (B) (recall that $\n{[c]_{a}} \defeq
+			concentration in these plots is $\n{[c]_a}=7$ in (A) and
+			$\n{[c]_a}= 40$ in (B) (recall that $\n{[c]_{a}} \defeq
 			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$,
@@
 	\begin{figure}%[h!]
-		\includegraphics{media/XapAB.eps}
 		\centering
+		\includegraphics{media/Fig6_xapAB.eps}
 		\caption{{\bf Phase portraits without XapA or XapB.}
@@
 		\centering
-		\includegraphics{media/FewerSites.eps}
+		\includegraphics{media/Fig7_coop.eps}
 		\caption{{\bf Phase portraits for less or no cooperativity.}
@@
 
+	Note that we neglect stochastic fluctuations in any of our parameters, in particular in the XapR copy number.
+The latter is on the order of 10, but because of the long life span and rare expression of proteins like XapR, we expect the influence of the simplification on our results to be rather small.
+Nevertheless, the overall stochastic fluctuations are expected to be larger in the real system than in our simulations.
+
 	\paragraph*{Bimodality and the extracellular xanthosine concentration.}
-	This stochastic approach results in the same bimodal distributions that were already
+	The stochastic approach results in the same bimodal distributions that were already
 	seen in the deterministic investigation and experimental studies.
@@
 	concentration. The parameters that were used are the same as in the previous
-	section (listed in Table~\ref{table1:nondim}). To obtain the distributions,
+	section (listed in Table~\ref{table1:nondim}). However, we also find good agreement of the simulations and the deterministic results for other sets of parameters. To obtain the distributions,
 	we ran the simulation 5000 times for a simulated time of $10^6 \unit{s}$
@@
 		\centering
-		\includegraphics{media/HistogramsC.eps}
+		\includegraphics{media/Fig8_distribution.eps}
 		\caption{{\bf Distributions from stochastic simulations and the corresponding phase portraits.}
@@
 	experiments: the mean numbers of mRNA, protein, and xanthosine in the
-	stochastic results are as predicted from the phase portraits. It does,
+	stochastic results are as predicted from the phase portraits.
+Also note the similarity to the shape of the experimental distributions that were presented in section ``experimental motivation''.
+It does,
 	however, become clear that the phase portraits do not tell whether the cells
@@
 	expression state during the run time of our simulations.
+This actually is a good finding: the biologically interesting feature is the switch-like behavior, whereas bimodality can be problematic for biology as this means that there is a large insecurity in the input-output behavior of a bacteria's reaction to xanthosine.
+Thus, the fact that bimodality only occurs in a small regime means that the circuit is a strong switch.
 
@@
 
-	\paragraph*{Time evolution and switching times.}
-	Fig~\ref{fig:StochT} shows the time evolution of the protein concentration for one exemplary run of the simulation.
+	\paragraph*{Time evolution and adaptation times.}
+	Fig~\ref{fig9:StochT} shows the time evolution of the protein concentration for one exemplary run of the simulation.
 	Again, the simulation was started with a mRNA, protein and intracellular
 	xanthosine count of 0 and was run for a simulated time of $5 \cdot 10^5
-	\unit{s}$. In this specific example, switching occurred after $10^5
+	\unit{s}$. In this specific example, the fixed point was reached after $10^5
 	\unit{s}$.
-
-	Comparing this to the experimentally expected timescales~\cite{Chure2019} is difficult,
-	because the switching time strongly depends on the extracellular xanthosine
+Of course, the system moves to its mean steady-state much faster when the extracellular xanthosine concentration is further away from the bimodal regime (in analogy to a ``critical point'').
+For comparison, the trajectory obtained from integrating the deterministic system is also shown in Fig~\ref{fig9:StochT}.
+It clearly agrees well with the simulation.
+There are some cases where the big adaptation step happens slightly earlier or later, but still at around $10^5 \unit{s}$.
+
+	\begin{figure}%[h!]
+		\centering
+		\includegraphics{media/Fig9_evolution.eps}
+		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB).}
+			Blue shows the results from one run of the stochastic simulation, orange shows the trajectory obtained from solving the deterministic ODE's.
+The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
+			at an mRNA, protein and intracellular xanthosine count of 0. In this
+			case the system reached the high-expression state after $10^5
+			\unit{s}$. The parameters that were used are the same as in
+			Table~\ref{table1:nondim}, the only exception being the
+			extracellular xanthosine concentration, which was chosen to be
+			$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
+			$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$) just as in
+			Fig~\subref{fig8:stochC}{(C)}.}
+		\label{fig9:StochT}
+	\end{figure}
+
+	Comparing this to experimentally expected timescales is difficult,
+	because the adaptation time strongly depends on the extracellular xanthosine
 	concentration. Experiments were always stopped after a few hours, and in
@@
 	bistability could lead to experimental bimodality if the experiment is
-	stopped too early. In this case, the observed switching time would be
+	stopped too early. In this case, the observed timescales would be
 	shorter, which makes the comparison to our simulations even harder. Thus, we
@@
 
-	Nevertheless, we do warn the reader that the timescales in the simulations
-	should be taken with reservation. Cell divisions are not considered here,
-	and neither is the burstiness of transcription and translation. This means
+	Nevertheless, we do warn the reader that the timescales in the simulations and even more so in the deterministic system
+	should be taken with reservation. Fluctuations in the parameters are not considered here, and neither are cell divisions or the burstiness of transcription and translation. This means
 	that stochasticity may be larger in the real system which should have an
-	influence on the timescales and probably shorten the time until switching
-	occurs.
+	influence on the timescales and may shorten the time until the fixed point is reached.
+However, including transcriptional bursts in the stochastic simulation changes little in the output: the qualitative behavior remains the same but fluctuations around the mean as well as in the adaptation time become larger and bimodality already occurs at lower $\mathrm{[c]_a}$ (see~\nameref{S1_Text}).
+Still, the fact that there are no big changes shows that the system is stable to stochastic perturbations and our assumptions should not be too significant.
+
+	Note that in a biologically realistic setting, the extracellular xanthosine concentration would probably be high enough to be in the regime where there only is the upper fixed point and thus no bistability or even bimodality.
+As mentioned before, the system reaches its mean steady-state faster in that regime, and hence, the bacteria can adapt more quickly.
+
+	\paragraph*{Bifurcation diagram and hysteresis.} 
+	In Figs~\ref{fig8:stochC} and~\ref{fig9:StochT}, the simulations were started at initial intracellular concentrations of 0 to investigate what happens if xanthosine is suddenly added to the cell's environment.
+We can now ask the opposite question: what happens when xanthosine is removed from the extracellular environment?
+To answer this, the simulation was started with initially fully	induced cells, i.e.
+at the mRNA, protein and intracellular xanthosine counts of the high fixed point in the corresponding phase portrait.
+
+	The distributions we obtained can be found in~\nameref{S1_Text}.
+Here, we instead present the results in form of a bifurcation diagram in Fig~\ref{fig10:Bifurcation}.
+In blue, the positions of the deterministic fixed points for the corresponding extracellular xanthosine concentration are shown.
+As explained before, there is only one fixed point for low and high values of $\mathrm{[c]_a}$, but in between, there are three.
+In yellow and orange, the results from the stochastic simulation are shown, where the mean of the resulting distribution was taken.
+For the orange points, the simulation was started at $\mathrm{[m]_a}=\mathrm{[p]_a}=\mathrm{[x]_a}=0$, leading to the distributions in Fig~\ref{fig8:stochC}.
+In contrast, the yellow points show the results from starting the simulation at the $\mathrm{[m]_a}$, $\mathrm{[p]_a}$, and $\mathrm{[x]_a}$ values of the high fixed point.
+Note that by taking the mean of each distribution, we hide the feature of bimodality.
+For that reason, we additionally show the position of the two peaks in the bimodal distribution by splitting it into two parts at its minimum and calculating the mean of each of the parts.
+Also note that it suffices to show the bifurcation diagram in one dimension (here: protein concentration) because the fixed points are uniquely determined by that (recall Fig~\ref{fig4:bistability}, the fixed points are unique points in the 3D system).
 
@@
 		\centering
-		\includegraphics{media/TimeEvolution.eps}
-		\caption{{\bf Time evolution of protein (XapA/XapB) from one run of the simulation.}
-			The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
-			at an mRNA, protein and intracellular xanthosine count of 0. In this
-			case the system switched into the high-expression state after $10^5
-			\unit{s}$. The parameters that were used are the same as in
-			Table~\ref{table1:nondim}, the only exception being the
-			extracellular xanthosine concentration, which was chosen to be
-			$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
-			$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$) just as in
-			Fig~\subref{fig8:stochC}{(B)}.}
-		\label{fig:StochT}
+		\includegraphics{media/Fig10_bifurcation.eps}
+		\caption{{\bf Bifurcation diagram showing the hysteresis loop in the stochastic system.}
+			The blue line shows the positions of deterministic fixed points, where the dashed line indicates the instability of the middle one.
+The orange and yellow points show the mean in the stochastic simulations when started at zero and at the high fixed point, respectively.
+The positions of the two peaks in the bimodal distributions are indicated by smaller points, connected by dotted lines.
+The arrows illustrate the hysteresis loop in the stochastic simulations.}
+		\label{fig10:Bifurcation}
 	\end{figure}
 
-	\paragraph*{Hysteresis.} 
-	In Figs~\ref{fig8:stochC} and~\ref{fig:StochT}, the simulations were started at initial
-	intracellular concentrations of 0 to investigate what happens if xanthosine
-	is suddenly added to the cell's environment. We can now ask the opposite
-	question: what happens when xanthosine is removed from the extracellular
-	environment? To answer this, the simulation was started with initially fully
-	induced cells, i.e. at the mRNA, protein and intracellular xanthosine counts
-	of the high fixed point in the corresponding phase portrait.
-
-	The resulting distributions can be found in~\nameref{S1_Text}. They show
-	clear hysteresis effects: there exist extracellular xanthosine concentrations where
-	initially uninduced cells remained uninduced and initially induced cells remain induced.
-	Only when the second and the third fixed point are very close can initially
-	induced cells switch to the uninduced state. Interestingly, this
-	behavior is symmetric to the ``switching on'' in the previous paragraphs,
-	where the lower two fixed points needed to be about as close as the upper
-	two need to be now for ``switching off''.
+	The figure clearly shows the hysteresis in the system: there exist extracellular xanthosine concentrations where
+	initially uninduced cells remain uninduced and initially induced cells remain induced.
+Only when the second and the third fixed point are very close can initially induced cells ``switch off'' to the uninduced state.
+This
+	behavior is symmetric to the ``switching on'' in the previous paragraphs. The arrows in Fig~\ref{fig10:Bifurcation} indicate the hysteresis loop.
 
@@
 	concentrations.
+
+	In addition to the hysteresis, Fig~\ref{fig10:Bifurcation} also illustrates the astonishingly good agreement of the stochastic simulation and the deterministic results, despite having copy numbers below 10 in some cases.
+Although one should be cautious about this because of higher stochasticity in the real biological system (addressed in the previous paragraph), the result does show that the switch-like feature in the circuit is strong and stable.
 
@@
 	concentration decreases, which explains previous experimental findings.
-
-	We have found that no bistability can emerge from the genetic circuit unless
+	Furthermore, our simulations show that the switch-like behavior is very robust.
+
+	However, we have found that no bistability can emerge from the genetic circuit unless
 	at least one component has two binding sites for its activator. Additional
@@
 	calculated from arguments made in~\cite{Cherry2000}. However, the
-	simulations showed that deterministic bistability does not mean that
-	bimodality occurs, which is why we have refrained from showing such
+	simulations showed that in the deterministically bistable regime, it is dependent on initial conditions which fixed points the system occupies, which is why we have refrained from showing such phase
 	diagrams. Furthermore, the timescales in the problem could be investigated
 	more thoroughly, for example the dependence of the switching time on
-	$\n{[c]_a}$, but such an analysis would probably need to
-	account for the burstiness of transcription and translation as well as for
-	cell divisions, which is anything but straightforward.
-
+	$\n{[c]_a}$, but such an analysis should probably account for	cell divisions and fluctuations in the parameters, which is not straightforward.
+Lastly, it could be interesting to investigate the magnitude of the fluctuations around the fixed point away from and near the bifurcations in $\mathrm{[c]_a}$.
+
+	Despite the small copy numbers at the lower fixed point, the stochastic simulations are in excellent agreement with deterministic predictions.
 All model parameters could be reasonably estimated despite the paucity of
 	experimental knowledge about the model system.
-The concentrations of mRNA, protein, and xanthosine at the fixed points as well as all qualitative features are as expected from similar systems and the few experiments on the \emph{xap} circuit, which suggests that the model captures the relevant components of the system correctly and is able to describe its
+The concentrations of mRNA, protein, and xanthosine at the fixed points as well as all qualitative features are as expected from similar systems and the few experiments on the \emph{xap} circuit.
+These points suggest that the model captures the relevant components of the system correctly and is able to describe its
 	dynamics. 
@@
 	in the model, parameter estimation, elaborations on the results, and the
-	chemical master equation of this circuit.
+	chemical master equation of this circuit. Experimental materials and methods.
 
@@
 	questions that arose, Nathan Belliveau for bringing similar systems to our
-	attention, and Jane Kondev and Jin Wang for interesting discussions.
+	attention, Patrick Lenggenhager for his help with \emph{Mathematica} issues, and Jane Kondev and Jin Wang for interesting discussions.
 	For assistance with illustrations, we also thank Nigel Orme and Griffin Chure.
@@
 
-		\bibitem{Chure2019}
-		Chure G, et~al.
-		\newblock Measurements on the \emph{xapABR} system in \emph{E. coli.};
-		Unpublished.
-
 		\bibitem{Marzen2013}

### 2020-02-19 s1_text/s1_text.tex
KL edits to SI re reviewer comments

@@
 \newpage
-\linenumbers
 
@@
 
-Furthermore, we do not expect our approximation to have a large influence: the interaction energy will make the doubly bound state dominate, and the polymerase-XapR interaction should be weak in the singly bound state, because very weak activation is observed in the experiment when one of the XapR binding sites is removed~\cite{Chure2019}.
+Furthermore, we do not expect our approximation to have a large influence: the interaction energy will make the doubly bound state dominate, and the polymerase-XapR interaction should be weak in the singly bound state, because very weak activation is observed in the experiment when one of the XapR binding sites is removed (see section~\ref{sec:exp} in this supporting information text).
 If $e^{\Delta \epsilon_{\n{X1}}}$ and $e^{\Delta \epsilon_{\n{X2}}}$ are small, we have $w_{\n{p}} w_{\n{X1}} e^{\Delta \epsilon_{\n{X1}}} \approx w_{\n{p}} w_{\n{X2}} e^{\Delta \epsilon_{\n{X2}}} \approx w_{\n{p}} w_{\n{X}}$.
@@
 It could, however, also be that the basal level of XapA is enough to obtain the small degradation effect without any other pathways.
+
+\FloatBarrier
+\newpage
 
@@
 
+\FloatBarrier
+\newpage
+
 \section{Additional plots and explanations of the results}
@@
 Here, several of the parameters from the 3D system could now be combined into one by defining $\rho \defeq \frac{\rho_{\n{m}} \rho_{\n{p}}}{\gamma_{\n{mp}}}$.
+\FloatBarrier
 
@@
 
+\paragraph*{Including transcriptional bursting.} To test the robustness of our results to larger stochasticity (which is expected in the real biological system), we included some burstiness in transcription in our stochastic simulations.
+We assumed a geometric distribution for the burst size $s$, i.e.
+$p(s) = (1/4)\cdot (1-1/4)^{s-1}$ with mean $4$, which is a bit larger than what usually occurs in biology \textcolor{red}{citation missing}.
+We then cut this distribution at a burst size of 10, meaning we did not consider larger bursts, leading to an actual mean of around $3.2$.
+The transcription rate was rescaled such that the mean number of mRNA's produced remained the same.
+
+We observe the same qualitative behavior but with stronger fluctuations around the mean and also larger variation in the time to the steady-state mean.
+As an example, Fig~\ref{evolbursts} shows the time evolution of two runs of the simulation with bursting for the same conditions (i.e.
+parameters and starting values) as the time evolution that was shown in the main text.
+Fig~\ref{distbursts} shows the distribution obtained for the same and an additional condition (i.e.
+parameters and starting values) as for the distributions shown in the main text.
+The variance in the timescales and distributions is larger than without bursting.
+Furthermore, switching already happens for lower $\mathrm{[c]_a}$: in \subref{distbursts}{C}, most cells are in the high expression state, whereas in the corresponding distribution in the main text, there were many uninduced cells, i.e.
+the first peak was higher.
+To obtain a bimodal distribution like in the main text, $\mathrm{[c]_a}$ now needs to be set to $17.5$ instead of $18.5$ as shown in \subref{distbursts}{B}.
+This was to be expected, since the larger fluctuations make it easier for the system to jump across the middle fixed point.
+
+Still, there are no big qualitative differences to the results without bursting.
+Thus, the assumptions we made that lead to less stochasticity than in the real biological system are not expected to have a very significant influence on our results, and the switch-like behavior seems to be stable to stochastic perturbations.
+\textcolor{red}{Fig 7 C not enough runs yet}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=1\textwidth]{FigSI6_7.pdf}
+	\caption{{\bf Time evolution from the stochastic simulation when transcriptional bursting is included.}
+		(A) and (B) each show the time evolution from one run of the simulation. The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
+		at an mRNA, protein and intracellular xanthosine count of 0. The parameters that were used are the same as in the
+		table in the main text, the only exception being the
+		extracellular xanthosine concentration, which was chosen to be
+		$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
+		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$).}
+	\label{evolbursts}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=1\textwidth]{FigSI6_7_2.pdf}
+	\caption{{\bf Distributions from the stochastic simulation when transcriptional bursting is included.}
+		(A) shows the distribution obtained from the simulation with $\n{[c]_a}=12$, (B), (C), and (D) show the same for $\n{[c]_a}=17.5$, $\mathrm{[c]_a}=18.5$, and $\mathrm{[c]_a}=25$, respectively (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
+		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$). All other parameters are as in the table in the main text, and the simulation
+		was run 1000 times for $10^6 \unit{s}$ each (simulated time) and
+		started at a mRNA, protein and intracellular xanthosine count of 0.}
+	\label{distbursts}
+\end{figure}
+
 \paragraph*{Hysteresis effects.}
@@
 
+\FloatBarrier
+\newpage
+
 \section{Chemical master equation}
@@
 \FloatBarrier
+\newpage
+
+\section{Experimental materials and methods} \label{sec:exp}
+
+\textcolor{red}{missing}
+
+\FloatBarrier
+\newpage
 
@@
 	\newblock doi:{10.1016/j.cels.2018.02.004}.
-
-	\bibitem{Chure2019}
-	Chure G, et~al.
-	\newblock Measurements on the \emph{xapABR} system in \emph{E. coli.};
-	Unpublished.
 

### 2020-02-22 main_text/main_text.tex
Another round of edits by KL

@@
 
-	\begin{figure}[h!]
+	\begin{figure}%[h!]
 		\centering
@@
 (A) shows the wild type circuit while in (B) the genes \emph{xapABR} and in (C) the genes \emph{xapAB} are removed.
-The wild type circuit behaves like a switch.}
+The wild type circuit behaves like a switch.
+\textcolor{red}{placeholder figure up to now}}
 		\label{fig1:data}
@@
 	Clearly, XapR regulates the \emph{xapAB} promoter.
-As mentioned in the introduction, there are two binding sites for XapR on the \emph{xapAB} promoter: one directly at the polymerase binding site, and one further upstream.
+As mentioned in the introduction, there are two binding sites for XapR on the \emph{xapAB} promoter: one directly at the polymerase binding site and one further upstream.
 When the upstream binding site is removed, bimodality is lost and the expression of \emph{xapAB} becomes several orders of magnitude weaker, as can be seen in Fig~\ref{}.
@@
 The construction and exploration of this model occupies the remainder of this work.
-We leave it for future work to quantitatively dissect and test the model, in the manner of, for example, \textcolor{red}{citations}.
+We leave it for future work to quantitatively dissect and test the model, in the manner of~\cite{Garcia2011,RazoMejia2018,Chure2019}, for example.
 
@@
 Nevertheless, in our case we find that stochastic simulations agree well with the deterministic results.
-Thus, deterministic phase portraits are a good starting point.
+Thus, deterministic phase portraits are a surprisingly good starting point.
 
@@
 	expression. In between, there are two deterministically stable expression
-	levels. Later on in this text, we show this result in form of a bifurcation diagram.
+	levels. Another means of visualizing this is with an bifurcation diagram, which we discuss in detail later.
 
@@
 	experiments: the mean numbers of mRNA, protein, and xanthosine in the
-	stochastic results are as predicted from the phase portraits.
-Also note the similarity to the shape of the experimental distributions that were presented in section ``experimental motivation''.
-It does,
+	stochastic results are as predicted from the phase portraits. It does,
 	however, become clear that the phase portraits do not tell whether the cells
@@
 	expression state during the run time of our simulations.
-This actually is a good finding: the biologically interesting feature is the switch-like behavior, whereas bimodality can be problematic for biology as this means that there is a large insecurity in the input-output behavior of a bacteria's reaction to xanthosine.
-Thus, the fact that bimodality only occurs in a small regime means that the circuit is a strong switch.
+This actually is a good finding: it implies that bistability, characterized by two separate \emph{stable} steady-states, largely persists even in the presence of stochastic fluctuations.
+Thus, the fact that bimodality only occurs in a small regime means that the circuit is a strong switch and the deterministic picture is sufficient except in a small range of parameter space.
+Whether or not that is where the system actually is remains a question for future experiments.
 
@@
 	transcription and translation bursts lead to higher stochasticity and cell
-	division leads to some discontinuity in the process.
+	division leads to some discontinuity in the process. The effect of bursts is addressed in~\nameref{S1_Text}.
 
@@
 	If, however, the mRNA number at the third (high) fixed point is too low, the
-	system will not be able to resolve the two points anymore. It is too low when the distance between the first and the third fixed point
-	becomes as low as the stochastic fluctuations in the system, which is around
-	3 mRNA molecules. In Fig~\ref{fig8:stochC}, the mRNA number at the low fixed
-	point was fluctuating between 0 and 2 with a mean significantly smaller than
-	1, and at the high fixed point it was varying between 2 and 6.
+	system may not be able to resolve the two points anymore.
+The tolerance to this is surprisingly large, though: In Fig~\ref{fig8:stochC}, the distance between the first and the third fixed point is around 3 mRNA molecules.
+While this is around the size of the fluctuations in Fig~\ref{fig8:stochC}, the latter become as large as 10 mRNA molecules when bursts are included in the simulation (see~\nameref{S1_Text}) and yet, the system is able to resolve the fixed points very well.
 
 	\paragraph*{Time evolution and adaptation times.}
-	Fig~\ref{fig9:StochT} shows the time evolution of the protein concentration for one exemplary run of the simulation.
+	Fig~\subref{fig9:StochT}{A} shows the time evolution of the protein concentration for one typical run of the simulation.
 	Again, the simulation was started with a mRNA, protein and intracellular
 	xanthosine count of 0 and was run for a simulated time of $5 \cdot 10^5
-	\unit{s}$. In this specific example, the fixed point was reached after $10^5
-	\unit{s}$.
-Of course, the system moves to its mean steady-state much faster when the extracellular xanthosine concentration is further away from the bimodal regime (in analogy to a ``critical point'').
-For comparison, the trajectory obtained from integrating the deterministic system is also shown in Fig~\ref{fig9:StochT}.
-It clearly agrees well with the simulation.
-There are some cases where the big adaptation step happens slightly earlier or later, but still at around $10^5 \unit{s}$.
+	\unit{s}$. In this specific example, the fixed point was reached after roughly $1.5 \cdot 10^5
+	\unit{s}$ (adaptation time).
+This time varies: Fig~\subref{fig9:StochT}{B} shows the distribution of adaptation times from 1000 runs of the simulation.
+In both figures, the trajectory obtained from integrating the deterministic system is shown for comparison.
+It clearly agree well with the simulation on average.
 
@@
 		\includegraphics{media/Fig9_evolution.eps}
-		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB).}
-			Blue shows the results from one run of the stochastic simulation, orange shows the trajectory obtained from solving the deterministic ODE's.
+		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
+			(A) Blue shows the results from one run of the stochastic simulation, orange shows the trajectory obtained from solving the deterministic ODE's.
 The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
-			at an mRNA, protein and intracellular xanthosine count of 0. In this
-			case the system reached the high-expression state after $10^5
-			\unit{s}$. The parameters that were used are the same as in
+			at an mRNA, protein and intracellular xanthosine count of 0. The parameters that were used are the same as in
 			Table~\ref{table1:nondim}, the only exception being the
@@
 			$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$) just as in
-			Fig~\subref{fig8:stochC}{(C)}.}
+			Fig~\subref{fig8:stochC}{(C)}.
+(B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 1000 runs of the simulation (same conditions as (A)).
+The orange line shows the corresponding deterministic time.
+The green dashed line right next to it indicated the mean of the blue distribution.}
 		\label{fig9:StochT}
@@
 However, including transcriptional bursts in the stochastic simulation changes little in the output: the qualitative behavior remains the same but fluctuations around the mean as well as in the adaptation time become larger and bimodality already occurs at lower $\mathrm{[c]_a}$ (see~\nameref{S1_Text}).
-Still, the fact that there are no big changes shows that the system is stable to stochastic perturbations and our assumptions should not be too significant.
-
-	Note that in a biologically realistic setting, the extracellular xanthosine concentration would probably be high enough to be in the regime where there only is the upper fixed point and thus no bistability or even bimodality.
-As mentioned before, the system reaches its mean steady-state faster in that regime, and hence, the bacteria can adapt more quickly.
+Still, the fact that there are no big changes shows that the system is stable to stochastic perturbations and our particular assumptions should not be too significant.
+
+	Of course, the system moves to its mean steady-state much faster when the extracellular xanthosine concentration is further away from the bimodal regime (in analogy to a ``critical point''). 
+	Note that it could well be that in reality, the extracellular xanthosine concentration is high enough to be in the regime where there only is the upper fixed point and thus no bistability or even bimodality.
+As the system reaches its mean steady-state faster in that regime, the bacteria could adapt more quickly.
 
@@
 	The distributions we obtained can be found in~\nameref{S1_Text}.
-Here, we instead present the results in form of a bifurcation diagram in Fig~\ref{fig10:Bifurcation}.
+Here, we instead present the results in the form of a bifurcation diagram in Fig~\ref{fig10:Bifurcation}.
 In blue, the positions of the deterministic fixed points for the corresponding extracellular xanthosine concentration are shown.
 As explained before, there is only one fixed point for low and high values of $\mathrm{[c]_a}$, but in between, there are three.
-In yellow and orange, the results from the stochastic simulation are shown, where the mean of the resulting distribution was taken.
+In yellow and orange, the results from the stochastic simulation are shown, where the mean of each distribution was taken.
 For the orange points, the simulation was started at $\mathrm{[m]_a}=\mathrm{[p]_a}=\mathrm{[x]_a}=0$, leading to the distributions in Fig~\ref{fig8:stochC}.
-In contrast, the yellow points show the results from starting the simulation at the $\mathrm{[m]_a}$, $\mathrm{[p]_a}$, and $\mathrm{[x]_a}$ values of the high fixed point.
+In contrast, the yellow points result from starting the simulation at the $\mathrm{[m]_a}$, $\mathrm{[p]_a}$, and $\mathrm{[x]_a}$ values of the high fixed point.
 Note that by taking the mean of each distribution, we hide the feature of bimodality.
-For that reason, we additionally show the position of the two peaks in the bimodal distribution by splitting it into two parts at its minimum and calculating the mean of each of the parts.
-Also note that it suffices to show the bifurcation diagram in one dimension (here: protein concentration) because the fixed points are uniquely determined by that (recall Fig~\ref{fig4:bistability}, the fixed points are unique points in the 3D system).
+For that reason, we additionally show the approximate position of the two peaks in the bimodal distribution.
+We do this by estimating the position of the minimum between the two peaks of the distribution, splitting it into two parts there, and calculating the mean of each of the parts.
+Also note that it suffices to show the bifurcation diagram in one dimension (here protein concentration) because this uniquely determines the fixed points (recall from Fig~\ref{fig4:bistability}, the fixed points are unique points in the 3D system).
 
@@
 		\caption{{\bf Bifurcation diagram showing the hysteresis loop in the stochastic system.}
-			The blue line shows the positions of deterministic fixed points, where the dashed line indicates the instability of the middle one.
+			The blue line shows the positions of deterministic fixed points, while the dashed line indicates the instability of the middle one.
 The orange and yellow points show the mean in the stochastic simulations when started at zero and at the high fixed point, respectively.
@@
 	In addition to the hysteresis, Fig~\ref{fig10:Bifurcation} also illustrates the astonishingly good agreement of the stochastic simulation and the deterministic results, despite having copy numbers below 10 in some cases.
-Although one should be cautious about this because of higher stochasticity in the real biological system (addressed in the previous paragraph), the result does show that the switch-like feature in the circuit is strong and stable.
+Although one should be cautious about this because of higher stochasticity in the real biological system (addressed above), the result does show that the switch-like feature in the circuit is strong and stable.
 
@@
 	calculated from arguments made in~\cite{Cherry2000}. However, the
-	simulations showed that in the deterministically bistable regime, it is dependent on initial conditions which fixed points the system occupies, which is why we have refrained from showing such phase
+	simulations showed that in the deterministically bistable regime, which fixed point(s) the system occupies is dependent on initial conditions, which is why we have refrained from showing such phase
 	diagrams. Furthermore, the timescales in the problem could be investigated
@@
 
+		\bibitem{Garcia2011}
+		Garcia HG, Phillips R.
+		\newblock Quantitative dissection of the simple repression input-output
+		function.
+		\newblock Proceedings of the National Academy of Sciences of the United States
+		of America. 2011;108 29:12173--8.
+
+		\bibitem{RazoMejia2018}
+		Razo-Mejia M, Barnes SL, Belliveau NM, Chure G, Einav T, Lewis M, et~al.
+		\newblock Tuning Transcriptional Regulation through Signaling: A Predictive
+		Theory of Allosteric Induction.
+		\newblock Cell Systems. 2018;6(4):456 -- 469.e10.
+		\newblock doi:{https://doi.org/10.1016/j.cels.2018.02.004}.
+
+		\bibitem{Chure2019}
+		Chure G, Kaczmarek ZA, Phillips R.
+		\newblock Physiological Adaptability and Parametric Versatility in a Simple
+		Genetic Circuit.
+		\newblock bioRxiv. 2019;doi:{10.1101/2019.12.19.878462}.
+
 		\bibitem{Marzen2013}
@@
 
-		\bibitem{RazoMejia2018}
-		Razo-Mejia M, Barnes SL, Belliveau NM, Chure G, Einav T, Lewis M, et~al.
-		\newblock Tuning Transcriptional Regulation through Signaling: A Predictive
-		Theory of Allosteric Induction.
-		\newblock Cell Systems. 2018;6(4):456 -- 469.e10.
-		\newblock doi:{10.1016/j.cels.2018.02.004}.
-
 	\end{thebibliography}

### 2020-02-22 s1_text/s1_text.tex
Another round of edits by KL

@@
 	\caption{{\bf A family of curves for increasing $\n{[c]_a}$.}
-		The blue curve is the protein nullcline, which does not change with $\n{[c]_a}$. The other lines are the xanthosine nullclines for different values of the extracellular xanthosine concentration.}
+		The blue, dashed curve is the protein nullcline, which does not change with $\n{[c]_a}$.
+The other curves are the xanthosine nullclines for different values of the extracellular xanthosine concentration.}
 	\label{figS5:incC}
@@
 
-\paragraph*{Including transcriptional bursting.} To test the robustness of our results to larger stochasticity (which is expected in the real biological system), we included some burstiness in transcription in our stochastic simulations.
+\paragraph*{Including transcriptional bursting.} To test the robustness of our results to larger stochasticity (which is expected in the real biological system), we explored how the addition of burstiness in transcription altered our stochastic simulations.
 We assumed a geometric distribution for the burst size $s$, i.e.
-$p(s) = (1/4)\cdot (1-1/4)^{s-1}$ with mean $4$, which is a bit larger than what usually occurs in biology \textcolor{red}{citation missing}.
-We then cut this distribution at a burst size of 10, meaning we did not consider larger bursts, leading to an actual mean of around $3.2$.
+$p(s|\theta) = \theta\cdot (1-\theta)^{s-1}$ with mean $\theta^{-1}=4$, which is comparable to or slightly larger than reported values~\cite{Golding2005}.
+We then truncate this distribution at a burst size of 10, meaning we did not consider larger bursts, leading to an actual mean of around $3.2$.
 The transcription rate was rescaled such that the mean number of mRNA's produced remained the same.
 
-We observe the same qualitative behavior but with stronger fluctuations around the mean and also larger variation in the time to the steady-state mean.
-As an example, Fig~\ref{evolbursts} shows the time evolution of two runs of the simulation with bursting for the same conditions (i.e.
+We observe the same qualitative behavior but with stronger fluctuations around the mean and larger variation in the time to reach the steady-state mean.
+As an example, Fig~\subref{evolbursts}{A} shows the time evolution of one run of the simulation with bursting for the same conditions (i.e.
 parameters and starting values) as the time evolution that was shown in the main text.
+In addition, the distribution of adaptation times can be seen in Fig~\subref{evolbursts}{B}.
 Fig~\ref{distbursts} shows the distribution obtained for the same and an additional condition (i.e.
@@
 Thus, the assumptions we made that lead to less stochasticity than in the real biological system are not expected to have a very significant influence on our results, and the switch-like behavior seems to be stable to stochastic perturbations.
-\textcolor{red}{Fig 7 C not enough runs yet}
 
@@
 	\includegraphics[width=1\textwidth]{FigSI6_7.pdf}
-	\caption{{\bf Time evolution from the stochastic simulation when transcriptional bursting is included.}
-		(A) and (B) each show the time evolution from one run of the simulation. The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
+	\caption{{\bf Time evolution and adaptation times from the stochastic simulation when transcriptional bursting is included.}
+		(A) shows the time evolution from one run of the simulation. The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
 		at an mRNA, protein and intracellular xanthosine count of 0. The parameters that were used are the same as in the
@@
 		$\n{[c]_a} = 25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
-		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$).}
+		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$).
+(B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 200 runs of the simulation (same conditions as (A)).
+The distribution from the main text (without bursting) is shown in orange for comparison.
+\textcolor{red}{will add some more runs, but the simulation takes forever}}
 	\label{evolbursts}
@@
 	\caption{{\bf Distributions from the stochastic simulation when transcriptional bursting is included.}
-		(A) shows the distribution obtained from the simulation with $\n{[c]_a}=12$, (B), (C), and (D) show the same for $\n{[c]_a}=17.5$, $\mathrm{[c]_a}=18.5$, and $\mathrm{[c]_a}=25$, respectively (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
+		Blue shows the distributions from the simulation with bursts, orange shows those without bursts from the main text for comparison.
+The following values of $\mathrm{[c]_a}$ were used: (A) $\n{[c]_a}=12$, (B) $\n{[c]_a}=17.5$ (not shown in main text), (C) $\mathrm{[c]_a}=18.5$, and (D) $\mathrm{[c]_a}=25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
 		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$). All other parameters are as in the table in the main text, and the simulation
@@
 As can be seen from the equation, none of the propensities depends explicitly on time and hence, we can use the hybrid algorithm between classical Gillespie and \texttau-leaping that we relied on.
+
+Note that in the case with bursts that is discussed above, the transcription term $r_{\n{m}} p_{\n{active}}(x) P(m-1,p,x)$ is changed to $r_{\n{m}} p_{\n{active}}(x) \sum_{k=1}^{10} P(m-k,p,x) \frac{\theta\cdot (1-\theta)^{k-1}}{Z}$ with $Z = \sum_{k=1}^{10} k \cdot \theta\cdot (1-\theta)^{k-1}$ and $\theta = 1/4$ in our example.
 
@@
 
+	\bibitem{Golding2005}
+	Golding I, Paulsson J, Zawilski SM, Cox EC.
+	\newblock Real-Time Kinetics of Gene Activity in Individual Bacteria.
+	\newblock Cell. 2005;123(6):1025--1036.
+	\newblock doi:{10.1016/j.cell.2005.09.031}.
+
 \end{thebibliography}

### 2020-02-22 main_text/main_text.tex
Add fig w/ real data from GC + minor edits from KL

@@
 	\section*{Experimental motivation} \label{sec:experiment}
-	Our work was motivated by the experimental observation of bimodality in the \emph{xap} circuit, which is shown in Fig~\ref{fig1:data}.
-We focus on the essential findings here and refer the reader to~\nameref{S1_Text} for more experimental details.
-Briefly, we placed a fluorescent reporter under the control of the wild type (wt) \emph{xapAB} promoter.
-This construct was placed in three different backgrounds: \textDelta\emph{xapABR}, \textDelta\emph{xapAB}, and wt.
-Fig~\subref{fig1:data}{A} shows that for increasing xanthosine concentration, expression in the wt-background increases in a switch-like way: it is nearly zero when there is little xanthosine (``uninduced state''), but increases drastically when there is more (``induced state'').
-In between, the aforementioned bimodal distribution is obtained, where some bacteria are in the uninduced expression state and others are induced.
-\textcolor{red}{maybe remove: In addition, we observed that the property of being induced seems to be passed on to daughter cells, suggesting that there is not merely bimodality, but bistability.}
+	Our work was motivated by the experimental observation of bimodality
+	in the \emph{xap} circuit, which is shown in Fig~\ref{fig1:data}. We
+	focus on the essential findings here and refer the reader
+	to~\nameref{S1_Text} for more experimental details. Briefly, we
+	placed a fluorescent reporter under the control of the wild type
+	(wt) \emph{xapAB} promoter. This construct was placed in three
+	different backgrounds: \textDelta\emph{xapABR},
+	\textDelta\emph{xapAB}, and wt. The right panel of
+	Fig~\subref{fig1:data}{A} shows that for increasing xanthosine
+	concentration, expression in the wt-background increases in a
+	switch-like way: it is nearly zero when there is little xanthosine
+	(``uninduced state''), but increases drastically when there is more
+	(``induced state''). In between, the aforementioned bimodal
+	distribution is obtained, where some bacteria are in the uninduced
+	expression state and others are induced.
 
@@
 		\centering
-		\includegraphics[width=0.4\textwidth]{media/Fig1_data.pdf}
-		\caption{{\bf Experimental data on the \emph{xap} circuit.} The expression of the \emph{xapAB} promoter was measured for different concentrations of xanthosine in the bacteria's environment.
-(A) shows the wild type circuit while in (B) the genes \emph{xapABR} and in (C) the genes \emph{xapAB} are removed.
-The wild type circuit behaves like a switch.
-\textcolor{red}{placeholder figure up to now}}
+		\includegraphics[width=1\textwidth]{media/Fig1_data.eps}
+		\caption{{\bf Experimental data on the \emph{xap} circuit.}
+		(A) The expression of the \emph{xapAB} promoter was measured for
+		different concentrations of xanthosine in the bacteria's
+		environment. The right panel shows the wild type circuit while
+		on the left the genes \emph{xapABR} and in the middle the genes
+		\emph{xapAB} are removed. The wild type circuit behaves like a
+		switch. Note that the fluorescence scales of the middle panel
+		and the other two are not comparable.
+		(B) The fold-change in gene expression upon addition of
+		xanthosine for the wt promoter and for the promoter with only
+		the XapR binding site right at the polymerase binding site.
+		}
 		\label{fig1:data}
@@
 
-	Clearly, XapR regulates the \emph{xapAB} promoter.
-As mentioned in the introduction, there are two binding sites for XapR on the \emph{xapAB} promoter: one directly at the polymerase binding site and one further upstream.
-When the upstream binding site is removed, bimodality is lost and the expression of \emph{xapAB} becomes several orders of magnitude weaker, as can be seen in Fig~\ref{}.
-	\textcolor{red}{data missing}
+	Clearly, XapR regulates the \emph{xapAB} promoter. As mentioned in
+	the introduction, there are two binding sites for XapR on the
+	\emph{xapAB} promoter: one directly at the polymerase binding site
+	and one further upstream. When the upstream binding site is removed,
+	bimodality is lost and the activation of the \emph{xapAB}-expression
+	by xanthosine (via XapR) becomes negligible, as can be seen in
+	Fig~\subref{fig1:data}{B}.
 
@@
 	\unit{s}$ (adaptation time).
-This time varies: Fig~\subref{fig9:StochT}{B} shows the distribution of adaptation times from 1000 runs of the simulation.
-In both figures, the trajectory obtained from integrating the deterministic system is shown for comparison.
-It clearly agree well with the simulation on average.
+	This time varies: Fig~\subref{fig9:StochT}{B} shows the distribution
+	of adaptation times from 1000 runs of the simulation (where we
+	neglected the around 10 runs with adaptation times larger than $6
+	\cdot 10^5 \unit{s}$). In both figures, the trajectory obtained from
+	integrating the deterministic system is shown for comparison. It
+	clearly agree well with the simulation on average.
 
@@
 		\caption{{\bf Bifurcation diagram showing the hysteresis loop in the stochastic system.}
-			The blue line shows the positions of deterministic fixed points, while the dashed line indicates the instability of the middle one.
-The orange and yellow points show the mean in the stochastic simulations when started at zero and at the high fixed point, respectively.
-The positions of the two peaks in the bimodal distributions are indicated by smaller points, connected by dotted lines.
-The arrows illustrate the hysteresis loop in the stochastic simulations.}
+			The blue line shows the positions of the deterministic fixed
+			points, while the dashed line indicates the instability of
+			the middle one. The orange and yellow points show the mean
+			of the stochastic simulations when started at zero and at
+			the high fixed point, respectively. The positions of the two
+			peaks in the bimodal distributions are indicated by smaller
+			points, connected by dotted lines. The arrows illustrate the
+			hysteresis loop in the stochastic simulations.
+			}
 		\label{fig10:Bifurcation}

### 2020-02-22 s1_text/s1_text.tex
Add fig w/ real data from GC + minor edits from KL

@@
 		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$).
-(B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 200 runs of the simulation (same conditions as (A)).
-The distribution from the main text (without bursting) is shown in orange for comparison.
-\textcolor{red}{will add some more runs, but the simulation takes forever}}
+		(B) Blue shows the time until 90\% of the protein concentration
+		at the upper fixed point was reached in 1000 runs of the
+		simulation (same conditions as (A)). The distribution from the
+		main text (without bursting) is shown in orange for comparison.
+		For both, the mean of the distribution is indicated by the line
+		in the corresponding color.
+		}
 	\label{evolbursts}

### 2020-02-22 main_text/main_text.tex
Small edits to main text by MJM

@@
 		(A) The expression of the \emph{xapAB} promoter was measured for
-		different concentrations of xanthosine in the bacteria's
-		environment. The right panel shows the wild type circuit while
+		different extracellular concentrations of xanthosine.
+		The right panel shows the wild type circuit while
 		on the left the genes \emph{xapABR} and in the middle the genes
 		\emph{xapAB} are removed. The wild type circuit behaves like a
-		switch. Note that the fluorescence scales of the middle panel
-		and the other two are not comparable.
+		switch. Note that the fluorescence scale of the middle panel
+		is not comparable with the other two,
+		and also that the chosen xanthosine concentrations are different.
 		(B) The fold-change in gene expression upon addition of
-		xanthosine for the wt promoter and for the promoter with only
-		the XapR binding site right at the polymerase binding site.
+		xanthosine for the wt promoter (left panel)
+		and for the promoter with only the XapR binding site adjacent
+		to the polymerase binding site (right panel).
+		Note that the fold-change used here differs from fold-change in,
+		e.g.,~\cite{Garcia2011,RazoMejia2018,Chure2019},
+		in that no subtraction of autofluorescence was carried out,
+		which is adequate for the qualitative comparison
+		of these two promoters.
 		}
@@
 
-	When all genes of the \emph{xapABR} circuit are removed, the xanthosine response of our reporter construct disappears (see Fig~\subref{fig1:data}{B}, where the last line is autofluorescence).
-This is the result of removing the transcription factor XapR that is induced by xanthosine.
-When only the genes \emph{xapAB} are removed but \emph{xapR} is kept, a response to the xanthosine concentration is regained but it is no longer switch-like (see Fig~\subref{fig1:data}{C}).
-Instead, the distribution remains unimodal and simply shifts to higher expression as the xanthosine concentration is increased.
-As we will see later on, this is because the circuit now lacks the positive feedback loop due to the xanthosine membrane transporter XapB.
-
-	Clearly, XapR regulates the \emph{xapAB} promoter. As mentioned in
+	When all genes of the \emph{xapABR} circuit are removed, the
+	xanthosine response of our reporter construct disappears (see
+	Fig~\subref{fig1:data}{A}, left panel).
+	This is the result of removing the transcription factor XapR that is
+	induced by xanthosine. When only the genes \emph{xapAB} are removed
+	but \emph{xapR} is kept, a response to the xanthosine concentration
+	is regained but it is no longer switch-like (see
+	Fig~\subref{fig1:data}{A}, middle panel). Instead, the distribution remains
+	unimodal and simply shifts to higher expression as the xanthosine
+	concentration is increased. As we will see later on, this is because
+	the circuit now lacks the positive feedback loop due to the
+	xanthosine membrane transporter XapB.
+
+	Can we gain further insight into how XapR regulates
+	the \emph{xapAB} promoter? As mentioned in
 	the introduction, there are two binding sites for XapR on the
 	\emph{xapAB} promoter: one directly at the polymerase binding site
-	and one further upstream. When the upstream binding site is removed,
-	bimodality is lost and the activation of the \emph{xapAB}-expression
-	by xanthosine (via XapR) becomes negligible, as can be seen in
-	Fig~\subref{fig1:data}{B}.
-
-	Already, we believe this data, combined with the prior knowledge of the field, clearly suggest a minimal model of the \emph{xapABR} system.
-The construction and exploration of this model occupies the remainder of this work.
-We leave it for future work to quantitatively dissect and test the model, in the manner of~\cite{Garcia2011,RazoMejia2018,Chure2019}, for example.
+	and one further upstream.
+	Working in the \textDelta\emph{xapAB} background,
+	we measured the expression level of our reporter when driven by two
+	constructs: the wild-type \emph{xapAB} promoter
+	(Fig~\subref{fig1:data}{B}, left panel), and the \emph{xapAB}
+	promoter with the upstream binding site removed
+	(Fig~\subref{fig1:data}{B}, right panel).
+	Clearly, removing one XapR binding site dramatically reduced the
+	responsiveness of the promoter to xanthosine (though not apparent
+	from Fig~\subref{fig1:data}{B}, this response remains detectable).
+
+	Already, we believe these data, combined with the prior knowledge of
+	the field, clearly suggest a minimal model of the \emph{xapABR}
+	system. The construction and exploration of this model occupies the
+	remainder of this work. We leave it for future work to
+	quantitatively dissect and test the model, in the manner
+	of~\cite{Garcia2011,RazoMejia2018,Chure2019}, for example.
 
@@
 	rate equal to zero. Our experiments show that the expression becomes very weak when one of
-	the XapR binding sites is removed from the promoter (see~\nameref{S1_Text}), suggesting that this
-	simplification is reasonable. Furthermore, we find that in
+	the XapR binding sites is removed from the promoter (see
+	Fig~\subref{fig1:data}{B} and~\nameref{S1_Text}), suggesting that
+	this simplification is reasonable. Furthermore, we find that in
 	the bistable parameter range, considering the single occupancy states as
@@
 	This time varies: Fig~\subref{fig9:StochT}{B} shows the distribution
-	of adaptation times from 1000 runs of the simulation (where we
-	neglected the around 10 runs with adaptation times larger than $6
-	\cdot 10^5 \unit{s}$). In both figures, the trajectory obtained from
+	of adaptation times from 1000 runs of the simulation.
+	In both figures, the trajectory obtained from
 	integrating the deterministic system is shown for comparison. It
-	clearly agree well with the simulation on average.
+	clearly agrees well with the simulation on average.
 
@@
 			Fig~\subref{fig8:stochC}{(C)}.
-(B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 1000 runs of the simulation (same conditions as (A)).
+			(B) In blue we plot the waiting time distribution for 1000
+			runs of the simulation (same conditions as (A)) to reach
+			the steady-state mean.
+			We define this time as the elapsed time when the protein
+			copy number first reaches 90\% of its value at the upper
+			fixed point.
+			To better visualize the bulk of the distribution, we
+			excluded from the plot $\sim10$ runs with adaptation times
+			larger than $6 \cdot 10^5 \unit{s}$.
+			The green dashed line indicates the mean of the blue distribution.
 			The orange line shows the corresponding deterministic time.
-The green dashed line right next to it indicated the mean of the blue distribution.}
+			}
 		\label{fig9:StochT}

### 2020-02-22 s1_text/s1_text.tex
Add few lines to SI on expt methods

@@
 \section{Experimental materials and methods} \label{sec:exp}
+Strains used in this work derive from \emph{E.\ coli} K12 MG1655 with
+the \emph{lac} operon removed and were constructed similarly to those
+used in~\cite{RazoMejia2018}. The YFP-expressing reporter constructs
+were cloned into a pZS25 backbone with kanamycin resistance and
+integrated at the \emph{galK} locus using recombineering.
+\textDelta\emph{xapAB} and \textDelta\emph{xapABR} knockouts were
+generated \textbf{???using P1 transduction from the Keio collection???}
 
-\textcolor{red}{missing}
+Flow cytometry measurements, as well as growth and preparation of cell
+cultures, were carried out as in~\cite{RazoMejia2018}.
+
+\textbf{??What else needs adding here??}
 

### 2020-02-23 rebuttal/rebuttal.tex
Add rebuttal letter + minor associated edits to main text

\documentclass[11pt,letterpaper]{article}
\usepackage[top=0.85in,left=0.85in,footskip=0.75in]{geometry}
\usepackage{mathpazo}
\usepackage{mdframed}
\usepackage{graphicx}
\usepackage{sectsty}
\allsectionsfont{\normalfont\sffamily\bfseries}
\subsectionfont{\normalfont\sffamily\bfseries}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{amsmath}

% To place figures exactly on a page
\usepackage{float}
\usepackage{graphicx}

% Bold the 'Figure #' in the caption and separate it from the title caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,font=small,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

% Strike out
\usepackage[normalem]{ulem}

% Custom BiBTeX style
\bibliographystyle{bibStyle}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{helvet}

% To define more useful LaTeX commands
\usepackage{xparse}
\usepackage{graphicx}

% Equation referencing
\DeclareDocumentCommand \eref{oooo} {\IfNoValueTF{#2}{Eq. R~(\ref{#1})}{\IfNoValueTF{#3}{Eqs.~(\ref{#1}) and (\ref{#2})}{\IfNoValueTF{#4}{Eqs.~(\ref{#1})-(\ref{#3})}{Eqs.~(\ref{#1})-(\ref{#4})}}}}

% Figure referencing
\DeclareDocumentCommand \fref{ooo} {\IfNoValueTF{#2}{Fig.R~\ref{#1}}{\IfNoValueTF{#3}{Figs.~\ref{#1} and \ref{#2}}{Figs.~\ref{#1}-\ref{#3}}}}
\renewcommand{\thefigure}{R\arabic{figure}} 
\renewcommand{\theequation}{\sffamily R\arabic{equation}} 
\renewcommand{\thetable}{\sffamily R\arabic{table}}

%% setup the response environment
\mdfdefinestyle{revstyle}{
outerlinewidth=2pt,
bottomline=true,
leftline=true,rightline=true,
topline=true,
font=\small,
skipabove=\baselineskip,
skipbelow=\baselineskip,
roundcorner=10pt,
innertopmargin=0.5em,
innerbottommargin=1em,
splittopskip=1em,
splitbottomskip=1em
}

\mdfdefinestyle{respstyle}{
outerlinewidth=2pt,
bottomline=false,
leftline=false,rightline=false,
topline=false,
font=\small,
backgroundcolor=gray!20,
skipabove=\baselineskip,
skipbelow=\baselineskip,
roundcorner=10pt,
innertopmargin=0.5em,
innerbottommargin=1em,
splittopskip=1em,
splitbottomskip=1em
}

\newmdenv[style=revstyle,frametitle=\sffamily \itshape Reviewer Comment]{review}
\newmdenv[style=respstyle,frametitle=\sffamily \itshape Author Response]{response}
\newmdenv[style=respstyle,frametitle=\sffamily \itshape Author Response: Executive Summary]{responseSummary}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% END MACROS SECTION%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\section*{Response to Reviewer for PONE-D-19-32846 ``Theoretical Investigation of a Genetic Switch for Metabolic Adaptation''}
Kathrin S. Laxhuber,$^{1}$ Muir J. Morrison,$^{2}$ Griffin Chure,$^{3}$ Rob Phillips$^{2, 3, \dagger}$
\vspace{10pt}

\noindent\tiny{$^1$ Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland,
$^2$ Department of Physics, California Institute of Technology, Pasadena, CA, USA,
$^3$ Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA, $^\dagger$ E-mail: \texttt{phillips@pboc.caltech.edu}}

\begin{responseSummary}
We thank the editor and reviewer for considering our manuscript and
bringing some important points to our attention. The reviewer's comments
significantly improved our work and inspired us to make very interesting
comparisons between the deterministic and stochastic picture of our
system. Furthermore, as rightly suggested by the reviewer, we have now
added experimental data to the manuscript and expanded the author list
to reflect this. Below, we address the reviewer's concerns one-by-one
and indicate where changes were made either to the main text or the
supporting information. Besides these changes, we made a few small
corrections to the text that had slipped our attention before. We
believe we have resolved the three issues brought up by the reviewer and
hope our manuscript is now ready for publication in PLOS ONE.
\end{responseSummary}

\subsection*{Reviewer \#1}
\begin{review}
In their manuscript "Theoretical investigation of a genetic switch for
metabolic adaptation", Laxhuber and colleagues present a very nice
theoretical analysis of the behavior of the E. coli xapABR system, using
both deterministic and stochastic simulations to demonstrate that the
system can show meaningful and important bistability driven by the
activity of the membrane transporter. In general the work is well done,
and it is technically of high quality, but at present I have two
substantial concerns that I think must be addressed in order for the
paper to live up to its potential.
\end{review}

\begin{response}
We thank the reviewer for their careful reading of our work and are
pleased that their impression of the manuscript is favorable overall.
\end{response}

\begin{review}
My first concern is that while the authors present their deterministic
simulations first, and draw many of their conclusions primarily from
those simulations, it seems to me that the use of such simulations is
contra-indicated by the authors' own findings. As both the number of
XapR molecules and mRNA molecules are small integers (likely 10 or
less), treatment of these species as continuous, deterministic
quantities can never be more than a poor approximation. I find this
troublesome for a couple of reasons. First, the authors still treat
their continuous treatment as an apparently 'good' way to model a
genetic switch like that at work at the xapAB promoter; while I find
their analysis insightful, in the end I cannot support such a
recommendation given the stochasticity intrinsic in the low protein and
mRNA copy numbers present here.
\end{review}

\begin{response}
We agree with the concern of the reviewer that in general, deterministic
approximations of systems with such low copy numbers should not be
trusted. We do not intend to give the general recommendation that a
continuous treatment alone is usually sufficient, and in order to avoid
this, we added some words of warning in the beginning of our
deterministic as well as our stochastic analysis.

Nevertheless, despite the low copy numbers,
for the system we investigate the deterministic model is a
surprisingly accurate description of the means of copy number and
waiting time distributions produced by the stochastic simulations.
To make this more apparent, we added some additional comparisons of the two
models. We address these in more detail in our responses below,
but to summarize: we added a bifurcation diagram
(Fig~\ref{bifucation}) which illustrates that the stochastic
distributions track the deterministic fixed points surprisingly well,
and we compared the stochastic timescales with the solution of the
deterministic model (Fig~\ref{adaptationtime}) and found excellent
agreement there too.
Even though stochastic fluctuations around the deterministic
trajectories may be large, the stochastic trajectories do still flow to
the fixed points as predicted by the deterministic model, except when
the system is close to either of the bifurcation points in
Fig~\ref{bifucation}, where it is unsurprising that fluctuations are
magnified in importance.
Hence, while we find the close agreement surprising, we believe it to be
appropriate to show and draw conclusions from deterministic phase
diagrams. The latter have the advantage of providing the reader with a
more intuitive understanding of the system than the plots of the results
from the stochastic simulation.
\end{response}

\begin{review}
Second, while the authors do technically recapitulate the bistable
behavior of the system in the stochastic version, the bistability seems
based on their description to be quite fragile and not very meaningful,
since it only occurs here for small distances between the lower and
upper fixed points
\end{review}

\begin{response}
We thank the reviewer for bringing this point to our attention, which we
believe to be a misunderstanding that resulted from a poor explanation
in our original text. We believe we have clarified this in the
manuscript, and allow us elaborate here too: Stochastic bimodality occurs
only when two of the three fixed points are close together, or in other
words, when the system is poised ``sufficiently near'' one of the
bifurcation points. This may become clearer when looking at the
bifurcation diagram that we added to our section about hysteresis at the
end of the main text. For convenience, we insert this plot below in
Fig~\ref{bifucation} which addresses what we believe the reviewer is
referring to: only when two of the three fixed points are close do the
stochastic trajectories occupy both fixed points, resulting in
bimodality.

Since this bimodality only occurs for a small region in parameter space,
this actually indicates strong and not fragile bistability. In other
words, the fact that the system does not stochastically ``jump'' to the
other fixed point means that the fixed points are very stable to
stochastic fluctuations. Thus, the switch-like feature is robust:
there are two distinct states that the system can occupy, and only near
the bifurcation points is it likely for stochastic trajectories to
transition between them. Put differently, deterministic bistability does
not need to result in stochastic bimodality, and in fact stochastic
bimodality is contrary to robust bistability.
\end{response}

\begin{figure}[h!]
	\centering
	\includegraphics{Fig10_bifurcation.eps}
	\caption{{\bf Bifurcation diagram showing the hysteresis loop in the stochastic system.}
		The blue line shows the positions of the deterministic fixed points, while the dashed line indicates the instability of the middle one. The orange and yellow points show the mean of the stochastic simulations when started at zero and at the high fixed point, respectively. The positions of the two peaks in the bimodal distributions are indicated by smaller points, connected by dotted lines. The arrows illustrate the hysteresis loop in the stochastic simulations.}
	\label{bifucation}
\end{figure}

\begin{review}
	(and seems to take an unreasonably long time to switch between states).
\end{review}

\begin{response}
To address the indeed valid point of unreasonable switching times, we
have compared the timescales in our stochastic simulations with those from the
numerical solution of the deterministic model. We find excellent
agreement, as shown in an updated figure in the main text and also below
in Fig~\ref{adaptationtime}. So, again perhaps surprisingly,
stochasticity does not materially alter the mean switching time as
computed from the deterministic picture. Even adding additional
stochasticity through transcriptional bursting (discussed more below)
does not alter the mean switching time, though it does broaden the
distribution of times.
It is possible that this broadening resolves the apparent paradox. Even
if the mean switching time is quite long, a non-negligible fraction of a
population of cells could nevertheless switch on a shorter time-scale,
especially with inclusion of extra stochasticity beyond what we have
explicitly considered.
\end{response}

	\begin{figure}[h!]
		\centering
		\includegraphics{Fig9_evolution.eps}
		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
			(A) Blue shows the results from one run of the stochastic simulation, orange shows the trajectory obtained from solving the deterministic ODE's. The simulation was started
			at an mRNA, protein and intracellular xanthosine count of 0. The
			extracellular xanthosine concentration was chosen to be
			$\mathrm{[c]_a} = 25$. (B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 1000 runs of the simulation (same conditions as (A)). The orange line shows the corresponding deterministic time. The green dashed line right next to it indicated the mean of the blue distribution.}
		\label{adaptationtime}
	\end{figure}

\begin{review}
I think the authors are likely correct in their assertion that they
underestimate the amount of noise in the system due to not including
transcriptional and translational bursting, and that the transition
rates would likely be faster, but this gives no idea of how *much*
faster, and thus it seems premature to simply declare victory and move
on.
\end{review}

\begin{response}
This is a very reasonable concern which we have investigated more
thoroughly now. We added rather strong transcriptional bursting to our
simulation and do not find qualitative changes in the results. In
particular, fluctuations around the mean become larger and, accordingly,
distributions broader. Furthermore, bimodality begins to appear for
lower values of the extracellular xanthosine concentration. Very
interestingly, the change in the \emph{mean} transition time between the
stable fixed points is negligible. However, the \emph{variance} in this
time across many runs of the simulation becomes larger and the
\emph{peak} of the distribution shifts to shorter times. We have
included two new figures in the SI discussing these results (Figs. 6 and
7), along with discussion there and in the main text.
%	Below, we also show the two new figures from S1 Text that illustrate the similarities and differences between the simulation with and without transcriptional bursting.
\end{response}

%\begin{figure}
%	\centering
%	\includegraphics[width=1\textwidth]{FigSI6_7.pdf}
%	\caption{{\bf Time evolution and adaptation times from the stochastic simulation when transcriptional bursting is included.}
%		(A) shows the time evolution from one run of the simulation. The simulation was started
%		at an mRNA, protein and intracellular xanthosine count of 0. The
%		extracellular xanthosine concentration was chosen to be
%		$\n{[c]_a} = 25$. (B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 1000 runs of the simulation (same conditions as (A)). The distribution from the main text (without bursting) is shown in orange for comparison.}
%	\label{evolbursts}
%\end{figure}
%
%\begin{figure}
%	\centering
%	\includegraphics[width=1\textwidth]{FigSI6_7_2.pdf}
%	\caption{{\bf Distributions from the stochastic simulation when transcriptional bursting is included.}
%		Blue shows the distributions from the simulation with bursts, orange shows those without bursts from the main text for comparison. The following values of $\mathrm{[c]_a}$ were used: (A) $\n{[c]_a}=12$, (B) $\n{[c]_a}=17.5$ (not shown in main text), (C) $\mathrm{[c]_a}=18.5$, and (D) $\mathrm{[c]_a}=25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
%		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$). All other parameters are as in the table in the main text, and the simulation
%		was run 1000 times for $10^6 \unit{s}$ each (simulated time) and
%		started at a mRNA, protein and intracellular xanthosine count of 0.}
%	\label{distbursts}
%\end{figure}

\begin{review}
My recommendation would be that the authors put in some more cautionary words about the applicability of the continuous model, and ideally spend a little more time characterizing the discrete model. in terms of switching times, parameters that give rise to bimodal behavior, effects of incorporating reasonable amounts of transcriptional bursting, etc. Some more direct comparisons of the behavior of the two models in similar ranges of parameters space would also likely be enlightening.
\end{review}

\begin{response}
We hope that we have addressed all these points in our responses above
and shortly summarize here: We put in some cautionary words about the
general usage of continuous models and demonstrated that for our system
and the conclusions we draw, the continuous model is surprisingly appropriate.
Furthermore, we extended our characterization of the stochastic model as
suggested by investigating the distribution of adaptation/switching
times as well as by incorporating transcriptional bursting in our
simulation. The deterministic and stochastic model are now directly
compared (for the same set of parameters) in the bifurcation diagram as
well as in terms of their adaptation/switching times.
\end{response}

\begin{review}
My other major difficulty here is that essentially all of the work is
premised on reference 22, which is 'unpublished data.' This is the only
experimental evidence given for the bistability of xapAB expression,
justification of many of the model parameters, experimentally expected
switching timescales, etc. It is impossible to properly interpret the
present work, either for the purposes of review or from the viewpoint of
an interested reader, without availability of those data, which either
ought to be their own independent publication or ought to be included
with this one.
\end{review}

\begin{response}
With further consideration, we agree with the reviewer that we relied
rather heavily on this unpublished data and that it is therefore
essential to make the experimental data available.
\textbf{Also address changes to author list once finalized.}
We have opted to include the data with this publication rather than
separately. We have added a new section, ``experimental motivation,'' in
the beginning of the main text which discusses the main features of the
data that are relevant for our model. Furthermore, we added a section on
the materials and experimental methods at the end of S1 Text (section
E).
The raw data is
publicly available at \textbf{!!still need to sort out
DOI/caltechData/ETHZ hosting/...??}.
\end{response}

%\begin{figure}[h]
%    \centering{
    %\includegraphics{figure_path}
%     \caption{\textbf{Figure caption.}}
%    \label{fig:figlabel}}
% \end{figure}

\bibliography{library}

\end{document}

### 2020-02-23 main_text/main_text.tex
Add rebuttal letter + minor associated edits to main text

@@
 	In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
-	expression state during the run time of our simulations.
-This actually is a good finding: it implies that bistability, characterized by two separate \emph{stable} steady-states, largely persists even in the presence of stochastic fluctuations.
-Thus, the fact that bimodality only occurs in a small regime means that the circuit is a strong switch and the deterministic picture is sufficient except in a small range of parameter space.
-Whether or not that is where the system actually is remains a question for future experiments.
-
-	We found that the two lower fixed points (marked as 1 and 2 in
+	expression state during the run time of our simulations. This result
+	implies that bistability, characterized by two separate
+	\emph{stable} steady-states, largely persists even in the presence
+	of stochastic fluctuations, suggesting that the deterministic
+	picture is surprisingly and unreasonably effective. In other words,
+	the fact that bimodality only occurs with some fine tuning of
+	parameters means that the circuit is a strong switch and the
+	deterministic picture is sufficient except in a small region of
+	parameter space. Whether or not the system is actually in that
+	region of parameter space remains a question for future experiments.
+
+	To elaborate on the preceeding statements,
+	we found that the two lower fixed points (marked as 1 and 2 in
 	Fig~\ref{fig4:bistability}) need to be very close like in
@@
 		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
-			(A) Blue shows the results from one run of the stochastic simulation, orange shows the trajectory obtained from solving the deterministic ODE's.
+			(A) Shown in blue is the result from one run of the stochastic simulation, and in orange the trajectory obtained from solving the deterministic ODE's.
 The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started

### 2020-02-23 main_text/main_text.tex
Update author list & acknowledgements

@@
 		Muir J. Morrison\textsuperscript{2},
+		Griffin Chure\textsuperscript{3},
 		Rob Phillips\textsuperscript{2,3$\ast$}
@@
 	\section*{Acknowledgments}
-	We thank Griffin Chure, Charlotte Strandkvist, and Kyle Naughton for
-	providing their data on the \emph{xap} system and pointing out the
-	bistability. We also thank Griffin Chure for answering the many biological
-	questions that arose, Nathan Belliveau for bringing similar systems to our
+	Portions of the experiments reported here were carried out at the
+	Physiology summer course at the Marine Biological Laboratory in Woods
+	Hole, MA, operated by the University of Chicago. We thank Charlotte
+	Strandkvist and Kyle Naughton for their work on the project during the
+	course. We also Nathan Belliveau for strain creation work and for
+	bringing similar systems to our
 	attention, Patrick Lenggenhager for his help with \emph{Mathematica} issues, and Jane Kondev and Jin Wang for interesting discussions.
-	For assistance with illustrations, we also thank Nigel Orme and Griffin Chure.
+	For assistance with illustrations, we also thank Nigel Orme.
 	This work was supported by 1R35 GM118043-01 Maximizing Investigators'

### 2020-03-01 code/figures/Fig2_flow_cytometry.py
Update of materials for flow figure

@@ -1,0 +1,143 @@
+#%%
+import numpy as np 
+import pandas as pd 
+import matplotlib.pyplot as plt
+import joypy 
+import glob
+import scipy.stats
+import xap.viz
+import xap.flow
+colors = xap.viz.pub_style()
+# %%
+# Parse all of hte files.
+files = glob.glob('../../data/*.csv')
+
+#%%
+dfs = []
+
+for f in files:
+    if '2018' in f:
+        date, _, strain, conc = f.split('/')[-1].split('_')
+        conc = float(conc.split('mg')[0])
+        promoter = 'wt'
+        location = 'MBL'
+    elif '2019' in f:
+        date, promoter, strain, _, conc = f.split('/')[-1].split('_') 
+        conc = float(conc.split('mg')[0])
+        location = 'CALTECH'
+
+    _df = pd.read_csv(f)
+    gated = xap.flow.gaussian_gate(_df, 0.4)
+    gated = gated[gated['gate']==1]
+    keys = _df.keys()
+    if 'GFP-A' in keys:
+        intensity = _df['GFP-A']
+    elif 'GFP' in keys:
+        intensity = _df['GFP']
+    elif 'FITC-H' in keys:
+        intensity = _df['FITC-H']
+
+    _df['intensity'] = np.log(intensity)
+    _df['date'] = date
+    _df['xan_mgml'] = conc
+    _df['strain'] = strain
+    _df['promoter'] = promoter
+    _df['location'] = location
+    _df = _df[['intensity', 'date', 'xan_mgml', 'strain', 'promoter', 'location']]
+    dfs.append(_df) 
+intensity_df = pd.concat(dfs, sort=False)
+
+# %%
+wt = intensity_df[(intensity_df['location']=='MBL') & 
+                  (intensity_df['strain']=='wt') &  
+                  (intensity_df['promoter']=='wt') & 
+                  (intensity_df['intensity'] > -2) & 
+                  (intensity_df['intensity'] < 10)]
+delta = intensity_df[(intensity_df['location']=='MBL') & 
+                     (intensity_df['strain']=='deltaABR') & 
+                     (intensity_df['promoter']=='wt') & 
+                     (intensity_df['intensity'] > -2) &
+                     (intensity_df['intensity'] < 10)]
[…]

### 2020-03-01 setup.py

@@ -1,0 +1,18 @@
+from setuptools import setup, find_packages
+from os import path
+
+__version__ = '0.0.01'
+
+setup(
+    name='xap',
+    version=__version__,
+    description='Various python utilities used for this project',
+    license='BSD',
+    classifiers=[
+        'Development Status :: 1 - Alpha',
+        'Intended Audience :: Developers',
+        'Programming Language :: Python :: 3',
+    ],
+    author="Griffin Chure",
+    author_email="gchure@caltech.edu"
+)

### 2020-03-01 xap/__init__.py

@@ -1,0 +1,12 @@
+# -*- coding: utf-8 -*-
+
+"""Top level package for MWC utilities"""
+
+
+# from . import mcmc
+from . import viz
+from . import flow
+
+__author__ = """Griffin Chure"""
+__email__ = """gchure@caltech.edu"""
+__version__ = '0.0.1'

### 2020-03-01 xap/flow.py

@@ -1,0 +1,206 @@
+import numpy as np
+import fcsparser
+import pandas as pd
+from ._fit_bivariate_normal_AstroML import fit_bivariate_normal
+import scipy.stats
+
+
+# #######################
+# Automated Gating
+# #######################
+def fit_2D_gaussian(df, x_val='FSC-A', y_val='SSC-A', log=False):
+    '''
+    This function hacks astroML fit_bivariate_normal to return the mean
+    and covariance matrix when fitting a 2D gaussian fuction to the data
+    contained in the x_val and y_val columns of the DataFrame df.
+
+    Parameters
+    ----------
+    df : DataFrame.
+        dataframe containing the data from which to fit the distribution
+    x_val, y_val : str.
+        name of the dataframe columns to be used in the function
+    log : bool.
+        indicate if the log of the data should be use for the fit or not
+
+    Returns
+    -------
+    mu : tuple.
+        (x, y) location of the best-fit bivariate normal
+    cov : 2 x 2 array
+        covariance matrix.
+        cov[0, 0] = variance of the x_val column
+        cov[1, 1] = variance of the y_val column
+        cov[0, 1] = cov[1, 0] = covariance of the data
+    '''
+    if log:
+        x = np.log10(df[x_val])
+        y = np.log10(df[y_val])
+    else:
+        x = df[x_val]
+        y = df[y_val]
+
+    # Fit the 2D Gaussian distribution using atroML function
+    mu, sigma_1, sigma_2, alpha = fit_bivariate_normal(x, y, robust=True)
+
+    # compute covariance matrix from the standar deviations and the angle
+    # that the fit_bivariate_normal function returns
+    sigma_xx = ((sigma_1 * np.cos(alpha)) ** 2 +
+                (sigma_2 * np.sin(alpha)) ** 2)
+    sigma_yy = ((sigma_1 * np.sin(alpha)) ** 2 +
+                (sigma_2 * np.cos(alpha)) ** 2)
+    sigma_xy = (sigma_1 ** 2 - sigma_2 ** 2) * np.sin(alpha) * np.cos(alpha)
+
+    # put elements of the covariance matrix into an actual matrix
+    cov = np.array([[sigma_xx, sigma_xy], [sigma_xy, sigma_yy]])
+
+    return mu, cov
+
+
+# #################
+def gauss_interval(df, mu, cov, x_val='FSC-A', y_val='SSC-A', log=False):
+    '''
+    Computes the of the statistic
+
+    (x - µx)'Σ(x - µx)
+
[…]

### 2020-03-01 xap/viz.py

@@ -1,0 +1,49 @@
+import bokeh.io
+import bokeh.plotting
+import bokeh.layouts
+import bokeh.palettes
+import bokeh.themes.theme
+import itertools
+import holoviews as hv
+import skimage.io
+import skimage.measure
+import seaborn as sns
+import numpy as np
+import pandas as pd
+import os
+import matplotlib
+import matplotlib.pyplot as plt
+from matplotlib.animation import FuncAnimation
+from scipy.signal import gaussian, convolve
+
+def pub_style(return_colors=True):
+    """
+    Sets the style to the publication style
+
+    Parameters
+    ----------
+    return_colors: Bool
+        If True, this will also return a dictionary
+
+    """
+    rc = {'axes.facecolor': '#ffffff',
+          'asxes.linewidth':0.75,
+          'axes.axisbelow':True,
+          'axes.edgecolor': 'black',
+          'xtick.color': 'black',
+          'ytick.color': 'black',
+          'font.family': 'Times New Roman',
+          'grid.linestyle': '-',
+          'grid.linewidth': 0.5,
+          'grid.alpha': 0.75,
+          'grid.color': '#FFFFFF',
+          'legend.frameon': True,
+          'legend.facecolor': '#FFFFFF',
+          'figure.dpi': 150}
+    plt.rc('text.latex', preamble=r'\usepackage{sfmath}')
+    plt.rc('mathtext', fontset='stixsans', sf='sans')
+    sns.set_style('darkgrid', rc=rc)
+    colors = ['#D56C55', '#738FC1', '#7AA974', '#D4C2D9', '#F1D4C9', '#C9D7EE',
+              '#DCECCB']
+    if return_colors:
+        return colors

### 2020-03-02 code/figures/NullclinePlots.m
Added all files for theoretical analysis
Mathematica notebooks for plots. Python scripts and input files for stochastic simulations. Data generated from simuations.

@@ -1,0 +1,314 @@
+(* ::Package:: *)
+
+(* ::Title:: *)
+(*NullclinePlots: Plotting of ODE's as Vectorfields, Showing Nullclines*)
+
+
+(* ::Text:: *)
+(*Author: Kathrin Laxhuber*)
+(*Created: 04.02.2019*)
+(*Last modified: 18.05.2019*)
+
+
+(* ::Section::Closed:: *)
+(*Begin Package*)
+
+
+BeginPackage["NullclinePlots`","MaTeX`"]
+
+
+(* ::Section::Closed:: *)
+(*Usage Definitions*)
+
+
+NullclinePlots::usage = "This package provides functions to create vector or stream plots of two ODE's and show the corresponding nullclines.";
+
+
+NullclinePlotSystem::usage="Plots a system of two ODE's as stream plot including the nullclines.";
+
+
+QuickNullclinePlotSystem::usage="Plots a system of two ODE's at vector or stream plots including the nullclines. Has less options than PlotSystem and plots faster.";
+
+
+VectorNullclinePlotSystem::usage="Plots a system of two ODE's as vector plot including the nullclines.";
+
+
+LogNullclinePlotSystem::usage="Plots a system of two ODE's as Log Log stream plot including the nullclines.";
+
+
+LogVectorNullclinePlotSystem::usage="Plots a system of two ODE's as Log Log vector plot including the nullclines.";
+
+
+(* ::Section:: *)
+(*Constants*)
+
+
+(* ::Subsection:: *)
+(*Basic / General*)
+
+
+plotOptionsBasic={Frame->True,FrameStyle->BlackFrame};
+
+
+(* ::Subsection:: *)
+(*LaTeX*)
+
+
+(* ::Input::Initialization:: *)
+texStyle={FontFamily->"Times",FontSize->11};
+plotSizesLaTeX=<|"Normal"->350,"Two"->250,"Three"->180|>;
+
+
+(* ::Subsection:: *)
+(*PowerPoint*)
+
+
+(* ::Input::Initialization:: *)
[…]

### 2020-03-02 code/stochastic_simulation/simulation.py

@@ -1,0 +1,226 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Feb 28 11:21:44 2019
+
+@author: Kathrin Laxhuber
+
+Theoretical investigation of a genetic switch for metabolic adaptation
+
+Xap System Simulation (3D)
+"""
+
+# %% Imports
+
+import numpy as np
+import time
+import csv
+import matplotlib.pyplot as plt
+import stochpy
+import os 
+import sys
+from datetime import datetime
+
+
+# %% Paths
+
+dir_path = os.path.dirname(os.path.realpath(__file__))
+
+
+# %% Function definitions
+
+def initialize(simulation,params,initials):
+    """
+    Sets parameter values and initial species numbers.
+    """
+    for i in range(len(initials)):
+        simulation.ChangeInitialSpeciesCopyNumber(initials[i][0],initials[i][1])
+    
+    for i in range(len(params)):
+        simulation.ChangeParameter(params[i][0],params[i][1])     
+
+def repeat(simulation,runs,endtime):
+    """
+    Perfoms multiple runs of the simulation.
+    
+    runs: number of runs to perform
+    simulation: the simulation to run
+    endtime: time until the simulation should run
+    """    
+    reslist = []  
+    
+    for i in range(runs):   
+        simulation.DoStochSim(mode="time",end=endtime)
+        res = simulation.data_stochsim.species[-1]
+        
+        reslist.append(res)
+        
+    reslist = np.array(reslist)
+        
+    return reslist
+    
+def species_to_index(species):
+    if species == "mrna":
+        index = 0
+    if species == "protein":
+        index = 1
+    elif species == "xanthosine":
+        index = 2
+    else:
+        sys.exit("Not a valid species.")
+        
+    return index
+
+def find_plot_distribution(simulation,params,initials,species,runs,endtime):
+    """
+    Finds, exports and plots the probability distribution of a given species.
+    
+    params: simulation parameters 
+    initials: initial values
+    species: the species to plot
+    runs: number of runs to perform for this
+    simulation: the simulation
+    endtime: time until the simulation should run
+    """
+    initialize(simulation,params,initials)
+    
+    index = species_to_index(species)
+    data = repeat(simulation,runs,endtime)[:,index]
+    
[…]

### 2020-03-02 code/stochastic_simulation/simulation_bursts.py

@@ -1,0 +1,225 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Feb 28 11:21:44 2019
+
+@author: Kathrin Laxhuber
+
+Theoretical investigation of a genetic switch for metabolic adaptation
+
+Xap System Simulation (3D)
+"""
+
+# %% Imports
+
+import numpy as np
+import time
+import csv
+import matplotlib.pyplot as plt
+import stochpy
+import os 
+import sys
+from datetime import datetime
+import random
+
+
+# %% Paths
+
+dir_path = os.path.dirname(os.path.realpath(__file__))
+
+
+# %% Function definitions
+
+def initialize(simulation,params,initials):
+    """
+    Sets parameter values and initial species numbers.
+    """
+    for i in range(len(initials)):
+        simulation.ChangeInitialSpeciesCopyNumber(initials[i][0],initials[i][1])
+    
+    for i in range(len(params)):
+        simulation.ChangeParameter(params[i][0],params[i][1])     
+
+def repeat(simulation,runs,endtime):
+    """
+    Perfoms multiple runs of the simulation.
+    
+    runs: number of runs to perform
+    simulation: the simulation to run
+    endtime: time until the simulation should run
+    """    
+    reslist = []  
+    
+    for i in range(runs):   
+        simulation.DoStochSim(mode="time",end=endtime)
+        res = simulation.data_stochsim.species[-1]
+        
+        reslist.append(res)
+        
+    reslist = np.array(reslist)
+        
+    return reslist
+    
+def species_to_index(species):
+    if species == "mrna":
+        index = 0
+    if species == "protein":
+        index = 1
+    elif species == "xanthosine":
+        index = 2
+    else:
+        sys.exit("Not a valid species.")
+        
+    return index
+
+def find_plot_distribution(simulation,params,initials,species,runs,endtime):
+    """
+    Finds, exports and plots the probability distribution of a given species.
+    
+    params: simulation parameters 
+    initials: initial values
+    species: the species to plot
+    runs: number of runs to perform for this
+    simulation: the simulation
+    endtime: time until the simulation should run
+    """
+    initialize(simulation,params,initials)
+    
+    index = species_to_index(species)
[…]

### 2020-03-02 rebuttal/rebuttal.tex
Partially included Rob's and Griffin's feedback
Rob: done. Griffin: up to experimental motivation.

@@
 
+{\Large\textcolor{red}{ToDo: change author list above and at the end, add correct github link (public repo!).}}
+
 \begin{responseSummary}
@@
 supporting information. Besides these changes, we made a few small
-corrections to the text that had slipped our attention before. We
+improvements and corrections to the text that had slipped our attention before. We
 believe we have resolved the three issues brought up by the reviewer and
@@
 metabolic adaptation", Laxhuber and colleagues present a very nice
-theoretical analysis of the behavior of the E. coli xapABR system, using
+theoretical analysis of the behavior of the \emph{E. coli} xapABR system, using
 both deterministic and stochastic simulations to demonstrate that the
@@
 We agree with the concern of the reviewer that in general, deterministic
-approximations of systems with such low copy numbers should not be
-trusted. We do not intend to give the general recommendation that a
+approximations of systems with such low copy numbers need to be verified. We do not intend to give the general recommendation that a
 continuous treatment alone is usually sufficient, and in order to avoid
@@
 
-Nevertheless, despite the low copy numbers,
-for the system we investigate the deterministic model is a
-surprisingly accurate description of the means of copy number and
-waiting time distributions produced by the stochastic simulations.
-To make this more apparent, we added some additional comparisons of the two
+Nevertheless, for our system we did successfully test the concordance of the stochastic simulations and the deterministic model.
+To make this more apparent, we included some additional comparisons of the two
 models. We address these in more detail in our responses below,
@@
 (Fig~\ref{bifucation}) which illustrates that the stochastic
-distributions track the deterministic fixed points surprisingly well,
+distributions track the deterministic fixed points very well,
 and we compared the stochastic timescales with the solution of the
@@
 agreement there too.
+
 Even though stochastic fluctuations around the deterministic
 trajectories may be large, the stochastic trajectories do still flow to
-the fixed points as predicted by the deterministic model, except when
-the system is close to either of the bifurcation points in
-Fig~\ref{bifucation}, where it is unsurprising that fluctuations are
-magnified in importance.
-Hence, while we find the close agreement surprising, we believe it to be
-appropriate to show and draw conclusions from deterministic phase
-diagrams. The latter have the advantage of providing the reader with a
-more intuitive understanding of the system than the plots of the results
-from the stochastic simulation.
+the fixed points as predicted by the deterministic model.
+Hence, the excellent agreement may be surprising or unintuitive at first, but the conclusions we draw from the deterministic phase diagrams turn out to be accurate.
+
+%, except when
+%the system is close to either of the bifurcation points in
+%Fig~\ref{bifucation}, where it is unsurprising that fluctuations are
+%magnified in importance.
+
+Given this concordance, we prefer to start by showing the deterministic case and illustrating our conclusions there before moving on to the stochastic picture.
+That is because the deterministic phase diagrams have the advantage of providing the reader with a more intuitive understanding of the system than the plots of the results from the stochastic simulation.
 \end{response}
@@
 The positions of the two peaks in the bimodal distributions are indicated by smaller points, connected by dotted lines.
+To make this clearer, the bimodal distributions themselves are shown in a light color (in a smoothed form).
 The arrows illustrate the hysteresis loop in the stochastic simulations.}
@@
 and shortly summarize here: We put in some cautionary words about the
-general usage of continuous models and demonstrated that for our system
-and the conclusions we draw, the continuous model is surprisingly appropriate.
+general usage of continuous models (beginning of sections on deterministic as well as stochastic results) and demonstrated that for our system
+and the conclusions we draw, the continuous model is surprisingly appropriate (bifurcation diagram and time evolution, end of results section).
 Furthermore, we extended our characterization of the stochastic model as
 suggested by investigating the distribution of adaptation/switching
-times as well as by incorporating transcriptional bursting in our
-simulation. The deterministic and stochastic model are now directly
+times (end of results section) as well as by incorporating transcriptional bursting in our
+simulation (end of section C in S1 Text). The deterministic and stochastic model are now directly
 compared (for the same set of parameters) in the bifurcation diagram as
@@
 essential to make the experimental data available.
-\textbf{Also address changes to author list once finalized.}
 We have opted to include the data with this publication rather than
-separately. We have added a new section, ``experimental motivation,'' in
+separately.
+For this reason, we have added Griffin Chure, Nathan Belliveau, Charlotte Strandkvist, and Kyle Naughton to the author list, who collected or were involved in collecting the data.
+We have added a new section, ``experimental motivation,'' in
 the beginning of the main text which discusses the main features of the
@@
 The raw data is
-publicly available at \textbf{!!still need to sort out
-DOI/caltechData/ETHZ hosting/...??}.
+publicly available at \url{https://github.com/RPGroup-PBoC/mwc_activation/}.
 \end{response}

### 2020-03-02 main_text/main_text.tex
Partially included Rob's and Griffin's feedback
Rob: done. Griffin: up to experimental motivation.

@@
 	\section*{Abstract}
-	Membrane transporters carry key metabolites through the cell membrane
-	and, from a resource standpoint, are hypothesized to be produced only when necessary. The
+	Membrane transporters carry key metabolites across the cell membrane
+	and, from a resource standpoint, are hypothesized to be produced when necessary. The
 	expression of membrane transporters in metabolic pathways is often
@@
 	generalizable physical model of the \emph{xapABR} genetic
-	circuit, which features a regulatory feedback loop through membrane
+	circuit, which features a regulatory feedback loop via membrane
 	transport (positive feedback) and enzymatic degradation (negative feedback)
@@
 	positive feedback
-	loop, for example in the well-known study of Novick and Weiner,
+	loop, for example in the well-known study by Novick and Weiner (1957),
 	among other works~\cite{Novick1957,Santillan2008,Ozbudak2004,Narang2008,Choi2008,Fritz2014,Jenkins2017,Siegele1997}.
@@
 	A key element of this mechanism is the presence of a transcription factor
-	which is always expressed at a low level (often at copy numbers
-	of order $\sim10$) and which binds to the transporter substrate.
+	which binds to the transporter substrate and which is often expressed at a low level (often at copy numbers
+	of order $\sim10$, \cite{Schmidt2015}).
 	This is resource efficient for the cell, as this low copy number
@@
 	circuit is made up of two operons: one that encodes for XapR and another
-	that encodes for XapA and XapB. XapR appears to be a transcription factor
+	that encodes for XapA and XapB. XapR is a transcription factor
 	that is induced by xanthosine and activates the \emph{xapAB} promoter,
@@
 	Experimentally, we found that the expression level of \emph{xapAB} among
-	cells follows a bimodal distribution and that the system seems to be
+	cells is bimodal and that the system seems to be
 	bistable (see next section). We aim to understand which of the circuit's
 	features are necessary for bistability and investigate its behavior in
-	different parameter regimes. After presenting some experimental background on the \emph{xap} genes, we will discuss the
+	different parameter regimes. After presenting some experimental background on the \emph{xap} genes, we discuss the
 	details of our model. Next, we estimate the free parameters and then present
@@
 	different backgrounds: \textDelta\emph{xapABR},
-	\textDelta\emph{xapAB}, and wt. The right panel of
+	\textDelta\emph{xapAB}, and wt. The left panel of
 	Fig~\subref{fig1:data}{A} shows that for increasing xanthosine
@@
 		\centering
-		\includegraphics[width=1\textwidth]{media/Fig1_data.eps}
+		\includegraphics[width=1\textwidth]{media/Fig1_experiment.eps}
 		\caption{{\bf Experimental data on the \emph{xap} circuit.}
@@
 		different extracellular concentrations of xanthosine.
-		The right panel shows the wild type circuit while
-		on the left the genes \emph{xapABR} and in the middle the genes
+		The left panel shows the wild type circuit while
+	in the middle the genes \emph{xapABR} and on the right the genes
 		\emph{xapAB} are removed. The wild type circuit behaves like a
@@
 		and also that the chosen xanthosine concentrations are different.
-		(B) The fold-change in gene expression upon addition of
+		(B) shows the fold-change in gene expression upon addition of
 		xanthosine for the wt promoter (left panel)
@@
 			peaks in the bimodal distributions are indicated by smaller
-			points, connected by dotted lines. The arrows illustrate the
+			points, connected by dotted lines. To make this clearer, the
+			bimodal distributions themselves are shown in a light color (in a smoothed form). The arrows illustrate the
 			hysteresis loop in the stochastic simulations.
@@
 	\section*{Acknowledgments}
-	Portions of the experiments reported here were carried out at the
-	Physiology summer course at the Marine Biological Laboratory in Woods
-	Hole, MA, operated by the University of Chicago. We thank Charlotte
-	Strandkvist and Kyle Naughton for their work on the project during the
-	course. We also Nathan Belliveau for strain creation work and for
-	bringing similar systems to our
-	attention, Patrick Lenggenhager for his help with \emph{Mathematica} issues, and Jane Kondev and Jin Wang for interesting discussions.
-	For assistance with illustrations, we also thank Nigel Orme.
+	We thank Patrick Lenggenhager for his help with \emph{Mathematica} issues, Jane Kondev and Jin Wang for interesting discussions, and Nigel Orme for assistance with illustrations.
+\textcolor{red}{Portions of the experiments reported here were carried out at the
+	Physiology Course at the Marine Biological Laboratory in Woods
+	Hole, MA, operated by the University of Chicago.}
 	This work was supported by 1R35 GM118043-01 Maximizing Investigators'

### 2020-03-03 main_text/main_text.tex
Implemented some of GC's edits

@@
 	It is quite conceivable that this auto-activation process is common to many
-	substances that a cell would want to accumulate. Such a switch enables the
+	substances that a cell would want to consume. Such a switch enables the
 	cell to sense and respond to its environment: if the substrate enters the
@@
 	which binds to the transporter substrate and which is often expressed at a low level (often at copy numbers
-	of order $\sim10$, \cite{Schmidt2015}).
+	of order $\sim10$, \cite{Schmidt2015, Li2014}).
 	This is resource efficient for the cell, as this low copy number
@@
 	different backgrounds: \textDelta\emph{xapABR},
-	\textDelta\emph{xapAB}, and wt. The left panel of
+	\textDelta\emph{xapAB}, and wt,
+	and expression as a function of extracellular xanthosine concentration
+	was measured using flow cytometetry, as has been described
+	previously~\cite{RazoMejia2018}. The left panel of
 	Fig~\subref{fig1:data}{A} shows that for increasing xanthosine
@@
 		(A) The expression of the \emph{xapAB} promoter was measured for
-		different extracellular concentrations of xanthosine.
+		different extracellular concentrations of xanthosine (vertical axis).
 		The left panel shows the wild type circuit while
-	in the middle the genes \emph{xapABR} and on the right the genes
-		\emph{xapAB} are removed. The wild type circuit behaves like a
+		the middle and right panels show the effect of deleting
+		the genes \emph{xapABR} and \emph{xapAB}, respectively.
+		The wild type circuit behaves like a
 		switch. Note that the fluorescence scale of the middle panel
@@
 		e.g.,~\cite{Garcia2011,RazoMejia2018,Chure2019},
-		in that no subtraction of autofluorescence was carried out,
+		in that no subtraction of autofluorescence was performed,
 		which is adequate for the qualitative comparison
@@
 
-	Can we gain further insight into how XapR regulates
-	the \emph{xapAB} promoter? As mentioned in
+	As mentioned in
 	the introduction, there are two binding sites for XapR on the
@@
 				into and out of the cell (discussed in more detail below).
-		\item XapR is induced by xanthosine.
-		\item The induced XapR binds to the \emph{xapAB} promoter, leading to transcription.
+		\item The presence of xanthosine shifts XapR's equilibrium from inactive to active.
+		\item Active XapR binds to the \emph{xapAB} promoter, leading to transcription.
 		\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
 		\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
-		\item Production of more XapA and XapB is balanced by decay or dilution through cell divisions.
+		\item Production of more XapA and XapB is balanced by degradation or dilution through cell divisions.
 	\end{itemize}
-	Because xanthosine induces the transcription factor XapR, we have positive
+	Because xanthosine activates the transcription factor XapR, we have positive
 	and negative feedback loops due to XapB and XapA, respectively. The remainder
@@
 	$\langle\n{[mRNA]}\rangle = \frac{r_{\n{m}}}{\gamma_{\n{m}}}
-	p_{\n{active}}$, which we will need in the next paragraph. The mean can also be found from the full chemical master equation, which is shown in~\nameref{S1_Text}.
+	p_{\n{active}}$, which we will need in the next paragraph. The mean can also be found from the full chemical master equation, which is provided in~\nameref{S1_Text}.
 
@@
 Nevertheless, in our case we find that stochastic simulations agree well with the deterministic results.
-Thus, deterministic phase portraits are a surprisingly good starting point.
+Thus, deterministic phase portraits are a valid starting point.
 
@@
 
-	As a remark we point out that we have not observed any oscillations in the
+	We point out that we have not observed any oscillations in the
 	system. Intuitively, they might be expected when the XapA rate is
@@
 	regime, does not even have a significant influence on the system. This can
-	be seen from the plots in figure~\ref{fig6:xapAB}.
+	be seen from the plots in Fig~\ref{fig6:xapAB}.
 	Degradation of xanthosine by XapA lowers the xanthosine and protein concentration at the upper
@@
 		survive, but grew extremely slowly, with xanthosine as the only carbon
-		source, which makes sense in light of Fig~\ref{fig6:xapAB}.
-		With \textit{xapB} removed, the switch never activates and the cells are
+		source. This suggests that low-affinity
+		import of xanthosine by NupC and NupG is sufficient to sustain slow growth,
+		but insufficient to serve as a stand-in for XapB.
+		Fig~\ref{fig6:xapAB} supports this supposition: in our model,
+		with \textit{xapB} removed, the switch never activates and the cells are
 		forced to survive with an extremely meager quantity of XapA to metabolize
@@
 	\paragraph*{The role of cooperativity.} 
-	The model has two cooperatively interacting binding sites for XapR on the
+	The model has two (putatively) cooperatively interacting binding sites for XapR on the
 	\emph{xapAB} promoter and two cooperative binding sites on XapR for
@@
 	we ran the simulation 5000 times for a simulated time of $10^6 \unit{s}$
-	each and started at a mRNA, protein and intracellular xanthosine count of 0.
+	each and started at an mRNA, protein and intracellular xanthosine count of 0.
 
@@
 	of stochastic fluctuations, suggesting that the deterministic
-	picture is surprisingly and unreasonably effective. In other words,
+	picture is remarkably effective. In other words,
 	the fact that bimodality only occurs with some fine tuning of
@@
 	We thank Patrick Lenggenhager for his help with \emph{Mathematica} issues, Jane Kondev and Jin Wang for interesting discussions, and Nigel Orme for assistance with illustrations.
-\textcolor{red}{Portions of the experiments reported here were carried out at the
+	Portions of the experiments reported here were carried out at the
 	Physiology Course at the Marine Biological Laboratory in Woods
-	Hole, MA, operated by the University of Chicago.}
+	Hole, MA, operated by the University of Chicago.
 	This work was supported by 1R35 GM118043-01 Maximizing Investigators'

### 2020-03-03 rebuttal/rebuttal.tex
More of GC's comments
Finished main_text GC comments

@@
 		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
-			(A) Blue shows the results from one run of the stochastic simulation, orange shows the trajectory obtained from solving the deterministic ODE's. The simulation was started
+			(A) Blue shows the results from one typical run of the stochastic simulation, light blue shows two more extreme runs, and orange shows the trajectory obtained from solving the deterministic ODE's.
+The simulation was started
 			at an mRNA, protein and intracellular xanthosine count of 0. The

### 2020-03-03 main_text/main_text.tex
More of GC's comments
Finished main_text GC comments

@@
 	Fig~\subref{fig8:stochC}{(B)} to give bimodality. For lower $\n{[c]_a}$,
-	meaning larger distance between the first and second fixed point,
+	meaning larger difference between the concentrations at the first and second fixed point,
 	almost no switching was observed. Of course, switching is also a matter of the
@@
 	describes the mean values and the simulation fluctuates around this mean.
-	If, however, the mRNA number at the third (high) fixed point is too low, the
+	However, if the mRNA number of the third (high) fixed point is so low that stochastic fluctuations are larger than the difference in concentrations between the first and the second fixed point, the
 	system may not be able to resolve the two points anymore.
@@
 		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
-			(A) Shown in blue is the result from one run of the stochastic simulation, and in orange the trajectory obtained from solving the deterministic ODE's.
-The simulation was run once for $5 \cdot 10^5 \unit{s}$ and started
+			(A) Shown in blue is the result from one typical run of the stochastic simulation, and in orange the trajectory obtained from solving the deterministic ODE's.
+In light blue, two more extreme runs of the simulation are shown for comparison.
+The simulation was run for $5 \cdot 10^5 \unit{s}$ each and started
 			at an mRNA, protein and intracellular xanthosine count of 0. The parameters that were used are the same as in
@@
 	\section*{Conclusion}
-	In this paper, we propose a simple model for genetic circuits containing
+	Here, we propose a simple model for genetic circuits containing
 	a membrane transporter whose gene expression is, directly or indirectly,
@@
 	dynamics. 
-	Furthermore, the modeling results let us, to some extent, understand why the biological system is constructed the way it is.
+	The modeling results let us, to some extent, understand how the biological system can achieve its function.
 By keeping the model as minimal as possible, but still modeling every part explicitly with an appropriate complexity, we can investigate the interesting features while still being able to understand the influence of all parameters and their interplay intuitively.
@@
 	\section*{Acknowledgments}
-	We thank Patrick Lenggenhager for his help with \emph{Mathematica} issues, Jane Kondev and Jin Wang for interesting discussions, and Nigel Orme for assistance with illustrations.
+	We thank Jane Kondev and Jin Wang for interesting discussions, Patrick Lenggenhager for his help with \emph{Mathematica} issues, and Nigel Orme for assistance with illustrations.
 	Portions of the experiments reported here were carried out at the

### 2020-03-04 main_text/main_text.tex
NB main text edits

@@
 		Griffin Chure\textsuperscript{3},
+		Nathan M. Belliveau\textsuperscript{4},
+		Charlotte Strandkvist\textsuperscript{5},
+		Kyle L. Naughton\textsuperscript{6},
 		Rob Phillips\textsuperscript{2,3$\ast$}
@@
 		\textbf{3} Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
+		\\
+		\textbf{4} Howard Hughes Medical Institute, University of Washington, Seattle, WA, USA
+		\\
+		\textbf{5} Department of Systems Biology, Harvard Medical School, Boston, MA, USA
+		\\
+		\textbf{6} Department of Physics and Astronomy, University of Southern California, Los Angeles, CA, USA
 		\\
@@
 	often encodes for a whole set of enzymes for this one metabolite, switches
-	``off'' again. Such a mechanism could be involved in various cases of
-	short-term adaptation.
+	``off'' again. Such a mechanism could plausibly be involved in various cases of
+	short-term adaptation such as, but not limited to, the \emph{lacZYA}, \emph{araFGH},
+	and \emph{xylFGH} examples mentioned above.
 
@@
 	different parameter regimes. After presenting some experimental background on the \emph{xap} genes, we discuss the
-	details of our model. Next, we estimate the free parameters and then present
+	details of our model. Lastly, we estimate the free parameters and then present
 	the observations we made through phase diagrams, followed by the results
@@
 	responsiveness of the promoter to xanthosine (though not apparent
-	from Fig~\subref{fig1:data}{B}, this response remains detectable).
+	from Fig~\subref{fig1:data}{B}, this response remains detectable
+	in other measurements covering a smaller dynamic range).
 
@@
 	\subsection*{Step by step modeling of the system}
-	In this section, we present our model of the \emph{xapAB} genetic circuit.
+	In this section, we present our model of the \emph{xapABR} genetic circuit.
 	Fig~\ref{fig2:model} shows an overview of this model.
@@
 		\item The presence of xanthosine shifts XapR's equilibrium from inactive to active.
-		\item Active XapR binds to the \emph{xapAB} promoter, leading to transcription.
+		\item Active XapR binds to the \emph{xapAB} promoter, increasing transcription.
 		\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
@@
 	\paragraph*{Transcription.}
-	Transcription and translation of the \emph{xapAB} gene, regulated by the
+	Transcription and translation of the \emph{xapAB} operon, regulated by the
 	induced XapR, produce the two proteins XapA and XapB. We start with
@@
 	\frac{\n{[p]}}{K_{\n{a}}}$, $\n{[x]_a} \defeq \frac{\n{[x]}}{K_{\n{a}}}$,
+	$\n{[c]_a} \defeq \frac{\n{[c]}}{K_{\n{a}}}$,
 	and $\tau \defeq \gamma_{\n{p}} t$. Using these definitions, the following
@@
 	stochastic simulations later on), but is low enough to lead to mean mRNA
-	numbers that are very reasonable (see~\cite{Milo2016} for the average mRNA
-	numbers in bacterial cells). The protein and xanthosine
+	numbers that are very reasonable (of order $10^{-2}$ to $10^{1}$
+	transcripts per gene per cell in \textit{E.\ coli}~\cite{Milo2016}).
+	The protein and xanthosine
 	concentrations followed from this, but with some variation in the parameters
@@
 	copy numbers, the number of reaction firings that must be simulated with
-	Gillespie's classical algorithm leads to an impractical computational
-	cost. This would make Gillespie's \texttau-leap algorithm ideal for the high
+	Gillespie's classical algorithm~\cite{Gillespie1976} leads to an impractical computational
+	cost. This would make Gillespie's \texttau-leap algorithm~\cite{Gillespie2001} ideal for the high
 	expression state. On the other hand, \texttau-leaping cannot be used for
@@
 		K-12.
-		\newblock Molecular and General Genetics MGG. 1980;179(2):341--348.
+		\newblock Molecular and General Genetics. 1980;179(2):341--348.
 
@@
 		involved in xanthosine catabolism in \emph{Escherichia coli}.
-		\newblock Journal of bacteriology. 1995;177(7559336):5506--5516.
+		\newblock Journal of Bacteriology. 1995;177(7559336):5506--5516.
 
@@
 			coli} regulatory protein XapR.
-		\newblock Journal of bacteriology. 1999;181(10400599):4397--4403.
+		\newblock Journal of Bacteriology. 1999;181(10400599):4397--4403.
 
@@
 		superfamily.
-		\newblock Journal of bacteriology. 2001;183(11466294):4900--4904.
+		\newblock Journal of Bacteriology. 2001;183(11466294):4900--4904.
 
@@
 		Milo R, Phillips R.
-		\newblock In: Cell biology by the numbers. Garland Science. 2016; 217.
+		\newblock In: Cell Biology by the Numbers. Garland Science. 2016; 217.
 
@@
 		\newblock Journal of Theoretical Biology. 2000;203(2):117--133.
+
+		\bibitem{Gillespie1976}
+		Gillespie, DT.
+		\newblock A General Method for Numerically Simulating the Stochastic
+		Time Evolution of Coupled Chemical Reactions.
+		\newblock Journal of Computational Physics. 1976;22(4):403--434.
+
+		\bibitem{Gillespie2001}
+		Gillespie, DT.
+		\newblock Approximate Accelerated Stochastic Simulation
+		of Chemically Reacting Systems.
+		\newblock J Chem Phys. 2001;115(4):1716--1733.
 

### 2020-03-04 main_text/main_text.tex
Add expt details

@@
 
+	\subsection*{Materials and Methods}
+	\paragraph*{Bacterial Strains.} Strains used in this work
+	derive from {\it E. coli} K12 MG1655 with the {\it lac} operon
+	removed and were constructed similarly to those used in
+	\cite{RazoMejia 2018, Belliveau2019}. $\Delta{xapAB}$ and
+	$\Delta{xapABR}$ knockouts were generated using the approach of
+	Datsenko and Wanner \cite{Datsenko2000}. For the fluorescent
+	reporter constructs the {\it xapAB} promoter region, with one or
+	both XapR binding sites, were extracted from {\it E. coli} K12
+	MG1655 by PCR. These were cloned into the YFP-expressing reporter
+	plasmid pZS25,  containing kanamycin resistance, and integrated at
+	the {\it galK} locus using $\lambda$ Red recombineering \cite{Sharan2009}.
+
+	\paragraph*{Flow Cytometry Measurements.}
+	Experimental measurements in Fig. 2B of the main text were obtained
+	via flow cytometry and were performed in a similar manner as
+	described in~\cite{RazoMejia2018}. Briefly, wild-type,
+	$\Delta{xapABR}$, and $\Delta{xapAB}$ were grown to saturation in
+	rich LB Miller (BD Medical) medium and were diluted 1:1000 into M9
+	minimal medium supplemented with 0.5\% (w/v) glucose and the
+	appropriate concentration of xanthosine. For measurements of the
+	$\Delta{xapAB}$ strain, chloramphenicol was also added to the growth
+	medium to ensure stability of the pZ3PN25-tetR plasmid along with 10
+	ng /mL of anhydro-tetracycline to maximize expression of
+	xapR-mCherry. Cultures were allowed to grow at 37$^\circ$ C for six
+	to eight hours to an OD$_{600nm}$ of $\approx$ 0.3 At this point,
+	the cultures were diluted 1:10 into M9 + 0.5\% glucose and were
+	immediately analyzed via a MacsQuant VYB Flow Cytometery
+	(MiltenyiBiotec). Measurement files were exported to CSV file
+	formats and analyzed as previously described
+	in~\cite{RazoMejia2018}.
+
 	\section*{Supporting information}
@@
 
+		\bibitem{Belliveau2019}
+		Belliveau, NM, Barnes, SL, Ireland, WT,
+		Jones, DL, Sweredoski, MJ, Moradian, A, et al.
+		\newblock Systematic approach for dissecting the molecular
+		mechanisms of transcriptional regulation in bacteria.
+		\newblock PNAS, 2018;115(21), E4796–E4805.
+		\newblock doi:{10.1073/pnas.1722055115}
+
+		\bibitem{Datsenko2000}
+		Datsenko, KA, Wanner, BL.
+		\newblock One-step inactivation of chromosomal genes
+		in Escherichia coli K-12 using PCR products.
+		\newblock PNAS, 2000;97(12), 6640–6645.
+		\newblock doi:{10.1073/pnas.120163297}
+
+		\bibitem{Sharan2009}
+		Sharan, SK, Thomason, LC, Kuznetsov, SG, Court, DL.
+		\newblock Recombineering: a homologous recombination-based
+		method of genetic engineering.
+		\newblock Nature Protocols, 2009;4(2), 206–223.
+		\newblock doi{10.1038/nprot.2008.227}
+
 	\end{thebibliography}

### 2020-03-04 s1_text/s1_text.tex
Update author list & acknowledgements

--- 2020-02-22 s1_text/s1_text.tex
@@
 	Muir J. Morrison\textsuperscript{2},
+	Griffin Chure\textsuperscript{3},
 	Rob Phillips\textsuperscript{2,3$\ast$}
@@
 \section{Experimental materials and methods} \label{sec:exp}
+\paragraph{Construction of $\Delta{xapABR}$ deletion strain.} Cells containing
+plasmid pSIM6 were diluted 1:100 in 50 ml LB media and grown to an OD600
+of $\approx 0.4$ at $30 ^\circ$C. The culture was immediately placed in
+a water bath shaker at 43$^\circ$C for 15 minutes and then cooled in an
+ice bath for 10 minutes. Cells were then spun down for 10 minutes
+(4,000~$g$, 4$^\circ$C) and resuspended on ice in 50~ml of chilled
+water. This was repeated three times before resuspending in 200~$\mu$L
+of chilled water to generate competent cells. Homologous primer
+extension sequences were obtained from Baba {\it et al.} \cite{Baba2006}
+and used to generate linear DNA containing a kanamycin resistance
+gene insert by PCR (amplified from plasmid pKD4), using primers that.
+contained homology for the region on the chromosome to be deleted
+\cite{Datsenko2000}. Electroporation was performed using 1~$\mu$L
+purified PCR product (about 100~ng DNA), mixed with 50~$\mu$L cells.
+Cells were resuspended in 750~$\mu$L SOC media and outgrowth was
+performed on a shaker at 30$^\circ$C for 90-120 minutes. Cells were then
+selected on an LB-agar plate containing kanamycin (30~$\mu$g/ml) and
+grown overnight at 30$^\circ$C. The deletions were confirmed by both
+colony PCR and sequencing. After confirmation, the deletion was
+transferred to a clean strain through P1 transduction and selection on
+kanamycin. Removal of kanamycin resistance was achieved by transforming
+cells with plasmid pCP20 (expressing FLP recombinase, carbenicillin
+resistance, on a temperature-sensitive replicon), with cells grown at
+30$^\circ$C. Lastly, several colonies were then growth colony-purified
+once non-selectively at 43$^\circ$C and then tested for loss of
+antibiotic resistance for both kanamycin and carbenicillin.
 
-\textcolor{red}{missing}
+\paragraph{Construction of $\Delta{xapAB}$ strain.}
+Further experiments are underway in which precise control of the xapR copy number are needed and the preliminary data from these experiments are shown in the middle panel of Fig.
+2A of the main text.
+This strain was constructed by creating a XapR-mCherry fusion protein expressed from a tetR regulated promoter.
+This fusion construct was integrated into the ycbN locus of the $\Delta{xapABR}$ strain making an effective $\Delta{xapAB}$ strain.
+A pZS3-PN25-TetR expression plasmid was also inserted into this genetic background which permits titratable control over the xapR-mCherry copy number.
+For the data presented in this work, expression was maximized using 10~ng/mL of the tetR inducer anhydro-tetracycline.
+Further information of a similar construct (LacI-mCherry) can be found in~\cite{Brewster2014} and~\cite{Chure2019}.
+
+\paragraph{Primers for deletion strains}
+Generation of $\Delta{xapABR}$:
+actatttccggcaattcctgtctcctcacctactgtgtcaatgcagccaaCATATGAATATCCTCCTTAG; $xapABR<>P2$
+(lower case is homology, upper case is pKD4 primer)
+
+agatacaacaggtatcgctaatatagaggaaagtgtcagagttaaaagatgttttttcatGTGTAGGCTGGAGCTGCTTC; $xapABR<>P1$ (lower case is homology, upper case is pKD4 primer)
+
+Primers for colony-PCR to confirm deletion of xapABR:
+(upstream of xapR) cgggtcgttagctcagttggta
+(downstream of XapA): ttcccagtcgaagaacccgtac
+
+\textbf{Finish reformatting to table, add other primers?}
 
@@
 
+	\bibitem{Baba2006} Baba T, Ara T, Hasegawa M,
+	Takai Y, Okumura Y, Baba, M, et al.
+	\newblock Construction of \textit{Escherichia coli} K-12 in-frame,
+	single-gene knockout mutants: the Keio collection.
+	\newblock Molecular Systems Biology, 2006;2(1), 2460.
+	\newblock doi:{10.1038/msb4100050}
+
+	\bibitem{Datsenko2000}
+	Datsenko, KA, Wanner, BL.
+	\newblock One-step inactivation of chromosomal genes
+	in Escherichia coli K-12 using PCR products.
+	\newblock PNAS, 2000;97(12), 6640–6645.
+	\newblock doi:{10.1073/pnas.120163297}
+
 \end{thebibliography}

### 2020-03-04 main_text/main_text.tex
Correct typos and new figure 10

@@
 			points, connected by dotted lines. To make this clearer, the
-			bimodal distributions themselves are shown in a light color (in a smoothed form). The arrows illustrate the
+			bimodal distributions themselves are shown on the right. The arrows illustrate the
 			hysteresis loop in the stochastic simulations.
@@
 	removed and were constructed similarly to those used in
-	\cite{RazoMejia 2018, Belliveau2019}. $\Delta{xapAB}$ and
+	\cite{RazoMejia2018, Belliveau2019}. $\Delta{xapAB}$ and
 	$\Delta{xapABR}$ knockouts were generated using the approach of
@@
 	xapR-mCherry. Cultures were allowed to grow at 37$^\circ$ C for six
-	to eight hours to an OD$_{600nm}$ of $\approx$ 0.3 At this point,
+	to eight hours to an OD$_{600nm}$ of $\approx$ 0.3. At this point,
 	the cultures were diluted 1:10 into M9 + 0.5\% glucose and were

### 2020-03-04 main_text/main_text.tex
GC comments and unit formatting
Finished GC S1 Text comments. Small typos/formatting corrections.

@@
 	derive from {\it E. coli} K12 MG1655 with the {\it lac} operon
-	removed and were constructed similarly to those used in
-	\cite{RazoMejia2018, Belliveau2019}. $\Delta{xapAB}$ and
+	removed and were constructed similarly to those used in~\cite{RazoMejia2018, Belliveau2019}. $\Delta{xapAB}$ and
 	$\Delta{xapABR}$ knockouts were generated using the approach of
-	Datsenko and Wanner \cite{Datsenko2000}. For the fluorescent
+	Datsenko and Wanner~\cite{Datsenko2000}. For the fluorescent
 	reporter constructs the {\it xapAB} promoter region, with one or
@@
 	plasmid pZS25, containing kanamycin resistance, and integrated at
-	the {\it galK} locus using $\lambda$ Red recombineering \cite{Sharan2009}.
+	the {\it galK} locus using $\lambda$ Red recombineering~\cite{Sharan2009}.
 
@@
 	ng /mL of anhydro-tetracycline to maximize expression of
-	xapR-mCherry. Cultures were allowed to grow at 37$^\circ$ C for six
-	to eight hours to an OD$_{600nm}$ of $\approx$ 0.3. At this point,
+	xapR-mCherry. Cultures were allowed to grow at $37\unit{^\circ C}$ for six
+	to eight hours to an OD$_{600\unit{nm}}$ of $\approx 0.3$. At this point,
 	the cultures were diluted 1:10 into M9 + 0.5\% glucose and were

### 2020-03-04 s1_text/s1_text.tex
GC comments and unit formatting
Finished GC S1 Text comments. Small typos/formatting corrections.

@@
 \usepackage{textgreek}
+\usepackage{upgreek}
 \usepackage{bm}
@@
 	\\
-	\textbf{3} Department of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
+	\textbf{3} Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
 	\\
@@
 The protein decay should be dominated by dilution through cell division, which is a discrete process.
-However, the cell grows before division, resulting in a continuous dilution.
+However, the cell grows before division, resulting in a continuous dilution in concentration.
 Furthermore, replication leads to a varying number of gene copies and thereby also a varying transcription rate.
@@
 \paragraph*{Kinetics of XapB, NupC, and NupG.}
-Entering the cell consists of two steps. In a first transport step, xanthosine passes the
+All statements in this paragraph are based on data
+from~\cite{Norholm2001}. Xanthosine entering the cell consists of two steps. In a first transport step, xanthosine passes the
 outer membrane through porins like Tsx, OmpF and OmpC. This does not seem to
@@
 xanthosine that then activate \emph{xapB}, and there appears to be no other
-significant way in which xanthosine can enter the cell. All statements in this paragraph are based on data
-from~\cite{Norholm2001}.
+significant way in which xanthosine can enter the cell.
 
@@
 \section{Experimental materials and methods} \label{sec:exp}
-\paragraph{Construction of $\Delta{xapABR}$ deletion strain.} Cells containing
-plasmid pSIM6 were diluted 1:100 in 50 ml LB media and grown to an OD600
-of $\approx 0.4$ at $30 ^\circ$C. The culture was immediately placed in
-a water bath shaker at 43$^\circ$C for 15 minutes and then cooled in an
+\paragraph{Construction of $\bm{\Delta{xapABR}}$ deletion strain.} Cells containing
+plasmid pSIM6 were diluted 1:100 in $50 \unit{mL}$ LB media and grown to an OD600
+of $\approx 0.4$ at $30^\circ$C. The culture was immediately placed in
+a water bath shaker at $43\unit{^\circ C}$ for 15 minutes and then cooled in an
 ice bath for 10 minutes. Cells were then spun down for 10 minutes
-(4,000~$g$, 4$^\circ$C) and resuspended on ice in 50~ml of chilled
-water. This was repeated three times before resuspending in 200~$\mu$L
+($4,000 \unit{g}$, $4\unit{^\circ C}$) and resuspended on ice in $50\unit{mL}$ of chilled
+water. This was repeated three times before resuspending in $200 \unit{\upmu L}$
 of chilled water to generate competent cells. Homologous primer
@@
 contained homology for the region on the chromosome to be deleted
-\cite{Datsenko2000}. Electroporation was performed using 1~$\mu$L
-purified PCR product (about 100~ng DNA), mixed with 50~$\mu$L cells.
-Cells were resuspended in 750~$\mu$L SOC media and outgrowth was
-performed on a shaker at 30$^\circ$C for 90-120 minutes. Cells were then
-selected on an LB-agar plate containing kanamycin (30~$\mu$g/ml) and
-grown overnight at 30$^\circ$C. The deletions were confirmed by both
+\cite{Datsenko2000}. Electroporation was performed using $1 \unit{\upmu L}$
+purified PCR product (about $100 \unit{ng}$ DNA), mixed with $50 \unit{\upmu L}$ cells.
+Cells were resuspended in $750 \unit{\upmu L}$ SOC media and outgrowth was
+performed on a shaker at $30\unit{^\circ C}$ for 90-120 minutes. Cells were then
+selected on an LB-agar plate containing kanamycin ($30 \unit{\upmu g/mL}$) and
+grown overnight at $30\unit{^\circ C}$. The deletions were confirmed by both
 colony PCR and sequencing. After confirmation, the deletion was
@@
 resistance, on a temperature-sensitive replicon), with cells grown at
-30$^\circ$C. Lastly, several colonies were then growth colony-purified
-once non-selectively at 43$^\circ$C and then tested for loss of
+$30\unit{^\circ C}$. Lastly, several colonies were then growth colony-purified
+once non-selectively at $43\unit{^\circ C}$ and then tested for loss of
 antibiotic resistance for both kanamycin and carbenicillin.
 
-\paragraph{Construction of $\Delta{xapAB}$ strain.}
+\paragraph{Construction of $\bm{\Delta{xapAB}}$ strain.}
 Further experiments are underway in which precise control of the xapR copy number are needed and the preliminary data from these experiments are shown in the middle panel of Fig.
@@
 A pZS3-PN25-TetR expression plasmid was also inserted into this genetic background which permits titratable control over the xapR-mCherry copy number.
-For the data presented in this work, expression was maximized using 10~ng/mL of the tetR inducer anhydro-tetracycline.
+For the data presented in this work, expression was maximized using $10 \unit{ng/mL}$ of the tetR inducer anhydro-tetracycline.
 Further information of a similar construct (LacI-mCherry) can be found in~\cite{Brewster2014} and~\cite{Chure2019}.
@@
 
+	\bibitem{Chure2019}
+	Chure G, Kaczmarek ZA, Phillips R.
+	\newblock Physiological Adaptability and Parametric Versatility in a Simple
+	Genetic Circuit.
+	\newblock bioRxiv. 2019;doi:{10.1101/2019.12.19.878462}.
+
 \end{thebibliography}

### 2020-03-04 rebuttal/rebuttal.tex
New Fig 10 and time evolution data

@@
 
-{\Large\textcolor{red}{ToDo: change author list above and at the end, add correct github link (public repo!).}}
+{\Large\textcolor{red}{ToDo: change author list above and at the very end of this.}}
 
@@
 	\includegraphics{Fig10_bifurcation.eps}
+	\vspace*{0.5em}
 	\caption{{\bf Bifurcation diagram showing the hysteresis loop in the stochastic system.}
@@
 The positions of the two peaks in the bimodal distributions are indicated by smaller points, connected by dotted lines.
-To make this clearer, the bimodal distributions themselves are shown in a light color (in a smoothed form).
+To make this clearer, the bimodal distributions themselves are shown on the right.
 The arrows illustrate the hysteresis loop in the stochastic simulations.}
@@
 		\includegraphics{Fig9_evolution.eps}
+		\vspace*{0.5em}
 		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
@@
 The raw data is
-publicly available at \url{https://github.com/RPGroup-PBoC/mwc_activation/}.
+publicly available at DOI: 10.5281/zenodo.3695049.
 \end{response}

### 2020-03-05 rebuttal/rebuttal.tex
Edits to expt details + author list & minor polish

@@
 \section*{Response to Reviewer for PONE-D-19-32846 ``Theoretical Investigation of a Genetic Switch for Metabolic Adaptation''}
-Kathrin S. Laxhuber,$^{1}$ Muir J. Morrison,$^{2}$ Griffin Chure,$^{3}$ Rob Phillips$^{2, 3, \dagger}$
+Kathrin S. Laxhuber\textsuperscript{1},
+Muir J. Morrison\textsuperscript{2},
+Griffin Chure\textsuperscript{3},
+Nathan M. Belliveau\textsuperscript{4},
+Charlotte Strandkvist\textsuperscript{5},
+Kyle L. Naughton\textsuperscript{6},
+Rob Phillips\textsuperscript{2,3$\dagger$}
 \vspace{10pt}
 
-\noindent\tiny{$^1$ Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland,
+\noindent\tiny{
+$^1$ Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland,
 $^2$ Department of Physics, California Institute of Technology, Pasadena, CA, USA,
-$^3$ Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA, $^\dagger$ E-mail: \texttt{phillips@pboc.caltech.edu}}
-
-{\Large\textcolor{red}{ToDo: change author list above and at the very end of this.}}
+$^3$ Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA,
+$^4$ Howard Hughes Medical Institute, University of Washington, Seattle, WA, USA,
+$^5$ Department of Systems Biology, Harvard Medical School, Boston, MA, USA,
+$^6$ Department of Physics and Astronomy, University of Southern California, Los Angeles, CA, USA, $^\dagger$ E-mail: \texttt{phillips@pboc.caltech.edu}}
 

### 2020-03-05 s1_text/s1_text.tex
Edits to expt details + author list & minor polish

@@
 	Griffin Chure\textsuperscript{3},
+	Nathan M. Belliveau\textsuperscript{4},
+	Charlotte Strandkvist\textsuperscript{5},
+	Kyle L. Naughton\textsuperscript{6},
 	Rob Phillips\textsuperscript{2,3$\ast$}
@@
 	\textbf{3} Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
+	\\
+	\textbf{4} Howard Hughes Medical Institute, University of Washington, Seattle, WA, USA
+	\\
+	\textbf{5} Department of Systems Biology, Harvard Medical School, Boston, MA, USA
+	\\
+	\textbf{6} Department of Physics and Astronomy, University of Southern California, Los Angeles, CA, USA
 	\\
@@
 and used to generate linear DNA containing a kanamycin resistance
-gene insert by PCR (amplified from plasmid pKD4), using primers that.
+gene insert by PCR (amplified from plasmid pKD4), using primers that
 contained homology for the region on the chromosome to be deleted
@@
 
-\paragraph{Primers for deletion strains}
-Generation of $\Delta{xapABR}$:
-actatttccggcaattcctgtctcctcacctactgtgtcaatgcagccaaCATATGAATATCCTCCTTAG; $xapABR<>P2$
-(lower case is homology, upper case is pKD4 primer)
-
-agatacaacaggtatcgctaatatagaggaaagtgtcagagttaaaagatgttttttcatGTGTAGGCTGGAGCTGCTTC; $xapABR<>P1$ (lower case is homology, upper case is pKD4 primer)
-
-Primers for colony-PCR to confirm deletion of xapABR:
-(upstream of xapR) cgggtcgttagctcagttggta
-(downstream of XapA): ttcccagtcgaagaacccgtac
-
-\textbf{Finish reformatting to table, add other primers?}
+\begin{table}
+\begin{tabular}{c p{5cm} p{4cm}}
+Primer & Sequence & Comments \\
+\hline
+$xapABR<>P2$ &
+actatttccggcaattcctgtctcctcac ctactgtgtcaatgcagccaaCATAT GAATATCCTCCTTAG &
+Forward primer for pKD4 with homology to \textit{xapABR} locus
+\\
+$xapABR<>P1$ &
+agatacaacaggtatcgctaatatagagg aaagtgtcagagttaaaagatgtttt ttcatGTGTAGGCTGGAGCTGCTTC &
+Reverse primer for pKD4 with homology to \textit{xapABR} locus
+\\
+xapABR-seq-fwd &
+cgggtcgttagctcagttggta &
+upstream of \textit{xapR}, to check \textit{xapABR} deletion
+\\
+xapABR-seq-rev &
+ttcccagtcgaagaacccgtac &
+downstream of \textit{xapA}, to check \textit{xapABR} deletion
+\end{tabular}
+\caption{\textbf{Primers for deletion strains.}
+Lower case sequences are homologous to a chromosomal locus.
+Upper case indicates sequences used for PCR amplification.
+}
+\end{table}
 
@@
 	\bibitem{Datsenko2000}
-	Datsenko, KA, Wanner, BL.
+	Datsenko KA, Wanner BL.
 	\newblock One-step inactivation of chromosomal genes
@@
 	\newblock doi:{10.1073/pnas.120163297}
+
+	\bibitem{Brewster2014}
+	Jones, DL, Brewster, RC, Phillips, R
+	\newblock Promoter architecture dictates cell-to-cell variability in gene expression.
+	\newblock Science, 2014;346(6216), 1533--1536.
+	\newblock doi:{10.1126/science.1255301}
 

### 2020-03-05 main_text/main_text.tex
Edits to expt details + author list & minor polish
More edits & polish on materials & methods

@@
 	xanthosine response of our reporter construct disappears (see
-	Fig~\subref{fig1:data}{A}, left panel).
+	Fig~\subref{fig1:data}{A}, middle panel).
 	This is the result of removing the transcription factor XapR that is
@@
 	is regained but it is no longer switch-like (see
-	Fig~\subref{fig1:data}{A}, middle panel). Instead, the distribution remains
+	Fig~\subref{fig1:data}{A}, right panel). Instead, the distribution remains
 	unimodal and simply shifts to higher expression as the xanthosine
@@
 	the introduction, there are two binding sites for XapR on the
-	\emph{xapAB} promoter: one directly at the polymerase binding site
+	\emph{xapAB} promoter: one partially overlapping the polymerase -35 site
 	and one further upstream.
@@
 	derive from {\it E. coli} K12 MG1655 with the {\it lac} operon
-	removed and were constructed similarly to those used in~\cite{RazoMejia2018, Belliveau2019}. $\Delta{xapAB}$ and
-	$\Delta{xapABR}$ knockouts were generated using the approach of
+	removed and were constructed similarly to those used in~\cite{RazoMejia2018, Belliveau2019}. 
+	The $\Delta{xapABR}$ knockout was generated using the approach of
 	Datsenko and Wanner~\cite{Datsenko2000}. For the fluorescent
@@
 	MG1655 by PCR. These were cloned into the YFP-expressing reporter
-	plasmid pZS25, containing kanamycin resistance, and integrated at
+	plasmid pZS25~\cite{Lutz1997},
+	containing kanamycin resistance, and integrated at
 	the {\it galK} locus using $\lambda$ Red recombineering~\cite{Sharan2009}.
+	A xapR-mCherry fusion was constructed in a pZS31 plasmid and was integrated
+	into the \textit{ybcN} locus, also using $\lambda$ Red recombineering.
+	Expression of the xapR-mCherry fusion was regulated by TetR expressed
+	extrachromosomally from a pZS3-PN25 plasmid.
+	Expression of the xapR-mCherry fusion was induced by the addition of
+	10~ng/mL anhydro-tetracycline.
+	A similar expression system was used in~\cite{Brewster2014}
+	More details can be found in~\nameref{S1_Text}.
 
 	\paragraph*{Flow Cytometry Measurements.}
-	Experimental measurements in Fig. 2B of the main text were obtained
+	Experimental measurements in Fig. 2 were obtained
 	via flow cytometry and were performed in a similar manner as
@@
 	minimal medium supplemented with 0.5\% (w/v) glucose and the
-	appropriate concentration of xanthosine. For measurements of the
-	$\Delta{xapAB}$ strain, chloramphenicol was also added to the growth
-	medium to ensure stability of the pZ3PN25-tetR plasmid along with 10
-	ng /mL of anhydro-tetracycline to maximize expression of
-	xapR-mCherry. Cultures were allowed to grow at $37\unit{^\circ C}$ for six
+	appropriate concentration of xanthosine.
+	Cultures were allowed to grow at $37\unit{^\circ C}$ for six
 	to eight hours to an OD$_{600\unit{nm}}$ of $\approx 0.3$. At this point,
@@
 
+		\bibitem{Lutz1997}
+		Lutz R, and Bujard, H
+		\newblock Independent and Tight Regulation of Transcriptional Units in \textit{Escherichia Coli} Via the {{LacR}}/{{O}}, the {{TetR}}/{{O}} and {{AraC}}/{{I1}}-{{I2}} Regulatory Elements.
+		\newblock Nucleic Acids Research, 1997;25(6), 1203--1210.
+		\newblock doi:{10.1093/nar/25.6.1203}
+
 		\bibitem{Sharan2009}
@@
 
+		\bibitem{Brewster2014}
+		Jones, DL, Brewster, RC, Phillips, R
+		\newblock Promoter architecture dictates cell-to-cell variability in gene expression.
+		\newblock Science, 2014;346(6216), 1533--1536.
+		\newblock doi:{10.1126/science.1255301}
+
 	\end{thebibliography}

### 2020-03-05 s1_text/s1_text.tex
More edits & polish on materials & methods

@@
 \paragraph{Construction of $\bm{\Delta{xapAB}}$ strain.}
-Further experiments are underway in which precise control of the xapR copy number are needed and the preliminary data from these experiments are shown in the middle panel of Fig.
-2A of the main text.
-This strain was constructed by creating a XapR-mCherry fusion protein expressed from a tetR regulated promoter.
-This fusion construct was integrated into the ycbN locus of the $\Delta{xapABR}$ strain making an effective $\Delta{xapAB}$ strain.
-A pZS3-PN25-TetR expression plasmid was also inserted into this genetic background which permits titratable control over the xapR-mCherry copy number.
-For the data presented in this work, expression was maximized using $10 \unit{ng/mL}$ of the tetR inducer anhydro-tetracycline.
-Further information of a similar construct (LacI-mCherry) can be found in~\cite{Brewster2014} and~\cite{Chure2019}.
+Further experiments are underway in which precise control of the xapR
+copy number are needed and the preliminary data from these experiments
+are shown in the right panel of Fig. 2A of the main text. This strain
+was constructed by creating a XapR-mCherry fusion protein expressed from
+a tetR regulated promoter. This fusion construct was integrated into the
+ycbN locus of the $\Delta{xapABR}$ strain making an effective
+$\Delta{xapAB}$ strain. A pZS3-PN25-TetR expression plasmid was also
+inserted into this genetic background which permits titratable control
+over the xapR-mCherry copy number. For the data presented in this work,
+expression was maximized using $10 \unit{ng/mL}$ of the TetR inducer
+anhydro-tetracycline. Further information of a similar construct
+(LacI-mCherry) can be found in~\cite{Brewster2014} and~\cite{Chure2019}.
 

### 2020-03-30 rebuttal/second_round/rebuttal2.tex
Added rebuttal letter for second round

--- 2020-03-05 rebuttal/rebuttal.tex
@@
 \begin{document}
-\section*{Response to Reviewer for PONE-D-19-32846 ``Theoretical Investigation of a Genetic Switch for Metabolic Adaptation''}
+\section*{Second Response to Reviewer for PONE-D-19-32846 ``Theoretical Investigation of a Genetic Switch for Metabolic Adaptation''}
 Kathrin S. Laxhuber\textsuperscript{1},
@@
 Nathan M. Belliveau\textsuperscript{4},
-Charlotte Strandkvist\textsuperscript{5},
+Charlotte Strand-kvist\textsuperscript{5},
 Kyle L. Naughton\textsuperscript{6},
@@
 \begin{responseSummary}
-We thank the editor and reviewer for considering our manuscript and
-bringing some important points to our attention. The reviewer's comments
-significantly improved our work and inspired us to make very interesting
-comparisons between the deterministic and stochastic picture of our
-system. Furthermore, as rightly suggested by the reviewer, we have now
-added experimental data to the manuscript and expanded the author list
-to reflect this. Below, we address the reviewer's concerns one-by-one
-and indicate where changes were made either to the main text or the
-supporting information. Besides these changes, we made a few small
-improvements and corrections to the text that had slipped our attention before. We
-believe we have resolved the three issues brought up by the reviewer and
-hope our manuscript is now ready for publication in PLOS ONE.
+We would like to thank the editor and reviewer for the thorough and very constructive review process. 
+We are happy about the positive overall feedback and hope that we have now resolved any remaining ambiguities.
 \end{responseSummary}
@@
 \begin{review}
-In their manuscript "Theoretical investigation of a genetic switch for
-metabolic adaptation", Laxhuber and colleagues present a very nice
-theoretical analysis of the behavior of the \emph{E. coli} xapABR system, using
-both deterministic and stochastic simulations to demonstrate that the
-system can show meaningful and important bistability driven by the
-activity of the membrane transporter. In general the work is well done,
-and it is technically of high quality, but at present I have two
-substantial concerns that I think must be addressed in order for the
-paper to live up to its potential.
+	In general the authors have done an admirable job addressing my comments.
 \end{review}
@@
 \begin{response}
-We thank the reviewer for their careful reading of our work and are
-pleased that their impression of the manuscript is favorable overall.
+	We thank the reviewer and are glad that we could clarify all previous issues.
 \end{response}
@@
 \begin{review}
-My first concern is that while the authors present their deterministic
-simulations first, and draw many of their conclusions primarily from
-those simulations, it seems to me that the use of such simulations is
-contra-indicated by the authors' own findings. As both the number of
-XapR molecules and mRNA molecules are small integers (likely 10 or
-less), treatment of these species as continuous, deterministic
-quantities can never be more than a poor approximation. I find this
-troublesome for a couple of reasons. First, the authors still treat
-their continuous treatment as an apparently 'good' way to model a
-genetic switch like that at work at the xapAB promoter; while I find
-their analysis insightful, in the end I cannot support such a
-recommendation given the stochasticity intrinsic in the low protein and
-mRNA copy numbers present here.
+	At this point, I find myself confused on one issue, which I hope the authors will address (unless I'm just being truly dense).
+In the right side of Figure 9, a series of phase portraits are displayed.
+I assume that these must have been from the deterministic system, since the authors make no reference to performing a suitable scan of conditions to obtain such diagrams for the stochastic system; I assume that these in fact arise from the deterministic system.
+However, the figure legend does not address this point at all, and the reader is left with the impression that those phase portraits arise from the stochastic system.
+The issue is not helped by the fact that the coloring and axis scaling for the right side of Figure 9 does not follow the same conventions used in earlier diagrams; for example, the right panel of Fig 9A looks a lot like Fig.
+5B, but the colors and axis scales are different.
 \end{review}
@@
 \begin{response}
-We agree with the concern of the reviewer that in general, deterministic
-approximations of systems with such low copy numbers need to be verified. We do not intend to give the general recommendation that a
-continuous treatment alone is usually sufficient, and in order to avoid
-this, we added some words of warning in the beginning of our
-deterministic as well as our stochastic analysis.
-
-Nevertheless, for our system we did successfully test the concordance of the stochastic simulations and the deterministic model.
-To make this more apparent, we included some additional comparisons of the two
-models. We address these in more detail in our responses below,
-but to summarize: we added a bifurcation diagram
-(Fig~\ref{bifucation}) which illustrates that the stochastic
-distributions track the deterministic fixed points very well,
-and we compared the stochastic timescales with the solution of the
-deterministic model (Fig~\ref{adaptationtime}) and found excellent
-agreement there too.
-
-Even though stochastic fluctuations around the deterministic
-trajectories may be large, the stochastic trajectories do still flow to
-the fixed points as predicted by the deterministic model.
-Hence, the excellent agreement may be surprising or unintuitive at first, but the conclusions we draw from the deterministic phase diagrams turn out to be accurate.
-
-%, except when
-%the system is close to either of the bifurcation points in
-%Fig~\ref{bifucation}, where it is unsurprising that fluctuations are
-%magnified in importance.
-
-Given this concordance, we prefer to start by showing the deterministic case and illustrating our conclusions there before moving on to the stochastic picture.
-That is because the deterministic phase diagrams have the advantage of providing the reader with a more intuitive understanding of the system than the plots of the results from the stochastic simulation.
+	This is a very good point that we had missed, and we thank the reviewer for reading our manuscript so carefully.
+The phase portraits are indeed from the deterministic system.
+We now addressed this in the figure legends and have changed the coloring in Fig 9.
 \end{response}
@@
 \begin{review}
-Second, while the authors do technically recapitulate the bistable
-behavior of the system in the stochastic version, the bistability seems
-based on their description to be quite fragile and not very meaningful,
-since it only occurs here for small distances between the lower and
-upper fixed points
+	I would also like to caution the authors (and editorial staff) that while the ``Revised Manuscript with Track Changes'' document was fully readable, the figure numbering of the main text figures seemed to be off in the submission system.
 \end{review}
@@
 \begin{response}
-We thank the reviewer for bringing this point to our attention, which we
-believe to be a misunderstanding that resulted from a poor explanation
-in our original text. We believe we have clarified this in the
-manuscript, and allow us elaborate here too: Stochastic bimodality occurs
-only when two of the three fixed points are close together, or in other
-words, when the system is poised ``sufficiently near'' one of the
-bifurcation points. This may become clearer when looking at the
-bifurcation diagram that we added to our section about hysteresis at the
-end of the main text. For convenience, we insert this plot below in
-Fig~\ref{bifucation} which addresses what we believe the reviewer is
-referring to: only when two of the three fixed points are close do the
-stochastic trajectories occupy both fixed points, resulting in
-bimodality.
-
-Since this bimodality only occurs for a small region in parameter space,
-this actually indicates strong and not fragile bistability. In other
-words, the fact that the system does not stochastically ``jump'' to the
-other fixed point means that the fixed points are very stable to
-stochastic fluctuations. Thus, the switch-like feature is robust:
-there are two distinct states that the system can occupy, and only near
-the bifurcation points is it likely for stochastic trajectories to
-transition between them. Put differently, deterministic bistability does
-not need to result in stochastic bimodality, and in fact stochastic
-bimodality is contrary to robust bistability.
+	The reason behind this apparent problem is that in the PDF, the figure numbering starts at one, while it starts at zero in the figure filenames.
+Thus, there is a shift of one between the figure filenames and the numbering in the PDF.
+However, the names and numbering in the original tex file code correspond to the figure filenames, such that we believe it should be clear to the editorial staff whenever they include the figures in the tex file.
 \end{response}
 
-\begin{figure}[h!]
-	\centering
-	\includegraphics{Fig10_bifurcation.eps}
-	\vspace*{0.5em}
-	\caption{{\bf Bifurcation diagram showing the hysteresis loop in the stochastic system.}
-		The blue line shows the positions of the deterministic fixed points, while the dashed line indicates the instability of the middle one.
-The orange and yellow points show the mean of the stochastic simulations when started at zero and at the high fixed point, respectively.
-The positions of the two peaks in the bimodal distributions are indicated by smaller points, connected by dotted lines.
-To make this clearer, the bimodal distributions themselves are shown on the right.
-The arrows illustrate the hysteresis loop in the stochastic simulations.}
-	\label{bifucation}
-\end{figure}
-
 \begin{review}
-	(and seems to take an unreasonably long time to switch between states).
+	As a final note, I would like to urge the authors to make the source code for their simulations publicly accessible, which would greatly help the community in adapting similar models for other systems.
 \end{review}
@@
 \begin{response}
-To address the indeed valid point of unreasonable switching times, we
-have compared the timescales in our stochastic simulations with those from the
-numerical solution of the deterministic model. We find excellent
-agreement, as shown in an updated figure in the main text and also below
-in Fig~\ref{adaptationtime}. So, again perhaps surprisingly,
-stochasticity does not materially alter the mean switching time as
-computed from the deterministic picture. Even adding additional
-stochasticity through transcriptional bursting (discussed more below)
-does not alter the mean switching time, though it does broaden the
-distribution of times.
-It is possible that this broadening resolves the apparent paradox. Even
-if the mean switching time is quite long, a non-negligible fraction of a
-population of cells could nevertheless switch on a shorter time-scale,
-especially with inclusion of extra stochasticity beyond what we have
-explicitly considered.
+	We fully agree with the reviewer and therefore have made all source code publicly available in the same repository as the raw data: \url{https://github.com/RPGroup-PBoC/xap_wt_paper}.
 \end{response}
 
-	\begin{figure}[h!]
-		\centering
-		\includegraphics{Fig9_evolution.eps}
-		\vspace*{0.5em}
-		\caption{{\bf Stochastic and deterministic time evolution of protein (XapA/XapB) and adaptation time.}
-			(A) Blue shows the results from one typical run of the stochastic simulation, light blue shows two more extreme runs, and orange shows the trajectory obtained from solving the deterministic ODE's.
-The simulation was started
-			at an mRNA, protein and intracellular xanthosine count of 0. The
-			extracellular xanthosine concentration was chosen to be
-			$\mathrm{[c]_a} = 25$.
-(B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 1000 runs of the simulation (same conditions as (A)).
-The orange line shows the corresponding deterministic time.
-The green dashed line right next to it indicated the mean of the blue distribution.}
-		\label{adaptationtime}
-	\end{figure}
-
-\begin{review}
-I think the authors are likely correct in their assertion that they
-underestimate the amount of noise in the system due to not including
-transcriptional and translational bursting, and that the transition
-rates would likely be faster, but this gives no idea of how *much*
-faster, and thus it seems premature to simply declare victory and move
-on.
-\end{review}
-
-\begin{response}
-This is a very reasonable concern which we have investigated more
-thoroughly now. We added rather strong transcriptional bursting to our
-simulation and do not find qualitative changes in the results. In
-particular, fluctuations around the mean become larger and, accordingly,
-distributions broader. Furthermore, bimodality begins to appear for
-lower values of the extracellular xanthosine concentration. Very
-interestingly, the change in the \emph{mean} transition time between the
-stable fixed points is negligible. However, the \emph{variance} in this
-time across many runs of the simulation becomes larger and the
-\emph{peak} of the distribution shifts to shorter times. We have
-included two new figures in the SI discussing these results (Figs. 6 and
-7), along with discussion there and in the main text.
-%	Below, we also show the two new figures from S1 Text that illustrate the similarities and differences between the simulation with and without transcriptional bursting.
-\end{response}
-
-%\begin{figure}
-%	\centering
-%	\includegraphics[width=1\textwidth]{FigSI6_7.pdf}
-%	\caption{{\bf Time evolution and adaptation times from the stochastic simulation when transcriptional bursting is included.}
-%		(A) shows the time evolution from one run of the simulation. The simulation was started
-%		at an mRNA, protein and intracellular xanthosine count of 0. The
-%		extracellular xanthosine concentration was chosen to be
-%		$\n{[c]_a} = 25$.
-(B) Blue shows the time until 90\% of the protein concentration at the upper fixed point was reached in 1000 runs of the simulation (same conditions as (A)).
-The distribution from the main text (without bursting) is shown in orange for comparison.}
-%	\label{evolbursts}
-%\end{figure}
-%
-%\begin{figure}
-%	\centering
-%	\includegraphics[width=1\textwidth]{FigSI6_7_2.pdf}
-%	\caption{{\bf Distributions from the stochastic simulation when transcriptional bursting is included.}
-%		Blue shows the distributions from the simulation with bursts, orange shows those without bursts from the main text for comparison.
-The following values of $\mathrm{[c]_a}$ were used: (A) $\n{[c]_a}=12$, (B) $\n{[c]_a}=17.5$ (not shown in main text), (C) $\mathrm{[c]_a}=18.5$, and (D) $\mathrm{[c]_a}=25$ (recall $\n{[c]_{a}} \defeq \frac{c}{K_{\n{a}}}$,
-%		$K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$). All other parameters are as in the table in the main text, and the simulation
-%		was run 1000 times for $10^6 \unit{s}$ each (simulated time) and
-%		started at a mRNA, protein and intracellular xanthosine count of 0.}
-%	\label{distbursts}
-%\end{figure}
-
-\begin{review}
-My recommendation would be that the authors put in some more cautionary words about the applicability of the continuous model, and ideally spend a little more time characterizing the discrete model.
-in terms of switching times, parameters that give rise to bimodal behavior, effects of incorporating reasonable amounts of transcriptional bursting, etc.
-Some more direct comparisons of the behavior of the two models in similar ranges of parameters space would also likely be enlightening.
-\end{review}
-
-\begin{response}
-We hope that we have addressed all these points in our responses above
-and shortly summarize here: We put in some cautionary words about the
-general usage of continuous models (beginning of sections on deterministic as well as stochastic results) and demonstrated that for our system
-and the conclusions we draw, the continuous model is surprisingly appropriate (bifurcation diagram and time evolution, end of results section).
-Furthermore, we extended our characterization of the stochastic model as
-suggested by investigating the distribution of adaptation/switching
-times (end of results section) as well as by incorporating transcriptional bursting in our
-simulation (end of section C in S1 Text). The deterministic and stochastic model are now directly
-compared (for the same set of parameters) in the bifurcation diagram as
-well as in terms of their adaptation/switching times.
-\end{response}
-
-\begin{review}
-My other major difficulty here is that essentially all of the work is
-premised on reference 22, which is 'unpublished data.' This is the only
-experimental evidence given for the bistability of xapAB expression,
-justification of many of the model parameters, experimentally expected
-switching timescales, etc. It is impossible to properly interpret the
-present work, either for the purposes of review or from the viewpoint of
-an interested reader, without availability of those data, which either
-ought to be their own independent publication or ought to be included
-with this one.
-\end{review}
-
-\begin{response}
-With further consideration, we agree with the reviewer that we relied
-rather heavily on this unpublished data and that it is therefore
-essential to make the experimental data available.
-We have opted to include the data with this publication rather than
-separately.
-For this reason, we have added Griffin Chure, Nathan Belliveau, Charlotte Strandkvist, and Kyle Naughton to the author list, who collected or were involved in collecting the data.
-We have added a new section, ``experimental motivation,'' in
-the beginning of the main text which discusses the main features of the
-data that are relevant for our model. Furthermore, we added a section on
-the materials and experimental methods at the end of S1 Text (section
-E).
-The raw data is
-publicly available at DOI: 10.5281/zenodo.3695049.
-\end{response}
-
-%\begin{figure}[h]
-%    \centering{
-    %\includegraphics{figure_path}
-%     \caption{\textbf{Figure caption.}}
-%    \label{fig:figlabel}}
-% \end{figure}
-
-\bibliography{library}
+%\bibliography{library}
 

### 2020-03-30 main_text/main_text.tex
Second revision according to reviewer feeback

@@
 		\includegraphics{media/Fig8_distribution.eps}
-		\caption{{\bf Distributions from stochastic simulations and the corresponding phase portraits.}
+		\caption{{\bf Distributions from stochastic simulations and the corresponding deterministic phase portraits.}
 			Apart from $\n{[c]_a}$, the parameters are the same as in
-			Table~\ref{table1:nondim}. For the distributions, the simulations
+			Table~\ref{table1:nondim}. The phase portraits were obtained from the deterministic system as those presented in the previous sections. For the distributions, the simulations
 			were run 5000 times for $10^6 \unit{s}$ each (simulated time) and
